---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")
require("ggvenn")
require("DESeq2")
require("STRINGdb")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-4"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
tmp_dir <- file.path(analysis_path, "temp")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)
dir.create(tmp_dir, recursive = T)


```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

candidate_genes_df <- readRDS(file.path(save_dir, "02-save-genes-where-the-variation-due-to-their-native-repeat-containing-intron-is-significant.rds"))
sig_result_list <- readRDS("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds")


```

- Perform gene set enrichment 

```{r}

significant_events <- sig_result_list$significant_events

string_db <- STRINGdb$new(version = "11.5", species = 9606, score_threshold = 200, input_directory = tmp_dir)

enrichment_df <- string_db$get_enrichment(candidate_genes_df$gene_id)
cell_types <- lapply(str_split(enrichment_df$inputGenes, pattern = ","), function(x){
    unique(significant_events[significant_events$gene_id %in% x, "tissue_type"])
})
enrichment_df$cell_type <- cell_types

enrichment_df <- enrichment_df[enrichment_df$fdr < 0.05,]

saveRDS(object = enrichment_df, file = file.path(save_dir, "02-save-gene-set-enrichment-for-repeat-containing-genes.rds"))
write.csv(x = enrichment_df[, -ncol(enrichment_df)], file = file.path(output_dir, "02-save-gene-set-enrichment-for-repeat-containing-genes.csv"))

```
