---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("ggplot2")
require("ggvenn")
require("STRINGdb")
require("parallel")
require("GenomicFeatures")
require("Biostrings")
require("BSgenome.Hsapiens.UCSC.hg38")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-4"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
tmp_dir <- file.path(analysis_path, "temp")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)
dir.create(tmp_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
all_gene_vs_lfcs_df <- readRDS(file.path(save_dir, "02-als-vs-control-genic-lfc-intronic-lfc-df-with-predictor-groups.rds"))
significant_de_event_list <- readRDS("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds")

```

- Select significantly affected genes


```{r}

significant_de_event_df <- significant_de_event_list$significant_events
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges
repeat_ranges <- repeat_ranges_list$repeat_ranges

strong_predictor_lfc_df <- all_gene_vs_lfcs_df[all_gene_vs_lfcs_df$predictor_group == "strong_predictors" & 
!is.na(all_gene_vs_lfcs_df$predictor_group) & 
all_gene_vs_lfcs_df$pathogenicity_pval < 0.05 & 
all_gene_vs_lfcs_df$genic_pval < 0.05 & 
all_gene_vs_lfcs_df$intron_pval < 0.05,]

```

- Number of introns which are in these genes

```{r}

feature_ranges_with_repeats$hash_id[
    feature_ranges_with_repeats$gene_id %in% strong_predictor_lfc_df$gene_id & 
    feature_ranges_with_repeats$feature_id %in% significant_de_event_df$feature_id[significant_de_event_df$region == "intron"]] %>% unique %>% length


```

- Plot LFC vs LFC plot for genes and introns


```{r}


ns_plot_df <- all_gene_vs_lfcs_df[all_gene_vs_lfcs_df$predictor_group == "strong_predictors" & 
!is.na(all_gene_vs_lfcs_df$predictor_group) & 
!(all_gene_vs_lfcs_df$pathogenicity_pval < 0.05 & 
all_gene_vs_lfcs_df$genic_pval < 0.05 & 
all_gene_vs_lfcs_df$intron_pval < 0.05),]
ns_plot_df$selection_group <- "not_selected"


sig_plot_df <- all_gene_vs_lfcs_df[all_gene_vs_lfcs_df$predictor_group == "strong_predictors" & 
!is.na(all_gene_vs_lfcs_df$predictor_group) & 
all_gene_vs_lfcs_df$pathogenicity_pval < 0.05 & 
all_gene_vs_lfcs_df$genic_pval < 0.05 & 
all_gene_vs_lfcs_df$intron_pval < 0.05,]
sig_plot_df$selection_group <- "selected"


plot_df <- rbind(sig_plot_df,ns_plot_df)



current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12, hjust = 1),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

Colours <- c("selected" = "#F36F48", "not_selected" = "#FFF9EC")
# Colours <- c("repeat_containing" = "#ff6b6c", "non_repeat_containing" = "#f3f9e3")
ThePlot <- ggplot(plot_df[plot_df$selection_group == "not_selected",], aes(x = genic_lfc, y = intron_lfc))
ThePlot <- ThePlot + geom_vline(xintercept = 0, linetype = "dashed")
ThePlot <- ThePlot + geom_hline(yintercept = 0, linetype = "dashed")
# ThePlot <- ThePlot + geom_density_2d_filled(alpha = 0.5, contour_var = "ndensity")
# ThePlot <- ThePlot + scale_fill_viridis(palette = "magma")
ThePlot <- ThePlot + geom_point(aes(fill = selection_group), shape = 21, alpha = 0.6, colour = "#000000")
ThePlot <- ThePlot + geom_point(data = plot_df[plot_df$selection_group == "selected",], aes(fill = selection_group), shape = 21, size = 1.5, colour = "#000000")
# ThePlot <- ThePlot + geom_smooth(method = "lm", formula = y ~ x, colour = "#000000")
# ThePlot <- ThePlot + geom_density(aes(colour = group))
# ThePlot <- ThePlot + geom_violin(aes(fill = group), colour = "#000000", 
#     lwd = 0.5, position = position_dodge(width = 0.9), scale = "width")
# ThePlot <- ThePlot + geom_boxplot(aes(fill = group), colour = "#000000", 
#     width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9), outlier.shape = NA)
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_continuous("log2 fold change gene expression (exon)")
ThePlot <- ThePlot + scale_y_continuous("log2 fold change of intron expression")
ThePlot <- ThePlot + scale_fill_manual("predictor classification", values = Colours)
# ThePlot <- ThePlot + scale_colour_manual("status", values = Colours)
# ThePlot <- ThePlot + scale_fill_brewer("predictor classification", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(status ~ .)
ThePlot <- ThePlot + ggtitle("read prop scatter plot")
ThePlot <- ThePlot + current_theme
ggsave(width = 5, height = 5, file = file.path(plot_dir, "03-02-plot-als-vs-control-lfc-plot-for-selected-strong-predictors-figure-4B.pdf"), ThePlot)


```

- Write repeat sequences for testing

```{r}

sorted_good_predictor_lfc_df <- strong_predictor_lfc_df[order(-log10(strong_predictor_lfc_df$pathogenicity_pval), decreasing = T),]

selected_introns <- feature_ranges_with_repeats[feature_ranges_with_repeats$gene_id %in% sorted_good_predictor_lfc_df$gene_id & 
    feature_ranges_with_repeats$type == "intron"]

selected_repeat_ranges <- repeat_ranges[selected_introns$which_repeat]
selected_repeat_ranges$hash_id <- selected_introns$hash_id
coord_start <- start(selected_repeat_ranges) - floor((100 - width(selected_repeat_ranges))/2)
coord_end <- end(selected_repeat_ranges) + floor((100 - width(selected_repeat_ranges))/2)

start(selected_repeat_ranges) <- coord_start
end(selected_repeat_ranges) <- coord_end

seqlevelsStyle(selected_repeat_ranges) <- "UCSC"
all_feature_sequences <- getSeq(BSgenome.Hsapiens.UCSC.hg38, selected_repeat_ranges)

names(all_feature_sequences) <- paste(selected_repeat_ranges$hash_id,names(selected_repeat_ranges), sep = ":")


writeXStringSet(x = all_feature_sequences, 
    filepath = file.path(output_dir, "01_features_containing_repeats_sig_intronic_foldchange_sig_genic_foldchange_ordered_by_pathogenicity.fa"),
    format = "fasta")

```

