---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("ggplot2")
require("ggvenn")
require("STRINGdb")
require("parallel")
require("GenomicFeatures")
require("Biostrings")
require("ggbeeswarm")
require("BSgenome.Hsapiens.UCSC.hg38")

genome <- BSgenome.Hsapiens.UCSC.hg38

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-4"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
tmp_dir <- file.path(analysis_path, "temp")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)
dir.create(tmp_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

feature_sequences <- readDNAStringSet(file.path(output_dir, "01_features_containing_repeats_sig_intronic_foldchange_sig_genic_foldchange_ordered_by_pathogenicity.fa"))
sequences_mapped_to_nereida <- read.table("final-figures/figure-4/input_files/files-from-nereida.tsv", header = F, sep = " ")
repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")

repeat_ranges <- repeat_ranges_list$repeat_ranges
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges

```

- Find the sequences which were validated by nereida

```{r}

names_split <- str_split(sequences_mapped_to_nereida[,2], pattern = ":", simplify = F)
repeat_coords <- vapply(names_split, function(x){
    paste(x[-1], collapse = ":")
},"1")

selected_sequence_ranges <- repeat_ranges[repeat_coords]
selected_sequence_ranges$nereida_ids <- sequences_mapped_to_nereida[,1]
selected_sequence_ranges$phase_separating <- F
selected_sequence_ranges$phase_separating[selected_sequence_ranges$nereida_ids %in% c("seq-2B", "seq-3B", "seq-4B", "seq-10G", "seq-11H", "seq-13J", "seq-14J", "seq-15J")] <- T
seqlevelsStyle(selected_sequence_ranges) <- "UCSC"

```
- get GC county

```{r}

CG_counts <- do.call(rbind, lapply(seq_along(selected_sequence_ranges), function(x){
    current_ranges <- selected_sequence_ranges[x]
    chr_string_set <- genome[[as.vector(seqnames(current_ranges))]]
    temp_ranges <- IRanges(start = start(current_ranges), end = end(current_ranges))
    sequences_in_windows <- extractAt(x = chr_string_set, at = temp_ranges)
    count_matrix <- DataFrame(letterFrequency(x = sequences_in_windows, 
            as.prob = FALSE, letters = "CG", OR = "|"))
    return(count_matrix)
}))

full_sequence_CG_counts <- do.call(rbind, lapply(sequences_mapped_to_nereida[,2], function(x){
    count_matrix <- DataFrame("full_sequence" = letterFrequency(x = feature_sequences[x], 
            as.prob = FALSE, letters = "CG", OR = "|"))
    return(count_matrix)
}))


repeat_unit_length <- c("trinucleotide" = 3, "quadnucleotide" = 4, "pentanucleotide" = 5, "hexanucleotide" = 6)
mcols(selected_sequence_ranges, use.names = F) <- cbind(mcols(selected_sequence_ranges, use.names = F), CG_counts, full_sequence_CG_counts)
selected_sequence_ranges$repeat_width <- width(selected_sequence_ranges) / repeat_unit_length[selected_sequence_ranges$span_group]
selected_sequence_ranges$gc_percent <- selected_sequence_ranges$C.G / width(selected_sequence_ranges)
selected_sequence_ranges$sequence_width <- width(feature_sequences[sequences_mapped_to_nereida[,2]])
selected_sequence_ranges$full_sequence_gc_percent <- selected_sequence_ranges$full_sequence.C.G / selected_sequence_ranges$sequence_width

```


- split the ranges


```{r}

ranges_split <- split(selected_sequence_ranges, selected_sequence_ranges$phase_separating)

selected_sequence_ranges$high_gc <- selected_sequence_ranges$full_sequence_gc_percent > 0.44

gc_pval <- vapply(seq(0.4, 0.5, by = 0.01), function(x){
    fisher.test(table(gc = selected_sequence_ranges$full_sequence_gc_percent > x, aggregate = selected_sequence_ranges$phase_separating), alternative = "greater")$p.value
},1)
names(gc_pval) <- seq(0.4, 0.5, by = 0.01)

plot_df <- data.frame(gc_percent = selected_sequence_ranges$full_sequence_gc_percent, 
    aggregate = selected_sequence_ranges$phase_separating)
vapply(split(plot_df, plot_df$aggregate), function(x){
    mean(x$gc_percent)
},1)


current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12, hjust = 1),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

Colours <- c("selected" = "#F36F48", "not_selected" = "#FFF9EC")
# Colours <- c("repeat_containing" = "#ff6b6c", "non_repeat_containing" = "#f3f9e3")
ThePlot <- ggplot(plot_df, aes(x = aggregate, y = gc_percent))
ThePlot <- ThePlot + geom_quasirandom(fill = "#FFF9EC", colour = "#000000", shape = 21)
ThePlot <- ThePlot + geom_boxplot(fill = "#FFF9EC", colour = "#000000", 
    width = 0.1, lwd = 0.5, alpha = 0.6, position = position_dodge(width = 0.9))
ThePlot <- ThePlot + scale_x_discrete("Aggregation property")
ThePlot <- ThePlot + scale_y_continuous("GC percent")
# ThePlot <- ThePlot + scale_fill_manual("predictor classification", values = Colours)
# ThePlot <- ThePlot + scale_colour_manual("status", values = Colours)
# ThePlot <- ThePlot + scale_fill_brewer("predictor classification", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(status ~ .)
ThePlot <- ThePlot + ggtitle("read prop scatter plot")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 7, file = file.path(plot_dir, "03-04-plot-GC-percent-of-repeat-sequences-supplementary-figure-15.pdf"), ThePlot)


```