---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

library("here")
message("Project root:", here(), "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("ggplot2")
require("ggvenn")
require("DESeq2")
require("STRINGdb")
require("parallel")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("final-figures", "figure-4")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
tmp_dir <- file.path(analysis_path, "temp")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)
dir.create(tmp_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS(here("final-figures","figure-1","rdata_files","00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds"))
pan_als_diff_object <- readRDS(here("13-03-public-data","rdata_files","02-02-01-save-deseq2-all-samples-als-vs-control-differential-object.rds"))
pan_als_contrast_list <- readRDS(here("13-03-public-data","rdata_files","03-01-02-assembled-contrasts-from-pan-als-comparison.rds"))
pan_als_gene_contrast_list <- readRDS(here("13-03-public-data","rdata_files","05-02-assembled-contrasts-from-pan-als-comparison.rds"))
candidate_genes_df <- readRDS(file.path(save_dir, "02-save-genes-where-the-variation-due-to-their-native-repeat-containing-intron-is-significant.rds"))
# diff_df <- readRDS(file.path(save_dir, "02-03-difference-of-median-distances-between-repeat-containing-and-non-repeat-containing-features.rds"))
hpo_terms_with_cluster_df <- readRDS(file.path(save_dir, "03-hpo-interesting-terms-with-group-assignment-and-plot-order.rds"))
significant_de_event_list <- readRDS(here("20-import-significant-events","rdata_files","02-significantly-differentially-expressed-events.rds"))
feature_ranges <- readRDS(here("05-simulated-reads-differential-profiling","rdata_files","01-all-merged-exonic-and-intronic-segments-in-genes.rds"))
model_output_list <- readRDS(file.path(save_dir, "01-save-model-output.rds"))

```


- Select terms to plot

```{r}


significant_de_event_df <- significant_de_event_list$significant_events
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges

significant_gene_df <- unique(significant_de_event_df[significant_de_event_df$pval < 0.05 & significant_de_event_df$region == "intron", c("gene_id", "status")])
status_split_by_genes <- split(significant_gene_df$status, significant_gene_df$gene_id)
genes_with_repeats <- names(status_split_by_genes[vapply(status_split_by_genes, function(x){
 "repeat_containing" %in% x
},T)])
unique_gene_ids <- unique(significant_gene_df$gene_id[significant_gene_df$gene_id %in% genes_with_repeats])
significant_gene_df <- significant_gene_df[significant_gene_df$gene_id %in% unique_gene_ids,]

feature_ranges$status <- "non_repeat_containing"
feature_ranges$status[feature_ranges$feature_id %in% feature_ranges_with_repeats$feature_id] <- "repeat_containing"


```

- Get counts for RCIs and non-RCIs


```{r}

norm_matrix <- counts(pan_als_diff_object, normalized = TRUE)
cpm_matrix <- t(t(norm_matrix) / colSums(norm_matrix))

coldata_df <- colData(pan_als_diff_object)
# keep_matrix <- norm_matrix[c(unique(percent_contribution_df$intron), unique(percent_contribution_df$exon)),]


feature_ranges_split <- split(feature_ranges, feature_ranges$gene_id)[unique(significant_gene_df$gene_id[significant_gene_df$status == "repeat_containing"])]
all_gene_level_intronic_counts_repeat <- do.call(rbind, mclapply(feature_ranges_split, function(x){
    get_ids <- rownames(cpm_matrix)[rownames(cpm_matrix) %in% x$hash_id[x$type == "intron" & x$status == "repeat_containing"]]
    if(length(get_ids) == 0){
        return(0)
    } 
    if(length(get_ids) == 1){
        return(mean(cpm_matrix[get_ids,], na.rm = T))
    }
    return(mean(colSums(cpm_matrix[get_ids,], na.rm = T)))
},mc.cores = 8))
all_gene_level_intronic_counts_nonrepeat <- do.call(rbind, mclapply(feature_ranges_split, function(x){
    get_ids <- rownames(cpm_matrix)[rownames(cpm_matrix) %in% x$hash_id[x$type == "intron" & x$status != "repeat_containing"]]
    if(length(get_ids) == 0){
        return(0)
    } 
    if(length(get_ids) == 1){
        return(mean(cpm_matrix[get_ids,], na.rm = T))
    }
    return(mean(colSums(cpm_matrix[get_ids,], na.rm = T)))
},mc.cores = 8))


```

- Get log fold changes and pvalues


```{r}


all_gene_level_intronic_counts_repeat_lfc <- do.call(rbind, mclapply(feature_ranges_split, function(x){
    current_contrasts_df <- pan_als_contrast_list[pan_als_contrast_list$feature_ids %in% x$hash_id[x$type == "intron" & x$status == "repeat_containing"],]
    return(mean(current_contrasts_df$logfc, na.rm = T))
},mc.cores = 8))

all_gene_level_intronic_counts_nonrepeat_lfc <- do.call(rbind, mclapply(feature_ranges_split, function(x){
    current_contrasts_df <- pan_als_contrast_list[pan_als_contrast_list$feature_ids %in% x$hash_id[x$type == "intron" & x$status != "repeat_containing"],]
    return(mean(current_contrasts_df$logfc, na.rm = T))
}, mc.cores = 8))

all_gene_level_intronic_counts_repeat_pval <- do.call(rbind, mclapply(feature_ranges_split, function(x){
    current_contrasts_df <- pan_als_contrast_list[pan_als_contrast_list$feature_ids %in% x$hash_id[x$type == "intron" & x$status == "repeat_containing"],]
    return(mean(current_contrasts_df$pval, na.rm = T))
},mc.cores = 8))

all_gene_level_intronic_counts_nonrepeat_pval <- do.call(rbind, mclapply(feature_ranges_split, function(x){
    current_contrasts_df <- pan_als_contrast_list[pan_als_contrast_list$feature_ids %in% x$hash_id[x$type == "intron" & x$status != "repeat_containing"],]
    return(mean(current_contrasts_df$pval, na.rm = T))
}, mc.cores = 8))

all_gene_level_intronic_counts_repeat_width <- do.call(rbind, mclapply(feature_ranges_split, function(x){
    current_ranges <- x[x$hash_id[x$type == "intron" & x$status == "repeat_containing"] %in% pan_als_contrast_list$feature_ids]
    return(mean(width(current_ranges)))
},mc.cores = 8))

all_gene_level_intronic_counts_nonrepeat_width <- do.call(rbind, mclapply(feature_ranges_split, function(x){
    current_ranges <- x[x$hash_id[x$type == "intron" & x$status != "repeat_containing"] %in% pan_als_contrast_list$feature_ids]
    return(mean(width(current_ranges)))
},mc.cores = 8))


```

- Get model output results

```{r}

percent_contribution_df <- model_output_list$percent_contribution
correlation_df <- model_output_list$correlation_df
pval_df <- model_output_list$pval_df


diff_df <- do.call(rbind, lapply(split(percent_contribution_df, percent_contribution_df$gene_id), function(temp_df){
    repeat_val <- temp_df$value[temp_df$group == "repeat_containing"]
    non_repeat_val <- temp_df$value[temp_df$group != "repeat_containing"]
    data.frame(gene_id = unique(temp_df$gene_id), perc_diff = median(repeat_val) - median(non_repeat_val))
}))
rownames(diff_df) <- NULL
diff_df$padj <- pval_df$padj[match(diff_df$gene_id, pval_df$gene_id)]
diff_df$group <- ifelse(diff_df$padj < 0.05, "Sig.", "NS")

pathogenic_potential_values_df <- do.call(rbind, lapply(split(percent_contribution_df, percent_contribution_df$gene_id), function(x){
    data.frame(gene_id = unique(x$gene_id),
        group = c("non_repeat_containing", "repeat_containing"),
        pathogenic_potential = c(median(x$value[x$group == "non_repeat_containing"]), median(x$value[x$group == "repeat_containing"])))
}))
rownames(pathogenic_potential_values_df) <- NULL

non_repeat_potential_df <- pathogenic_potential_values_df[pathogenic_potential_values_df$group == "non_repeat_containing",]
repeat_potential_df <- pathogenic_potential_values_df[pathogenic_potential_values_df$group == "repeat_containing",]

```

- Aggregate all data together

```{r}


gene_vs_non_repeat_df <- data.frame(gene_id = rownames(all_gene_level_intronic_counts_nonrepeat_lfc),
    status = "non_repeat_containing",
    intron_lfc = all_gene_level_intronic_counts_nonrepeat_lfc,
    genic_lfc = pan_als_gene_contrast_list[match(rownames(all_gene_level_intronic_counts_nonrepeat_lfc), pan_als_gene_contrast_list$gene_ids),"logfc"],
    intron_pval = all_gene_level_intronic_counts_nonrepeat_pval,
    intronic_width = all_gene_level_intronic_counts_nonrepeat_width,
    intron_cpm = all_gene_level_intronic_counts_nonrepeat[rownames(all_gene_level_intronic_counts_nonrepeat_lfc),] * 1e6,
    genic_pval = pan_als_gene_contrast_list[match(rownames(all_gene_level_intronic_counts_nonrepeat_lfc), pan_als_gene_contrast_list$gene_ids),"pval"],
    pathogenic_potential = non_repeat_potential_df[match(rownames(all_gene_level_intronic_counts_nonrepeat_lfc), non_repeat_potential_df$gene_id), "pathogenic_potential"])

gene_vs_repeat_df <- data.frame(gene_id = rownames(all_gene_level_intronic_counts_repeat_lfc),
    status = "repeat_containing",
    intron_lfc = all_gene_level_intronic_counts_repeat_lfc,
    genic_lfc = pan_als_gene_contrast_list[match(rownames(all_gene_level_intronic_counts_repeat_lfc), pan_als_gene_contrast_list$gene_ids),"logfc"],
    intron_pval = all_gene_level_intronic_counts_repeat_pval,
    intronic_width = all_gene_level_intronic_counts_repeat_width,
    intron_cpm = all_gene_level_intronic_counts_repeat[rownames(all_gene_level_intronic_counts_repeat_lfc),] * 1e6,
    genic_pval = pan_als_gene_contrast_list[match(rownames(all_gene_level_intronic_counts_repeat_lfc), pan_als_gene_contrast_list$gene_ids),"pval"],
    pathogenic_potential = repeat_potential_df[match(rownames(all_gene_level_intronic_counts_nonrepeat_lfc), repeat_potential_df$gene_id), "pathogenic_potential"])


all_gene_vs_lfcs_df <- rbind(gene_vs_non_repeat_df, gene_vs_repeat_df)
all_gene_vs_lfcs_df$pathogenicity_pval <- diff_df[match(all_gene_vs_lfcs_df$gene_id, diff_df$gene_id), "padj"]
all_gene_vs_lfcs_df$significance <- ifelse(all_gene_vs_lfcs_df$pathogenicity_pval < 0.05, "fdr < 0.05", "NS")


```


- Add predictor groups 

```{r}

gene_split <- lapply(split(str_split(hpo_terms_with_cluster_df$inputGenes, pattern = ","), hpo_terms_with_cluster_df$group), function(x){
    unique(do.call(c, x))
})
gene_split[["moderate_predictors"]] <- gene_split[["moderate_predictors"]][!(gene_split[["moderate_predictors"]] %in% gene_split[["strong_predictors"]])]
gene_split[["modest_predictors"]] <- gene_split[["modest_predictors"]][
    !(gene_split[["modest_predictors"]] %in% gene_split[["moderate_predictors"]]) & 
    !(gene_split[["modest_predictors"]] %in% gene_split[["strong_predictors"]])]

all_gene_vs_lfcs_df$predictor_group <- NA
all_gene_vs_lfcs_df$predictor_group[all_gene_vs_lfcs_df$gene_id %in% gene_split$strong_predictors] <- "strong_predictors"
all_gene_vs_lfcs_df$predictor_group[all_gene_vs_lfcs_df$gene_id %in% gene_split$moderate_predictors] <- "moderate_predictors"
all_gene_vs_lfcs_df$predictor_group[all_gene_vs_lfcs_df$gene_id %in% gene_split$modest_predictors] <- "modest_predictors"

saveRDS(object = all_gene_vs_lfcs_df, file = file.path(save_dir, "02-als-vs-control-genic-lfc-intronic-lfc-df-with-predictor-groups.rds"))

```

