---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")
require("ggvenn")
require("DESeq2")
require("STRINGdb")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-4"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
tmp_dir <- file.path(analysis_path, "temp")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)
dir.create(tmp_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
repeat_containing_ranges <- repeat_ranges_list$repeat_containing_ranges
enrichment_df <- readRDS(file.path(save_dir, "02-save-gene-set-enrichment-for-repeat-containing-genes.rds"))
sig_result_list <- readRDS("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds")
significant_events <- sig_result_list$significant_events
candidate_genes_df <- readRDS(file.path(save_dir, "02-save-genes-where-the-variation-due-to-their-native-repeat-containing-intron-is-significant.rds"))

```


- Select terms to plot

```{r}


enrichment_df$prop_repeats <- vapply(str_split(enrichment_df$inputGenes, pattern = ","), function(x){
    length(intersect(x, candidate_genes_df$gene_id))/length(unique(candidate_genes_df$gene_id))
},1)

strong_predictors <- c("EFO:0004713", "HP:0003808", "HP:0001252",
    "EFO:0004314", "HP:0011804", "HP:0003011", "EFO:0006941", "HP:0002167", 
    "EFO:0005921", "EFO:0003892", "HP:0012638", "HP:0011442", "HP:0002817",
    "HP:0002814")
moderate_predictors <- c("EFO:0003925", "EFO:0008354", 
    "HP:0000708", "EFO:0007998", "EFO:0007710", "EFO:0007803",
    "EFO:0000618", "HP:0100543", "EFO:0005774",
    "EFO:0010144", "EFO:0009386", "HP:0001288", "HP:0001290", "EFO:0008475",
    "EFO:0007680", "EFO:0003926", "HP:0100022", "EFO:0004529", "EFO:0005105",
    "EFO:0004732")
modest_predictors <- c("EFO:0003926", "EFO:0004875", "EFO:0004731", "EFO:0004337", 
    "EFO:0008393", "EFO:0005229", "EFO:0005301", 
    "HP:0000736", "HP:0004408", "HP:0002020", "HP:0001251", 
    "EFO:0009700", "EFO:0007699", "HP:0008765", "HP:0000738", 
    "HP:0000486", "HP:0000364", "HP:0000508",  
    "EFO:0004874", "EFO:0004365", "EFO:0004555", "HP:0001249",
    "EFO:0007911", "EFO:0004530", "EFO:0005110", 
    "EFO:0004346", "HP:0031466", "EFO:0009594", "EFO:0005427", 
    "EFO:0004825", "HP:0002168", "EFO:0011015")

hpo_interesting_df <- enrichment_df[enrichment_df$term %in% c(strong_predictors, moderate_predictors, modest_predictors),]
hpo_interesting_df$group <- "modest_predictors"
hpo_interesting_df$group[hpo_interesting_df$term %in% moderate_predictors] <- "moderate_predictors"
hpo_interesting_df$group[hpo_interesting_df$term %in% strong_predictors] <- "strong_predictors"
hpo_interesting_df$size <- 1 
hpo_interesting_df$size[hpo_interesting_df$group == "strong_predictors"] <- 3
hpo_interesting_df$size[hpo_interesting_df$group == "moderate_predictors"] <- 2


```

- Plot HPO gene set enrichment


```{r}

hpo_interesting_df$description <- factor(hpo_interesting_df$description, levels = hpo_interesting_df$description[order(hpo_interesting_df$fdr, decreasing = F)])

hpo_split_by_cluster <- split(hpo_interesting_df, hpo_interesting_df$group)
hpo_split_by_cluster <- hpo_split_by_cluster[c("strong_predictors", "moderate_predictors", "modest_predictors")]
corrected_plot_order_hpo_df <- do.call(rbind, lapply(seq_along(hpo_split_by_cluster), function(x){
    offset <- 0
    if(x > 1){
        offset <- sum(vapply(seq(from = 1, to = x - 1), function(y){
            nrow(hpo_split_by_cluster[[y]])
        },1))
    }
    temp_df <- hpo_split_by_cluster[[x]]
    temp_df$plot_order <- seq(from = 1, to = nrow(temp_df)) + offset
    temp_df
}))

vline_df <- do.call(rbind, lapply(seq_along(hpo_split_by_cluster), function(x){
    offset <- 0
    if(x > 1){
        offset <- sum(vapply(seq(from = 1, to = x - 1), function(y){
            nrow(hpo_split_by_cluster[[y]])
        },1))
    }
    temp_df <- data.frame(xaxis = nrow(hpo_split_by_cluster[[x]]) + 0.5 + offset)
    temp_df
}))
vline_df <- rbind(vline_df, data.frame(xaxis = 0.5))

breaks <- c(corrected_plot_order_hpo_df$plot_order, max(corrected_plot_order_hpo_df$plot_order) + 1)
break_lab <- c(as.character(corrected_plot_order_hpo_df$description), "")

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 8, angle = 90, hjust = 1),
    axis.text.y = element_text(size = 8),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

Colours <- c("strong_predictors" = "#f36f48", "moderate_predictors" = "#5688C7", "modest_predictors" = "#FDE2E2")
ThePlot <- ggplot(corrected_plot_order_hpo_df, aes(x = plot_order, y = -log10(fdr)))
# ThePlot <- ThePlot + geom_vline(xintercept = 0, linetype = "dashed", colour = "#000000")
ThePlot <- ThePlot + geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "#000000")
ThePlot <- ThePlot + geom_vline(data = vline_df, aes(xintercept = xaxis), linetype = "solid", colour = "#000000")
ThePlot <- ThePlot + geom_point(aes(fill = as.character(group)), colour = "#000000", shape = 21)
ThePlot <- ThePlot + scale_x_continuous("description", breaks = breaks, expand = c(0,0), labels = break_lab)
ThePlot <- ThePlot + scale_y_continuous("-log10(fdr)")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + scale_fill_manual("Predictors", values = Colours)
# ThePlot <- ThePlot + facet_grid(. ~ cluster, scales = "free_x")
ThePlot <- ThePlot + ggtitle("gene set enrichment for repeat containing genes")
ThePlot <- ThePlot + current_theme
ggsave(height = 5, width = 10, file = file.path(plot_dir, "03-01-plot-HPO-gene-set-enrichment-figure-4A.pdf"), ThePlot)

saveRDS(object = corrected_plot_order_hpo_df, file = file.path(save_dir, "03-hpo-interesting-terms-with-group-assignment-and-plot-order.rds"))

```
