---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the scripts required to generate Supplementary figure 18C and 18D. We model proportion of intronic signal assigned to R-loops as a function of intronic length assigned to R-loops. We then validate this model using a monte carlo approach.


### 1.0 Package requirements

- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

library("here")
message("Project root:", here(), "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("ggplot2")
require("ggbeeswarm")
require("ggvenn")
require("STRINGdb")
require("parallel")
require("GenomicFeatures")
require("Biostrings")


```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("final-figures", "figure-4")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
tmp_dir <- file.path(analysis_path, "temp")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)
dir.create(tmp_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}



repeat_ranges_list <- readRDS(here("final-figures","figure-1","rdata_files","00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds"))
sig_result_list <- readRDS(here("20-import-significant-events","rdata_files","02-significantly-differentially-expressed-events.rds"))
feature_ranges <- readRDS(here("05-simulated-reads-differential-profiling","rdata_files","01-all-merged-exonic-and-intronic-segments-in-genes.rds"))
annotation_gtf <- here("Genomes","Homo_sapiens.GRCh38.104.chr_patch_hapl_scaff.gtf")
rloop_ranges <- readRDS(here("30-sequence-analysis-of-introns","rdata_files","90-01-merged-rloop-ranges-in-differentially-expressed-introns.rds"))
dripseq_samplesheet_df <- read.csv("../sporadic-pbmc-drip-fero-et-al-2024/sample-details/sampleinfo.csv")
dripseq_aln_dir <- file.path("../sporadic-pbmc-drip-fero-et-al-2024/output_files/bigwig/")
dripseq_list <- readRDS("34-dripseq-analysis/rdata_files/02-01-list-of-total-rloop-coverage-matrix-and-library-sizes.rds")


```

- Fetch features containing repeats

```{r}

dripseq_samplesheet_df$bigwig <- vapply(dripseq_samplesheet_df$SampleName, function(x){
    list.files(path = dripseq_aln_dir, pattern = x) 
},"1")

```

- Generate signal matrices

```{r}

library_sizes <- dripseq_list$library_sizes
repeat_signal_matrix <- dripseq_list$repeat_signal_matrix
reduced_rloop_signal_matrix <- dripseq_list$reduced_rloop_signal_matrix
intronic_signal_matrix <- dripseq_list$feature_signal_matrix

reduced_rloop_ranges <- dripseq_list$reduced_rloop_ranges
rloop_width_per_intron <- vapply(split(width(reduced_rloop_ranges), reduced_rloop_ranges$feature_id), sum, 1)

```

- Get differentially expressed features

```{r}

repeat_ranges <- repeat_ranges_list$repeat_ranges
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges
sig_result_df <- sig_result_list$significant_events
sig_result_df <- sig_result_df[sig_result_df$region == "intron",]

de_feature_ranges <- feature_ranges[feature_ranges$feature_id %in% sig_result_df$feature_ids,]
de_feature_ranges_with_repeats <- feature_ranges_with_repeats[
    feature_ranges_with_repeats$feature_id %in% sig_result_df$feature_ids & 
    feature_ranges_with_repeats$type == "intron"]

```

- Fetch the following signal matrices
    - library size normalised matrices for reduced R-loops
    - library size normalised matrices for introns
    - library size normalised matrices for repeat regions

```{r}

norm_reduced_rloop_signal_matrix <- t(t(reduced_rloop_signal_matrix) / (library_sizes["total_coverage",colnames(reduced_rloop_signal_matrix)] / 1e6))

norm_intronic_signal_matrix <- t(t(intronic_signal_matrix) / (library_sizes["total_coverage",colnames(intronic_signal_matrix)] / 1e6))

norm_repeat_signal_matrix <- t(t(repeat_signal_matrix) / (library_sizes["total_coverage",colnames(repeat_signal_matrix)] / 1e6))

```

- Add quantitative columns to the differentially expressed features 


```{r}

de_feature_ranges$intronic_signal <- rowMeans(norm_intronic_signal_matrix)[de_feature_ranges$feature_id]
de_feature_ranges$rloop_signal <- rowMeans(norm_reduced_rloop_signal_matrix)[de_feature_ranges$feature_id]
de_feature_ranges$rloop_signal_prop <- de_feature_ranges$rloop_signal / de_feature_ranges$intronic_signal
de_feature_ranges$rloop_width <- rloop_width_per_intron[de_feature_ranges$feature_id]
de_feature_ranges$rloop_width_prop <- de_feature_ranges$rloop_width / width(de_feature_ranges)

```

- create regressor modelling proportion of intronic signal assigned to r-loop as a function of proportion of intronic length assigned to r-loop

```{r}

non_na_de_feature_ranges <- de_feature_ranges[!is.na(de_feature_ranges$rloop_signal_prop) & de_feature_ranges$type == "intron"]

model_df <- data.frame(proportion_of_signal = non_na_de_feature_ranges$rloop_signal_prop, 
    proportion_of_length = non_na_de_feature_ranges$rloop_width_prop)

a_model <- lm(data = model_df, formula = proportion_of_signal ~ proportion_of_length)

# Call:
# lm(formula = proportion_of_signal ~ proportion_of_length, data = model_df)

# Residuals:
#      Min       1Q   Median       3Q      Max
# -0.91772 -0.03606 -0.00409  0.03331  0.75913

# Coefficients:
#                       Estimate Std. Error t value Pr(>|t|)
# (Intercept)          0.0007896  0.0006664   1.185    0.236
# proportion_of_length 0.9696989  0.0017870 542.637   <2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

# Residual standard error: 0.1158 on 51852 degrees of freedom
# Multiple R-squared:  0.8503,    Adjusted R-squared:  0.8503
# F-statistic: 2.945e+05 on 1 and 51852 DF,  p-value: < 2.2e-16

```

- Plot the model

```{r}


ThePlot <- ggplot(model_df, aes(x = proportion_of_length, y = proportion_of_signal))
ThePlot <- ThePlot + geom_point(colour = "#000000", fill = "#FFF8F0", shape = 21)
ThePlot <- ThePlot + geom_smooth(method = "lm", formula = y ~ x, shape = 21)
ThePlot <- ThePlot + scale_x_continuous("Proportion of intronic width in R-loops")
ThePlot <- ThePlot + scale_y_continuous("Proportion of DRIP-seq signal in R-loops")
# ThePlot <- ThePlot + scale_fill_manual("classes", values = Colours)
# ThePlot <- ThePlot + scale_colour_manual("status", values = Colours)
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(class ~ ., scales = "free")
ThePlot <- ThePlot + ggtitle("frequency of repeats")
ThePlot <- ThePlot + theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))
ggsave(width = 10, height = 10, file = file.path(plot_dir, "04-05-01-model-proportion-of-intronic-signal-in-rloop-as-a-function-of-intronic-length-in-rloop-Supplementary-figure-18C.pdf"), ThePlot)

```


- Create the monte carlo simulation

```{r}

# select 75% of the data for training and keep 25% back for validation

select_n <- floor(nrow(model_df) * 0.75)
iter <- 10000
all_rows <- seq_len(nrow(model_df))
sampled_rows_list <- lapply(seq(1, iter), function(x){
    training_set <- sample(x = all_rows, size = select_n)
    return(list(training = training_set, validation = all_rows[!(all_rows %in% training_set)]))
})


all_model_list <- lapply(sampled_rows_list, function(x){
    a_model <- lm(data = model_df[x$training,], formula = proportion_of_signal ~ proportion_of_length)
    predict_df <- data.frame(proportion_of_length = model_df[x$validation,"proportion_of_length"])
    predict_df$predicted_signal_in_repeats <- predict(object = a_model, newdata = predict_df)
    predict_df$observed_signal_in_repeats <- model_df[x$validation,"proportion_of_signal"]
    return(list(prediction = predict_df, model = a_model))
})


all_model_rsq_df <- do.call(rbind, lapply(seq_along(all_model_list), function(x){
    stats <- summary(all_model_list[[x]]$model)
    data.frame(iter = x, rsq = stats$r.squared)
}))


```

- Compute the model statistics

```{r}

statistics_df <- do.call(rbind, lapply(seq_along(all_model_list), function(x){
    current_prediction_df <- all_model_list[[x]]$prediction
    correlation <- cor(current_prediction_df$predicted_signal_in_repeats, current_prediction_df$observed_signal_in_repeats)
    
    residual_sum_of_squares <- sum((current_prediction_df$observed_signal_in_repeats - current_prediction_df$predicted_signal_in_repeats)^2)
    total_sum_of_squares <- sum((current_prediction_df$observed_signal_in_repeats - mean(current_prediction_df$observed_signal_in_repeats))^2)
    rsquared <- 1 - (residual_sum_of_squares / total_sum_of_squares)

    root_mean_squared_error <- sqrt(mean(residual_sum_of_squares))

    mean_absolute_error <- mean(abs(current_prediction_df$observed_signal_in_repeats - current_prediction_df$predicted_signal_in_repeats))

    bias <- mean(current_prediction_df$predicted_signal_in_repeats - current_prediction_df$observed_signal_in_repeats)
    data.frame(iter = x, label = c("rsq", "rmse", "mae", "bias"), value = c(rsquared, root_mean_squared_error, mean_absolute_error, bias))
}))

lapply(unique(statistics_df$label), function(x){
    summary(statistics_df$value[statistics_df$label == x])
})



ThePlot <- ggplot(statistics_df, aes(x = value))
# ThePlot <- ThePlot + geom_violin(aes(fill = condition), colour = "#000000", 
#     lwd = 0.5, position = position_dodge(width = 0.9))
ThePlot <- ThePlot + geom_density(colour = "#000000", fill = "#FFF8F0")
ThePlot <- ThePlot + scale_x_continuous("quality statistics")
ThePlot <- ThePlot + scale_y_continuous("density")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_wrap(label ~ ., scales = "free")
ThePlot <- ThePlot + ggtitle("frequency of repeats")
ThePlot <- ThePlot + theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))
ggsave(width = 10, height = 10, file = file.path(plot_dir, "04-05-02-perform-monte-carlo-crossvalidation-75-25-on-regression-model-supplementary-figure-18D.pdf"), ThePlot)



```