---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}


library("here")
message("Project root:", here(), "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("ggplot2")
require("ggbeeswarm")
require("ggvenn")
require("STRINGdb")
require("parallel")
require("GenomicFeatures")
require("Biostrings")


```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("final-figures", "figure-4")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
tmp_dir <- file.path(analysis_path, "temp")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)
dir.create(tmp_dir, recursive = T)


```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS(here("final-figures","figure-1","rdata_files","00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds"))
sig_result_list <- readRDS(here("20-import-significant-events","rdata_files","02-significantly-differentially-expressed-events.rds"))
feature_ranges <- readRDS(here("05-simulated-reads-differential-profiling","rdata_files","01-all-merged-exonic-and-intronic-segments-in-genes.rds"))
annotation_gtf <- here("Genomes","Homo_sapiens.GRCh38.104.chr_patch_hapl_scaff.gtf")
rloop_ranges <- readRDS(here("30-sequence-analysis-of-introns","rdata_files","90-01-merged-rloop-ranges-in-differentially-expressed-introns.rds"))

```

- Select significantly affected genes

```{r}

repeat_ranges <- repeat_ranges_list$repeat_ranges
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges
sig_result_df <- sig_result_list$significant_events
sig_result_df <- sig_result_df[sig_result_df$region == "intron",]

```


- Isolate differentially expressed features and change seqlevels to ucsc.

```{r}


seqlevelsStyle(feature_ranges) <- "UCSC"
de_feature_ranges <- feature_ranges[feature_ranges$feature_id %in% sig_result_df$feature_id,]
de_feature_ranges$status <- ifelse(de_feature_ranges$feature_id %in% feature_ranges_with_repeats$feature_id, "repeat_containing", "non_repeat_containing")


```


- Compute R-loop density


```{r}


rloop_split <- split(rloop_ranges, rloop_ranges$feature_id)
rloop_density_df <- do.call(rbind, mclapply(rloop_split, function(x){
    message(unique(x$feature_id))
    ret_df <- data.frame(feature_id = unique(x$feature_id),
        rloop_length = sum(width(reduce(x))),
        total_rloop_length = sum(width(x)),
        rloop_frequency = length(x))
    return(ret_df)
},mc.cores = 6))

rloop_density_df$status <- ifelse(rloop_density_df$feature_id %in% feature_ranges_with_repeats$feature_id, "repeat_containing", "non_repeat_containing")

rloop_frequency_df <- as.data.frame(table(rloop_density_df$status))
colnames(rloop_frequency_df) <- c("status", "frequency")
rloop_frequency_df$proportion <- rloop_frequency_df$frequency / table(de_feature_ranges$status)


contingency_matrix <- matrix(data = c(length(de_feature_ranges[de_feature_ranges$status == "repeat_containing" & de_feature_ranges$feature_id %in% rloop_density_df$feature_id]),
    length(de_feature_ranges[de_feature_ranges$status != "repeat_containing" & de_feature_ranges$feature_id %in% rloop_density_df$feature_id]),
    length(de_feature_ranges[de_feature_ranges$status == "repeat_containing" & !(de_feature_ranges$feature_id %in% rloop_density_df$feature_id)]),
    length(de_feature_ranges[de_feature_ranges$status != "repeat_containing" & !(de_feature_ranges$feature_id %in% rloop_density_df$feature_id)])),
    nrow = 2, ncol = 2)

rownames(contingency_matrix) <- c("rci", "non-rci")
colnames(contingency_matrix) <- c("has_rloop", "no_rloop")

#         has_rloop no_rloop
# rci          3320     1728
# non-rci     48828   207186

fisher.test(contingency_matrix)
#         Fisher's Exact Test for Count Data

# data:  contingency_matrix
# p-value < 2.2e-16
# alternative hypothesis: true odds ratio is not equal to 1
# 95 percent confidence interval:
#  7.683461 8.652134
# sample estimates:
# odds ratio
#   8.153039

chisq.test(contingency_matrix)$stdres
#         has_rloop  no_rloop
# rci      82.17554 -82.17554
# non-rci -82.17554  82.17554

```

- Plot Proportion of introns containing repeats


```{r}

Colours <- c("repeat_containing" = "#ff6b6c", "non_repeat_containing" = "#f3f9e3")
ThePlot <- ggplot(rloop_frequency_df, aes(x = status, y = proportion))
ThePlot <- ThePlot + geom_col(aes(fill = status), colour = "#000000")
# ThePlot <- ThePlot + geom_hline(yintercept = log10(100), colour = "#000000",linetype = "dashed")
ThePlot <- ThePlot + scale_x_discrete("Intron group")
ThePlot <- ThePlot + scale_y_continuous("proportion of differentially expressed introns containing rloops")
ThePlot <- ThePlot + scale_fill_manual("Intron groups", values = Colours)
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("proportion of introns containing R-loop")
ThePlot <- ThePlot + theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))
ggsave(width = 5, height = 10, file = file.path(plot_dir, "04-01-proportion-of-introns-containing-rloops-figure-5B.pdf"), ThePlot)

```
