---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

library("here")
message("Project root:", here(), "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("ggplot2")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("final-figures","figure-2")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS(here("final-figures","figure-1","rdata_files","00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds"))
feature_ranges <- readRDS(here("05-simulated-reads-differential-profiling","rdata_files","01-all-merged-exonic-and-intronic-segments-in-genes.rds"))
sig_result_list <- readRDS(here("20-import-significant-events","rdata_files","02-significantly-differentially-expressed-events.rds"))
answerals_object_list <- readRDS(here("39-answerals-complete","rdata_files","01-02-list_of_counts_stat_samplesheet_from_answerals.rds"))
sampleinfo_df <- read.csv(here("39-answerals-complete","input_files","metadata","aals_dataportal_datatable.csv"))

# cell_line_markers_csv <- read.csv("13-01-answerals/input_files/Baxi-et-al-IF-raw-data-file-for-Koustav-Pal-confirmed-data.csv")

participant_info_df <- read.csv("39-answerals-complete","input_files","metadata","aals_participants.csv")
external_controls_df <- read.csv("39-answerals-complete","input_files","metadata","external_controls.csv") 
family_history_log <- read.csv("39-answerals-complete","input_files","metadata","clinical","Family_History_Log.csv")
mutation_df <- read.csv("39-answerals-complete","input_files","metadata","clinical","ALS_Gene_Mutations.csv")
nfl_df <- read.csv("39-answerals-complete","input_files","metadata","NfL","NfL_Seimens_Atellica_Immunoassay_Proteomics_2022_07.csv")

```

- Load repeat and rci object

```{r}

repeat_ranges <- unique(repeat_ranges_list$repeat_ranges)
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges

```

### Analysis


- Generate TPM matrix

```{r}

sampleinfo_df$sample_name <- str_replace_all(sampleinfo_df$participant_id, pattern = "-", replacement = "_")

sig_result_df <- sig_result_list$significant_events

repeat_containing_introns_df <- sig_result_df[sig_result_df$status == "repeat_containing" & 
    sig_result_df$region == "intron",]

repeat_containing_introns_df$cohort <- "public-data"
repeat_containing_introns_df$cohort[grepl(x = repeat_containing_introns_df$study, pattern = "nygc")] <- "nygc"
repeat_containing_introns_df$cohort[grepl(x = repeat_containing_introns_df$study, pattern = "answerals")] <- "answerals"

repeat_containing_introns_df$tissue_type <- "iPSCs"
repeat_containing_introns_df$tissue_type[grepl(x = repeat_containing_introns_df$study, pattern = "nygc")] <- "postmortem"

counts_matrix <- answerals_object_list$counts
patient_df <- answerals_object_list$sample_df
temp_matrix <- counts_matrix / (width(feature_ranges)[match(rownames(counts_matrix), feature_ranges$hash_id)] / 1e3)
tpm_matrix <- t(t(temp_matrix) / (colSums(temp_matrix) / 1e6))

keep_hash_ids <- feature_ranges_with_repeats$hash_id[match(unique(repeat_containing_introns_df$feature_ids), feature_ranges_with_repeats$feature_id)]
keep_tpm_matrix <- tpm_matrix[keep_hash_ids,colnames(tpm_matrix) %in% patient_df$column_id]

repeat_lengths <- feature_ranges_with_repeats$repeat_width[match(rownames(keep_tpm_matrix), feature_ranges_with_repeats$hash_id)]

```

- Generate liability data


```{r}

# threshold for reliable express 
threshold <- 0.1

patient_liability_df <- do.call(rbind, lapply(seq_len(nrow(patient_df)), function(x1){
    temp_df <- patient_df[x1,]
    return_df <- temp_df
    current_columns <- temp_df$column_id
    
    current_tpm_vector <- Reduce("+", lapply(current_columns, function(x){
        keep_tpm_matrix[,x]  
    })) / length(current_columns)

    all_introns <- feature_ranges_with_repeats 
    all_introns$signal <- current_tpm_vector[all_introns$hash_id]
    all_introns$weighted.signal <- all_introns$signal / (table(all_introns$hash_id)[all_introns$hash_id])
    all_introns$weighted_signal_per_expansion <- all_introns$weighted.signal / all_introns$expansion_size
    all_introns <- all_introns[!is.na(all_introns$signal)]

    expected_split <- split(all_introns$weighted.signal, all_introns$span_group)
    total_expected_signal <- vapply(expected_split, sum, 1)
    prop_expected <- total_expected_signal / sum(total_expected_signal)

    return_df$n_of_liability <- length(current_tpm_vector[current_tpm_vector > threshold])
    return_df$repeat_intron_expr_tpm <- sum(current_tpm_vector[current_tpm_vector > threshold])
    return_df$median_repeat_intron_expr_tpm <- median(current_tpm_vector[current_tpm_vector > threshold])
    if(any(is.na(return_df$repeat_intron_expr_tpm))){
        stop(x1)
    }
    return_df$median_repeat_intron_expr_tpm <- median(current_tpm_vector[current_tpm_vector > threshold])

    current_introns <- all_introns[all_introns$hash_id %in% names(current_tpm_vector[current_tpm_vector > threshold])]

    width_split <- split(current_introns$repeat_width, current_introns$span_group)
    total_width <- sum(current_introns$repeat_width)
    prop_width <- vapply(width_split, sum, 1) / total_width

    signal_split <- split(current_introns$weighted.signal, current_introns$span_group)

    observed_proportion <- vapply(c("trinucleotide", "quadnucleotide", "pentanucleotide", "hexanucleotide"), function(x){
        if(x %in% names(signal_split)){
            return(sum(signal_split[[x]]) / sum(current_introns$weighted.signal))
        }else{
            return(0)
        }
    },1)
    enrichment <- observed_proportion / prop_expected[names(observed_proportion)]

    overall_mean_signal <- mean(current_introns$weighted.signal)
    mean_weighted_signal <- vapply(c("trinucleotide", "quadnucleotide", "pentanucleotide", "hexanucleotide"), function(x){
        if(x %in% current_introns$span_group){
            return(mean(current_introns$weighted.signal[current_introns$span_group == x]))
        }else{
            return(0)
        }
    },1)

    share_of_signal <- vapply(c("trinucleotide", "quadnucleotide", "pentanucleotide", "hexanucleotide"), function(x){
        if(x %in% current_introns$span_group){
            prop <- sum(width(current_introns[current_introns$span_group == x])) / sum(width(current_introns))
            return(prop  * (mean_weighted_signal[x] / overall_mean_signal))
        }else{
            return(0)
        }
    },1)

    within_group_theil <- vapply(c("trinucleotide", "quadnucleotide", "pentanucleotide", "hexanucleotide"), function(x){
        if(x %in% current_introns$span_group){
            current_signal <- current_introns$weighted.signal[current_introns$span_group == x]
            inner_ratio <- current_signal / mean_weighted_signal[x]
            return(sum(inner_ratio * log(inner_ratio)))
        }else{
            return(0)
        }
    },1)

    overall_ratios <- vapply(c("trinucleotide", "quadnucleotide", "pentanucleotide", "hexanucleotide"), function(x){
        if(x %in% current_introns$span_group){
            return(mean_weighted_signal[x] / overall_mean_signal)
        }else{
            return(0)
        }
    },1)
    all_between_repeat_component <- log(overall_ratios) * share_of_signal
    between_repeat_component <- sum(all_between_repeat_component)


    theil_t <- sum(share_of_signal * within_group_theil) + between_repeat_component

    names(enrichment) <- paste(names(enrichment), "enrichment", sep = "_")
    names(observed_proportion) <- paste(names(observed_proportion), "observed", sep = "_")
    names(all_between_repeat_component) <- paste(names(all_between_repeat_component), "between_theil", sep = "_")

    new_return_df <- return_df
    new_return_df$entropy <- theil_t
    return(cbind(new_return_df[1,], t(as.matrix(enrichment)), t(as.matrix(observed_proportion)), t(as.matrix(all_between_repeat_component))))
}))

```


- Attach metadata

```{r}

patient_liability_df$Participant_ID <- vapply(str_split(patient_liability_df$sample_name, pattern = "-"), function(x){
    paste(x[c(1,2)], collapse = "-")
},"1")

sample_name_matches <- match(patient_liability_df$Participant_ID, sampleinfo_df$Participant_ID)

patient_liability_df$di_m_ns_line_name <- sampleinfo_df$diMNs_line[sample_name_matches]


patient_liability_df[,"SMI32"] <- sampleinfo_df[sample_name_matches,"PCT_SMI32"]
patient_liability_df[,"ISL1"] <- sampleinfo_df[sample_name_matches,"PCT_ISL1"]
patient_liability_df[,"NKX6.1"] <- sampleinfo_df[sample_name_matches,"PCT_NKX61"]
patient_liability_df[,"TUJ1"] <- sampleinfo_df[sample_name_matches,"PCT_TUJ1"]
patient_liability_df[,"S100b"] <- sampleinfo_df[sample_name_matches,"PCT_S100b"]
patient_liability_df$nygc_cgnd <- sampleinfo_df$NYGC_CGND_ID[sample_name_matches]
patient_liability_df$progression <- sampleinfo_df$ALSFRS_R_PROGRESSION_SLOPE[sample_name_matches]
patient_liability_df$age_onset <- sampleinfo_df$AGE_AT_SYMPTOM_ONSET[sample_name_matches]
patient_liability_df$age_at_collection <- sampleinfo_df$Age_at_First_PBMC_Collection[sample_name_matches]
patient_liability_df$age_at_collection[is.na(patient_liability_df$age_at_collection)] <- external_controls_df[match(patient_liability_df$Participant_ID[is.na(patient_liability_df$age_at_collection)], external_controls_df$Participant_ID),"Age_Sample_Collection_Cedars"]
patient_liability_df$Number_of_Visits <- sampleinfo_df$Number_of_Visits[sample_name_matches]
# patient_liability_df$early_late_onset <- sampleinfo_df$early_late_onset[sample_name_matches]
# patient_liability_df$early_late_death <- sampleinfo_df$early_late_death[sample_name_matches]
patient_liability_df$family_history <- family_history_log$famrel[match(patient_liability_df$Participant_ID, family_history_log$Participant_ID)]
patient_liability_df$mnd_group <- participant_info_df$Cohort[match(patient_liability_df$Participant_ID, participant_info_df$Participant_ID)]
patient_liability_df$mnd_group[grepl(x = patient_liability_df$Participant_ID, "CTRL")] <- "Healthy Control"
patient_liability_df$mnd_group[patient_liability_df$mnd_group == "Healthy Control"] <- "healthy"
patient_liability_df$mnd_group[patient_liability_df$mnd_group == "ALS"] <- "als"
patient_liability_df$mnd_group[patient_liability_df$mnd_group == "Non-ALS MND"] <- "other_mnd"
patient_liability_df$mnd_group[patient_liability_df$mnd_group == "Asymptomatic ALS Gene carrier"] <- "carrier"
patient_liability_df$mutation <- vapply(patient_liability_df$Participant_ID, function(x){
    mutations <- c("ang", "c9orf72", "fus", "mutot", "progran", "setx", "sod1", "tau", "tdp43", "vapb", "vcp")
    current_mutation_df <- mutation_df[mutation_df$Participant_ID == x,mutations]
    non_na <- colnames(current_mutation_df)[!is.na(current_mutation_df) & current_mutation_df != 2]
    if(length(non_na) > 0){
        return(paste(non_na, collapse = "|"))
    }
    return("NA")
},"1")

patient_liability_df$tested_for_c9orf72_repeat_expansion <- sampleinfo_df[match(patient_liability_df$Participant_ID, sampleinfo_df$Participant_ID),"EH_C9orf72"] == "Yes"
patient_liability_df$c9orf72_repeat_length <- sampleinfo_df[match(patient_liability_df$Participant_ID, sampleinfo_df$Participant_ID),"C9orf72_repeat_length"]

patient_liability_df$tested_for_atxn2_repeat_expansion <- sampleinfo_df[match(patient_liability_df$Participant_ID, sampleinfo_df$Participant_ID),"EH_ATXN2"] == "Yes"
patient_liability_df$atxn2_repeat_length <- sampleinfo_df[match(patient_liability_df$Participant_ID, sampleinfo_df$Participant_ID),"ATXN2_repeat_length"]

patient_liability_df$expansion_group <- "Not tested"
patient_liability_df$expansion_group[(patient_liability_df$tested_for_c9orf72_repeat_expansion | patient_liability_df$tested_for_atxn2_repeat_expansion) & 
    (patient_liability_df$c9orf72_repeat_length >= 24 | patient_liability_df$atxn2_repeat_length >= 24) & 
    !is.na(patient_liability_df$c9orf72_repeat_length) & !is.na(patient_liability_df$atxn2_repeat_length)] <- "Tested and contains a C9ORF72 or ATXN2 repeat length than 24"
patient_liability_df$expansion_group[(patient_liability_df$tested_for_c9orf72_repeat_expansion | patient_liability_df$tested_for_atxn2_repeat_expansion) & 
    (patient_liability_df$c9orf72_repeat_length < 24 & patient_liability_df$atxn2_repeat_length < 24) & 
    !is.na(patient_liability_df$c9orf72_repeat_length) & !is.na(patient_liability_df$atxn2_repeat_length)] <- "Tested and lacks a C9ORF72 or ATXN2 pathological repeat"

```

- Save AnswerALS liability data

```{r}

saveRDS(object = patient_liability_df, file = file.path(save_dir, "02-save-answerals-rci-expression-summary.rds"))

```
