---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-2"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
feature_ranges <- readRDS("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds")
sig_result_list <- readRDS("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds")
answerals_object_list <- readRDS("39-answerals-complete/rdata_files/01-02-list_of_counts_stat_samplesheet_from_answerals.rds")
sampleinfo_df <- read.csv("39-answerals-complete/input_files/metadata/aals_dataportal_datatable.csv")
cell_line_markers_csv <- read.csv("13-01-answerals/input_files/Baxi-et-al-IF-raw-data-file-for-Koustav-Pal-confirmed-data.csv")
patient_liability_df <- readRDS(file.path(save_dir, "02-save-answerals-rci-expression-summary.rds"))

participant_info_df <- read.csv("39-answerals-complete/input_files/metadata/aals_participants.csv")
external_controls_df <- read.csv("39-answerals-complete/input_files/metadata/external_controls.csv") 
family_history_log <- read.csv("39-answerals-complete/input_files/metadata/clinical/Family_History_Log.csv")
mutation_df <- read.csv("39-answerals-complete/input_files/metadata/clinical/ALS_Gene_Mutations.csv")
nfl_df <- read.csv("39-answerals-complete/input_files/metadata/NfL/NfL_Seimens_Atellica_Immunoassay_Proteomics_2022_07.csv")

```

- Load repeat and rci object

```{r}

repeat_ranges <- unique(repeat_ranges_list$repeat_ranges)
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges

```

### Analysis


- Generate TPM matrix

```{r}

# sampleinfo_df$sample_name <- str_replace_all(sampleinfo_df$participant_id, pattern = "-", replacement = "_")

sig_result_df <- sig_result_list$significant_events

repeat_containing_introns_df <- sig_result_df[sig_result_df$status == "repeat_containing" & 
    sig_result_df$region == "intron",]

repeat_containing_introns_df$cohort <- "public-data"
repeat_containing_introns_df$cohort[grepl(x = repeat_containing_introns_df$study, pattern = "nygc")] <- "nygc"
repeat_containing_introns_df$cohort[grepl(x = repeat_containing_introns_df$study, pattern = "answerals")] <- "answerals"

repeat_containing_introns_df$tissue_type <- "iPSCs"
repeat_containing_introns_df$tissue_type[grepl(x = repeat_containing_introns_df$study, pattern = "nygc")] <- "postmortem"

counts_matrix <- answerals_object_list$counts
temp_matrix <- counts_matrix / (width(feature_ranges)[match(rownames(counts_matrix), feature_ranges$hash_id)] / 1e3)
tpm_matrix <- t(t(temp_matrix) / (colSums(temp_matrix) / 1e6))

keep_hash_ids <- feature_ranges_with_repeats$hash_id[match(unique(repeat_containing_introns_df$feature_ids), feature_ranges_with_repeats$feature_id)]
keep_tpm_matrix <- tpm_matrix[keep_hash_ids,colnames(tpm_matrix) %in% patient_liability_df$column_id]

repeat_lengths <- feature_ranges_with_repeats$repeat_width[match(rownames(keep_tpm_matrix), feature_ranges_with_repeats$hash_id)]

```


- Generate penetrant RCIs


```{r}

# threshold for reliable express 
threshold <- 0.1
patient_penetrance_matrix <- do.call(rbind, lapply(seq_len(nrow(patient_liability_df)), function(x1){
    temp_df <- patient_liability_df[x1,]
    return_df <- temp_df
    current_columns <- temp_df$column_id
    
    current_tpm_vector <- Reduce("+", lapply(current_columns, function(x){
        keep_tpm_matrix[,x]  
    })) / length(current_columns)

    a_vector <- rep(0, length(unique(feature_ranges_with_repeats$hash_id)))
    names(a_vector) <- unique(feature_ranges_with_repeats$hash_id)
    a_vector[names(current_tpm_vector[current_tpm_vector > threshold])] <- 1
    return(a_vector)
}))
patient_liability_df$penetrance_row <- seq_len(nrow(patient_penetrance_matrix))

```

- Save AnswerALS liability data
 
```{r}

saveRDS(object = list(liability_df = patient_liability_df, penetrance_matrix = patient_penetrance_matrix), file = file.path(save_dir, "03-save-answerals-rci-expression-summary-with-penetrance-matrix.rds"))

```
