---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

library("here")
message("Project root:", here(), "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("ggplot2")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "39-answerals-complete"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

liability_with_residuals_list <- readRDS(file.path(save_dir, "02-01-modelled-liability-df-with-residuals.rds"))
liability_df <- liability_with_residuals_list$liability
residuals_df <- liability_with_residuals_list$residuals
residuals_df$subject_id <- liability_df$subject_id[match(residuals_df$Participant_ID, liability_df$Participant_ID)]
participant_info_df <- read.csv(here("39-answerals-complete","input_files","metadata","aals_participants.csv"))
external_controls_df <- read.csv(here("39-answerals-complete","input_files","metadata","external_controls.csv"))
nfl_df <- read.csv(here("39-answerals-complete","input_files","metadata","NfL","NfL_Seimens_Atellica_Immunoassay_Proteomics_2022_07.csv"))
sampleinfo_df <- read.csv("39-answerals-complete","input_files","metadata","aals_dataportal_datatable.csv")

```

- select the nfl value at the lowest timepoint

```{r}

nfl_values <- vapply(split(nfl_df, nfl_df$SubjID), function(x){
    non_nas <- x$Result.Value[!is.na(x$Result.Value)]
    ret_val <- as.numeric(non_nas[which.min(non_nas)])
    if(length(ret_val) == 0){
        return(NA)
    }
    return(ret_val)
},1)

liability_df$nfl <- nfl_values[liability_df$subject_id]
liability_df$sex <- participant_info_df$Sex[match(liability_df$Participant_ID, participant_info_df$Participant_ID)]
liability_df$sex[is.na(liability_df$sex)] <- external_controls_df$SEX[match(liability_df$Participant_ID[is.na(liability_df$sex)], external_controls_df$Participant_ID)]
residuals_df$nfl <- nfl_values[residuals_df$subject_id]
liability_df <- liability_df[liability_df$mnd_group %in% c("healthy", "als"),]

```

- Investigate association between nfl and median expression of als spinal cord specific rcis

```{r}

wilcox.test(liability_df[liability_df$mnd_group %in% c("als"),"intron_expression_rate"],
    liability_df[liability_df$mnd_group %in% c("healthy"),"intron_expression_rate"])
# W = 30679, p-value = 0.009144

fetch_residuals <- function(liability_df, var_x, var_y = "nfl", group_name, return_model = F, return_summary = F){
    liability_df$var_x <- liability_df[,var_x]
    liability_df$var_y <- liability_df[,var_y]
    a_model <- lm(log10(var_y) ~ age_at_collection + sex + mnd_group + var_x:mnd_group, data = liability_df)
    if(return_model){
        return(a_model)
    }
    if(return_summary){
        return(summary(a_model))
    }
    residual_df <- data.frame(group = group_name,
    Participant_ID = liability_df$Participant_ID,
    mnd_group = liability_df$mnd_group,
    rci_liability = liability_df$var_x,
    median_expr = liability_df$var_y,
    fitted_values = a_model$fitted.values, 
    residuals = a_model$residuals)
    return(residual_df)
}

cor(liability_df[liability_df$mnd_group %in% c("healthy", "als") & 
    !is.na(liability_df$nfl),"nfl"], 
    liability_df[liability_df$mnd_group %in% c("healthy", "als") & 
    !is.na(liability_df$nfl),"repeat_intron_expr_tpm"])
# -0.05666871

cor(liability_df[liability_df$mnd_group %in% c("healthy", "als") & 
    !is.na(liability_df$nfl),"nfl"], 
    liability_df[liability_df$mnd_group %in% c("healthy", "als") & 
    !is.na(liability_df$nfl),"intron_expression_rate"])
# 0.05149547

cor.test(liability_df[liability_df$mnd_group %in% c("healthy", "als"),"repeat_intron_expr_tpm"], 
    liability_df[liability_df$mnd_group %in% c("healthy", "als"),"intron_expression_rate"])
# -0.6560285



cor.test(log10(liability_df$intron_expression_rate), as.numeric(liability_df$mnd_group))
#         Pearson's product-moment correlation

# data:  log10(liability_df$intron_expression_rate) and as.numeric(liability_df$mnd_group)
# t = -2.2265, df = 591, p-value = 0.02636
# alternative hypothesis: true correlation is not equal to 0
# 95 percent confidence interval:
#  -0.17046643 -0.01076571
# sample estimates:
#         cor
# -0.09120243

```

Investigate correlation between expansion length and rci emission rate

```{r}

# summary(lm(data = liability_df[liability_df$mnd_group %in% c("healthy", "als"),], formula = log10(nfl) ~ age_at_collection + sex + mnd_group + log10(intron_expression_rate):mnd_group))
cor.test(liability_df$c9orf72_repeat_length[liability_df$expansion_group == "Tested and contains a C9ORF72 or ATXN2 repeat length than 24"], liability_df$intron_expression_rate[liability_df$expansion_group == "Tested and contains a C9ORF72 or ATXN2 repeat length than 24"])
#         Pearson's product-moment correlation

# data:  liability_df$c9orf72_repeat_length[liability_df$expansion_group == "Tested and contains a C9ORF72 or ATXN2 repeat length than 24"] and liability_df$intron_expression_rate[liability_df$expansion_group == "Tested and contains a C9ORF72 or ATXN2 repeat length than 24"]
# t = 1.0788, df = 48, p-value = 0.2861
# alternative hypothesis: true correlation is not equal to 0
# 95 percent confidence interval:
#  -0.1300555  0.4144597
# sample estimates:
#       cor
# 0.1538622

cor.test(liability_df$atxn2_repeat_length[liability_df$expansion_group == "Tested and contains a C9ORF72 or ATXN2 repeat length than 24"], liability_df$intron_expression_rate[liability_df$expansion_group == "Tested and contains a C9ORF72 or ATXN2 repeat length than 24"])
#         Pearson's product-moment correlation

# data:  liability_df$atxn2_repeat_length[liability_df$expansion_group == "Tested and contains a C9ORF72 or ATXN2 repeat length than 24"] and liability_df$intron_expression_rate[liability_df$expansion_group == "Tested and contains a C9ORF72 or ATXN2 repeat length than 24"]
# t = -1.2325, df = 48, p-value = 0.2238
# alternative hypothesis: true correlation is not equal to 0
# 95 percent confidence interval:
#  -0.4324111  0.1084930
# sample estimates:
#        cor
# -0.1751438

# 0.009168731

```

- Prepare model 

```{r}

model_1 <- lm(data = liability_df[liability_df$mnd_group %in% c("healthy", "als") & !is.na(liability_df$nfl),], formula = log10(nfl) ~ age_at_collection + sex + mnd_group + intron_expression_rate:mnd_group)
summary(model_1)
# summary(model_1)
# Call:
# lm(formula = log10(nfl) ~ age_at_collection + sex + mnd_group +
#     intron_expression_rate:mnd_group, data = liability_df[liability_df$mnd_group %in%
#     c("healthy", "als") & !is.na(liability_df$nfl), ])

# Residuals:
#      Min       1Q   Median       3Q      Max
# -0.80871 -0.17266  0.01443  0.18415  0.79297

# Coefficients:
#                                           Estimate Std. Error t value Pr(>|t|)
# (Intercept)                              7.163e-01  2.647e-01   2.706  0.00708
# age_at_collection                        6.300e-03  1.267e-03   4.973 9.46e-07
# sexMale                                 -9.057e-02  2.905e-02  -3.118  0.00194
# mnd_groupals                             4.418e-01  2.678e-01   1.650  0.09968
# mnd_grouphealthy:intron_expression_rate -6.668e-07  6.348e-05  -0.011  0.99162
# mnd_groupals:intron_expression_rate      4.453e-05  2.481e-05   1.795  0.07339

# (Intercept)                             **
# age_at_collection                       ***
# sexMale                                 **
# mnd_groupals                            .
# mnd_grouphealthy:intron_expression_rate
# mnd_groupals:intron_expression_rate     .
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

# Residual standard error: 0.29 on 440 degrees of freedom
#   (13 observations deleted due to missingness)
# Multiple R-squared:  0.402,     Adjusted R-squared:  0.3952
# F-statistic: 59.16 on 5 and 440 DF,  p-value: < 2.2e-16

require(car)
# > vif(model_1, type = "predictor")
# GVIFs computed for predictors
#                            GVIF Df GVIF^(1/(2*Df))         Interacts With
# age_at_collection      1.058812  1        1.028986                   --
# sex                    1.099720  1        1.048675                   --
# mnd_group              1.132753  3        1.020993 intron_expression_rate
# intron_expression_rate 1.132753  3        1.020993 intron_expression_rate
#                                                            Other Predictors
# age_at_collection                    sex, mnd_group, intron_expression_rate
# sex                    age_at_collection, mnd_group, intron_expression_rate
# mnd_group                                            age_at_collection, sex
# intron_expression_rate                               age_at_collection, sex

```

- Plot observed nfl values for each covariate

```{r}

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))


selected_liability_df <- liability_df[liability_df$mnd_group %in% c("healthy", "als") & !is.na(liability_df$nfl),]

colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
ThePlot <- ggplot(selected_liability_df, aes(x = age_at_collection, y = nfl))
ThePlot <- ThePlot + geom_point(aes(fill = mnd_group), colour = "#000000", shape = 21)
ThePlot <- ThePlot + geom_smooth(method = "lm", aes(colour = mnd_group), formula = y ~ x)
ThePlot <- ThePlot + scale_x_continuous("Age at collection")
ThePlot <- ThePlot + scale_y_continuous("Observed NFL levels in serum")
ThePlot <- ThePlot + scale_colour_manual("MND status", values = colours)
ThePlot <- ThePlot + scale_fill_manual("MND status", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_wrap(. ~ mnd_group, scales = "free", nrow = 1, ncol = 2)
ThePlot <- ThePlot + ggtitle("sex vs nfl")
ThePlot <- ThePlot + current_theme
ggsave(width = 10, height = 5, file = file.path(plot_dir, "03-03-01-plot-age-at-collection-vs-observed-nfl-in-healthy-and-ALS.pdf"), ThePlot)

require(ggbeeswarm)
colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
ThePlot <- ggplot(selected_liability_df, aes(x = mnd_group, y = nfl))
# ThePlot <- ThePlot + geom_density(aes(fill = sex), alpha = 0.4, colour = "#000000")
ThePlot <- ThePlot + geom_quasirandom(aes(fill = mnd_group), shape = 21, colour = "#000000")
ThePlot <- ThePlot + geom_boxplot(aes(fill = mnd_group), colour = "#000000", alpha = 0.6,
    width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9), outlier.shape = NA)
ThePlot <- ThePlot + scale_x_discrete("MND group")
ThePlot <- ThePlot + scale_y_continuous("Observed NFL levels in serum")
ThePlot <- ThePlot + scale_colour_manual("MND status", values = colours)
ThePlot <- ThePlot + scale_fill_manual("MND status", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(. ~ mnd_group)
ThePlot <- ThePlot + ggtitle("sex vs nfl")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 10, file = file.path(plot_dir, "03-03-02-plot-mnd-group-vs-observed-nfl-in-healthy-and-ALS.pdf"), ThePlot)


require(ggbeeswarm)
colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
ThePlot <- ggplot(selected_liability_df, aes(x = sex, y = nfl))
# ThePlot <- ThePlot + geom_density(aes(fill = sex), alpha = 0.4, colour = "#000000")
ThePlot <- ThePlot + geom_quasirandom(aes(fill = mnd_group), dodge.width = 0.9, shape = 21, colour = "#000000")
ThePlot <- ThePlot + geom_boxplot(aes(fill = mnd_group), colour = "#000000", alpha = 0.6,
    width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9), outlier.shape = NA)
ThePlot <- ThePlot + scale_x_discrete("Sex")
ThePlot <- ThePlot + scale_y_continuous("Observed NFL levels in serum")
ThePlot <- ThePlot + scale_colour_manual("MND status", values = colours)
ThePlot <- ThePlot + scale_fill_manual("MND status", values = colours)
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("sex vs nfl")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 10, file = file.path(plot_dir, "03-03-03-plot-sex-vs-observed-nfl-in-healthy-and-ALS.pdf"), ThePlot)


require(ggbeeswarm)
colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
ThePlot <- ggplot(selected_liability_df, aes(x = log10(intron_expression_rate), y = log10(nfl)))
ThePlot <- ThePlot + geom_point(aes(fill = mnd_group), colour = "#000000", shape = 21)
ThePlot <- ThePlot + geom_smooth(method = "lm", aes(colour = mnd_group), formula = y ~ x)
ThePlot <- ThePlot + scale_x_continuous("RCI emission rate")
ThePlot <- ThePlot + scale_y_continuous("Observed NFL levels in serum")
ThePlot <- ThePlot + scale_colour_manual("MND status", values = colours)
ThePlot <- ThePlot + scale_fill_manual("MND status", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_wrap(. ~ mnd_group, scales = "free", nrow = 1, ncol = 2)
ThePlot <- ThePlot + ggtitle("sex vs nfl")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 7, file = file.path(plot_dir, "03-03-04-plot-intron-expression-rate-vs-observed-nfl-in-healthy-and-ALS.pdf"), ThePlot)


```


- Define function for predicting the model 

```{r}

predict_nfl_values <- function(pred_df, column_var, a_model){
    column_values <- quantile(pred_df[,column_var], seq(0,1,0.01))
    prediction_list <- lapply(seq_along(column_values), function(x){
        observed_nfl <- pred_df$nfl
        observed_values <- pred_df[,column_var]
        observed_mnd_group <- pred_df$mnd_group
        # pred_df$mnd_group <- mnd_status
        pred_df[,column_var] <- column_values[x]
        current_prediction <- predict(a_model, pred_df)
        predicted_nfl <- 10^current_prediction
        summary_stats <- data.frame(
            # mnd_group = mnd_status,
            threshold_name = names(column_values)[x],
            threshold = column_values[x],
            Mean_NFL = mean(predicted_nfl, na.rm = T),
            Median_NFL = median(predicted_nfl, na.rm = T),
            SD_NFL = sd(predicted_nfl, na.rm = T))
        pred_data <- data.frame(
            subject_id = pred_df$subject_id, 
            mnd_group = pred_df$mnd_group,
            threshold_name = names(column_values)[x],
            threshold = column_values[x],
            predicted_nfl = predicted_nfl,
            observed_nfl = observed_nfl,
            current_column = observed_values)
        colnames(pred_data) <- c(colnames(pred_data)[-ncol(pred_data)], column_var)
        return(list(summary = summary_stats, prediction = pred_data))
    })
    summary_df <- do.call(rbind, lapply(prediction_list, function(x){
        x$summary
    }))
    prediction_df <- do.call(rbind, lapply(prediction_list, function(x){
        x$prediction
    }))
    prediction_df$distance <- sqrt((prediction_df$observed_nfl - prediction_df$predicted_nfl)^2 + (prediction_df[,column_var] - prediction_df$threshold)^2)
    sample_wise_min_dist_df <- do.call(rbind, lapply(split(prediction_df, prediction_df$subject_id), function(x){
        x[which.min(x$distance),]
    }))
    return(list(summary = summary_df, prediction = prediction_df, best = sample_wise_min_dist_df))
}

```

- Predict the effect of intron expression rate on nfl

```{r}

selected_liability_df <- liability_df[liability_df$mnd_group %in% c("healthy", "als") & !is.na(liability_df$nfl),]
pred_df <- selected_liability_df

prediction_list <- predict_nfl_values(pred_df, 
    column_var = "intron_expression_rate",
    a_model = model_1)
best_rci_emission_prediction_df <- prediction_list$best

```

- Plot the association between nfl and intron expression rate

```{r}

require(ggbeeswarm)
colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
ThePlot <- ggplot(best_rci_emission_prediction_df, aes(x = log10(threshold), y = log10(predicted_nfl)))
ThePlot <- ThePlot + geom_point(aes(fill = mnd_group), colour = "#000000", shape = 21)
ThePlot <- ThePlot + geom_smooth(method = "lm", aes(colour = mnd_group), formula = y ~ x)
ThePlot <- ThePlot + scale_x_continuous("RCI emission rate")
ThePlot <- ThePlot + scale_y_continuous("Predicted NFL levels by MND status (after adjusting for covariates)")
ThePlot <- ThePlot + scale_colour_manual("MND status", values = colours)
ThePlot <- ThePlot + scale_fill_manual("MND status", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_wrap(. ~ mnd_group, scales = "free", nrow = 1, ncol = 2)
ThePlot <- ThePlot + ggtitle("sex vs nfl")
ThePlot <- ThePlot + current_theme
ggsave(width = 10, height = 5, file = file.path(plot_dir, "03-02-plot-intron-expression-rate-vs-nfl-in-healthy-and-ALS.pdf"), ThePlot)

```

- Plot predicted NFL level vs age at collection

```{r}


selected_liability_df <- liability_df[liability_df$mnd_group %in% c("healthy", "als") & 
    !is.na(liability_df$nfl) & !is.na(liability_df$age_at_collection),]
pred_df <- selected_liability_df

prediction_list <- predict_nfl_values(pred_df, 
    column_var = "age_at_collection",
    a_model = model_1)
best_age_based_prediction_df <- prediction_list$best

require(ggbeeswarm)
colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
ThePlot <- ggplot(best_age_based_prediction_df, aes(x = age_at_collection, y = predicted_nfl))
ThePlot <- ThePlot + geom_point(aes(fill = mnd_group), colour = "#000000", shape = 21)
ThePlot <- ThePlot + geom_smooth(method = "lm", aes(colour = mnd_group), formula = y ~ x)
ThePlot <- ThePlot + scale_x_continuous("Age at collection")
ThePlot <- ThePlot + scale_y_continuous("Predicted NFL levels by MND status (after adjusting for covariates)")
ThePlot <- ThePlot + scale_colour_manual("MND status", values = colours)
ThePlot <- ThePlot + scale_fill_manual("MND status", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_wrap(. ~ mnd_group, scales = "free", nrow = 1, ncol = 2)
ThePlot <- ThePlot + ggtitle("sex vs nfl")
ThePlot <- ThePlot + current_theme
ggsave(width = 10, height = 5, file = file.path(plot_dir, "03-03-plot-age-at-collection-vs-nfl-in-healthy-and-ALS.pdf"), ThePlot)


```

- Plot predicted NFL level by MND group 

```{r}

selected_liability_df <- liability_df[liability_df$mnd_group %in% c("healthy", "als") & 
    !is.na(liability_df$nfl) & !is.na(liability_df$age_at_collection),]
pred_df <- selected_liability_df
pred_df$mnd_group <- "healthy"
pred_healthy <- predict(model_1, pred_df)

pred_df$mnd_group <- "als"
pred_als <- predict(model_1, pred_df)


nfl_healthy <- 10^pred_healthy
nfl_als <- 10^pred_als

summary_stats <- data.frame(
   Group = c("healthy", "als"),
   Mean_NFL = c(mean(nfl_healthy), mean(nfl_als)),
   Median_NFL = c(median(nfl_healthy), median(nfl_als)),
   SD_NFL = c(sd(nfl_healthy), sd(nfl_als)))

pred_data <- data.frame(
  mnd_group = rep(c("healthy", "als"), each = length(nfl_healthy)),
  predicted_nfl = c(nfl_healthy, nfl_als))

pred_data$mnd_group <- factor(pred_data$mnd_group, levels = c("healthy", "als"))

require(ggbeeswarm)
colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
ThePlot <- ggplot(pred_data, aes(x = mnd_group, y = predicted_nfl))
# ThePlot <- ThePlot + geom_density(aes(fill = sex), alpha = 0.4, colour = "#000000")
ThePlot <- ThePlot + geom_quasirandom(aes(fill = mnd_group), shape = 21, colour = "#000000")
ThePlot <- ThePlot + geom_boxplot(aes(fill = mnd_group), colour = "#000000", alpha = 0.6,
    width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9), outlier.shape = NA)
ThePlot <- ThePlot + scale_x_discrete("MND group")
ThePlot <- ThePlot + scale_y_continuous("Predicted NFL levels by MND status (after adjusting for covariates)")
ThePlot <- ThePlot + scale_colour_manual("MND status", values = colours)
ThePlot <- ThePlot + scale_fill_manual("MND status", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(. ~ mnd_group)
ThePlot <- ThePlot + ggtitle("sex vs nfl")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 10, file = file.path(plot_dir, "03-04-plot-mnd-group-vs-nfl-in-healthy-and-ALS.pdf"), ThePlot)

```

- Plot predicted NFL by MND group for each sex

```{r}

selected_liability_df <- liability_df[!is.na(liability_df$nfl) & !is.na(liability_df$age_at_collection),]
pred_df <- selected_liability_df
pred_df$sex <- "Male"
pred_male <- predict(model_1, pred_df)

pred_df$sex <- "Female"
pred_female <- predict(model_1, pred_df)

nfl_male <- 10^pred_male
nfl_female <- 10^pred_female

summary_stats <- data.frame(
   Sex = c("Male", "Female"),
   Mean_NFL = c(mean(nfl_male), mean(nfl_female)),
   Median_NFL = c(median(nfl_male), median(nfl_female)),
   SD_NFL = c(sd(nfl_male), sd(nfl_female)))

pred_data <- data.frame(
  sex = rep(c("Male", "Female"), each = length(nfl_male)),
  mnd_group = rep(pred_df$mnd_group, times = 2),
  predicted_nfl = c(nfl_male, nfl_female))

pred_data$sex <- factor(pred_data$sex, levels = c("Male", "Female"))

require(ggbeeswarm)
colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
ThePlot <- ggplot(pred_data, aes(x = sex, y = predicted_nfl))
# ThePlot <- ThePlot + geom_density(aes(fill = sex), alpha = 0.4, colour = "#000000")
ThePlot <- ThePlot + geom_quasirandom(aes(fill = mnd_group), dodge.width = 0.9, shape = 21, colour = "#000000")
ThePlot <- ThePlot + geom_boxplot(aes(fill = mnd_group), colour = "#000000", alpha = 0.6,
    width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9), outlier.shape = NA)
ThePlot <- ThePlot + scale_x_discrete("Sex")
ThePlot <- ThePlot + scale_y_continuous("Predicted NFL levels (after adjusting for covariates)")
ThePlot <- ThePlot + scale_colour_manual("MND status", values = colours)
ThePlot <- ThePlot + scale_fill_manual("MND status", values = colours)
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("sex vs nfl")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 10, file = file.path(plot_dir, "03-04-plot-sex-vs-nfl-in-healthy-and-ALS.pdf"), ThePlot)

```

- save object

```{r}

saveRDS(object = model_1, file = file.path(save_dir, "02-02-save-model-for-predicting-nfl-levels-based-on-intron-expression-rate-without-expansion-groups.rds"))

```

