---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

library("here")
message("Project root:", here(), "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("ggplot2")
require("ggbeeswarm")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("final-figures","figure-2")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```


### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS(here("final-figures","figure-1","rdata_files","00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds"))
feature_ranges <- readRDS(here("05-simulated-reads-differential-profiling","rdata_files","01-all-merged-exonic-and-intronic-segments-in-genes.rds"))
patient_liability_df <- readRDS(here("final-figures","figure-2","rdata_files","01-save-nygc-rci-liabilities-in-patient-samples.rds"))
non_rci_patient_liability_df <- readRDS(here("final-figures","figure-2","rdata_files","01-save-nygc-non-rci-liabilities-in-patient-samples.rds"))
patient_liability_df <- cbind(patient_liability_df, non_repeat_intron_expr_tpm = non_rci_patient_liability_df$non_repeat_intron_expr_tpm)

```

### Analysis

- Summarise liability values by each patient 

```{r}

mean(patient_liability_df[patient_liability_df$tissue_group == "brain" & patient_liability_df$condition %in% c("als", "ctrl"),"repeat_intron_expr_tpm"])
# 265.6693
sd(patient_liability_df[patient_liability_df$tissue_group == "brain" & patient_liability_df$condition %in% c("als", "ctrl"),"repeat_intron_expr_tpm"])
# 95.8088

mean(patient_liability_df[patient_liability_df$tissue_group == "spinal_cord" & patient_liability_df$condition %in% c("als", "ctrl"),"repeat_intron_expr_tpm"])
# 246.2109
sd(patient_liability_df[patient_liability_df$tissue_group == "spinal_cord" & patient_liability_df$condition %in% c("als", "ctrl"),"repeat_intron_expr_tpm"])
# 46.92009


summary_df <- do.call(rbind, lapply(split(patient_liability_df, paste(patient_liability_df$subject_id, patient_liability_df$tissue_group)), function(temp_df){
    return_df <- unique(temp_df[,c("subject_id", "tissue_group", "condition")])
    return_df$repeat_intron_expr_tpm <- mean(temp_df$repeat_intron_expr_tpm)
    return_df$non_repeat_intron_expr_tpm <- mean(temp_df$non_repeat_intron_expr_tpm)
    return(return_df)
}))

table(unique(summary_df[,c("subject_id", "condition")])$condition)


```

- Plot liability values

```{r}

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

colours <- c("ctrl" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
fill_colours <- c("ctrl" = "#F6F4F3", 
    "c9orf72" = "#169873", "sporadic" = "#F49FBC",
    "tardbp" = "#FFD3BA", "mutation" = "#9EBD6E",
    "none" = "#805D93")

summary_df$condition <- factor(summary_df$condition, levels = c("ctrl", "als", "als+ftd"))

ThePlot <- ggplot(summary_df[summary_df$condition != "als+ftd",], 
    aes(x = condition, y = repeat_intron_expr_tpm))
ThePlot <- ThePlot + geom_violin(aes(fill = condition), colour = "#000000")
ThePlot <- ThePlot + geom_boxplot(aes(fill = condition), colour = "#000000", 
    width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9))
# ThePlot <- ThePlot + geom_smooth(method = "lm", formula = y ~ x, aes(colour = als_group))
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_discrete("Anatomical region")
ThePlot <- ThePlot + scale_y_continuous("RCI liability score")
# ThePlot <- ThePlot + scale_colour_manual("Condition colour", values = colours)
ThePlot <- ThePlot + scale_fill_manual("condition", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_wrap(. ~ tissue_group, scales = "free")
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression across anatomical regions")
ThePlot <- ThePlot + current_theme
ggsave(width = 4, height = 7, file = file.path(plot_dir, "01-00-plot-RCI-liability-violins-in-brain-and-spinal-cord-stratify-condition.pdf"), ThePlot)


ThePlot <- ggplot(summary_df[summary_df$condition != "als+ftd",], 
    aes(x = condition, y = log10(non_repeat_intron_expr_tpm)))
ThePlot <- ThePlot + geom_violin(aes(fill = condition), colour = "#000000")
ThePlot <- ThePlot + geom_boxplot(aes(fill = condition), colour = "#000000", 
    width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9))
# ThePlot <- ThePlot + geom_smooth(method = "lm", formula = y ~ x, aes(colour = als_group))
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_discrete("Anatomical region")
ThePlot <- ThePlot + scale_y_continuous("RCI liability score")
# ThePlot <- ThePlot + scale_colour_manual("Condition colour", values = colours)
ThePlot <- ThePlot + scale_fill_manual("condition", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_wrap(. ~ tissue_group, scales = "free")
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression across anatomical regions")
ThePlot <- ThePlot + current_theme
ggsave(width = 4, height = 7, file = file.path(plot_dir, "01-00-plot-non-RCI-liability-violins-in-brain-and-spinal-cord-stratify-condition.pdf"), ThePlot)





```

- Perform tests

```{r}

test_df <- summary_df[summary_df$condition != "als+ftd",]
test_df$condition <- as.character(test_df$condition)

mean(test_df[test_df$tissue_group == "brain","repeat_intron_expr_tpm"])
# 276.6796
sd(test_df[test_df$tissue_group == "brain","repeat_intron_expr_tpm"])
# 83.96082

mean(test_df[test_df$tissue_group == "spinal_cord","repeat_intron_expr_tpm"])
# 246.0596
sd(test_df[test_df$tissue_group == "spinal_cord","repeat_intron_expr_tpm"])
# 43.17391


mean(test_df[test_df$tissue_group == "brain","non_repeat_intron_expr_tpm"])
# 276.6796
sd(test_df[test_df$tissue_group == "brain","non_repeat_intron_expr_tpm"])
# 83.96082

mean(test_df[test_df$tissue_group == "spinal_cord","non_repeat_intron_expr_tpm"])
# 76042.19
sd(test_df[test_df$tissue_group == "spinal_cord","non_repeat_intron_expr_tpm"])
# 2815.688





wilcox.test(test_df[test_df$tissue_group == "brain","repeat_intron_expr_tpm"], 
    test_df[test_df$tissue_group == "spinal_cord","repeat_intron_expr_tpm"])

lapply(split(test_df, test_df$tissue_group),function(temp_df){
    wilcox.test(temp_df$non_repeat_intron_expr_tpm[temp_df$condition == "als"], temp_df$non_repeat_intron_expr_tpm[temp_df$condition == "ctrl"])
})



lapply(split(test_df, test_df$tissue_group),function(temp_df){
    wilcox.test(temp_df$repeat_intron_expr_tpm[temp_df$condition == "als"], temp_df$repeat_intron_expr_tpm[temp_df$condition == "ctrl"])
})


lapply(split(test_df, test_df$condition),function(temp_df){
    wilcox.test(temp_df$repeat_intron_expr_tpm[temp_df$tissue_group == "spinal_cord"], temp_df$repeat_intron_expr_tpm[temp_df$tissue_group == "brain"])
})


```

- Plot RCI liabilities in individual tissue samples

```{r}


patient_liability_df$condition <- factor(patient_liability_df$condition, levels = c("ctrl", "als", "als+ftd"))

ThePlot <- ggplot(patient_liability_df[patient_liability_df$condition != "als+ftd",], 
    aes(x = tissue, y = repeat_intron_expr_tpm))
ThePlot <- ThePlot + geom_violin(aes(fill = condition), colour = "#000000", 
    lwd = 0.5, position = position_dodge(width = 0.9), scale = "width")
ThePlot <- ThePlot + geom_boxplot(aes(fill = condition), colour = "#000000", 
    width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9))
# ThePlot <- ThePlot + geom_point(aes(fill = condition), shape = 21, colour = "#000000", position = position_jitterdodge(dodge.width = 0.9))
# ThePlot <- ThePlot + geom_smooth(method = "lm", formula = y ~ x, aes(colour = als_group))
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_discrete("Tissues")
ThePlot <- ThePlot + scale_y_continuous("RCI liability")
# ThePlot <- ThePlot + scale_colour_manual("", values = colours)
ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_wrap(. ~ tissue_group, scales = "free")
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression across tissues")
ThePlot <- ThePlot + current_theme
ggsave(width = 10, height = 5, file = file.path(plot_dir, "01-01-plot-RCI-liability-in-tissue-samples.pdf"), ThePlot)


n_patient_df <- as.data.frame(table(patient_liability_df[,c("tissue", "condition")]))
n_patient_df <- n_patient_df[n_patient_df$condition != "als+ftd",]
n_patient_df$tissue_group <- "brain"
n_patient_df$tissue_group[grepl(x = n_patient_df$tissue, pattern = "Spinal")] <- "spinal_cord"

ThePlot <- ggplot(n_patient_df, 
    aes(x = tissue, y = log10(Freq), group = condition))
ThePlot <- ThePlot + geom_col(aes(fill = condition), position = position_dodge(width = 0.9), colour = "#000000")
ThePlot <- ThePlot + geom_text(aes(label = Freq), position = position_dodge(width = 0.9))
# ThePlot <- ThePlot + geom_point(aes(fill = condition), shape = 21, colour = "#000000", position = position_jitterdodge(dodge.width = 0.9))
# ThePlot <- ThePlot + geom_smooth(method = "lm", formula = y ~ x, aes(colour = als_group))
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_discrete("Tissues")
ThePlot <- ThePlot + scale_y_continuous("N patients")
# ThePlot <- ThePlot + scale_colour_manual("", values = colours)
ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_wrap(. ~ tissue_group, scales = "free")
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression across tissues")
ThePlot <- ThePlot + current_theme
ggsave(width = 10, height = 5, file = file.path(plot_dir, "01-01-plot-patient-Ns-in-tissue-samples.pdf"), ThePlot)


```

- How many patients had provided both brain and spinal cord tissue segments


```{r}

# number_of_regions_per_patient <- table(summary_df$subject_id)
# patients_with_both_brain_and_spinal_cord_segments <- number_of_regions_per_patient[number_of_regions_per_patient > 1]

# table(summary_df$condition[summary_df$subject_id %in% names(patients_with_both_brain_and_spinal_cord_segments)])

# patients_with_both_tissues_df <- summary_df[summary_df$subject_id %in% names(patients_with_both_brain_and_spinal_cord_segments),]

# patients_df_split <- split(patients_with_both_tissues_df, patients_with_both_tissues_df$subject_id)

# patient_pairwise_scatter_df <- do.call(rbind, lapply(patients_df_split, function(temp_df){
#     data.frame(subject_id = unique(temp_df$subject_id),
#         condition = unique(temp_df$condition),
#         brain_liability = temp_df$repeat_intron_expr_tpm[temp_df$tissue_group == "brain"],
#         spinal_cord_liability = temp_df$repeat_intron_expr_tpm[temp_df$tissue_group == "spinal_cord"])
# }))

```

- compute linear model


```{r}

# summary(lm(data = patient_pairwise_scatter_df, formula = spinal_cord_liability ~ brain_liability + brain_liability:condition))

# Call:
# lm(formula = brain_liability ~ spinal_cord_liability + spinal_cord_liability:condition,
#     data = patient_pairwise_scatter_df)

# Residuals:
#     Min      1Q  Median      3Q     Max
# -123.07  -51.26  -10.35   33.55  230.02

# Coefficients:
#                                         Estimate Std. Error t value Pr(>|t|)
# (Intercept)                            226.15429   27.50436   8.222 2.59e-14
# spinal_cord_liability                    0.37297    0.14617   2.552 0.011476
# spinal_cord_liability:conditionals      -0.29317    0.07688  -3.813 0.000183
# spinal_cord_liability:conditionals+ftd  -0.21595    0.10434  -2.070 0.039781

# (Intercept)                            ***
# spinal_cord_liability                  *
# spinal_cord_liability:conditionals     ***
# spinal_cord_liability:conditionals+ftd *
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

# Residual standard error: 66.35 on 198 degrees of freedom
# Multiple R-squared:  0.0703,    Adjusted R-squared:  0.05622
# F-statistic: 4.991 on 3 and 198 DF,  p-value: 0.002331

# colours <- c("ctrl" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
# ThePlot <- ggplot(patient_pairwise_scatter_df, 
#     aes(x = brain_liability, y = spinal_cord_liability))
# ThePlot <- ThePlot + geom_point(aes(fill = condition), colour = "#000000", shape = 21)
# # ThePlot <- ThePlot + geom_point(aes(fill = condition), shape = 21, colour = "#000000", position = position_jitterdodge(dodge.width = 0.9))
# ThePlot <- ThePlot + geom_smooth(method = "lm", formula = y ~ x, aes(colour = condition))
# # ThePlot <- ThePlot + stat_summary(aes(colour = condition))
# ThePlot <- ThePlot + scale_x_continuous("spinal cord RCI liability")
# ThePlot <- ThePlot + scale_y_continuous("brain RCI liability")
# ThePlot <- ThePlot + scale_colour_manual("Condition", values = colours)
# ThePlot <- ThePlot + scale_fill_manual("Condition", values = colours)
# # ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
# ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_wrap(. ~ condition, scales = "free")
# ThePlot <- ThePlot + ggtitle("Repeat containing intron expression across tissues")
# ThePlot <- ThePlot + current_theme
# ggsave(width = 10, height = 5, file = file.path(plot_dir, "01-03-plot-pairwise-scatterplot-for-spinal-cord-vs-brain-rci-liability.pdf"), ThePlot)


```
