---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

library("here")
message("Project root:", here(), "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("ggplot2")
require("ggbeeswarm")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("final-figures","figure-2")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS(here("final-figures","figure-1","rdata_files","00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds"))
feature_ranges <- readRDS(here("05-simulated-reads-differential-profiling","rdata_files","01-all-merged-exonic-and-intronic-segments-in-genes.rds"))
sig_result_list <- readRDS(here("20-import-significant-events","rdata_files","02-significantly-differentially-expressed-events.rds"))
answerals_object_list <- readRDS(here("13-01-answerals","rdata_files","01-02-list_of_counts_stat_samplesheet_from_answerals.rds"))
patient_liability_df <- readRDS(file.path(save_dir, "02-save-answerals-rci-expression-summary.rds"))

```

### Analysis

- Plot density distribution of RCI liability

```{r}

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")

mean(patient_liability_df$repeat_intron_expr_tpm)
sd(patient_liability_df$repeat_intron_expr_tpm)

ThePlot <- ggplot(patient_liability_df[patient_liability_df$mnd_group %in% c("healthy", "als"),], 
    aes(x = repeat_intron_expr_tpm))
ThePlot <- ThePlot + geom_density(fill = "#F6F4F3")
ThePlot <- ThePlot + geom_vline(xintercept = 200, lty = "dashed", colour = "#000000")
ThePlot <- ThePlot + scale_x_continuous("RCI liability")
ThePlot <- ThePlot + scale_y_continuous("density")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("RCI liability density distribution")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 7, file = file.path(plot_dir, "02-01-supplementary-figure-6B.pdf"), ThePlot)


```

- Compare RCI liability between conditions

```{r}

patient_liability_df$mnd_group <- factor(patient_liability_df$mnd_group, levels = c("healthy", "als", "other_mnd", "carrier"))
ThePlot <- ggplot(patient_liability_df[patient_liability_df$mnd_group %in% c("als", "healthy"),], 
    aes(x = mnd_group, y = repeat_intron_expr_tpm))
ThePlot <- ThePlot + geom_boxplot(aes(fill = mnd_group), colour = "#000000", 
    width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9))
ThePlot <- ThePlot + geom_quasirandom(aes(fill = mnd_group), colour = "#000000", shape = 21)
ThePlot <- ThePlot + scale_x_discrete("mnd group")
ThePlot <- ThePlot + scale_y_continuous("RCI liability")
ThePlot <- ThePlot + scale_colour_manual("Condition colour", values = colours)
ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_wrap(. ~ tissue, scales = "free")
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression in AnswerALS iPSCs")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 10, file = file.path(plot_dir, "02-02-plot-RCI-liability-in-healthy-and-ALS-figure-2C.pdf"), ThePlot)


wilcox.test(patient_liability_df$repeat_intron_expr_tpm[patient_liability_df$mnd_group == "als"], patient_liability_df$repeat_intron_expr_tpm[patient_liability_df$mnd_group == "healthy"])

#         Wilcoxon rank sum test with continuity correction

# data:  patient_liability_df$repeat_intron_expr_tpm[patient_liability_df$mnd_group == "als"] and patient_liability_df$repeat_intron_expr_tpm[patient_liability_df$mnd_group == "healthy"]
# W = 46195, p-value = 0.001516
# alternative hypothesis: true location shift is not equal to 0


t.test(patient_liability_df$repeat_intron_expr_tpm[patient_liability_df$mnd_group == "als"], patient_liability_df$repeat_intron_expr_tpm[patient_liability_df$mnd_group == "healthy"])

# t = 3.3931, df = 379.45, p-value = 0.0007639
# alternative hypothesis: true difference in means is not equal to 0
# 95 percent confidence interval:
#   3.036565 11.405449
# sample estimates:
# mean of x mean of y
#  159.0595  151.8385

```

- Test the effect of tails 

```{r}

select_q <- rev(seq(0.90, 1, by = 0.01))
vapply(select_q, function(x){
    wilcox.test(patient_liability_df$repeat_intron_expr_tpm[
            patient_liability_df$mnd_group == "als" & 
            patient_liability_df$repeat_intron_expr_tpm <= quantile(patient_liability_df$repeat_intron_expr_tpm, x)], 
        patient_liability_df$repeat_intron_expr_tpm[patient_liability_df$mnd_group == "healthy" & 
            patient_liability_df$repeat_intron_expr_tpm <= quantile(patient_liability_df$repeat_intron_expr_tpm, x)], alternative = "greater")$p.value
},1)

```
