---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

library("here")
message("Project root:", here(), "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("ggplot2")
require("ggvenn")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("final-figures","figure-2")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS(here("final-figures","figure-1","rdata_files","00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds"))
patient_penetrance_list <- readRDS(file.path(save_dir, "03-save-answerals-rci-expression-summary-with-penetrance-matrix.rds"))
nygc_penetrance_list <- readRDS(file.path(save_dir, "02-save-nygc-rci-expression-summary-with-penetrance-matrix.rds"))

patient_liability_df <- patient_penetrance_list$liability_df
penetrance_matrix <- patient_penetrance_list$penetrance_matrix
rownames(penetrance_matrix) <- patient_liability_df$sample_name

nygc_patient_liability_df <- nygc_penetrance_list$liability_df
nygc_penetrance_matrix <- nygc_penetrance_list$penetrance_matrix


```

- Find AnswerALS penetrant RCIs

```{r}

patient_df_split <- split(patient_liability_df, patient_liability_df$mnd_group)
ordered_patient_penetrance_matrix_list <- lapply(patient_df_split[c("healthy", "als")], function(temp_df){
    penetrance_matrix[temp_df$sample_name[order(temp_df$repeat_intron_expr_tpm, decreasing = F)],]
})

answerals_penetrant_rcis <- lapply(ordered_patient_penetrance_matrix_list, function(x){
	temp_vector <- colSums(x)
	names(temp_vector[temp_vector > 0])
})

```

- Plot intersection between healthy and als

```{r}

colours <- c("#F6F4F3", "#F87575")

pdf(file.path(plot_dir, "06-01-intersection-between-healthy-and-als-in-answerals-supplementary-figure-7A.pdf"), width = 7, height = 7)

ggvenn(answerals_penetrant_rcis, c("healthy", "als"), fill_color = colours)

dev.off()

```

- Find penetrant RCIs in NYGC samples

```{r}

nygc_patient_df_split <- split(nygc_patient_liability_df, paste(nygc_patient_liability_df$condition, nygc_patient_liability_df$tissue_group))
nygc_ordered_patient_penetrance_matrix_list <- lapply(nygc_patient_df_split, function(temp_df){
    nygc_penetrance_matrix[temp_df$penetrance_row[order(temp_df$rci_liability, decreasing = F)],]
})

nygc_penetrant_rcis <- lapply(nygc_ordered_patient_penetrance_matrix_list, function(x){
    temp_vector <- colSums(x)
    names(temp_vector[temp_vector > 0])
})


```

- Plot intersection between NYGC penetrant rcis in each tissue region


```{r}

# colours <- c("#F6F4F3", "#F87575", "#FFA9A3")
colours <- c("#F6F4F3", "#F87575")
brain_rcis <- nygc_penetrant_rcis[grepl(x = names(nygc_penetrant_rcis), pattern = "brain")]
names(brain_rcis) <- str_replace_all(string = names(brain_rcis), pattern = " brain", replacement = "")
pdf(file.path(plot_dir, "06-02-intersection-between-nygc-brain-penetrant-rcis-supplementary-figure-7A.pdf"), width = 7, height = 7)
ggvenn(brain_rcis, c("ctrl", "als"), fill_color = colours)
dev.off()


# colours <- c("#F6F4F3", "#F87575", "#FFA9A3")
colours <- c("#F6F4F3", "#F87575")
spinal_cord_rcis <- nygc_penetrant_rcis[grepl(x = names(nygc_penetrant_rcis), pattern = "spinal_cord")]
names(spinal_cord_rcis) <- str_replace_all(string = names(spinal_cord_rcis), pattern = " spinal_cord", replacement = "")
pdf(file.path(plot_dir, "06-03-intersection-between-nygc-spinal_cord-penetrant-rcis-supplementary-figure-7A.pdf"), width = 7, height = 7)
ggvenn(spinal_cord_rcis, c("ctrl", "als"), fill_color = colours)
dev.off()


```


- Plot intersection between AnswerALS and NYGC


```{r}

intersect(unique(do.call(c, answerals_penetrant_rcis)), unique(do.call(c, spinal_cord_rcis))) %>% length
# 96
intersect(unique(do.call(c, answerals_penetrant_rcis)), unique(do.call(c, brain_rcis))) %>% length
# 1409
# 
nygc_answerals_intersect_df <- do.call(rbind, lapply(names(answerals_penetrant_rcis), function(x){
    answerals_rcis <- answerals_penetrant_rcis[[x]]
    nygc_name_split <- str_split(names(nygc_penetrant_rcis), pattern = " ", simplify = T)
    ji_list <- vapply(nygc_penetrant_rcis, function(y){
        length(intersect(answerals_rcis, y)) / length(union(answerals_rcis, y))
    },1)
    len_intersect <- vapply(nygc_penetrant_rcis, function(y){
        length(intersect(answerals_rcis, y)) 
    },1)
    prop_intersect_answerals <- vapply(nygc_penetrant_rcis, function(y){
        length(intersect(answerals_rcis, y)) / length(answerals_rcis)
    },1)
    prop_intersect_nygc <- vapply(nygc_penetrant_rcis, function(y){
        length(intersect(answerals_rcis, y)) / length(y)
    },1)
    data.frame(answerals_condition = x, 
        nygc_condition = nygc_name_split[,1], 
        nygc_tissue = nygc_name_split[,2],
        length_intersect = len_intersect,
        prop_intersect_answerals = prop_intersect_answerals,
        prop_intersect_nygc = prop_intersect_nygc)
}))

```

- Plot heatmap of intersections

```{r}

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")

nygc_answerals_intersect_df$nygc_condition <- factor(nygc_answerals_intersect_df$nygc_condition, levels = c("ctrl", "als", "als+ftd"))
nygc_answerals_intersect_df$answerals_condition <- factor(nygc_answerals_intersect_df$answerals_condition, levels = c("healthy", "als"))

ThePlot <- ggplot(nygc_answerals_intersect_df[nygc_answerals_intersect_df$nygc_condition != "als+ftd",], 
    aes(x = answerals_condition, y = nygc_condition))
ThePlot <- ThePlot + geom_tile(aes(fill = log10(prop_intersect_nygc)))
ThePlot <- ThePlot + geom_text(aes(label = round(prop_intersect_nygc, 2)))
ThePlot <- ThePlot + scale_x_discrete("AnswerALS iPSC-MN MND groups")
ThePlot <- ThePlot + scale_y_discrete("NYGC PM RCIs")
ThePlot <- ThePlot + scale_fill_distiller("Proportion intersection", palette = "Reds", direction = 1)
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_grid(. ~ nygc_tissue)
ThePlot <- ThePlot + ggtitle("Intersection between AnswerALS iPSC-MN data and NYGC PM tissue samples")
ThePlot <- ThePlot + current_theme
ggsave(width = 10, height = 10, file = file.path(plot_dir, "06-04-plot-proportion-of-nygc-pm-rcis-captured-between-answerals-ipsc-mn-and-nygc-pm-supplementary-figure-7B.pdf"), ThePlot)

```

