---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

library("here")
message("Project root:", here(), "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("ggplot2")
library(rlang)
library(riskRegression)
library(survival)
library(pammtools)
library(gganimate)
library(transformr)
library(plotly)
library(reshape2)
library(knitr)
library(rmarkdown)

require("contsurvplot")
require("RColorBrewer")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("final-figures","figure-2")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load penetrance data

```{r}

nygc_penetrance_list <- readRDS(here("final-figures","figure-2","rdata_files","02-save-nygc-rci-expression-summary-with-penetrance-matrix.rds"))
nygc_patient_liability_df <- nygc_penetrance_list$liability_df
nygc_penetrance_matrix <- nygc_penetrance_list$penetrance_matrix

```

- Prepare data for survival analysis

```{r}

nygc_patient_liability_df$type <- factor(nygc_patient_liability_df$type, levels = c("control", "c9orf72-ALS", "other_mutation-fALS", "sporadic"))
nygc_patient_liability_df$type_group <- as.character(nygc_patient_liability_df$type)
nygc_patient_liability_df$type_group[nygc_patient_liability_df$type %in% c("other_mutation-fALS", "sporadic")] <- "sporadic+other_fALS"

patient_surv_df <- do.call(rbind, lapply(split(nygc_patient_liability_df, paste(nygc_patient_liability_df$subject_id, nygc_patient_liability_df$tissue)), function(temp_df){
    message(unique(temp_df$subject_id))
    data.frame(subject_id = unique(temp_df$subject_id),
        tissue = unique(temp_df$tissue),
        tissue_group = unique(temp_df$tissue_group),
        age_at_onset = unique(temp_df$age_at_symptom_onset),
        age_at_death = unique(temp_df$age_at_death),
        disease_duration_in_months = unique(temp_df$disease_duration_in_months),
        sex_genotype = unique(temp_df$sex_genotype),
        mnd_with_ftd = unique(temp_df$mnd_with_ftd),
        condition = unique(temp_df$condition),
        mnd_with_dementia = unique(temp_df$mnd_with_dementia),
        type_group = unique(temp_df$type_group),
        rci_liability = max(temp_df$rci_liability),
        intron_expression_rate = mean(temp_df$intron_expression_rate))
}))

patient_surv_df$age_at_death[patient_surv_df$age_at_death == "90 or Older"] <- "90"
patient_surv_df$age_at_death <- as.numeric(patient_surv_df$age_at_death)
patient_surv_df$age_at_onset <- as.numeric(patient_surv_df$age_at_onset)
patient_surv_df$disease_duration <- as.numeric(patient_surv_df$disease_duration_in_months)
patient_surv_df$condition <- factor(patient_surv_df$condition, levels = c("ctrl", "als", "als+ftd"))


summary(lm(data = patient_surv_df[patient_surv_df$tissue == "Spinal_Cord_Cervical",], formula = intron_expression_rate ~ condition + condition + condition:rci_liability))
summary(lm(data = patient_surv_df[patient_surv_df$tissue == "Spinal_Cord_Thoracic",], formula = intron_expression_rate ~ condition + condition + condition:rci_liability))
summary(lm(data = patient_surv_df[patient_surv_df$tissue == "Spinal_Cord_Lumbar",], formula = intron_expression_rate ~ condition + condition + condition:rci_liability))
summary(lm(data = patient_surv_df[patient_surv_df$tissue == "Cortex",], formula = intron_expression_rate ~ condition + condition + condition:rci_liability))
summary(lm(data = patient_surv_df[patient_surv_df$tissue == "Cortex_Motor",], formula = intron_expression_rate ~ condition + condition + condition:rci_liability))
summary(lm(data = patient_surv_df[patient_surv_df$tissue == "Cerebellum",], formula = intron_expression_rate ~ condition + condition + condition:rci_liability))

```

- Compare RCI liability vs Intron expression rate

```{r}

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

colours <- c("ctrl" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
ThePlot <- ggplot(patient_surv_df[patient_surv_df$condition %in% c("ctrl", "als"),], 
    aes(x = log10(rci_liability), y = log10(intron_expression_rate)))
ThePlot <- ThePlot + geom_point(aes(fill = condition), colour = "#000000", shape = 21)
ThePlot <- ThePlot + geom_smooth(method = "lm", aes(colour = condition), formula = y ~ x)
ThePlot <- ThePlot + scale_x_continuous("RCI liability")
ThePlot <- ThePlot + scale_y_continuous("RCI emission rate")
ThePlot <- ThePlot + scale_colour_manual("Condition colour", values = colours)
ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_wrap(. ~ tissue, scales = "free")
ThePlot <- ThePlot + ggtitle("RCI liability ~ RCI emission rate")
ThePlot <- ThePlot + current_theme
ggsave(width = 10, height = 5, file = file.path(plot_dir, "06-07-plot-RCI-liability-vs-emission-rate-figure-3D.pdf"), ThePlot)


summary(lm(data = patient_surv_df[patient_surv_df$tissue == "Spinal_Cord_Cervical",], formula = intron_expression_rate ~ age_at_onset))
summary(lm(data = patient_surv_df[patient_surv_df$tissue == "Spinal_Cord_Thoracic",], formula = intron_expression_rate ~ age_at_onset))
summary(lm(data = patient_surv_df[patient_surv_df$tissue == "Spinal_Cord_Lumbar",], formula = intron_expression_rate ~ age_at_onset))
summary(lm(data = patient_surv_df[patient_surv_df$tissue == "Cortex",], formula = intron_expression_rate ~ age_at_onset))
summary(lm(data = patient_surv_df[patient_surv_df$tissue == "Cortex_Motor",], formula = intron_expression_rate ~ age_at_onset))
summary(lm(data = patient_surv_df[patient_surv_df$tissue == "Cerebellum",], formula = intron_expression_rate ~ age_at_onset))

```
- Plot RCI emission rate versus age of symptom onset.

```{r}

colours <- c("ctrl" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
ThePlot <- ggplot(patient_surv_df[patient_surv_df$condition == "als",], 
    aes(x = age_at_onset, y = intron_expression_rate))
ThePlot <- ThePlot + geom_point(aes(fill = condition), colour = "#000000", shape = 21)
ThePlot <- ThePlot + geom_smooth(method = "lm", aes(colour = condition), formula = y ~ x)
ThePlot <- ThePlot + scale_x_continuous("Age of symptom onset")
ThePlot <- ThePlot + scale_y_continuous("RCI emission rate")
ThePlot <- ThePlot + scale_colour_manual("Condition colour", values = colours)
ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_wrap(. ~ tissue, scales = "free")
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression in AnswerALS iPSCs")
ThePlot <- ThePlot + current_theme
ggsave(width = 10, height = 10, file = file.path(plot_dir, "06-08-plot-RCI-emission-rate-vs-age-of-symptom-onset.pdf"), ThePlot)

```