---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

We want to understand if Microsatellites are expressed and incorporated into mature mRNAs. Towards this end, we are performing a complete genome-wide investigation of micro-satellite repeats. 


### 1.0 Package requirements


- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

library(here)
i_am("00-analysis-markdowns/final_figures/figure-1/01-prepare-figure-1A.rmd")
message("Project root:", here(), "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("final-figures","figure-1")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Import all predicted micro-satellites

- Import all repeats

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_sequence_files <- c(trinucleotide = here("01-repeat-characterisation","input_files","02-00-nonoverlapping_genic_sequences_chr1-22_trinucleotide.fa.tsv"),
    quadnucleotide = here("01-repeat-characterisation","input_files","02-00-nonoverlapping_genic_sequences_chr1-22_quad.sense.fa.tsv"),
    pentanucleotide = here("01-repeat-characterisation","input_files","02-00-nonoverlapping_genic_sequences_chr1-22_penta.sense.fa.tsv"),
    hexanucleotide_GP = here("01-repeat-characterisation","input_files","02-00-nonoverlapping_genic_sequences_chr1-22_GP.sense.fa.tsv"),
    hexanucleotide_GA = here("01-repeat-characterisation","input_files","02-00-nonoverlapping_genic_sequences_chr1-22_GA.sense.fa.tsv"),
    hexanucleotide_GR = here("01-repeat-characterisation","input_files","02-00-nonoverlapping_genic_sequences_chr1-22_GR.sense.fa.tsv"),
    hexanucleotide_PA = here("01-repeat-characterisation","input_files","02-00-nonoverlapping_genic_sequences_chr1-22_PA.sense.fa.tsv"),
    hexanucleotide_PR = here("01-repeat-characterisation","input_files","02-00-nonoverlapping_genic_sequences_chr1-22_PR.sense.fa.tsv"),
    hexanucleotide_GL = here("01-repeat-characterisation","input_files","02-00-nonoverlapping_genic_sequences_chr1-22_GL.sense.fa.tsv"),
    hexanucleotide_PS = here("01-repeat-characterisation","input_files","02-00-nonoverlapping_genic_sequences_chr1-22_PS.sense.fa.tsv"),
    dodecanucleotide = here("01-repeat-characterisation","input_files","02-00-nonoverlapping_genic_sequences_chr1-22_dodeca.sense.fa.tsv"))


repeat_sequence_df <- do.call(rbind, lapply(names(repeat_sequence_files), function(x){
    message(x)
    temp_df <- data.table::fread(repeat_sequence_files[x], data.table = F)
    if(nrow(temp_df) > 0){
        temp_df$nucleotide_group <- x
        return(temp_df)
    }
    return(NULL)    
}))

coord_split_matrix <- str_split(repeat_sequence_df[,1], pattern = ":", simplify = T)

sense_repeat_coord_ranges <- Brick_make_ranges(chr = coord_split_matrix[,2], 
    start = as.numeric(coord_split_matrix[,3]) + repeat_sequence_df[,4],
    end = as.numeric(coord_split_matrix[,3]) + repeat_sequence_df[,4] + (str_length(repeat_sequence_df[,2])*repeat_sequence_df[,3]) - 1,
    strand = coord_split_matrix[,5])
sense_repeat_coord_ranges$ssr <- repeat_sequence_df$SSR
sense_repeat_coord_ranges$span_group <- repeat_sequence_df$nucleotide_group
sense_repeat_coord_ranges <- sense_repeat_coord_ranges[strand(sense_repeat_coord_ranges) == "+"]

antisense_repeat_coord_ranges <- Brick_make_ranges(chr = coord_split_matrix[,2], 
    start = as.numeric(coord_split_matrix[,4]) - repeat_sequence_df[,4] - (str_length(repeat_sequence_df[,2])*repeat_sequence_df[,3]) + 1,
    end = as.numeric(coord_split_matrix[,4]) - repeat_sequence_df[,4],
    strand = coord_split_matrix[,5])
antisense_repeat_coord_ranges$ssr <- repeat_sequence_df$SSR
antisense_repeat_coord_ranges$span_group <- repeat_sequence_df$nucleotide_group
antisense_repeat_coord_ranges <- antisense_repeat_coord_ranges[strand(antisense_repeat_coord_ranges) == "-"]

repeat_coord_ranges <- unique(c(sense_repeat_coord_ranges, antisense_repeat_coord_ranges))

repeat_coord_ranges$span_group[grepl(x = repeat_coord_ranges$span_group, pattern = "hexanucleotide")] <- "hexanucleotide"


```

### 3.1.0 Import exonic ranges

- Import exonic and intronic ranges

```{r}

feature_ranges <- readRDS(here("05-simulated-reads-differential-profiling","rdata_files","01-all-merged-exonic-and-intronic-segments-in-genes.rds"))

```


### 3.1.2 Report the number of repeats overlapping introns or exons

- Overlap repeats with 

```{r}

seqlevelsStyle(repeat_coord_ranges) <- seqlevelsStyle(feature_ranges)
ol_object <- findOverlaps(repeat_coord_ranges, feature_ranges, ignore.strand = T, type = "any")
repeats_overlapping_features <- repeat_coord_ranges[unique(queryHits(ol_object))]
ol_split <- split(feature_ranges$type[subjectHits(ol_object)], queryHits(ol_object))
all_ol_types <- lapply(ol_split, function(x){
    unique(x)
})


```

```{r}


mapped_repeat_ranges <- repeat_coord_ranges[queryHits(ol_object)]
mapped_repeat_ranges$repeat_id <- names(mapped_repeat_ranges)
mapped_repeat_ranges$feature_hash_id <- feature_ranges$hash_id[subjectHits(ol_object)]
names(mapped_repeat_ranges) <- NULL
write.table(as.data.frame(mapped_repeat_ranges), file = here("final-figures", "figure-1", "output_files", "01-01-genomic-coordinates-of-repeat-regions.txt"), sep = "\t")

mapped_feature_ranges <- feature_ranges[subjectHits(ol_object)]
names(mapped_feature_ranges) <- NULL
mapped_feature_ranges$repeat_id <- mapped_repeat_ranges$repeat_id
elementMetadata(mapped_feature_ranges) <- elementMetadata(mapped_feature_ranges)[!(names(elementMetadata(mapped_feature_ranges)) %in% 
    c("tx_id", "tx_name", "exon_id", "exon_name", "exon_rank"))]
write.table(as.data.frame(mapped_feature_ranges), file = here("final-figures", "figure-1", "output_files", "01-02-genomic-coordinates-of-features-containing-any-repeat-overlap.txt"), sep = "\t")

```

- What the total number of repeats that are overlapping exons or introns?

```{r}


length(unique(repeats_overlapping_features))
# [1] 8432

total_exonic <- length(all_ol_types[vapply(all_ol_types, function(x){
    all(x == "exon")
},T)])
total_intron <- length(all_ol_types[vapply(all_ol_types, function(x){
    all(x == "intron")
},T)])
total_shared <- length(all_ol_types[vapply(all_ol_types, function(x){
    all("intron" %in% x & "exon" %in% x)
},T)]) 

frequency_df <- data.frame(
    group = c("exon", "intron", "shared"),
    frequency = c(total_exonic, total_intron, total_shared))

frequency_df$percentage <- frequency_df$frequency / length(unique(repeats_overlapping_features))

require(dplyr)
plot_freq_df <- frequency_df %>% 
group_by(group) %>% 
# arrange(desc(subgroup)) %>%
mutate(label_y = cumsum(frequency) - 0.5 * frequency) %>%
select(group,  frequency, percentage, label_y) %>%
as.data.frame()

ThePlot <- ggplot(plot_freq_df,aes(x = group, y = frequency))
ThePlot <- ThePlot + geom_col(fill = "grey", colour = "#000000")
ThePlot <- ThePlot + geom_label(aes(y = label_y, label = round(percentage * 100, 2)))
ThePlot <- ThePlot + scale_x_discrete("feature type")
ThePlot <- ThePlot + scale_y_continuous("proportion of total microsatellites")
ThePlot <- ThePlot + scale_fill_brewer("overlap type", palette = "Pastel1")
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(. ~ region)
ThePlot <- ThePlot + ggtitle("Frequency of micro-satellites in exons/introns")
ThePlot <- ThePlot + theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))
ggsave(ThePlot, file = file.path(plot_dir, "01-figure-1A.pdf"), width = 3, height = 7)

```

- Make the feature ranges containing repeats


```{r}

overlaps_object <- findOverlaps(repeat_coord_ranges, feature_ranges)

feature_ranges_with_repeats <- feature_ranges[subjectHits(overlaps_object)]
feature_ranges_with_repeats$which_repeat <- queryHits(overlaps_object)
feature_ranges_with_repeats$repeat_width <- width(repeat_coord_ranges)[feature_ranges_with_repeats$which_repeat]
feature_ranges_with_repeats$repeat_unit_size <- str_length(repeat_coord_ranges$ssr)[feature_ranges_with_repeats$which_repeat]
feature_ranges_with_repeats$span_group <- repeat_coord_ranges$span_group[feature_ranges_with_repeats$which_repeat]
feature_ranges_with_repeats$expansion_size <- feature_ranges_with_repeats$repeat_width / feature_ranges_with_repeats$repeat_unit_size
feature_ranges_with_repeats$ssr <- repeat_coord_ranges$ssr[feature_ranges_with_repeats$which_repeat]


```



- Save the repeat ranges


```{r}

saveRDS(object = list(repeat_ranges = repeat_coord_ranges, repeat_containing_ranges = feature_ranges_with_repeats), 
    file = file.path(save_dir, "00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds"))

```

- How many repeats does an intron or exon host? 


```{r}


intronic_frequencies_df <- as.data.frame(table(table(feature_ranges_with_repeats$feature_id[feature_ranges_with_repeats$type == "intron"])))
intronic_frequencies_df$region <- "intron"
intronic_frequencies_df$proportion <- intronic_frequencies_df$Freq / sum(intronic_frequencies_df$Freq)
exonic_frequencies_df <- as.data.frame(table(table(feature_ranges_with_repeats$feature_id[feature_ranges_with_repeats$type == "exon"])))
exonic_frequencies_df$region <- "exon"
exonic_frequencies_df$proportion <- exonic_frequencies_df$Freq / sum(exonic_frequencies_df$Freq)

plot_df <- rbind(exonic_frequencies_df, intronic_frequencies_df)

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12, hjust = 1),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

colours <- c("intron" = "#F24236", "exon" = "#F7F7F2")
ThePlot <- ggplot(plot_df, aes(x = region, y = proportion))
ThePlot <- ThePlot + geom_bar(aes(fill = log10(as.numeric(Var1))), stat = "identity", colour = "#000000")
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_discrete("region")
ThePlot <- ThePlot + scale_y_continuous("proportion of features with k repeats")
ThePlot <- ThePlot + scale_fill_distiller("number of repeats", palette = "Reds", direction = 1)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("density of repeat occurence")
ThePlot <- ThePlot + current_theme
ggsave(width = 3, height = 5, file = file.path(plot_dir, "01-02-supplementary-figure-1A.pdf"), ThePlot)

```




