---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

We want to understand if Microsatellites are expressed and incorporated into mature mRNAs. Towards this end, we are performing a complete genome-wide investigation of micro-satellite repeats. 


### 1.0 Package requirements



- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

library("here")
i_am("00-analysis-markdowns/final_figures/figure-1/01-plot-supplementary-figure-1B.rmd")
message("Project root:", here(), "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
require("GenomicFeatures")
require("ggplot2")
require("dplyr")

```

### 2.0 Include experiment matrix

- Setup directories

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("final-figures","figure-1")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
source_dir <- file.path(analysis_path, "src")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

# Function declaration


```{r}

require("rtracklayer")
require("parallel")
source(file.path())
phylop_cons_tracks <- file.path(input_dir, "hg38.phyloP7way.bw")
save_file <- file.path(save_dir, "01-save-conservation-tracks.rds")
source(file.path(source_dir, "source.R"))

```


### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS(file.path(save_dir, "00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds"))
repeat_coord_ranges <- repeat_ranges_list$repeat_ranges
feature_ranges <- readRDS(here("05-simulated-reads-differential-profiling", "rdata_files", "01-all-merged-exonic-and-intronic-segments-in-genes.rds"))

```

- Select repeats which are at least 100bp within introns

```{r}

ol_object <- findOverlaps(repeat_coord_ranges, feature_ranges, type = "within")
repeats_within_features <- repeat_coord_ranges[queryHits(ol_object)]
repeats_within_features$which_feature <- subjectHits(ol_object)
repeats_within_features$type <- feature_ranges$type[repeats_within_features$which_feature]

intronic_repeat_ranges <- repeats_within_features[repeats_within_features$type == "intron"]
intronic_repeat_ranges$dist_from_start <- start(intronic_repeat_ranges) - start(feature_ranges)[intronic_repeat_ranges$which_feature]
intronic_repeat_ranges$dist_from_end <- end(feature_ranges)[intronic_repeat_ranges$which_feature] - end(intronic_repeat_ranges)

selected_repeats <- unique(names(intronic_repeat_ranges)[intronic_repeat_ranges$dist_from_start > 100 & intronic_repeat_ranges$dist_from_end > 100])

selected_repeat_ranges <- repeat_coord_ranges[selected_repeats]
seqlevelsStyle(selected_repeat_ranges) <- "UCSC"

```

```{r}

extend_ranges <- selected_repeat_ranges
start(extend_ranges) <- start(extend_ranges) - 100
end(extend_ranges) <- end(extend_ranges) + 100
wig_ranges <- import.bw(con = phylop_cons_tracks, which = extend_ranges)


```

- Compute Conservation of repeat and upstream flanking regions 

```{r}

save_file <- file.path(save_dir, "01-03-conservation-plot-object.rds")

if(!file.exists(save_file)){
    all_repeat_matrix_list <- fetch_bigwig_signal(
        repeat_ranges = selected_repeat_ranges, 
        index_start = 1, 
        index_end = length(selected_repeat_ranges),
        bw_signal = wig_ranges)

    rownames(all_repeat_matrix_list) <- names(selected_repeat_ranges)
    sense_matrix <- all_repeat_matrix_list[as.vector(strand(selected_repeat_ranges) == "+"),]
    antisense_matrix <- all_repeat_matrix_list[as.vector(strand(selected_repeat_ranges) == "-"),ncol(all_repeat_matrix_list):1]
    all_matrix <- rbind(sense_matrix, antisense_matrix)

    conservation_plot_df <- reshape2::melt(all_matrix)
    colnames(conservation_plot_df) <- c("repeat_id", "position", "phylop")
    conservation_plot_df$ssr <- selected_repeat_ranges$ssr[conservation_plot_df$repeat_id]
    conservation_plot_df$span_group <- selected_repeat_ranges$span_group[conservation_plot_df$repeat_id]

    saveRDS(object = conservation_plot_df, file = save_file)    
}else{
    conservation_plot_df <- readRDS(save_file)
}


plot_df <- conservation_plot_df %>%
group_by(position) %>%
summarise(mean_phylop = mean(phylop, na.rm = T)) %>%
as.data.frame()


current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

ThePlot <- ggplot(plot_df, 
    aes(x = position, y = mean_phylop))
ThePlot <- ThePlot + geom_vline(xintercept = c(11,20), linetype = "dashed", colour = "#000000")
ThePlot <- ThePlot + geom_hline(yintercept = 0, linetype = "dashed", colour = "#000000")
ThePlot <- ThePlot + geom_line(colour = "#FF0000")
# ThePlot <- ThePlot + geom_smooth(method = "loess", formula = y ~ x, colour = "#FF0000")
# ThePlot <- ThePlot + geom_vline(xintercept = 0.187415205, linetype = "dashed")
# ThePlot <- ThePlot + geom_line()
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_continuous("conservation", breaks = c(1,11,20,30), labels = c("-100","repeat_start","repeat_end","+100"))
ThePlot <- ThePlot + scale_y_continuous("mean phylop")
# ThePlot <- ThePlot + scale_fill_distiller("fisher pval", palette = "YlOrRd", na.value = "grey")
# ThePlot <- ThePlot + scale_fill_manual("DE direction", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(. ~ region)
ThePlot <- ThePlot + ggtitle("Conservation score distribution of regions")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 7, file = file.path(plot_dir, "01-03-supplementary-figure-1B.pdf"), ThePlot)

```





