---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

We want to understand if Microsatellites are expressed and incorporated into mature mRNAs. Towards this end, we are performing a complete genome-wide investigation of micro-satellite repeats. 


### 1.0 Package requirements

- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

library("here")
message("Project root:", here(), "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
require("GenomicFeatures")
require("ggplot2")
require("dplyr")

```

### 2.0 Include experiment matrix

- Setup directories

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("final-figures","figure-1")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
source_dir <- file.path(analysis_path, "src")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Declare functions


```{r}

summarise_for_barplot <- function(events_df, total_length){
    require("dplyr")
    events_df_split <- split(events_df, paste(events_df$study, events_df$contrasts, sep = "_"))
    frequency_df <- do.call(rbind, lapply(events_df_split, function(temp_df){
        events_up <- length(unique(temp_df$feature_ids[temp_df$logfc > 0]))
        events_down <- length(unique(temp_df$feature_ids[temp_df$logfc < 0]))
        return_df <- data.frame(study = unique(temp_df$study),
            contrast = unique(temp_df$contrasts),
            direction = c("up", "down"),
            frequency = c(events_up, events_down))
        return_df$proportion <- return_df$frequency / total_length
        return_df <- return_df %>% 
            arrange(desc(direction)) %>% 
            mutate(prop_lab_y = cumsum(proportion) - (0.5 * proportion),
                freq_lab_y = cumsum(frequency) - (0.5 * frequency)) %>%
            as.data.frame()
        return(return_df)
    }))
    rownames(frequency_df) <- NULL
    return(frequency_df)
}

```

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

sig_results_list <- readRDS(here("20-import-significant-events", "rdata_files", "02-significantly-differentially-expressed-events.rds"))
feature_ranges <- readRDS(here("05-simulated-reads-differential-profiling","rdata_files","01-all-merged-exonic-and-intronic-segments-in-genes.rds"))
repeat_ranges_list <- readRDS(here("final-figures","figure-1","rdata_files","00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds"))
sig_result_df <- sig_results_list$significant_events
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges

```

### 4.1 Plot n of significant differentially expressed events

- Plot density of differential expression

```{r}

all_introns_df <- sig_result_df[sig_result_df$region == "intron",]

all_intron_frequency_df <- summarise_for_barplot(all_introns_df, total_length = length(unique(feature_ranges$feature_id[feature_ranges$type == "intron"])))

all_intron_frequency_df$cohort <- "public-data"
all_intron_frequency_df$cohort[grepl(x = all_intron_frequency_df$study, pattern = "nygc")] <- "nygc"
all_intron_frequency_df$cohort[grepl(x = all_intron_frequency_df$study, pattern = "answerals")] <- "answerals"
all_intron_frequency_df$tissue_type <- "iPSCs"
all_intron_frequency_df$tissue_type[all_intron_frequency_df$cohort == "nygc"] <- "postmortem"

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12, hjust = 1),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))


all_totals_df <- all_intron_frequency_df %>% 
group_by(study, contrast) %>%
mutate(total = sum(frequency)) %>% ungroup() %>%
dplyr::select(cohort, study, tissue_type, contrast, total) %>%
unique() %>% as.data.frame

mean_expr_per_cohort <- all_totals_df %>% 
group_by(cohort) %>%
mutate(means = mean(total)) %>%
dplyr::select(cohort, tissue_type, means) %>%
unique %>% as.data.frame

#        cohort tissue_type     means
# 1   answerals       iPSCs   8901.50
# 2 public-data       iPSCs  12347.55
# 3        nygc  postmortem 190311.00

median_expr_per_cohort <- all_totals_df %>% 
group_by(tissue_type) %>%
mutate(means = median(total)) %>%
dplyr::select(tissue_type, means) %>%
unique %>% as.data.frame

#   tissue_type    means
# 1       iPSCs  10679.5
# 2  postmortem 191702.5

ThePlot <- ggplot(all_intron_frequency_df, 
    aes(x = log10(frequency)))
ThePlot <- ThePlot + geom_density(aes(fill = tissue_type), alpha = 0.6, colour = "#000000")
ThePlot <- ThePlot + geom_vline(data = median_expr_per_cohort, aes(xintercept = log10(means)), linetype = "dashed", colour = "#000000")
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_continuous("log10 frequency of differential expression")
ThePlot <- ThePlot + scale_y_continuous("density")
ThePlot <- ThePlot + scale_fill_brewer("tissue type", palette = "Set1")
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("density of events identified")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 7, file = file.path(plot_dir, "02-04-supplementary-figure-1E.pdf"), ThePlot)


```

- Show significant differentially expressed events under pval < 0.05

```{r}

all_introns_df <- sig_result_df[sig_result_df$region == "intron",]

frequency_df <- summarise_for_barplot(all_introns_df, total_length = length(unique(feature_ranges$feature_id[feature_ranges$type == "intron"])))

plot_frequency_df <- frequency_df
plot_frequency_df$cohort <- "public-data"
plot_frequency_df$cohort[grepl(x = plot_frequency_df$study, pattern = "nygc")] <- "nygc"
plot_frequency_df$cohort[grepl(x = plot_frequency_df$study, pattern = "answerals")] <- "answerals"
plot_frequency_df$tissue_type <- "iPSCs"
plot_frequency_df$tissue_type[grepl(x = plot_frequency_df$study, pattern = "nygc")] <- "postmortem"

public_data_frequency_df <- plot_frequency_df[plot_frequency_df$cohort == "public-data",]
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_c9orf72_vs_ctrl"] <- "c9orf72"
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_fus_vs_ctrl"] <- "fus"
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_c9orf72_vs_iso"] <- "c9orf72"
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_sod1_vs_ctrl"] <- "sod1"
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_sod1_vs_iso"] <- "sod1"
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_vcp_vs_ctrl"] <- "vcp"
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_d35_mut_vs_wt"] <- "vcp"
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_tardbp_vs_wt"] <- "tardbp"
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_tardbp_vs_ctrl"] <- "tardbp"
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_vcp_vs_wt"] <- "vcp"
public_data_frequency_df$xaxis <- paste(public_data_frequency_df$study, public_data_frequency_df$contrast, sep = "_")

answerals_frequency_df <- plot_frequency_df[plot_frequency_df$cohort == "answerals",]
answerals_frequency_df$contrast[answerals_frequency_df$contrast == "fals_vs_control"] <- "familial"
answerals_frequency_df$contrast[answerals_frequency_df$contrast == "sals_vs_control"] <- "sporadic"
answerals_frequency_df$xaxis <- paste(answerals_frequency_df$study, answerals_frequency_df$contrast, sep = "_")


nygc_frequency_df <- plot_repeat_frequency_df[plot_repeat_frequency_df$cohort == "nygc",]
nygc_frequency_df$contrast[nygc_frequency_df$contrast == "als_vs_control"] <- "als"
nygc_frequency_df$contrast[nygc_frequency_df$contrast == "als+ftd_vs_control"] <- "als+ftd"

nygc_frequency_df$xaxis <- paste(nygc_frequency_df$study, nygc_frequency_df$contrast, sep = "_")


answerals_frequency_df$xaxis <- factor(answerals_frequency_df$xaxis, 
    levels = c("answerals_familial", "answerals_sporadic"))

public_data_frequency_df$xaxis <- factor(public_data_frequency_df$xaxis, 
    levels = c("c9orf72-sareen-2013_c9orf72", "c9orf72-sterneckert-2020_c9orf72", "c9orf72-dafinca-2020_c9orf72", 
        "c9orf72-catanese-2021_c9orf72", "c9orf72-casteli-2021_c9orf72", "c9orf72-sommer-2022_c9orf72", 
        "tardbp-melamed-2019_tardbp", "tardbp-dafinca-2020_tardbp", "tardbp-smith-2021_tardbp", "sod1-tardbp-dash-2022_tardbp", "vcp-ziff-2023_tardbp",
        "sod1-moccia-2014_sod1", "sod1-tardbp-dash-2022_sod1", "sod1-bhinge-2017_sod1", "sod1-wang-2017_sod1",
        "fus-kapeli-2016_fus", "fus-desantis-2017_fus", "fus-hawkins-2022_fus", "c9orf72-catanese-2021_fus",
        "vcp-luisier-2018_vcp", "vcp-tyzack-2021_vcp", "vcp-ziff-2023_vcp"))

append_row_df <- public_data_frequency_df[1,]
append_row_df$xaxis <- "Boo_1"
append_row_df$xaxis <- factor(append_row_df$xaxis, levels = c("Boo_1"))
ipsc_frequency_df <- rbind(public_data_frequency_df, append_row_df, answerals_frequency_df)

```

- Plot data


```{r}


current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12, angle = 90, hjust = 1),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))


append_row_df <- public_data_frequency_df[1,]
append_row_df$xaxis <- "Boo_2"
append_row_df$xaxis <- factor(append_row_df$xaxis, levels = c("Boo_2"))
all_frequency_df <- rbind(ipsc_frequency_df, append_row_df, plot_nygc_frequency_df)

# colours <- c("public-data" = "#DF2935", "answerals" = "#51BBFE", "nygc" = "#FDCA40")
colours <- c("up" = "#F24236", "down" = "#F7F7F2")
ThePlot <- ggplot(all_frequency_df, 
    aes(x = xaxis, y = frequency))
ThePlot <- ThePlot + geom_col(aes(fill = direction), colour = "#000000")
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_discrete("study")
ThePlot <- ThePlot + scale_y_continuous("n of DE")
ThePlot <- ThePlot + scale_fill_manual("DE direction", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("Proportion of repeat containing introns DE")
ThePlot <- ThePlot + current_theme
ggsave(width = 10, height = 7, file = file.path(plot_dir, "02-02-supplementary-figure-1C.pdf"), ThePlot)


```

### 4.2 Plot n of significant differentially expressed repeat containing events


- Show significant differentially expressed repeat containing events under pval < 0.05

```{r}

repeat_containing_introns_df <- sig_result_df[sig_result_df$status == "repeat_containing" & 
    sig_result_df$region == "intron",]

repeat_frequency_df <- summarise_for_barplot(repeat_containing_introns_df, total_length = length(unique(feature_ranges_with_repeats$feature_id[feature_ranges_with_repeats$type == "intron"])))


plot_repeat_frequency_df <- repeat_frequency_df

plot_repeat_frequency_df$cohort <- "public-data"
plot_repeat_frequency_df$cohort[grepl(x = plot_repeat_frequency_df$study, pattern = "nygc")] <- "nygc"
plot_repeat_frequency_df$cohort[grepl(x = plot_repeat_frequency_df$study, pattern = "answerals")] <- "answerals"

plot_repeat_frequency_df$tissue_type <- "iPSCs"
plot_repeat_frequency_df$tissue_type[grepl(x = plot_repeat_frequency_df$study, pattern = "nygc")] <- "postmortem"

public_data_frequency_df <- plot_repeat_frequency_df[plot_repeat_frequency_df$cohort == "public-data",]
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_c9orf72_vs_ctrl"] <- "c9orf72"
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_fus_vs_ctrl"] <- "fus"
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_c9orf72_vs_iso"] <- "c9orf72"
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_sod1_vs_ctrl"] <- "sod1"
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_sod1_vs_iso"] <- "sod1"
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_vcp_vs_ctrl"] <- "vcp"
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_d35_mut_vs_wt"] <- "vcp"
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_tardbp_vs_wt"] <- "tardbp"
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_tardbp_vs_ctrl"] <- "tardbp"
public_data_frequency_df$contrast[public_data_frequency_df$contrast == "diff_expr_vcp_vs_wt"] <- "vcp"
public_data_frequency_df$xaxis <- paste(public_data_frequency_df$study, public_data_frequency_df$contrast, sep = "_")

answerals_frequency_df <- plot_repeat_frequency_df[plot_repeat_frequency_df$cohort == "answerals",]
answerals_frequency_df$contrast[answerals_frequency_df$contrast == "fals_vs_control"] <- "familial"
answerals_frequency_df$contrast[answerals_frequency_df$contrast == "sals_vs_control"] <- "sporadic"
answerals_frequency_df$xaxis <- paste(answerals_frequency_df$study, answerals_frequency_df$contrast, sep = "_")

nygc_frequency_df <- plot_repeat_frequency_df[plot_repeat_frequency_df$cohort == "nygc",]
nygc_frequency_df$contrast[nygc_frequency_df$contrast == "als_vs_control"] <- "als"
nygc_frequency_df$contrast[nygc_frequency_df$contrast == "als+ftd_vs_control"] <- "als+ftd"


nygc_frequency_df$xaxis <- paste(nygc_frequency_df$study, nygc_frequency_df$contrast, sep = "_")


answerals_frequency_df$xaxis <- factor(answerals_frequency_df$xaxis, 
    levels = c("answerals_familial", "answerals_sporadic"))

public_data_frequency_df$xaxis <- factor(public_data_frequency_df$xaxis, 
    levels = c("c9orf72-sareen-2013_c9orf72", "c9orf72-sterneckert-2020_c9orf72", "c9orf72-dafinca-2020_c9orf72", 
        "c9orf72-catanese-2021_c9orf72", "c9orf72-casteli-2021_c9orf72", "c9orf72-sommer-2022_c9orf72", 
        "tardbp-melamed-2019_tardbp", "tardbp-dafinca-2020_tardbp", "tardbp-smith-2021_tardbp", "sod1-tardbp-dash-2022_tardbp", "vcp-ziff-2023_tardbp",
        "sod1-moccia-2014_sod1", "sod1-tardbp-dash-2022_sod1", "sod1-bhinge-2017_sod1", "sod1-wang-2017_sod1",
        "fus-kapeli-2016_fus", "fus-desantis-2017_fus", "fus-hawkins-2022_fus", "c9orf72-catanese-2021_fus",
        "vcp-luisier-2018_vcp", "vcp-tyzack-2021_vcp", "vcp-ziff-2023_vcp"))

append_row_df <- public_data_frequency_df[1,]
append_row_df$xaxis <- "Boo_1"
append_row_df$xaxis <- factor(append_row_df$xaxis, levels = c("Boo_1"))
ipsc_frequency_df <- rbind(public_data_frequency_df, append_row_df, answerals_frequency_df)


plot_nygc_frequency_df <- nygc_frequency_df[nygc_frequency_df$contrast != "none",]
plot_nygc_frequency_df$xaxis <- factor(plot_nygc_frequency_df$xaxis, 
    levels = c("nygc_Cortex_als", "nygc_Cortex_als+ftd", 
        "nygc_Cortex_Motor_als", "nygc_Cortex_Motor_als+ftd",
        "nygc_Cerebellum_als", "nygc_Cerebellum_als+ftd",
        "nygc_Spinal_Cord_Cervical_als", "nygc_Spinal_Cord_Cervical_als+ftd",
        "nygc_Spinal_Cord_Thoracic_als", "nygc_Spinal_Cord_Thoracic_als+ftd",
        "nygc_Spinal_Cord_Lumbar_als", "nygc_Spinal_Cord_Lumbar_als+ftd"))


append_row_df <- public_data_frequency_df[1,]
append_row_df$xaxis <- "Boo_2"
append_row_df$xaxis <- factor(append_row_df$xaxis, levels = c("Boo_2"))
all_frequency_df <- rbind(ipsc_frequency_df, append_row_df, plot_nygc_frequency_df)

```

- Plot data


```{r}


current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12, angle = 90, hjust = 1),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

# colours <- c("public-data" = "#DF2935", "answerals" = "#51BBFE", "nygc" = "#FDCA40")
colours <- c("up" = "#F24236", "down" = "#F7F7F2")
ThePlot <- ggplot(all_frequency_df, 
    aes(x = xaxis, y = proportion))
ThePlot <- ThePlot + geom_col(aes(fill = direction),  colour = "#000000")
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_discrete("study")
ThePlot <- ThePlot + scale_y_continuous("proportion")
ThePlot <- ThePlot + scale_fill_manual("DE direction", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("Proportion of repeat containing introns DE")
ThePlot <- ThePlot + current_theme
ggsave(width = 10, height = 7, file = file.path(plot_dir, "02-03-supplementary-figure-1D.pdf"), ThePlot)


```
