---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

We are investigating the differential expression profiling statistics of DE introns. 


### 1.0 Package requirements

- change to current project root directory and include required packages 


```{r include = TRUE, message = FALSE, warning = FALSE}

library("here")
message("Project root:", here(), "\n")
require("parallel")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("Biostrings")
require("rtracklayer")
require("GenomeInfoDb")
require("GenomicFeatures")
require("ggplot2")

```

### 2.0 Include experiment matrix

- Import data

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("final-figures","figure-1")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

feature_ranges <- readRDS(here("05-simulated-reads-differential-profiling","rdata_files","01-all-merged-exonic-and-intronic-segments-in-genes.rds"))
sig_result_list <- readRDS(here("20-import-significant-events","rdata_files","02-significantly-differentially-expressed-events.rds"))
repeat_ranges_list <- readRDS(here("final-figures","figure-1","rdata_files","00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds"))

```

### 3.1.2 Report the number of repeats overlapping introns or exons

```{r}

feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges
sig_results_df <- sig_result_list$significant_events
sig_results_df <- sig_results_df[sig_results_df$region == "intron",]

sig_results_df$tissue_type[sig_results_df$tissue_type != "iPSCs"] <- "postmortem"

unique_intron_df <- unique(sig_results_df[,c("tissue_type", "feature_ids", "status")])
frequency_df <- as.data.frame(table(unique_intron_df[,"tissue_type"]))
colnames(frequency_df) <- c("tissue_type", "frequency")
frequency_df$tissue_type <- factor(frequency_df$tissue_type, levels = c("iPSCs", "postmortem"))

```

- Plot frequency profile for all de introns.

```{r}

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12, angle = 90, hjust = 1),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

# "AnswerALS" = "#51BBFE"
colours <- c("up" = "#F24236", "down" = "#F7F7F2")
colours <- c("iPSCs" = "#DF2935" , "postmortem" = "#FDCA40")
ThePlot <- ggplot(frequency_df, 
    aes(x = tissue_type, y = frequency))
ThePlot <- ThePlot + geom_col(aes(fill = tissue_type), colour = "#000000")
ThePlot <- ThePlot + geom_text(aes(label = frequency), colour = "#000000")
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_discrete("tissue_type")
ThePlot <- ThePlot + scale_y_continuous("frequency")
ThePlot <- ThePlot + scale_fill_manual("tissue_type", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("Number of differentially expressed introns")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 7, file = file.path(plot_dir, "02-01-figure-1B.pdf"), ThePlot)


```


- Plot frequency profile for repeat containing de introns.

```{r}


frequency_df <- as.data.frame(table(unique_intron_df[unique_intron_df$status == "repeat_containing","tissue_type"]))
frequency_df$proportion <- frequency_df$Freq / length(unique(feature_ranges_with_repeats$hash_id))
colnames(frequency_df) <- c("tissue_type", "frequency", "proportion")
frequency_df$tissue_type <- factor(frequency_df$tissue_type, levels = c("iPSCs", "postmortem"))

colours <- c("up" = "#F24236", "down" = "#F7F7F2")
colours <- c("iPSCs" = "#DF2935" , "postmortem" = "#FDCA40")
ThePlot <- ggplot(frequency_df, 
    aes(x = tissue_type, y = proportion))
ThePlot <- ThePlot + geom_col(aes(fill = tissue_type), colour = "#000000")
ThePlot <- ThePlot + geom_text(aes(label = frequency), colour = "#000000")
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_discrete("tissue_type")
ThePlot <- ThePlot + scale_y_continuous("proportion")
ThePlot <- ThePlot + scale_fill_manual("tissue_type", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("Number of differentially expressed introns")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 7, file = file.path(plot_dir, "02-02-figure-1C.pdf"), ThePlot)


```

- Write output listing all DE RCIs and the repeats within these RCIs

```{r}

reporting_df <- unique(sig_results_df[,c("feature_ids", "gene_id", "gene_name", "status", "tissue_type")])

unique_report_df <- do.call(rbind, lapply(split(reporting_df, reporting_df$feature_ids), function(x){
    return_df <- unique(x[,c(1,2,3,4)])
    return_df$de_in_ipsc <- "iPSCs" %in% x$tissue_type
    return_df$de_in_postmortem <- "postmortem" %in% x$tissue_type
    return(return_df)
}))
rownames(unique_report_df) <- NULL
write.table(unique_report_df, file = file.path(output_dir, "01-03-differentially-expressed-introns.txt"), sep = "\t")

repeat_ranges <- repeat_ranges_list$repeat_ranges

de_rci_ranges <- feature_ranges_with_repeats[feature_ranges_with_repeats$feature_id %in% sig_results_df$feature_ids]
elementMetadata(de_rci_ranges) <- elementMetadata(de_rci_ranges)[!(names(elementMetadata(de_rci_ranges)) %in% 
    c("tx_id", "tx_name", "exon_id", "exon_name", "exon_rank"))]
de_repeat_ranges <- repeat_ranges[de_rci_ranges$which_repeat]
de_repeat_ranges$hash_id <- de_rci_ranges$hash_id
de_rci_ranges$repeat_id <- names(de_repeat_ranges)
names(de_repeat_ranges) <- NULL

write.table(as.data.frame(de_rci_ranges), file = file.path(output_dir, "01-04-differentially-expressed-repeat-containing-introns.txt"), sep = "\t")
write.table(as.data.frame(de_repeat_ranges), file = file.path(output_dir, "01-05-repeats-inside-differentially-expressed-repeat-containing-introns.txt"), sep = "\t")

```
