---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

Investigate length distribution of microsatellites


### 1.0 Package requirements



- change to current working directory `/camp/home/palk/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
require("GenomicFeatures")
require("ggplot2")
require("dplyr")

```

### 2.0 Include experiment matrix

- Setup directories

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-1"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
source_dir <- file.path(analysis_path, "src")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

# Function declaration


```{r}

require("rtracklayer")
require("GenomicRanges")
require("parallel")
require("stringr")
# source(file.path())
# phylop_cons_tracks <- file.path(input_dir, "hg38.phyloP7way.bw")
# save_file <- file.path(save_dir, "01-save-conservation-tracks.rds")
# source(file.path(source_dir, "source.R"))

```


### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
feature_ranges <- readRDS(file.path("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds"))
sig_result_list <- readRDS(file.path("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds"))

repeat_colour_groups <- c("trinucleotide" = "#EEB902", "quadnucleotide" = "#97CC04", 
    "pentanucleotide" = "#F45D01", "hexanucleotide" = "#2D7DD2")

```

- Select repeat containing introns

```{r}

repeat_ranges <- unique(repeat_ranges_list$repeat_ranges)
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges


unique_feature_id_df <- unique(data.frame(feature_id = feature_ranges_with_repeats$feature_id,
    repeat_id = names(repeat_ranges)[feature_ranges_with_repeats$which_repeat],
    type = feature_ranges_with_repeats$type))
unique_feature_id_df$span_group <- repeat_ranges[unique_feature_id_df$repeat_id]$span_group
unique_feature_id_df$ssr <- repeat_ranges[unique_feature_id_df$repeat_id]$ssr
unique_feature_id_df$repeat_width <- width(repeat_ranges[unique_feature_id_df$repeat_id])
unique_feature_id_df$repeat_unit_length <- str_length(unique_feature_id_df$ssr)
unique_feature_id_df$expansion_size <- unique_feature_id_df$repeat_width / unique_feature_id_df$repeat_unit_length


```

- Plot proportion of repeats in introns

```{r}

sig_results_df <- sig_result_list$significant_events
sig_results_df <- sig_results_df[sig_results_df$region == "intron",]

repeat_containing_introns_df <- sig_results_df[sig_results_df$status == "repeat_containing",]

unique_feature_id_df$differentially_expressed <- unique_feature_id_df$feature_id %in% repeat_containing_introns_df$feature_ids
rci_de_intron_df <- unique_feature_id_df[unique_feature_id_df$differentially_expressed & unique_feature_id_df$type == "intron",]

class_freq_df <- as.data.frame(table(rci_de_intron_df$span_group))
class_freq_df$proportion <- class_freq_df$Freq / sum(class_freq_df$Freq)
class_freq_df$type <- "intron"

#              Var1 Freq proportion   type
# 1  hexanucleotide 2081 0.33521263 intron
# 2 pentanucleotide  186 0.02996134 intron
# 3  quadnucleotide  222 0.03576031 intron
# 4   trinucleotide 3719 0.59906572 intron


current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

ThePlot <- ggplot(class_freq_df, 
    aes(x = type, y = proportion))
ThePlot <- ThePlot + geom_col(aes(fill = Var1), colour = "#000000")
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_discrete("Type")
ThePlot <- ThePlot + scale_y_continuous("proportion")
ThePlot <- ThePlot + scale_fill_manual("Repeat colours", values = repeat_colour_groups)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("Proportion of repeats in each repeat class")
ThePlot <- ThePlot + current_theme
ggsave(width = 3, height = 7, file = file.path(plot_dir, "05-01-Supplementary-figure-4A.pdf"), ThePlot)

```

- Plot proportion of repeat width in introns

```{r}

total_repeat_width <- sum(rci_de_intron_df$repeat_width)
class_specific_width <- vapply(split(rci_de_intron_df$repeat_width, rci_de_intron_df$span_group), sum, 1)
class_specific_expansion_width <- vapply(split(rci_de_intron_df$expansion_size, rci_de_intron_df$span_group), sum, 1)

width_prop_df <- data.frame(span_group = names(class_specific_width),
    width = class_specific_width)
width_prop_df$prop <- width_prop_df$width / total_repeat_width
width_prop_df$type <- "intron"

#                      span_group width       prop   type
# hexanucleotide   hexanucleotide 46062 0.32639150 intron
# pentanucleotide pentanucleotide  6335 0.04488928 intron
# quadnucleotide   quadnucleotide  4968 0.03520283 intron
# trinucleotide     trinucleotide 83760 0.59351639 intron


ThePlot <- ggplot(width_prop_df, 
    aes(x = type, y = prop))
ThePlot <- ThePlot + geom_col(aes(fill = span_group), colour = "#000000")
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_discrete("Type")
ThePlot <- ThePlot + scale_y_continuous("proportion")
ThePlot <- ThePlot + scale_fill_manual("Repeat colours", values = repeat_colour_groups)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("Proportion of repeat width in each class")
ThePlot <- ThePlot + current_theme
ggsave(width = 3, height = 7, file = file.path(plot_dir, "05-02-supplementary-figure-4B.pdf"), ThePlot)

```

- Plot proportion of expansion width in introns

```{r}

total_expansion_width <- sum(rci_de_intron_df$expansion_size)
expansion_prop_df <- data.frame(span_group = names(class_specific_expansion_width),
    width = class_specific_expansion_width)
expansion_prop_df$prop <- expansion_prop_df$width / total_expansion_width
expansion_prop_df$type <- "intron"

#                      span_group width       prop   type
# hexanucleotide   hexanucleotide  7677 0.20146434 intron
# pentanucleotide pentanucleotide  1267 0.03324936 intron
# quadnucleotide   quadnucleotide  1242 0.03259329 intron
# trinucleotide     trinucleotide 27920 0.73269301 intron

ThePlot <- ggplot(expansion_prop_df, 
    aes(x = type, y = prop))
ThePlot <- ThePlot + geom_col(aes(fill = span_group), colour = "#000000")
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_discrete("Type")
ThePlot <- ThePlot + scale_y_continuous("proportion")
ThePlot <- ThePlot + scale_fill_manual("Repeat colours", values = repeat_colour_groups)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("Proportion of expansion width in each class")
ThePlot <- ThePlot + current_theme
ggsave(width = 3, height = 7, file = file.path(plot_dir, "05-03-supplementary-figure-4C.pdf"), ThePlot)

```

- Plot repeat length of repeats

```{r}

ThePlot <- ggplot(rci_de_intron_df, 
    aes(x = log10(repeat_width)))
ThePlot <- ThePlot + geom_density(aes(fill = span_group), colour = "#000000", alpha = 0.6)
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_continuous("repeat length")
ThePlot <- ThePlot + scale_y_continuous("density")
ThePlot <- ThePlot + scale_fill_manual("Repeat colours", values = repeat_colour_groups)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("Proportion of repeat containing introns DE")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 7, file = file.path(plot_dir, "05-04-supplementary-figure-4D.pdf"), ThePlot)


```





