---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

Investigate length distribution of microsatellites


### 1.0 Package requirements



- change to current working directory `/camp/home/palk/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
require("GenomicFeatures")
require("ggplot2")
require("dplyr")

```

### 2.0 Include experiment matrix

- Setup directories

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-1"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
source_dir <- file.path(analysis_path, "src")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

# Function declaration


```{r}

require("rtracklayer")
require("GenomicRanges")
require("parallel")
# source(file.path())
# phylop_cons_tracks <- file.path(input_dir, "hg38.phyloP7way.bw")
# save_file <- file.path(save_dir, "01-save-conservation-tracks.rds")
# source(file.path(source_dir, "source.R"))

```


### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
feature_ranges <- readRDS(file.path("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds"))
sig_result_list <- readRDS(file.path("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds"))
sequence_complexity_ranges <- readRDS(file.path("30-sequence-analysis-of-introns/rdata_files/07-02-save-sequence-complexity-ranges-of-all-differentially-expressed-sequences.rds"))

repeat_colour_groups <- c("trinucleotide" = "#EEB902", "quadnucleotide" = "#97CC04", 
    "pentanucleotide" = "#F45D01", "hexanucleotide" = "#2D7DD2")

```

- clean up the sequence complexity ranges

```{r}

sequence_complexity_ranges <- sequence_complexity_ranges[!vapply(sequence_complexity_ranges, is.null, T)]
sequence_complexity_ranges_names <- vapply(sequence_complexity_ranges, function(x){
    unique(as.vector(seqnames(x)))
},"1")
names(sequence_complexity_ranges) <- sequence_complexity_ranges_names
sequence_complexity <- vapply(sequence_complexity_ranges, function(x){
    mean(x$complexity)
},1)
# 4261 were too small

```

- Assign sequence complexity values

```{r}

sig_results_df <- sig_result_list$significant_events
de_introns_df <- sig_results_df[sig_results_df$region == "intron",]

unique_plot_df <- unique(de_introns_df[,c("feature_ids", "status", "region")])
unique_plot_df$complexity <- sequence_complexity[unique_plot_df$feature_ids]
unique_plot_df <- unique_plot_df[!is.na(unique_plot_df$complexity),]
unique_plot_df$sequence_length <- width(feature_ranges)[match(unique_plot_df$feature_ids, feature_ranges$feature_id)]
# non_repeat_containing     repeat_containing
#                251756                  5045

```

- Plot sequence complexity scores


```{r}


current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

Colours <- c("repeat_containing" = "#ff6b6c", "non_repeat_containing" = "#f3f9e3")
ThePlot <- ggplot(unique_plot_df, 
    aes(x = complexity))
ThePlot <- ThePlot + geom_density(aes(fill = status), colour = "#000000", alpha = 0.6)
# ThePlot <- ThePlot + geom_boxplot(aes(fill = status), colour = "#000000", 
#     width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9))
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_continuous("log sequence complexity")
ThePlot <- ThePlot + scale_y_continuous("density")
ThePlot <- ThePlot + scale_fill_manual("Repeat colours", values = Colours)
# ThePlot <- ThePlot + facet_grid(. ~ region)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("sequence complexity of differentially expressed regions")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 7, file = file.path(plot_dir, "05-05-Supplementary-figure-2C.pdf"), ThePlot)


```


- Investigate association between sequence length and sequence complexity


```{r}

len_quantiles <- quantile(log10(unique_plot_df$sequence_length), seq(0.1,1,0.1))
unique_plot_df$len_q <- vapply(log10(unique_plot_df$sequence_length), function(x){
    names(which(len_quantiles >= x)[1])
},"1")
unique_plot_df$len_q <- factor(unique_plot_df$len_q, levels = names(len_quantiles))

Colours <- c("repeat_containing" = "#ff6b6c", "non_repeat_containing" = "#f3f9e3")
ThePlot <- ggplot(unique_plot_df, 
    aes(x = len_q, y = complexity))
ThePlot <- ThePlot + geom_violin(aes(fill = status), colour = "#000000", 
    lwd = 0.5, position = position_dodge(width = 0.9), scale = "width")
ThePlot <- ThePlot + geom_boxplot(aes(fill = status), colour = "#000000", 
    width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9))
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_discrete("log10 sequence length")
ThePlot <- ThePlot + scale_y_continuous("sequence complexity")
ThePlot <- ThePlot + scale_fill_manual("Repeat content", values = Colours)
ThePlot <- ThePlot + facet_grid(. ~ status)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("sequence complexity vs sequence length")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 5, file = file.path(plot_dir, "05-06-Supplementary-figure-2D.pdf"), ThePlot)


cor(log10(unique_plot_df$sequence_length)[unique_plot_df$status == "repeat_containing"], unique_plot_df$complexity[unique_plot_df$status == "repeat_containing"], method = "pearson")
# 0.4798721

summary(lm(data = unique_plot_df[unique_plot_df$status == "repeat_containing",], formula = complexity ~ log10(sequence_length)))
# Call:
# lm(formula = complexity ~ log10(sequence_length), data = unique_plot_df[unique_plot_df$status ==
#     "repeat_containing", ])

# Residuals:
#      Min       1Q   Median       3Q      Max
# -0.32069 -0.00869  0.00166  0.01271  0.11658

# Coefficients:
#                         Estimate Std. Error t value Pr(>|t|)
# (Intercept)            0.7298135  0.0021394  341.12   <2e-16 ***
# log10(sequence_length) 0.0195644  0.0005037   38.84   <2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

# Residual standard error: 0.02599 on 5043 degrees of freedom
# Multiple R-squared:  0.2303,    Adjusted R-squared:  0.2301
# F-statistic:  1509 on 1 and 5043 DF,  p-value: < 2.2e-16



cor(log10(unique_plot_df$sequence_length)[unique_plot_df$status != "repeat_containing"], unique_plot_df$complexity[unique_plot_df$status != "repeat_containing"], method = "pearson")

summary(lm(data = unique_plot_df[unique_plot_df$status != "repeat_containing",], formula = complexity ~ log10(sequence_length)))


```

- Construct repeat containing ranges

```{r}

repeat_ranges <- unique(repeat_ranges_list$repeat_ranges)
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges

```


- Investigate variance explained by repeat component


```{r}

omega_sq <- function(aovm){
    sum_stats <- summary(aovm)[[1]]
    SSm <- sum_stats[["Sum Sq"]][1]
    SSr <- sum_stats[["Sum Sq"]][2]
    DFm <- sum_stats[["Df"]][1]
    MSr <- sum_stats[["Mean Sq"]][2]
    W2 <- (SSm-DFm*MSr)/(SSm+SSr+MSr)
    return(W2)
}

stats_df <- unique_plot_df[unique_plot_df$status == "repeat_containing",]
stats_df$repeat_length <- vapply(seq_len(nrow(stats_df)), function(x){
    current_ranges <- feature_ranges_with_repeats[feature_ranges_with_repeats$feature_id == stats_df$feature_id[x]]
    current_repeats <- reduce(repeat_ranges[current_ranges$which_repeat])
    return(sum(width(GenomicRanges::intersect(current_repeats, current_ranges))))
},1)


stats_df$variance_explained <- vapply(seq_len(nrow(stats_df)),function(x){
    message(x)
    current_df <- stats_df[x,]
    current_complexity_ranges <- sequence_complexity_ranges[[current_df$feature_id]]
    current_ranges <- feature_ranges_with_repeats[feature_ranges_with_repeats$feature_id == current_df$feature_id]
    seqlevels(current_complexity_ranges) <- unique(as.vector(seqnames(current_ranges)))
    current_repeat <- repeat_ranges[current_ranges$which_repeat]
    ol_object <- findOverlaps(current_complexity_ranges, current_repeat)
    current_complexity_ranges$status <- "non_repeat_containing"
    current_complexity_ranges$status[queryHits(ol_object)] <- "repeat_containing"
    temp_df <- data.frame(complexity = current_complexity_ranges$complexity, 
        status = factor(current_complexity_ranges$status, levels = c("repeat_containing","non_repeat_containing")))
    if(length(unique(temp_df$status)) == 1){
        return(NA)
    }
    model_object <- aov(formula = complexity ~ status, data = temp_df)
    return(omega_sq(model_object))
},1)
stats_df$variance_explained[stats_df$variance_explained < 0] <- 0

```

- Plot variance explained by repeats


```{r}


current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

plot_stats_df <- stats_df[!is.na(stats_df$variance_explained),]
variance_density <- density(log10(plot_stats_df$variance_explained))

which_max <- variance_density$x[which.max(variance_density$y)]
# 10^which_max

nrow(plot_stats_df[plot_stats_df$variance_explained > 0.01,]) / nrow(plot_stats_df)
# [1] 0.1153846

Colours <- c("repeat_containing" = "#ff6b6c", "non_repeat_containing" = "#f3f9e3")
ThePlot <- ggplot(stats_df, 
    aes(x = log10(variance_explained)))
ThePlot <- ThePlot + geom_density(fill = "#ff6b6c", colour = "#000000")
ThePlot <- ThePlot + geom_vline(xintercept = which_max, lty = "dashed")
ThePlot <- ThePlot + geom_vline(xintercept = -2, lty = "dashed")
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_continuous("log10 percentage variance attributed to repeat component in introns")
ThePlot <- ThePlot + scale_y_continuous("density")
# ThePlot <- ThePlot + scale_fill_manual("Repeat content", values = Colours)
# ThePlot <- ThePlot + facet_grid(. ~ status)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("sequence complexity vs sequence length")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 7, file = file.path(plot_dir, "05-07-Supplementary-figure-2E.pdf"), ThePlot)


```



- Plot variance explained by repeats


```{r}

# gt_1perc_contrib_df <- plot_stats_df[plot_stats_df$variance_explained >= 10^which_max,]


Colours <- c("repeat_containing" = "#ff6b6c", "non_repeat_containing" = "#f3f9e3")
ThePlot <- ggplot(plot_stats_df[plot_stats_df$variance_explained > 0,], 
    aes(x = log10(sequence_length), y = log10(variance_explained)))
ThePlot <- ThePlot + geom_point(fill = "#ff6b6c", colour = "#000000", shape = 21 )
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_continuous("log10 intron length")
ThePlot <- ThePlot + scale_y_continuous("log10 percentage variance attributed to repeat component in introns")
# ThePlot <- ThePlot + scale_fill_manual("Repeat content", values = Colours)
# ThePlot <- ThePlot + facet_grid(. ~ status)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("sequence complexity vs sequence length")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 7, file = file.path(plot_dir, "05-08-Supplementary-figure-2F.pdf"), ThePlot)

cor(log10(plot_stats_df[plot_stats_df$variance_explained > 0,"variance_explained"]), log10(plot_stats_df[plot_stats_df$variance_explained > 0,"sequence_length"]))

# summary(lm(data = plot_stats_df[plot_stats_df$variance_explained > 0,], formula = log10(variance_explained) ~ log10(sequence_length)))
# Call:
# lm(formula = log10(variance_explained) ~ log10(sequence_length),
#     data = plot_stats_df[plot_stats_df$variance_explained > 0,
#         ])

# Residuals:
#     Min      1Q  Median      3Q     Max
# -3.7021 -0.4330  0.0768  0.5230  3.0171

# Coefficients:
#                        Estimate Std. Error t value Pr(>|t|)
# (Intercept)             0.79791    0.06666   11.97   <2e-16 ***
# log10(sequence_length) -0.92107    0.01565  -58.87   <2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

# Residual standard error: 0.7385 on 4266 degrees of freedom
# Multiple R-squared:  0.4482,    Adjusted R-squared:  0.4481
# F-statistic:  3465 on 1 and 4266 DF,  p-value: < 2.2e-16

```