---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("ggplot2")
require("ggvenn")
require("STRINGdb")
require("parallel")
require("GenomicFeatures")
require("Biostrings")
require("BSgenome.Hsapiens.UCSC.hg38")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-5"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
tmp_dir <- file.path(analysis_path, "temp")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)
dir.create(tmp_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
significant_de_event_list <- readRDS("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds")
final_rloop_ranges <- readRDS("30-sequence-analysis-of-introns/rdata_files/90-01-merged-rloop-ranges-in-differentially-expressed-introns.rds")
feature_ranges <- readRDS("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds")

```

- Select significantly affected genes

```{r}

significant_de_event_df <- significant_de_event_list$significant_events
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges
repeat_ranges <- repeat_ranges_list$repeat_ranges

rloop_split <- split(final_rloop_ranges, final_rloop_ranges$feature_id)
rloop_density_df <- do.call(rbind, mclapply(rloop_split, function(x){
    message(unique(x$feature_id))
    ret_df <- data.frame(feature_id = unique(x$feature_id),
        rloop_length = sum(width(reduce(x))),
        total_rloop_length = sum(width(x)),
        rloop_frequency = length(x))
    return(ret_df)
},mc.cores = 6))

rloop_density_df$mean_rloop_length <- rloop_density_df$total_rloop_length / rloop_density_df$rloop_frequency
rloop_density_df$density <- rloop_density_df$rloop_length / (width(feature_ranges)[match(rloop_density_df$feature_id, feature_ranges$feature_id)] / 1e3)
rloop_density_df$status <- ifelse(rloop_density_df$feature_id %in% feature_ranges_with_repeats$feature_id, "repeat_containing", "non_repeat_containing")

saveRDS(object = rloop_density_df, file = file.path(save_dir, "01-rloop-stats-dataframe.rds"))

```
