---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
source("src/colours.R")
source("../useful-functions/basepair_coverage_functions.R")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("ggplot2")
require("ggvenn")
require("STRINGdb")
require("parallel")
require("GenomicFeatures")
require("Biostrings")
require("BSgenome.Hsapiens.UCSC.hg38")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-5"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
tmp_dir <- file.path(analysis_path, "temp")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)
dir.create(tmp_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
significant_de_event_list <- readRDS("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds")
rloop_ranges <- readRDS("30-sequence-analysis-of-introns/rdata_files/90-01-merged-rloop-ranges-in-differentially-expressed-introns.rds")
feature_ranges <- readRDS("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds")
rloop_density_df <- readRDS("final-figures/figure-5/rdata_files/01-rloop-stats-dataframe.rds")

```

- Select significantly affected genes

```{r}

significant_de_event_df <- significant_de_event_list$significant_events
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges
repeat_ranges <- repeat_ranges_list$repeat_ranges

```

- Plot r-loops near repeats

```{r}

de_feature_ranges <- feature_ranges[feature_ranges$feature_id %in% significant_de_event_df$feature_id,]
de_feature_ranges_with_repeats <- feature_ranges_with_repeats[feature_ranges_with_repeats$feature_id %in% significant_de_event_df$feature_id,]

de_repeat_ranges <- repeat_ranges[de_feature_ranges_with_repeats$which_repeat]
de_repeat_ranges$which_intron <- seq_along(de_feature_ranges_with_repeats)
de_repeat_ranges$feature_id <- de_feature_ranges_with_repeats$feature_id[de_repeat_ranges$which_intron]
de_repeat_ranges <- de_repeat_ranges[de_repeat_ranges$feature_id %in% rloop_ranges$feature_id]
seqlevelsStyle(de_repeat_ranges) <- "UCSC"
start(de_repeat_ranges) <- start(de_repeat_ranges) - 1000
end(de_repeat_ranges) <- end(de_repeat_ranges) + 1000

rci_rloop_ranges <- rloop_ranges[rloop_ranges$feature_id %in% de_repeat_ranges$feature_id]



ol_object <- findOverlaps(rci_rloop_ranges, de_repeat_ranges)

rloop_density_df <- data.frame(rloop_index = queryHits(ol_object),
    repeat_index = subjectHits(ol_object),
    span_group = de_repeat_ranges$span_group[subjectHits(ol_object)])

rloop_freq_df <- as.data.frame(table(unique(rloop_density_df[,c("span_group", "repeat_index")])$span_group))
colnames(rloop_freq_df) <- c("span_group", "frequency")
rloop_freq_df$proportion <- rloop_freq_df$frequency / table(de_repeat_ranges$span_group)
# rloop_freq_df <- rbind(rloop_freq_df, data.frame(span_group = "undetected",
#   frequency = sum(table(de_repeat_ranges$span_group)) - sum(rloop_freq_df$frequency),
#   proportion = 1 - sum(rloop_freq_df$proportion)))
# rloop_freq_df$region <- "intron"

rloop_freq_df$span_group <- factor(rloop_freq_df$span_group, levels = c("trinucleotide", "quadnucleotide", "pentanucleotide", "hexanucleotide"))
# repeat_colour_groups <- c("undetected" = "#C9CAD9", repeat_colour_groups)
ThePlot <- ggplot(rloop_freq_df, aes(x = span_group, y = proportion))
ThePlot <- ThePlot + geom_col(aes(fill = span_group), colour = "#000000")
ThePlot <- ThePlot + scale_x_discrete("Intron group")
ThePlot <- ThePlot + scale_y_continuous("proportion of repeats containing predicted r-loops within 1kb")
ThePlot <- ThePlot + scale_fill_manual("Intron groups", values = repeat_colour_groups)
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("R-loop density in introns")
ThePlot <- ThePlot + theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))
ggsave(width = 5, height = 10, file = file.path(plot_dir, "02-06-plot-proportion-of-rloops-within-1kb-of-repeat.pdf"), ThePlot)

```

- Plot distance to nearest repeat 

```{r}


dist_object <- distanceToNearest(de_repeat_ranges, rci_rloop_ranges)

rloop_dist_df <- data.frame(
    distance = elementMetadata(dist_object)[["distance"]],
    span_group = de_repeat_ranges$span_group[queryHits(dist_object)])
rloop_dist_df$dist_binned <- ceiling(rloop_dist_df$distance / 1e3)
cumsum_dist_df <- do.call(rbind, lapply(split(rloop_dist_df, rloop_dist_df$span_group), function(temp_df){
    freq_dist_df <- as.data.frame(table(temp_df$dist_binned))
    freq_dist_df[,1] <- as.numeric(as.character(freq_dist_df[,1]))
    freq_dist_df$cumsum <- cumsum(freq_dist_df[,2])
    freq_dist_df$prop <- freq_dist_df$cumsum / max(freq_dist_df$cumsum)
    freq_dist_df$distance <- as.numeric(freq_dist_df[,1]) * 1000
    freq_dist_df$span_group <- unique(temp_df$span_group)
    return(freq_dist_df)
}))

cumsum_dist_df_split <- split(cumsum_dist_df, cumsum_dist_df$span_group)
max_occurence <- vapply(cumsum_dist_df_split, function(x){
    max(x[,3])
},1)


ThePlot <- ggplot(cumsum_dist_df[cumsum_dist_df$distance > 0,], aes(x = log10(distance), y = prop))
ThePlot <- ThePlot + geom_line(aes(colour = span_group))
ThePlot <- ThePlot + geom_vline(xintercept = c(log10(1e3), log10(1e4)), linetype = "dashed")
ThePlot <- ThePlot + scale_x_continuous("Distance to closest repeat")
ThePlot <- ThePlot + scale_y_continuous("proportion of rloops within distance")
ThePlot <- ThePlot + scale_colour_manual("Intron groups", values = repeat_colour_groups)
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("R-loop distance distribution to repeats")
ThePlot <- ThePlot + theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))
ggsave(width = 5, height = 10, file = file.path(plot_dir, "02-07-plot-distance-of-rloops-to-closest-repeats-in-introns.pdf"), ThePlot)



```
