---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("ggplot2")
require("ggbeeswarm")
require("ggvenn")
require("STRINGdb")
require("parallel")
require("GenomicFeatures")
require("Biostrings")
source("../useful-functions/plot-utils.R")
source("../useful-functions/basepair_coverage_functions.R")
source("src/colours.R")


```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-5"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
tmp_dir <- file.path(analysis_path, "temp")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)
dir.create(tmp_dir, recursive = T)


```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
sig_result_list <- readRDS("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds")
feature_ranges <- readRDS("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds")
annotation_gtf <- "/nemo/lab/patanir/home/users/palk/Genomes/ENSEMBL_hg38/Homo_sapiens.GRCh38.104.chr_patch_hapl_scaff.gtf"
rloop_ranges <- readRDS("30-sequence-analysis-of-introns/rdata_files/90-01-merged-rloop-ranges-in-differentially-expressed-introns.rds")

```

- Select significantly affected genes

```{r}

repeat_ranges <- repeat_ranges_list$repeat_ranges
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges
sig_result_df <- sig_result_list$significant_events
sig_result_df <- sig_result_df[sig_result_df$region == "intron",]

```


- Isolate differentially expressed features and change seqlevels to ucsc.

```{r}

de_feature_ranges <- feature_ranges[feature_ranges$feature_id %in% sig_result_df$feature_ids,]

de_feature_ranges$feature_type <- ifelse(de_feature_ranges$feature_id %in% feature_ranges_with_repeats$feature_id, "repeat_containing", "non_repeat_containing")

rloop_ranges$feature_type <- de_feature_ranges$feature_type[match(rloop_ranges$feature_id, de_feature_ranges$feature_id)]
repeat_containing_rloop_ranges <- rloop_ranges[rloop_ranges$feature_type == "repeat_containing"]
repeat_containing_rloop_ranges$which_repeat <- feature_ranges_with_repeats$which_repeat[match(repeat_containing_rloop_ranges$feature_id, feature_ranges_with_repeats$feature_id)]

de_feature_ranges_with_repeats <- feature_ranges_with_repeats[
    feature_ranges_with_repeats$feature_id %in% sig_result_df$feature_ids & 
    feature_ranges_with_repeats$type == "intron"]
# de_repeat_ranges <- repeat_ranges[de_feature_ranges_with_repeats$which_repeat]

dist_from_start <- start(repeat_ranges)[de_feature_ranges_with_repeats$which_repeat] - start(de_feature_ranges_with_repeats)
dist_from_end <- end(de_feature_ranges_with_repeats) - end(repeat_ranges)[de_feature_ranges_with_repeats$which_repeat]

de_feature_ranges_with_repeats$within <- dist_from_start > 0 & dist_from_end > 0

within_de_feature_ranges <- de_feature_ranges_with_repeats[de_feature_ranges_with_repeats$within]



```

- Create R-loop metaprofile

```{r}


repeat_start_matrix_list <- lapply(seq_along(within_de_feature_ranges), function(x){
    message(x)
    current_feature_ranges <- within_de_feature_ranges[x]
    current_repeat_ranges <- repeat_ranges[current_feature_ranges$which_repeat]
    if((start(current_repeat_ranges) - 1000) < start(current_feature_ranges) | (end(current_repeat_ranges) + 1000) > end(current_feature_ranges)){
        return(NULL)
    }
    current_ranges <- current_repeat_ranges
    if(as.vector(strand(current_ranges)) == "+"){
        end(current_ranges) <- start(current_ranges)
        start(current_ranges) <- start(current_ranges) - 1000
        start_vector <- .return_overlap_weighted_score_between_ranges(current_ranges, bin_to = 100, repeat_containing_rloop_ranges)
    }
    if(as.vector(strand(current_ranges)) == "-"){
        start(current_ranges) <- end(current_ranges)
        end(current_ranges) <- end(current_ranges) + 1000
        start_vector <- rev(.return_overlap_weighted_score_between_ranges(current_ranges, bin_to = 100, repeat_containing_rloop_ranges))
    }
    return(start_vector)
})
names(repeat_start_matrix_list) <- names(repeat_ranges)[within_de_feature_ranges$which_repeat]
repeat_start_matrix <- do.call(rbind, repeat_start_matrix_list)

repeat_start_means <- colMeans(repeat_start_matrix)
repeat_start_df <- data.frame(position = seq(-99, 0), value = repeat_start_means)


repeat_end_matrix_list <- lapply(seq_along(within_de_feature_ranges), function(x){
    message(x)
    current_feature_ranges <- within_de_feature_ranges[x]
    current_repeat_ranges <- repeat_ranges[current_feature_ranges$which_repeat]
    if((start(current_repeat_ranges) - 1000) < start(current_feature_ranges) | (end(current_repeat_ranges) + 1000) > end(current_feature_ranges)){
        return(NULL)
    }
    current_ranges <- current_repeat_ranges
    if(as.vector(strand(current_ranges)) == "+"){
        start(current_ranges) <- end(current_ranges)
        end(current_ranges) <- end(current_ranges) + 1000
        end_vector <- .return_overlap_weighted_score_between_ranges(current_ranges, bin_to = 100, repeat_containing_rloop_ranges)
    }
    if(as.vector(strand(current_ranges)) == "-"){
        end(current_ranges) <- start(current_ranges)
        start(current_ranges) <- start(current_ranges) - 1000
        end_vector <- rev(.return_overlap_weighted_score_between_ranges(current_ranges, bin_to = 100, repeat_containing_rloop_ranges))
    }
    return(end_vector)
})
names(repeat_end_matrix_list) <- names(repeat_ranges)[within_de_feature_ranges$which_repeat]
repeat_end_matrix <- do.call(rbind, repeat_end_matrix_list)


```

- Plot R-loop metaprofile


```{r}


upstream_df <- do.call(rbind, lapply(rownames(repeat_start_matrix), function(x){
    data.frame(repeat_id = x, 
        type = "upstream",
        repeat_class = repeat_ranges[x]$span_group,
        repeat_type = repeat_ranges[x]$ssr,
        position = seq(-100, -1),
        values = repeat_start_matrix[x,])
}))

downstream_df <- do.call(rbind, lapply(rownames(repeat_end_matrix), function(x){
    data.frame(repeat_id = x, 
        type = "downstream",
        repeat_class = repeat_ranges[x]$span_group,
        repeat_type = repeat_ranges[x]$ssr,
        position = seq(1, 100),
        values = repeat_end_matrix[x,])
}))


all_df <- rbind(upstream_df, downstream_df)



ThePlot <- ggplot(all_df, aes(x = position, y = values))
ThePlot <- ThePlot + stat_summary(aes(y = values, colour = repeat_class), fun.y = mean, geom="line") 
# ThePlot <- ThePlot + geom_smooth(method = "lm", formula = y ~ splines::bs(x, 5), aes(colour = repeat_class))
ThePlot <- ThePlot + scale_x_continuous("Position")
ThePlot <- ThePlot + scale_y_continuous("average r-loop profile")
ThePlot <- ThePlot + scale_colour_manual("Intron groups", values = repeat_colour_groups)
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_grid(. ~ type, scales = "free")
ThePlot <- ThePlot + ggtitle("Frequency of gquads in introns")
ThePlot <- ThePlot + theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))
ggsave(width = 7, height = 7, file = file.path(plot_dir, "04-02-plot-rloops-coverage-near-repeats-in-introns-figure-5C.pdf"), ThePlot)


```
