---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("ggplot2")
require("ggbeeswarm")
require("ggvenn")
require("STRINGdb")
require("parallel")
require("GenomicFeatures")
require("Biostrings")


```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-5"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
tmp_dir <- file.path(analysis_path, "temp")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)
dir.create(tmp_dir, recursive = T)


```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
sig_result_list <- readRDS("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds")
feature_ranges <- readRDS("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds")
annotation_gtf <- "/nemo/lab/patanir/home/users/palk/Genomes/ENSEMBL_hg38/Homo_sapiens.GRCh38.104.chr_patch_hapl_scaff.gtf"
rloop_ranges <- readRDS("30-sequence-analysis-of-introns/rdata_files/90-01-merged-rloop-ranges-in-differentially-expressed-introns.rds")
dripseq_samplesheet_df <- read.csv("/nemo/project/proj-luscombn-patani/working/public-data/sporadic-pbmc-drip-fero-et-al-2024/sample-details/sampleinfo.csv")
dripseq_aln_dir <- file.path("/nemo/project/proj-luscombn-patani/working/public-data/sporadic-pbmc-drip-fero-et-al-2024/output_files/bigwig/")
dripseq_list <- readRDS("34-dripseq-analysis/rdata_files/02-01-list-of-total-rloop-coverage-matrix-and-library-sizes.rds")

```

- Select significantly affected genes

```{r}

repeat_ranges <- repeat_ranges_list$repeat_ranges
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges

library_sizes <- dripseq_list$library_sizes
intronic_signal_matrix <- dripseq_list$feature_signal_matrix
libnorm_signal <- t(t(intronic_signal_matrix) / (library_sizes[1,] / 1e6))
mean_signal <- rowMeans(libnorm_signal)

reduced_rloop_signal_matrix <- dripseq_list$reduced_rloop_signal_matrix
prop_rloop_signal_matrix <- reduced_rloop_signal_matrix / intronic_signal_matrix[rownames(reduced_rloop_signal_matrix),]
prop_rloop_signal_matrix[is.na(prop_rloop_signal_matrix)] <- 0
mean_prop_signal <- rowMeans(prop_rloop_signal_matrix)

```

- Generate signal matrices

```{r}

reduced_rloop_ranges <- dripseq_list$reduced_rloop_ranges

num_rloops <- table(rloop_ranges$feature_id)
reduced_rloop_ranges_split <- split(width(reduced_rloop_ranges), reduced_rloop_ranges$feature_id)
width_in_rloop <- do.call(c, mclapply(reduced_rloop_ranges_split, sum, mc.cores = 8))
prop_width_in_rloop <- width_in_rloop / width(feature_ranges)[match(names(width_in_rloop), feature_ranges$feature_id)]


```

- Prepare data.frame for CDF computation

```{r}


feature_signal_df <- data.frame(feature_id = rownames(intronic_signal_matrix), 
    prop_width = 0, frequency = 0, prop_signal = 1)
feature_signal_df$prop_width[match(names(prop_width_in_rloop), feature_signal_df$feature_id)] <- prop_width_in_rloop
feature_signal_df$frequency[match(names(num_rloops), feature_signal_df$feature_id)] <- num_rloops
feature_signal_df$prop_signal[match(names(mean_prop_signal), feature_signal_df$feature_id)] <- mean_prop_signal

feature_signal_df$hash_id <- feature_ranges$hash_id[match(feature_signal_df$feature_id, feature_ranges$feature_id)]
feature_signal_df$signal <- mean_signal[feature_signal_df$feature_id]
feature_signal_df$signal_scaled_to_tot <- feature_signal_df$signal / sum(feature_signal_df$signal)
rownames(feature_signal_df) <- NULL

intron_feature_signal_df <- feature_signal_df[grepl(x = feature_signal_df$feature_id, pattern = "intron"),]

intron_feature_signal_df$status <- ifelse(intron_feature_signal_df$hash_id %in% feature_ranges_with_repeats$hash_id, 
    "repeat_containing","non_repeat_containing")
intron_feature_signal_df$pred_group <- ifelse(intron_feature_signal_df$prop_width > 0, "pred_rloop", "no_rloop")

```


- Get CDF


```{r}

introns_with_predicted_rloops_df <- intron_feature_signal_df[intron_feature_signal_df$signal > 0,]
quantiles <- quantile(log10(introns_with_predicted_rloops_df$signal), seq(0,1,0.01))
return_df_l1 <- do.call(rbind, lapply(unique(introns_with_predicted_rloops_df$status), function(x){
    return_df_l2 <- do.call(rbind, lapply(unique(introns_with_predicted_rloops_df$pred_group), function(y){
        cumulative_values <- vapply(quantiles, function(z){
            cumulative_val <- nrow(introns_with_predicted_rloops_df[
                    introns_with_predicted_rloops_df$status == x &
                    introns_with_predicted_rloops_df$pred_group == y &
                    log10(introns_with_predicted_rloops_df$signal) <= z,]) / nrow(introns_with_predicted_rloops_df[introns_with_predicted_rloops_df$status == x &
                        introns_with_predicted_rloops_df$pred_group == y,])
        },1)
        cdf_df <- data.frame(signal = quantiles, 
            cdf = cumulative_values, 
            status = x, pred_group = y)
        return(cdf_df)
    }))
}))


# ks test to compare the two curves

do.call(rbind, lapply(unique(introns_with_predicted_rloops_df$status), function(x){
    current_df <- return_df_l1[return_df_l1$status == x,]
    pval <- ks.test(x = current_df$cdf[current_df$pred_group == "no_rloop"],
        y = current_df$cdf[current_df$pred_group == "pred_rloop"])$p.value
    data.frame(status = x,
        p.value = pval)
}))
#                  status      p.value
# 1 non_repeat_containing 0.0001474968
# 2     repeat_containing 0.2152773326

```

- Plot CDF of repeats


```{r}

# intron_feature_signal_df$pred_group <- ifelse(intron_feature_signal_df$prop_width > 0, "pred_rloop", "no_rlopp")
Colours <- c("pred_rloop" = "#F05D5E", "no_rloop" = "#058ED9")
ThePlot <- ggplot(return_df_l1, aes(x = signal, y = cdf))
ThePlot <- ThePlot + geom_step(aes(colour = pred_group))
ThePlot <- ThePlot + scale_x_continuous("log10 mean library size normalised rloop signal within introns")
ThePlot <- ThePlot + scale_y_continuous("CDF")
ThePlot <- ThePlot + scale_colour_manual("Rloop prediction", values = Colours)
# ThePlot <- ThePlot + scale_colour_manual("status", values = Colours)
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_grid(. ~ status, scales = "free")
ThePlot <- ThePlot + ggtitle("library size normalised R-loop signal")
ThePlot <- ThePlot + theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))
ggsave(width = 12, height = 7, file = file.path(plot_dir, "04-03-plot-cdf-of-rloop-signal-within-introns-containing-predicted-vs-non-predicted-rloops-figure-5D.pdf"), ThePlot)


```
