---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("ggplot2")
require("ggbeeswarm")
require("ggvenn")
require("STRINGdb")
require("parallel")
require("GenomicFeatures")
require("Biostrings")


```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-5"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
tmp_dir <- file.path(analysis_path, "temp")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)
dir.create(tmp_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
sig_result_list <- readRDS("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds")
feature_ranges <- readRDS("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds")
annotation_gtf <- "/nemo/lab/patanir/home/users/palk/Genomes/ENSEMBL_hg38/Homo_sapiens.GRCh38.104.chr_patch_hapl_scaff.gtf"
rloop_ranges <- readRDS("30-sequence-analysis-of-introns/rdata_files/90-01-merged-rloop-ranges-in-differentially-expressed-introns.rds")
dripseq_samplesheet_df <- read.csv("/nemo/project/proj-luscombn-patani/working/public-data/sporadic-pbmc-drip-fero-et-al-2024/sample-details/sampleinfo.csv")
dripseq_aln_dir <- file.path("/nemo/project/proj-luscombn-patani/working/public-data/sporadic-pbmc-drip-fero-et-al-2024/output_files/bigwig/")
dripseq_list <- readRDS("34-dripseq-analysis/rdata_files/02-01-list-of-total-rloop-coverage-matrix-and-library-sizes.rds")

```

- Select significantly affected genes


```{r}


repeat_ranges <- repeat_ranges_list$repeat_ranges
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges

sig_result_df <- sig_result_list$significant_events
sig_result_df <- sig_result_df[sig_result_df$region == "intron",]

de_feature_ranges <- feature_ranges[feature_ranges$feature_id %in% sig_result_df$feature_ids,]
de_feature_ranges_with_repeats <- feature_ranges_with_repeats[
    feature_ranges_with_repeats$feature_id %in% sig_result_df$feature_ids & 
    feature_ranges_with_repeats$type == "intron"]

```

- Generate signal matrices

```{r}

library_sizes <- dripseq_list$library_sizes
intronic_signal_matrix <- dripseq_list$feature_signal_matrix
libnorm_signal <- t(t(intronic_signal_matrix) / (library_sizes[1,] / 1e6))
mean_signal <- rowMeans(libnorm_signal)

reduced_rloop_signal_matrix <- dripseq_list$reduced_rloop_signal_matrix
prop_rloop_signal_matrix <- reduced_rloop_signal_matrix / intronic_signal_matrix[rownames(reduced_rloop_signal_matrix),]
prop_rloop_signal_matrix[is.na(prop_rloop_signal_matrix)] <- 0
mean_prop_signal <- rowMeans(prop_rloop_signal_matrix)

repeat_signal_matrix <- dripseq_list$repeat_signal_matrix
libnorm_repeat_signal_matrix <- t(t(repeat_signal_matrix) / (library_sizes[1,] / 1e6))
mean_repeat_signal <- rowMeans(libnorm_repeat_signal_matrix)


```

- Prepare data.frame for CDF computation

```{r}


reduced_rloop_ranges <- dripseq_list$reduced_rloop_ranges
repeats_in_de_intron <- repeat_ranges[de_feature_ranges_with_repeats$which_repeat]

ol_object <- findOverlaps(repeats_in_de_intron, reduced_rloop_ranges)

feature_signal_df <- data.frame(repeat_id = unique(names(repeats_in_de_intron)), signal = 0)
feature_signal_df$signal <- mean_repeat_signal[feature_signal_df$repeat_id]
feature_signal_df$repeat_status <- "outside_rloops"
feature_signal_df$repeat_status[feature_signal_df$repeat_id %in% names(repeats_in_de_intron)[queryHits(ol_object)]] <- "overlapping_rloops"

```


- Get CDF


```{r}

introns_with_predicted_rloops_df <- feature_signal_df[feature_signal_df$signal > 0,]
quantiles <- quantile(log10(introns_with_predicted_rloops_df$signal), seq(0,1,0.01))
return_df_l1 <- do.call(rbind, lapply(unique(introns_with_predicted_rloops_df$repeat_status), function(x){
        cumulative_values <- vapply(quantiles, function(z){
            cumulative_val <- nrow(introns_with_predicted_rloops_df[
                    introns_with_predicted_rloops_df$repeat_status == x &
                    log10(introns_with_predicted_rloops_df$signal) <= z,]) / nrow(introns_with_predicted_rloops_df[introns_with_predicted_rloops_df$repeat_status == x,])
        },1)
        cdf_df <- data.frame(signal = quantiles, 
            cdf = cumulative_values, 
            repeat_status = x)
        return(cdf_df)
}))


ks.test(x = return_df_l1$cdf[return_df_l1$repeat_status == "outside_rloops"],
        y = return_df_l1$cdf[return_df_l1$repeat_status == "overlapping_rloops"])$p.value
# [1] 0.8174604
```

- Plot CDF of repeats


```{r}

Colours <- c("overlapping_rloops" = "#F05D5E", "outside_rloops" = "#058ED9")
ThePlot <- ggplot(return_df_l1, aes(x = signal, y = cdf))
ThePlot <- ThePlot + geom_step(aes(colour = repeat_status))
ThePlot <- ThePlot + scale_x_continuous("log10 mean library size normalised rloop signal within introns")
ThePlot <- ThePlot + scale_y_continuous("CDF")
ThePlot <- ThePlot + scale_colour_manual("Rloop prediction", values = Colours)
# ThePlot <- ThePlot + scale_colour_manual("status", values = Colours)
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(. ~ status, scales = "free")
ThePlot <- ThePlot + ggtitle("library size normalised R-loop signal")
ThePlot <- ThePlot + theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))
ggsave(width = 7, height = 7, file = file.path(plot_dir, "04-04-figure-5E.pdf"), ThePlot)


```
