---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the scripts required to generate Supplementary figure 18C and 18D. We model proportion of intronic signal assigned to R-loops as a function of intronic length assigned to R-loops. We then validate this model using a monte carlo approach.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
source("../useful-functions/plot-utils.R")
source("../src/basepair_coverage_functions.R")
source("src/colours.R")
require("cowplot")
require("ggplot2")
require("GenomicRanges")
require("HiCBricks")
require("stringr")
require("RColorBrewer")
require("rslurm")
require("dplyr")
require("rslurm")
require("rtracklayer")
require("parallel")
require("matrixStats")
require("ComplexHeatmap")
require("circlize")
require("viridis")


```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-5"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
tmp_dir <- file.path(analysis_path, "temp")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)
dir.create(tmp_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
sig_result_list <- readRDS("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds")
feature_ranges <- readRDS("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds")
annotation_gtf <- "/nemo/lab/patanir/home/users/palk/Genomes/ENSEMBL_hg38/Homo_sapiens.GRCh38.104.chr_patch_hapl_scaff.gtf"
rloop_ranges <- readRDS("30-sequence-analysis-of-introns/rdata_files/90-01-merged-rloop-ranges-in-differentially-expressed-introns.rds")
dripseq_samplesheet_df <- read.csv("/nemo/project/proj-luscombn-patani/working/public-data/sporadic-pbmc-drip-fero-et-al-2024/sample-details/sampleinfo.csv")
dripseq_aln_dir <- file.path("/nemo/project/proj-luscombn-patani/working/public-data/sporadic-pbmc-drip-fero-et-al-2024/output_files/bigwig/")
dripseq_list <- readRDS("34-dripseq-analysis/rdata_files/02-01-list-of-total-rloop-coverage-matrix-and-library-sizes.rds")

```

- Fetch features containing repeats

```{r}

dripseq_samplesheet_df$bigwig <- vapply(dripseq_samplesheet_df$SampleName, function(x){
    list.files(path = dripseq_aln_dir, pattern = x) 
},"1")

```

- Generate signal matrices

```{r}

library_sizes <- dripseq_list$library_sizes
repeat_signal_matrix <- dripseq_list$repeat_signal_matrix
reduced_rloop_signal_matrix <- dripseq_list$reduced_rloop_signal_matrix
intronic_signal_matrix <- dripseq_list$feature_signal_matrix

reduced_rloop_ranges <- dripseq_list$reduced_rloop_ranges
rloop_width_per_intron <- vapply(split(width(reduced_rloop_ranges), reduced_rloop_ranges$feature_id), sum, 1)


```

- Get differentially expressed features

```{r}


repeat_ranges <- repeat_ranges_list$repeat_ranges
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges
sig_result_df <- sig_result_list$significant_events
sig_result_df <- sig_result_df[sig_result_df$region == "intron",]

de_feature_ranges <- feature_ranges[feature_ranges$feature_id %in% sig_result_df$feature_ids,]
de_feature_ranges_with_repeats <- feature_ranges_with_repeats[
    feature_ranges_with_repeats$feature_id %in% sig_result_df$feature_ids & 
    feature_ranges_with_repeats$type == "intron"]

```

- Fetch the following signal matrices
    - library size normalised matrices for reduced R-loops
    - library size normalised matrices for introns
    - library size normalised matrices for repeat regions

```{r}

norm_reduced_rloop_signal_matrix <- t(t(reduced_rloop_signal_matrix) / (library_sizes["total_coverage",colnames(reduced_rloop_signal_matrix)] / 1e6))

norm_intronic_signal_matrix <- t(t(intronic_signal_matrix) / (library_sizes["total_coverage",colnames(intronic_signal_matrix)] / 1e6))

norm_repeat_signal_matrix <- t(t(repeat_signal_matrix) / (library_sizes["total_coverage",colnames(repeat_signal_matrix)] / 1e6))

```

- Add quantitative columns to the differentially expressed features 


```{r}

de_feature_ranges$intronic_signal <- rowMeans(norm_intronic_signal_matrix)[de_feature_ranges$feature_id]
de_feature_ranges$rloop_signal <- rowMeans(norm_reduced_rloop_signal_matrix)[de_feature_ranges$feature_id]
de_feature_ranges$rloop_signal_prop <- de_feature_ranges$rloop_signal / de_feature_ranges$intronic_signal
de_feature_ranges$rloop_width <- rloop_width_per_intron[de_feature_ranges$feature_id]
de_feature_ranges$rloop_width_prop <- de_feature_ranges$rloop_width / width(de_feature_ranges)

```

- create regressor modelling proportion of intronic signal assigned to r-loop as a function of proportion of intronic length assigned to r-loop

```{r}

non_na_de_feature_ranges <- de_feature_ranges[!is.na(de_feature_ranges$rloop_signal_prop) & de_feature_ranges$type == "intron"]

model_df <- data.frame(proportion_of_signal = non_na_de_feature_ranges$rloop_signal_prop, 
    proportion_of_length = non_na_de_feature_ranges$rloop_width_prop)

a_model <- lm(data = model_df, formula = proportion_of_signal ~ proportion_of_length)

# Call:
# lm(formula = proportion_of_signal ~ proportion_of_length, data = model_df)

# Residuals:
#      Min       1Q   Median       3Q      Max
# -0.91772 -0.03606 -0.00409  0.03331  0.75913

# Coefficients:
#                       Estimate Std. Error t value Pr(>|t|)
# (Intercept)          0.0007896  0.0006664   1.185    0.236
# proportion_of_length 0.9696989  0.0017870 542.637   <2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

# Residual standard error: 0.1158 on 51852 degrees of freedom
# Multiple R-squared:  0.8503,    Adjusted R-squared:  0.8503
# F-statistic: 2.945e+05 on 1 and 51852 DF,  p-value: < 2.2e-16

```

- Predict the expected value


```{r}

de_feature_ranges_with_repeats$intronic_signal <- rowMeans(norm_intronic_signal_matrix)[de_feature_ranges_with_repeats$feature_id]
de_feature_ranges_with_repeats$mean_repeat_dripseq_signal <- rowMeans(norm_repeat_signal_matrix)[names(repeat_ranges)[de_feature_ranges_with_repeats$which_repeat]]
de_feature_ranges_with_repeats$repeat_prop_width <- de_feature_ranges_with_repeats$repeat_width / width(de_feature_ranges_with_repeats)


predict_df <- data.frame(proportion_of_length = de_feature_ranges_with_repeats$repeat_prop_width)
de_feature_ranges_with_repeats$expected_proportion_of_signal_in_repeats <- predict(object = a_model, newdata = predict_df)

de_feature_ranges_with_repeats$expected_signal_in_repeats <- de_feature_ranges_with_repeats$expected_proportion_of_signal_in_repeats * de_feature_ranges_with_repeats$intronic_signal
de_feature_ranges_with_repeats$repeat_foldchange <- de_feature_ranges_with_repeats$mean_repeat_dripseq_signal / de_feature_ranges_with_repeats$expected_signal_in_repeats


```

- Perform tests of significance on the foldchanges


```{r}


repeat_foldchange_matrix <- norm_repeat_signal_matrix / de_feature_ranges_with_repeats$expected_signal_in_repeats[match(rownames(norm_repeat_signal_matrix), names(repeat_ranges)[de_feature_ranges_with_repeats$which_repeat])]
colnames(repeat_foldchange_matrix) <- dripseq_samplesheet_df$characteristics_patient_code[match(colnames(norm_repeat_signal_matrix), dripseq_samplesheet_df$SampleName)]


repeat_fold_change_df <- reshape2::melt(repeat_foldchange_matrix)
colnames(repeat_fold_change_df) <- c("repeat_id", "patient_id", "foldchange")
repeat_fold_change_df$condition <- "ALS"
repeat_fold_change_df$condition[!grepl(x = repeat_fold_change_df$patient_id, pattern = "ALS")] <- "Control"

foldchange_df <- do.call(rbind, lapply(split(repeat_fold_change_df, paste(repeat_fold_change_df$repeat_id, repeat_fold_change_df$condition)), function(temp_df){
    return_df <- unique(temp_df[,c("repeat_id", "condition")])
    return_df$percent_zero <- length(temp_df$foldchange[temp_df$foldchange == 0]) / length(temp_df$foldchange)
    return_df$foldchange <- mean(temp_df$foldchange, na.rm = T)
    return(return_df)
}))

foldchange_df_split <- split(foldchange_df, foldchange_df$repeat_id)
non_na_foldchange_df <- do.call(rbind, foldchange_df_split[vapply(foldchange_df_split, function(x){
    !is.na(x$foldchange[x$condition == "ALS"]) & !is.na(x$foldchange[x$condition == "Control"])
},T)])
non_na_foldchange_df$ssr <- repeat_ranges[non_na_foldchange_df$repeat_id]$ssr
non_na_foldchange_df$span_group <- repeat_ranges[non_na_foldchange_df$repeat_id]$span_group


wilcox.test(non_na_foldchange_df$foldchange[non_na_foldchange_df$condition == "ALS"], non_na_foldchange_df$foldchange[non_na_foldchange_df$condition == "Control"])


#         Wilcoxon rank sum test with continuity correction

# data:  non_na_foldchange_df$foldchange[non_na_foldchange_df$condition == "ALS"] and non_na_foldchange_df$foldchange[non_na_foldchange_df$condition == "Control"]
# W = 24343276, p-value < 2.2e-16
# alternative hypothesis: true location shift is not equal to 0


```

- Plot the global distribution in ALS and control

```{r}

colours <- c("Control" = "#F6F4F3", "ALS" = "#F87575")
ThePlot <- ggplot(non_na_foldchange_df, aes(x = condition, y = log2(foldchange+1)))
# ThePlot <- ThePlot + geom_violin(aes(fill = condition), colour = "#000000", 
#     lwd = 0.5, position = position_dodge(width = 0.9))
ThePlot <- ThePlot + geom_violin(aes(fill = condition), colour = "#000000", position = position_dodge(width = 0.9), outlier.size = NA)
ThePlot <- ThePlot + geom_boxplot(aes(fill = condition), colour = "#000000", 
    width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9))
ThePlot <- ThePlot + scale_x_discrete("Condition")
ThePlot <- ThePlot + scale_y_continuous("log2 (observed/expected + 1)")
ThePlot <- ThePlot + scale_fill_manual("condition", values = colours)
# ThePlot <- ThePlot + scale_colour_manual("status", values = Colours)
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(class ~ ., scales = "free")
ThePlot <- ThePlot + ggtitle("distribution of fold change over regression based expected drip-seq signal")
ThePlot <- ThePlot + theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))
ggsave(width = 5, height = 10, file = file.path(plot_dir, "04-06-plot-distribution-of-observed-over-expected-dripseq-signal-for-als-and-control-figure-5F.pdf"), ThePlot)

```

- Compare the foldchanges between conditions grouped by motif class

```{r}


foldchange_df_split_by_motif <- split(non_na_foldchange_df, non_na_foldchange_df$ssr)
min_group_size_gt_5_df <- do.call(rbind, foldchange_df_split_by_motif[vapply(foldchange_df_split_by_motif, function(x){
    length(unique(x$repeat_id)) > 5
},T)])

rownames(min_group_size_gt_5_df) <- NULL


get_ci <- function(x, condition, var = "foldchange"){
    temp_df <- x[x$condition == condition,]
    margin <- qt(0.975,df=nrow(temp_df)-1) * sd(temp_df[,"foldchange"]) / sqrt(nrow(temp_df))
    return_df <- data.frame(condition = condition, 
        var = var,
        lower = mean(temp_df[,"foldchange"]) - margin,
        mean = mean(temp_df[,"foldchange"]),
        upper = mean(temp_df[,"foldchange"]) + margin)
    return(return_df)
}


all_significance_df <- do.call(rbind, lapply(split(min_group_size_gt_5_df, min_group_size_gt_5_df$ssr), function(x){
    test_result <- wilcox.test(x$foldchange[x$condition == "ALS"], x$foldchange[x$condition == "Control"])
    als_df <- get_ci(x, condition = "ALS")
    control_df <- get_ci(x, condition = "Control")
    return_df <- rbind(control_df, als_df)
    return_df$motif <- unique(x$ssr)
    return_df$n <- length(unique(x$repeat_id))
    return_df$pvalue <- test_result$p.value
    return(return_df)
}))

all_significance_df$sig <- "NS"
all_significance_df$sig[all_significance_df$pvalue < 0.05] <- "*"
all_significance_df$sig[all_significance_df$pvalue < 0.01] <- "**"
all_significance_df$sig[all_significance_df$pvalue < 0.001] <- "***"

```

- Plot the distribution for each motif

```{r}

pvalues_by_ssr <- split(all_significance_df[all_significance_df$condition == "ALS","pvalue"], all_significance_df[all_significance_df$condition == "ALS","motif"])
ssr_levels <- names(pvalues_by_ssr)[order(-log10(do.call(c, pvalues_by_ssr)), decreasing = T)]


all_significance_df$motif <- factor(all_significance_df$motif, levels = ssr_levels)

colours <- c("Control" = "#F6F4F3", "ALS" = "#F87575")
ThePlot <- ggplot(all_significance_df[all_significance_df$sig != "NS",], aes(x = motif, y = log2(mean+1)))
# ThePlot <- ThePlot + geom_violin(aes(fill = condition), colour = "#000000", 
#     lwd = 0.5, position = position_dodge(width = 0.9))
ThePlot <- ThePlot + geom_pointrange(aes(ymin = log2(lower+1), ymax = log2(upper+1), fill = condition), shape = 21, colour = "#000000", position = position_dodge(width = 0.9))
ThePlot <- ThePlot + scale_x_discrete("Repeat motifs")
ThePlot <- ThePlot + scale_y_continuous("log2 observed/expected")
ThePlot <- ThePlot + scale_fill_manual("condition", values = colours)
# ThePlot <- ThePlot + scale_colour_manual("status", values = Colours)
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(class ~ ., scales = "free")
ThePlot <- ThePlot + ggtitle("distribution of fold change over regression based expected drip-seq signal")
ThePlot <- ThePlot + theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12, angle = 90, hjust = 1, vjust = 0.5),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))
ggsave(width = 10, height = 5, file = file.path(plot_dir,  "04-07-plot-distribution-of-observed-over-expected-dripseq-signal-for-each-motif-figure-5G.pdf"), ThePlot)




significant_df <- all_significance_df[all_significance_df$sig != "NS",]
significant_df$motif <- droplevels(significant_df$motif)
table(vapply(split(significant_df, significant_df$motif), function(x){
    x$mean[x$condition == "ALS"] > x$mean[x$condition == "Control"]
},1))


compare_fc_matrix <- do.call(rbind, lapply(split(significant_df, significant_df$motif), function(x){
    cbind(x$mean[x$condition == "ALS"], x$mean[x$condition == "Control"])
}))
rownames(compare_fc_matrix) <- names(split(significant_df, significant_df$motif))
percent_diff <- (compare_fc_matrix[,1] -  compare_fc_matrix[,2]) / compare_fc_matrix[,2]

compare_fc_matrix[order(percent_diff, decreasing = T),]


colours <- c("Control" = "#F6F4F3", "ALS" = "#F87575")
ThePlot <- ggplot(all_significance_df[all_significance_df$condition == "ALS" & all_significance_df$sig != "NS",], aes(x = motif, y = var))
# ThePlot <- ThePlot + geom_violin(aes(fill = condition), colour = "#000000", 
#     lwd = 0.5, position = position_dodge(width = 0.9))
ThePlot <- ThePlot + geom_tile(aes(fill = -log10(pvalue)))
ThePlot <- ThePlot + scale_x_discrete("Repeat motifs")
ThePlot <- ThePlot + scale_y_discrete("foldchange")
ThePlot <- ThePlot + scale_fill_distiller("-log10 pvalue", palette = "Reds", direction = 1)
# ThePlot <- ThePlot + scale_colour_manual("status", values = Colours)
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(class ~ ., scales = "free")
ThePlot <- ThePlot + ggtitle("pvalue distribution for individual motifs")
ThePlot <- ThePlot + theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12, angle = 90, hjust = 1, vjust = 0.5),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))
ggsave(width = 10, height = 5, file = file.path(plot_dir, "04-07-plot-pvalue-distribution-of-observed-over-expected-dripseq-signal-for-each-motif-figure-5G.pdf"), ThePlot)

```