---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-3"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
patient_penetrance_list <- readRDS(file.path(save_dir, "03-save-answerals-rci-expression-summary-with-penetrance-matrix.rds"))
answerals_counts_list <- readRDS("13-01-answerals/rdata_files/01-02-list_of_counts_stat_samplesheet_from_answerals.rds")

counts_matrix <- answerals_counts_list$counts
patient_liability_df <- patient_penetrance_list$liability_df
penetrance_matrix <- patient_penetrance_list$penetrance_matrix
rownames(penetrance_matrix) <- patient_liability_df$sample_name

```

- Intersection between control and ALS

```{r}

patient_df_split <- split(patient_liability_df, patient_liability_df$mnd_group)
ordered_patient_penetrance_matrix_list <- lapply(patient_df_split[c("healthy", "als")], function(temp_df){
    penetrance_matrix[temp_df$sample_name[order(temp_df$repeat_intron_expr_tpm, decreasing = F)],]
})

penetrant_rcis <- lapply(ordered_patient_penetrance_matrix_list, function(x){
	temp_vector <- colSums(x)
	names(temp_vector[temp_vector > 0])
})

# intersection between control and ALS 

intersecting_rcis <- intersect(penetrant_rcis[["als"]], penetrant_rcis[["healthy"]])
all_rcis <- union(penetrant_rcis[["als"]], penetrant_rcis[["healthy"]])

# > length(intersecting_rcis) / length(all_rcis)
# [1] 0.709478
# > length(penetrant_rcis[["healthy"]]) / length(all_rcis)
# [1] 0.7163462
# > length(penetrant_rcis[["als"]]) / length(all_rcis)
# [1] 0.9931319

```

- Similarity between ipsc lines 

```{r}

library(factoextra)
library(NbClust)

current_penetrance_matrix <- penetrance_matrix[patient_liability_df$sample_name[patient_liability_df$mnd %in% c("healthy", "als")],]
pair_matrix <- cbind(rep(seq_len(nrow(current_penetrance_matrix)), each = nrow(current_penetrance_matrix)),
    rep(seq_len(nrow(current_penetrance_matrix)), times = nrow(current_penetrance_matrix)))

similarity_df <- do.call(rbind, lapply(seq_len(nrow(pair_matrix)), function(x){
    current_x_row <- current_penetrance_matrix[pair_matrix[x,1],all_rcis]
    current_x <- names(current_x_row)[current_x_row == 1]
    current_y_row <- current_penetrance_matrix[pair_matrix[x,2],all_rcis]
    current_y <- names(current_y_row)[current_y_row == 1]

    jaccard_index <- length(intersect(current_x, current_y)) / length(union(current_x, current_y))
    return(data.frame(x = rownames(current_penetrance_matrix)[pair_matrix[x,1]],
                y = rownames(current_penetrance_matrix)[pair_matrix[x,2]],
                ji = jaccard_index))
}))

similarity_matrix <- reshape2::acast(data = similarity_df, formula = x ~ y, value.var = "ji")
dissimilarity_matrix <- 1 - similarity_matrix
distance_matrix <- as.dist(dissimilarity_matrix)

elbow_plot <- fviz_nbclust(dissimilarity_matrix, FUNcluster = kmeans, method = "gap_stat", k.max = 15, nboot = 500) +
    geom_vline(xintercept = 5, linetype = 2)+
  labs(subtitle = "Elbow method")
ggsave(width = 10, height = 10, file = file.path(plot_dir, "05-01-plot-optimal-number-of-clusters-of-patients-supplementary-figure-7B.pdf"), elbow_plot)

kclustering <- kmeans(x = dissimilarity_matrix, centers = 7)

```


- Determine how many clusters are purely control or als

```{r}

clusters <- kclustering$cluster
clusters_split <- split(names(clusters), clusters)

vapply(clusters_split, function(x){
    current_matrix <- similarity_matrix[x,x]
    mean(current_matrix)
},1)


```

- Compute progressive gain in jaccard

```{r}

all_rcis <- colnames(penetrance_matrix)[colSums(penetrance_matrix) > 0]
progressive_gain_in_ji_df <- do.call(rbind, lapply(ordered_patient_penetrance_matrix_list, function(a_sub_matrix){
    i <- 1
    a_list <- list()
    df_list <- list()
    while(i <= nrow(a_sub_matrix)){
        a_row <- a_sub_matrix[i,all_rcis]
        a_list[[i]] <- names(a_row)[a_row == 1]
        if(i == 1){
            df_list[[i]] <- data.frame(rank = i, 
                sample_name = rownames(a_sub_matrix)[i], 
                jaccard_index = length(unique(do.call(c, a_list))) / length(all_rcis),
                proportion_over_total = length(unique(a_list[[i]])) / length(all_rcis))
        }else{
            prior <- do.call(c, a_list[seq(from = 1, to = i - 1)])
            intersect_hash <- intersect(prior, do.call(c, a_list))
            jaccard <- length(intersect_hash) / length(all_rcis)
            current_jaccard <- length(unique(do.call(c, a_list))) / length(all_rcis)

            df_list[[i]] <- data.frame(rank = i, 
                sample_name = rownames(a_sub_matrix)[i],
                jaccard_index = current_jaccard,
                proportion_over_total = length(unique(a_list[[i]])) / length(all_rcis))
        }
        i <- i+1
    }
    return_df <- do.call(rbind, df_list)
    return_df$rci_liability <- patient_liability_df$repeat_intron_expr_tpm[match(return_df$sample_name, patient_liability_df$sample_name)]
    return_df$mnd_group <- patient_liability_df$mnd_group[match(return_df$sample_name, patient_liability_df$sample_name)]
    return_df
}))

```

- Plot progressive gain in jaccard index

```{r}

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")

ThePlot <- ggplot(progressive_gain_in_ji_df[progressive_gain_in_ji_df$mnd_group %in% c("healthy", "als"),], 
    aes(x = rci_liability, y = jaccard_index, group = mnd_group))
# ThePlot <- ThePlot + geom_violin(aes(fill = condition), colour = "#000000", 
#     lwd = 0.5, position = position_dodge(width = 0.9), scale = "width")
# ThePlot <- ThePlot + geom_line(aes(colour = mnd_group))
# ThePlot <- ThePlot + geom_boxplot(aes(fill = mnd_group), colour = "#000000", 
#     width = 0.1, lwd = 0.5, alpha = 0.6, position = position_dodge(width = 0.9))
# ThePlot <- ThePlot + geom_point(aes(fill = condition), shape = 21, colour = "#000000", position = position_jitterdodge(dodge.width = 0.9))
ThePlot <- ThePlot + geom_smooth(method = "loess", formula = y ~ x, aes(colour = mnd_group))
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_continuous("RCI liability")
ThePlot <- ThePlot + scale_y_continuous("jaccard index")
ThePlot <- ThePlot + scale_colour_manual("condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(. ~ predictor_group, scales = "free")
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression across tissues")
ThePlot <- ThePlot + current_theme
ggsave(width = 10, height = 10, file = file.path(plot_dir, "05-01-plot-RCI-liability-vs-jaccard-index-supplementary-figure-7B.pdf"), ThePlot)

ThePlot <- ggplot(progressive_gain_in_ji_df[progressive_gain_in_ji_df$mnd_group %in% c("healthy", "als"),], 
    aes(x = rci_liability, y = proportion_over_total, group = mnd_group))
# ThePlot <- ThePlot + geom_violin(aes(fill = condition), colour = "#000000", 
#     lwd = 0.5, position = position_dodge(width = 0.9), scale = "width")
# ThePlot <- ThePlot + geom_line(aes(colour = mnd_group))
# ThePlot <- ThePlot + geom_boxplot(aes(fill = mnd_group), colour = "#000000", 
#     width = 0.1, lwd = 0.5, alpha = 0.6, position = position_dodge(width = 0.9))
# ThePlot <- ThePlot + geom_point(aes(fill = condition), shape = 21, colour = "#000000", position = position_jitterdodge(dodge.width = 0.9))
ThePlot <- ThePlot + geom_smooth(method = "loess", formula = y ~ x, aes(colour = mnd_group))
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_continuous("RCI liability")
ThePlot <- ThePlot + scale_y_continuous("Proportion of RCIs driving liability")
ThePlot <- ThePlot + scale_colour_manual("condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(. ~ predictor_group, scales = "free")
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression across tissues")
ThePlot <- ThePlot + current_theme
ggsave(width = 10, height = 10, file = file.path(plot_dir, "05-01-plot-RCI-liability-vs-proportion-of-penetrant-rcis-supplementary-figure-7B.pdf"), ThePlot)


```