---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")
require("ggbeeswarm")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-3"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}


repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
feature_ranges <- readRDS("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds")
sig_result_list <- readRDS("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds")
repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
patient_penetrance_list <- readRDS(file.path(save_dir, "03-save-answerals-rci-expression-summary-with-penetrance-matrix.rds"))
answerals_object_list <- readRDS("13-01-answerals/rdata_files/01-02-list_of_counts_stat_samplesheet_from_answerals.rds")
sampleinfo_df <- read.csv("/nemo/project/proj-luscombn-patani/working/answerals/sample-details/samplesheet.csv")
nygc_penetrance_list <- readRDS(file.path("final-figures/figure-2/rdata_files/02-save-nygc-rci-expression-summary-with-penetrance-matrix.rds"))

patient_liability_df <- patient_penetrance_list$liability_df
penetrance_matrix <- patient_penetrance_list$penetrance_matrix
rownames(penetrance_matrix) <- patient_liability_df$sample_name

nygc_patient_liability_df <- nygc_penetrance_list$liability_df
nygc_penetrance_matrix <- nygc_penetrance_list$penetrance_matrix


```

- Load repeat and rci object

```{r}

repeat_ranges <- unique(repeat_ranges_list$repeat_ranges)
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges

```

- Generate TPM matrix

```{r}

sampleinfo_df$sample_name <- str_replace_all(sampleinfo_df$participant_id, pattern = "-", replacement = "_")

sig_result_df <- sig_result_list$significant_events

repeat_containing_introns_df <- sig_result_df[sig_result_df$status == "repeat_containing" & 
    sig_result_df$region == "intron",]

repeat_containing_introns_df$cohort <- "public-data"
repeat_containing_introns_df$cohort[grepl(x = repeat_containing_introns_df$study, pattern = "nygc")] <- "nygc"
repeat_containing_introns_df$cohort[grepl(x = repeat_containing_introns_df$study, pattern = "answerals")] <- "answerals"

repeat_containing_introns_df$tissue_type <- "iPSCs"
repeat_containing_introns_df$tissue_type[grepl(x = repeat_containing_introns_df$study, pattern = "nygc")] <- "postmortem"

counts_matrix <- answerals_object_list$counts
patient_df <- answerals_object_list$sample_df
temp_matrix <- counts_matrix / (width(feature_ranges)[match(rownames(counts_matrix), feature_ranges$hash_id)] / 1e3)
tpm_matrix <- t(t(temp_matrix) / (colSums(temp_matrix) / 1e6))

keep_hash_ids <- feature_ranges_with_repeats$hash_id[match(unique(repeat_containing_introns_df$feature_ids), feature_ranges_with_repeats$feature_id)]
keep_tpm_matrix <- tpm_matrix[keep_hash_ids,colnames(tpm_matrix) %in% patient_df$column_id]

repeat_lengths <- feature_ranges_with_repeats$repeat_width[match(rownames(keep_tpm_matrix), feature_ranges_with_repeats$hash_id)]

```

- Find penetrant RCIs in NYGC samples

```{r}

nygc_patient_df_split <- split(nygc_patient_liability_df, paste(nygc_patient_liability_df$condition, nygc_patient_liability_df$tissue_group))
nygc_ordered_patient_penetrance_matrix_list <- lapply(nygc_patient_df_split, function(temp_df){
    nygc_penetrance_matrix[temp_df$penetrance_row[order(temp_df$repeat_intron_expr_tpm, decreasing = F)],]
})

nygc_penetrant_rcis <- lapply(nygc_ordered_patient_penetrance_matrix_list, function(x){
    temp_vector <- colSums(x)
    names(temp_vector[temp_vector > 0])
})


```

- Find PM ALS spinal cord specific RCIs

```{r}

pm_als_spinal_cord_rcis <- setdiff(nygc_penetrant_rcis[["als spinal_cord"]], nygc_penetrant_rcis[["ctrl spinal_cord"]])
pm_ctrl_spinal_cord_rcis <- setdiff(nygc_penetrant_rcis[["ctrl spinal_cord"]], nygc_penetrant_rcis[["als spinal_cord"]])

```

- Find PM ALS brain specific RCIs

```{r}

pm_als_brain_rcis <- setdiff(nygc_penetrant_rcis[["als brain"]], nygc_penetrant_rcis[["ctrl brain"]])
pm_ctrl_brain_rcis <- setdiff(nygc_penetrant_rcis[["ctrl brain"]], nygc_penetrant_rcis[["als brain"]])

```




- Generate liability data for PM spinal cord specific RCIs

```{r}

threshold <- 0.1

pm_spinal_cord_specific_liability_df <- do.call(rbind, lapply(seq_len(nrow(patient_df)), function(x1){
    temp_df <- patient_df[x1,]
    return_df <- temp_df
    current_columns <- temp_df$column_id
    
    current_tpm_vector <- Reduce("+", lapply(current_columns, function(x){
        keep_tpm_matrix[,x]  
    })) / length(current_columns)

    gt_tpm <- current_tpm_vector[current_tpm_vector > threshold]
    return_df$median_pm_als_spinal_cord_specific_expr_tpm <- median(current_tpm_vector[pm_als_spinal_cord_rcis])
    return_df$median_pm_ctrl_spinal_cord_specific_expr_tpm <- median(current_tpm_vector[pm_ctrl_spinal_cord_rcis])
    # return_df$proportion_of_pm_specific_rcis <- length(pm_als_spinal_cord_rcis[pm_als_spinal_cord_rcis %in% names(gt_tpm)]) / length(gt_tpm)
    # return_df$proportion_liability_from_pm_specific_rcis <- sum(gt_tpm[names(gt_tpm) %in% pm_als_spinal_cord_rcis]) / sum(gt_tpm)

    return(return_df)
}))

patient_liability_df$median_pm_als_spinal_cord_specific_expr_tpm <- pm_spinal_cord_specific_liability_df$median_pm_als_spinal_cord_specific_expr_tpm[match(patient_liability_df$sample_name, pm_spinal_cord_specific_liability_df$sample_name)]
patient_liability_df$median_pm_ctrl_spinal_cord_specific_expr_tpm <- pm_spinal_cord_specific_liability_df$median_pm_ctrl_spinal_cord_specific_expr_tpm[match(patient_liability_df$sample_name, pm_spinal_cord_specific_liability_df$sample_name)]
# patient_liability_df$proportion_of_pm_specific_rcis <- pm_spinal_cord_specific_liability_df$proportion_of_pm_specific_rcis[match(patient_liability_df$sample_name, pm_spinal_cord_specific_liability_df$sample_name)]
# patient_liability_df$proportion_liability_from_pm_specific_rcis <- pm_spinal_cord_specific_liability_df$proportion_liability_from_pm_specific_rcis[match(patient_liability_df$sample_name, pm_spinal_cord_specific_liability_df$sample_name)]


patient_liability_df$mnd_group <- factor(patient_liability_df$mnd_group, levels = c("healthy", "als", "carrier", "other_mnd"))

summary(lm(data = patient_liability_df[patient_liability_df$mnd_group %in% c("healthy", "als") & patient_liability_df$repeat_intron_expr_tpm <= 200,], formula = median_pm_als_spinal_cord_specific_expr_tpm ~ repeat_intron_expr_tpm + mnd_group + repeat_intron_expr_tpm:mnd_group))


# Call:
# lm(formula = median_pm_spinal_cord_specific_expr_tpm ~ repeat_intron_expr_tpm +
#     mnd_group + repeat_intron_expr_tpm:mnd_group, data = patient_liability_df[patient_liability_df$mnd_group %in%
#     c("healthy", "als") & patient_liability_df$repeat_intron_expr_tpm <=
#     200, ])

# Residuals:
#       Min        1Q    Median        3Q       Max
# -0.060047 -0.013590 -0.001559  0.013290  0.062299

# Coefficients:
#                                       Estimate Std. Error t value Pr(>|t|)
# (Intercept)                         -0.0087859  0.0214122  -0.410   0.6819
# repeat_intron_expr_tpm               0.0005732  0.0001359   4.218 3.41e-05 ***
# mnd_groupals                        -0.0468193  0.0238365  -1.964   0.0506 .
# repeat_intron_expr_tpm:mnd_groupals  0.0003075  0.0001522   2.020   0.0444 *
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

# Residual standard error: 0.01931 on 260 degrees of freedom
# Multiple R-squared:  0.4135,    Adjusted R-squared:  0.4067
# F-statistic:  61.1 on 3 and 260 DF,  p-value: < 2.2e-16

# 34.9% faster with increasing liability

summary(lm(data = patient_liability_df[patient_liability_df$mnd_group %in% c("healthy", "als") & patient_liability_df$repeat_intron_expr_tpm <= 200,], formula = median_pm_ctrl_spinal_cord_specific_expr_tpm ~ repeat_intron_expr_tpm + mnd_group + repeat_intron_expr_tpm:mnd_group))







```

- Compare RCI liability vs gain in signal in als specific rcis

```{r}

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
patient_liability_df$mnd_group <- factor(patient_liability_df$mnd_group, levels = c("healthy", "als", "other_mnd", "carrier"))
ThePlot <- ggplot(patient_liability_df[patient_liability_df$mnd_group %in% c("als", "healthy") & patient_liability_df$repeat_intron_expr_tpm <= 200,], 
    aes(x = repeat_intron_expr_tpm, y = median_pm_als_spinal_cord_specific_expr_tpm))
ThePlot <- ThePlot + geom_point(aes(fill = mnd_group), colour = "#000000", shape = 21)
ThePlot <- ThePlot + geom_smooth(method = "lm", aes(colour = mnd_group), formula = y ~ x)
ThePlot <- ThePlot + scale_x_continuous("ALS specific rci expression")
ThePlot <- ThePlot + scale_y_continuous("Median PM spinal cord specific RCI expression in iPSC-MNs")
ThePlot <- ThePlot + scale_colour_manual("Condition colour", values = colours)
ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(. ~ mnd_group)
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression in AnswerALS iPSCs")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 10, file = file.path(plot_dir, "06-05-plot-RCI-liability-vs-median-pm-als-spinal-cord-specific-rci-expression-in-healthy-and-ALS-figure-2E.pdf"), ThePlot)

# colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
# patient_liability_df$mnd_group <- factor(patient_liability_df$mnd_group, levels = c("healthy", "als", "other_mnd", "carrier"))
# ThePlot <- ggplot(patient_liability_df[patient_liability_df$mnd_group %in% c("als", "healthy") & patient_liability_df$repeat_intron_expr_tpm <= 200,], 
#     aes(x = repeat_intron_expr_tpm, y = median_pm_ctrl_spinal_cord_specific_expr_tpm))
# ThePlot <- ThePlot + geom_point(aes(fill = mnd_group), colour = "#000000", shape = 21)
# ThePlot <- ThePlot + geom_smooth(method = "lm", aes(colour = mnd_group), formula = y ~ x)
# ThePlot <- ThePlot + scale_x_continuous("ALS specific rci expression")
# ThePlot <- ThePlot + scale_y_continuous("Median PM spinal cord specific RCI expression in iPSC-MNs")
# ThePlot <- ThePlot + scale_colour_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# # ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
# ThePlot <- ThePlot + theme_bw()
# # ThePlot <- ThePlot + facet_grid(. ~ mnd_group)
# ThePlot <- ThePlot + ggtitle("Repeat containing intron expression in AnswerALS iPSCs")
# ThePlot <- ThePlot + current_theme
# ggsave(width = 7, height = 10, file = file.path(plot_dir, "06-05-plot-RCI-liability-vs-median-pm-control-spinal-cord-specific-rci-expression-in-healthy-and-ALS-figure-2E.pdf"), ThePlot)

```

- Generate liability data for PM brain specific RCIs

```{r}

threshold <- 0.1

pm_brain_specific_liability_df <- do.call(rbind, lapply(seq_len(nrow(patient_df)), function(x1){
    temp_df <- patient_df[x1,]
    return_df <- temp_df
    current_columns <- temp_df$column_id
    
    current_tpm_vector <- Reduce("+", lapply(current_columns, function(x){
        keep_tpm_matrix[,x]  
    })) / length(current_columns)

    gt_tpm <- current_tpm_vector[current_tpm_vector > threshold]
    return_df$median_pm_als_brain_specific_expr_tpm <- median(current_tpm_vector[pm_als_brain_rcis])
    return_df$median_pm_ctrl_brain_specific_expr_tpm <- median(current_tpm_vector[pm_ctrl_brain_rcis])
    # return_df$proportion_of_pm_specific_rcis <- length(pm_als_brain_rcis[pm_als_brain_rcis %in% names(gt_tpm)]) / length(gt_tpm)
    # return_df$proportion_liability_from_pm_specific_rcis <- sum(gt_tpm[names(gt_tpm) %in% pm_als_brain_rcis]) / sum(gt_tpm)

    return(return_df)
}))

patient_liability_df$median_pm_als_brain_specific_expr_tpm <- pm_brain_specific_liability_df$median_pm_als_brain_specific_expr_tpm[match(patient_liability_df$sample_name, pm_brain_specific_liability_df$sample_name)]
patient_liability_df$median_pm_ctrl_brain_specific_expr_tpm <- pm_brain_specific_liability_df$median_pm_ctrl_brain_specific_expr_tpm[match(patient_liability_df$sample_name, pm_brain_specific_liability_df$sample_name)]
# patient_liability_df$proportion_of_pm_specific_rcis <- pm_brain_specific_liability_df$proportion_of_pm_specific_rcis[match(patient_liability_df$sample_name, pm_brain_specific_liability_df$sample_name)]
# patient_liability_df$proportion_liability_from_pm_specific_rcis <- pm_brain_specific_liability_df$proportion_liability_from_pm_specific_rcis[match(patient_liability_df$sample_name, pm_brain_specific_liability_df$sample_name)]


patient_liability_df$mnd_group <- factor(patient_liability_df$mnd_group, levels = c("healthy", "als", "carrier", "other_mnd"))

summary(lm(data = patient_liability_df[patient_liability_df$mnd_group %in% c("healthy", "als") & patient_liability_df$repeat_intron_expr_tpm <= 200,], formula = median_pm_als_brain_specific_expr_tpm ~ repeat_intron_expr_tpm + mnd_group + repeat_intron_expr_tpm:mnd_group))

# Call:
# lm(formula = median_pm_brain_specific_expr_tpm ~ repeat_intron_expr_tpm +
#     mnd_group + repeat_intron_expr_tpm:mnd_group, data = patient_liability_df[patient_liability_df$mnd_group %in%
#     c("healthy", "als") & patient_liability_df$repeat_intron_expr_tpm <=
#     200, ])

# Residuals:
#        Min         1Q     Median         3Q        Max
# -0.0059943 -0.0011583 -0.0002387  0.0010505  0.0061078

# Coefficients:
#                                       Estimate Std. Error t value Pr(>|t|)
# (Intercept)                         -5.280e-03  1.882e-03  -2.805  0.00541 **
# repeat_intron_expr_tpm               9.905e-05  1.195e-05   8.291 6.08e-15 ***
# mnd_groupals                        -4.707e-04  2.095e-03  -0.225  0.82246
# repeat_intron_expr_tpm:mnd_groupals  1.612e-06  1.338e-05   0.120  0.90422
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

# Residual standard error: 0.001698 on 260 degrees of freedom
# Multiple R-squared:  0.5757,    Adjusted R-squared:  0.5709
# F-statistic: 117.6 on 3 and 260 DF,  p-value: < 2.2e-16

summary(lm(data = patient_liability_df[patient_liability_df$mnd_group %in% c("healthy", "als") & patient_liability_df$repeat_intron_expr_tpm <= 200,], formula = median_pm_ctrl_brain_specific_expr_tpm ~ repeat_intron_expr_tpm + mnd_group + repeat_intron_expr_tpm:mnd_group))

```

- Compare RCI liability vs gain in signal in als brain specific rcis

```{r}

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
patient_liability_df$mnd_group <- factor(patient_liability_df$mnd_group, levels = c("healthy", "als", "other_mnd", "carrier"))
ThePlot <- ggplot(patient_liability_df[patient_liability_df$mnd_group %in% c("als", "healthy") & patient_liability_df$repeat_intron_expr_tpm <= 200,], 
    aes(x = repeat_intron_expr_tpm, y = median_pm_als_brain_specific_expr_tpm))
ThePlot <- ThePlot + geom_point(aes(fill = mnd_group), colour = "#000000", shape = 21)
ThePlot <- ThePlot + geom_smooth(method = "lm", aes(colour = mnd_group), formula = y ~ x)
ThePlot <- ThePlot + scale_x_continuous("ALS specific rci expression")
ThePlot <- ThePlot + scale_y_continuous("Median PM spinal cord specific RCI expression in iPSC-MNs")
ThePlot <- ThePlot + scale_colour_manual("Condition colour", values = colours)
ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(. ~ mnd_group)
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression in AnswerALS iPSCs")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 10, file = file.path(plot_dir, "06-06-plot-RCI-liability-vs-median-pm-als-brain-specific-rci-expression-in-healthy-and-ALS-figure-2E.pdf"), ThePlot)

ThePlot <- ggplot(patient_liability_df[patient_liability_df$mnd_group %in% c("als", "healthy") & patient_liability_df$repeat_intron_expr_tpm <= 200,], 
    aes(x = repeat_intron_expr_tpm, y = median_pm_ctrl_brain_specific_expr_tpm))
ThePlot <- ThePlot + geom_point(aes(fill = mnd_group), colour = "#000000", shape = 21)
ThePlot <- ThePlot + geom_smooth(method = "lm", aes(colour = mnd_group), formula = y ~ x)
ThePlot <- ThePlot + scale_x_continuous("ALS specific rci expression")
ThePlot <- ThePlot + scale_y_continuous("Median PM spinal cord specific RCI expression in iPSC-MNs")
ThePlot <- ThePlot + scale_colour_manual("Condition colour", values = colours)
ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(. ~ mnd_group)
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression in AnswerALS iPSCs")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 10, file = file.path(plot_dir, "06-06-plot-RCI-liability-vs-median-pm-ctrl-brain-specific-rci-expression-in-healthy-and-ALS-figure-2E.pdf"), ThePlot)

```

