---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

library("here")
message("Project root:", here(), "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("ggplot2")
library(rlang)
library(riskRegression)
library(survival)
library(survminer)
library(pammtools)
library(gganimate)
library(transformr)
library(plotly)
library(reshape2)
library(knitr)
library(rmarkdown)

require("contsurvplot")
require("RColorBrewer")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("final-figures","figure-3")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load penetrance data

```{r}

nygc_penetrance_list <- readRDS(here("final-figures", "figure-2", "rdata_files", "02-save-nygc-rci-expression-summary-with-penetrance-matrix.rds"))
clinical_metadata_df <- read.csv(here("13-nygc-consortium", "input_files", "samplesheet.csv"))
nygc_patient_liability_df <- nygc_penetrance_list$liability_df
nygc_penetrance_matrix <- nygc_penetrance_list$penetrance_matrix

```

### Analysis


- Get the genetic group 

```{r}

genetic_test_df <- unique(clinical_metadata_df[,c("external_subject_id", "genetic_test")])
genetic_test_df$C9orf72_status <- "Unknown"
genetic_test_df$C9orf72_status[grepl(x = genetic_test_df$genetic_test, pattern = "Positive for C9orf72 Repeat Expansion")] <- "Positive"
genetic_test_df$C9orf72_status[grepl(x = genetic_test_df$genetic_test, pattern = "Positive for C9orf72 Repeat Expansion") & grepl(x = genetic_test_df$genetic_test, pattern = "conflicts")] <- "Negative"
genetic_test_df$C9orf72_status[grepl(x = genetic_test_df$genetic_test, pattern = "Negative for C9orf72 Repeat Expansion")] <- "Negative"
genetic_test_df$C9orf72_status[grepl(x = genetic_test_df$genetic_test, pattern = "Negative for C9orf72 Repeat Expansion") & grepl(x = genetic_test_df$genetic_test, pattern = "conflicts")] <- "Negative"


```


- Prepare data for brain and spinal cord age at death


```{r}

nygc_patient_liability_df$type <- factor(nygc_patient_liability_df$type, levels = c("control", "c9orf72-ALS", "other_mutation-fALS", "sporadic"))
nygc_patient_liability_df$type_group <- as.character(nygc_patient_liability_df$type)
nygc_patient_liability_df$type_group[nygc_patient_liability_df$type %in% c("other_mutation-fALS", "sporadic")] <- "sporadic+other_fALS"
nygc_patient_liability_df$tissue_group <- "spinal_cord"
nygc_patient_liability_df$tissue_group[nygc_patient_liability_df$tissue %in% c("Cortex", "Cortex_Motor", "Cerebellum")] <- "brain"

patient_surv_df <- do.call(rbind, lapply(split(nygc_patient_liability_df, paste(nygc_patient_liability_df$subject_id, nygc_patient_liability_df$tissue)), function(temp_df){
    message(unique(temp_df$subject_id))
    data.frame(subject_id = unique(temp_df$subject_id),
        tissue = unique(temp_df$tissue),
        # tissue_group = unique(temp_df$tissue_group),
        age_at_onset = unique(temp_df$age_at_symptom_onset),
        age_at_death = unique(temp_df$age_at_death),
        disease_duration_in_months = unique(temp_df$disease_duration_in_months),
        sex_genotype = unique(temp_df$sex_genotype),
        mnd_with_ftd = unique(temp_df$mnd_with_ftd),
        condition = unique(temp_df$condition),
        mnd_with_dementia = unique(temp_df$mnd_with_dementia),
        type_group = unique(temp_df$type_group),
        rci_liability = mean(temp_df$rci_liability),
        intron_expression_rate = mean(temp_df$intron_expression_rate))
}))

patient_surv_df$age_at_death[patient_surv_df$age_at_death == "90 or Older"] <- "90"
patient_surv_df$age_at_death <- as.numeric(patient_surv_df$age_at_death)
patient_surv_df$age_at_onset <- as.numeric(patient_surv_df$age_at_onset)
patient_surv_df$disease_duration <- as.numeric(patient_surv_df$disease_duration_in_months)

patient_surv_df$expansion_status <- genetic_test_df$C9orf72_status[match(patient_surv_df$subject_id, genetic_test_df$external_subject_id)]
patient_surv_df$expansion_status <- factor(patient_surv_df$expansion_status, levels = c("Unknown", "Positive", "Negative"))

```

- Perform modelling of the age of onset survival function using rci liability

```{r}

require(patchwork)
als_surv_df <- patient_surv_df[!(patient_surv_df$condition %in% c("als+ftd")),]
als_surv_df$condition <- factor(als_surv_df$condition, levels = c("ctrl", "als"))
als_surv_df$sex_genotype <- factor(als_surv_df$sex_genotype, levels = c("XX", "XY"))
als_surv_df$status <- 1
als_surv_df$status[is.na(als_surv_df$age_at_death)] <- 0

onset_vs_liability_model_plots <- lapply(unique(als_surv_df$tissue), function(x){
    current_als_surv_df <- als_surv_df[als_surv_df$tissue == x,]
    current_als_surv_df$rci_liability_scaled <- scale(current_als_surv_df$rci_liability)
    model <- coxph(Surv(age_at_death, status) ~ condition + sex_genotype + rci_liability_scaled, 
        data=current_als_surv_df, x=TRUE)
    ggforest(model, data = current_als_surv_df)
})
names(onset_vs_liability_model_plots) <- unique(als_surv_df$tissue)

all_onset_plots <- wrap_plots(onset_vs_liability_model_plots[
    c("Spinal_Cord_Cervical", "Spinal_Cord_Lumbar", "Spinal_Cord_Thoracic")])

ggsave(file = file.path(plot_dir, "07-04-plot-cox-model-output-for-age-at-onset.pdf"),
all_onset_plots, width = 10, height = 5)

```

- Perform modelling of the disease duration survival function using rci liability

```{r}

require(patchwork)
als_surv_df <- patient_surv_df[(patient_surv_df$condition %in% c("als")),]
als_surv_df$sex_genotype <- factor(als_surv_df$sex_genotype, levels = c("XX", "XY"))
als_surv_df$status <- 1
als_surv_df$status[als_surv_df$age_atis.na(als_surv_df$disease_duration)] <- 0

duration_vs_emission_model_plots <- lapply(unique(als_surv_df$tissue), function(x){
    current_als_surv_df <- als_surv_df[als_surv_df$tissue == x,]
    current_als_surv_df$rci_emission_scaled <- scale(current_als_surv_df$intron_expression_rate)
    model <- coxph(Surv(disease_duration, status) ~ age_at_onset + sex_genotype + rci_emission_scaled, 
        data = current_als_surv_df, x=TRUE)
    ggforest(model, data = current_als_surv_df)
})
names(duration_vs_emission_model_plots) <- unique(als_surv_df$tissue)

all_duration_plots <- wrap_plots(duration_vs_emission_model_plots[
    c("Spinal_Cord_Cervical", "Spinal_Cord_Lumbar", "Spinal_Cord_Thoracic")])

ggsave(file = file.path(plot_dir, "07-05-forest-plot-of-cox-proportional-hazards-model-for-disease-duration.pdf"), all_duration_plots, width = 10, height = 5)

```

