---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")
library(rlang)
library(riskRegression)
library(survival)
library(survminer)
library(pammtools)
library(gganimate)
library(transformr)
library(plotly)
library(reshape2)
library(knitr)
library(rmarkdown)

require("contsurvplot")
require("RColorBrewer")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-3"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load penetrance data

```{r}

nygc_penetrance_list <- readRDS(file.path("final-figures/figure-2/rdata_files/02-save-nygc-rci-expression-summary-with-penetrance-matrix.rds"))
clinical_metadata_df <- read.csv("0103-NYGC-consortium/input_files/samplesheet.csv")
nygc_patient_liability_df <- nygc_penetrance_list$liability_df
nygc_penetrance_matrix <- nygc_penetrance_list$penetrance_matrix

```

### Analysis


- Get the genetic group 

```{r}

genetic_test_df <- unique(clinical_metadata_df[,c("external_subject_id", "genetic_test")])
genetic_test_df$C9orf72_status <- "Unknown"
genetic_test_df$C9orf72_status[grepl(x = genetic_test_df$genetic_test, pattern = "Positive for C9orf72 Repeat Expansion")] <- "Positive"
genetic_test_df$C9orf72_status[grepl(x = genetic_test_df$genetic_test, pattern = "Positive for C9orf72 Repeat Expansion") & grepl(x = genetic_test_df$genetic_test, pattern = "conflicts")] <- "Negative"
genetic_test_df$C9orf72_status[grepl(x = genetic_test_df$genetic_test, pattern = "Negative for C9orf72 Repeat Expansion")] <- "Negative"
genetic_test_df$C9orf72_status[grepl(x = genetic_test_df$genetic_test, pattern = "Negative for C9orf72 Repeat Expansion") & grepl(x = genetic_test_df$genetic_test, pattern = "conflicts")] <- "Negative"


```


- Prepare data for brain and spinal cord age at death


```{r}

nygc_patient_liability_df$type <- factor(nygc_patient_liability_df$type, levels = c("control", "c9orf72-ALS", "other_mutation-fALS", "sporadic"))
nygc_patient_liability_df$type_group <- as.character(nygc_patient_liability_df$type)
nygc_patient_liability_df$type_group[nygc_patient_liability_df$type %in% c("other_mutation-fALS", "sporadic")] <- "sporadic+other_fALS"
nygc_patient_liability_df$tissue_group <- "spinal_cord"
nygc_patient_liability_df$tissue_group[nygc_patient_liability_df$tissue %in% c("Cortex", "Cortex_Motor", "Cerebellum")] <- "brain"

patient_surv_df <- do.call(rbind, lapply(split(nygc_patient_liability_df, paste(nygc_patient_liability_df$subject_id, nygc_patient_liability_df$tissue_group)), function(temp_df){
    message(unique(temp_df$subject_id))
    data.frame(subject_id = unique(temp_df$subject_id),
        tissue_group = unique(temp_df$tissue_group),
        age_at_onset = unique(temp_df$age_at_symptom_onset),
        age_at_death = unique(temp_df$age_at_death),
        disease_duration_in_months = unique(temp_df$disease_duration_in_months),
        sex_genotype = unique(temp_df$sex_genotype),
        mnd_with_ftd = unique(temp_df$mnd_with_ftd),
        condition = unique(temp_df$condition),
        mnd_with_dementia = unique(temp_df$mnd_with_dementia),
        type_group = unique(temp_df$type_group),
        rci_liability = mean(temp_df$rci_liability),
        intron_expression_rate = mean(temp_df$intron_expression_rate))
}))

patient_surv_df$age_at_death[patient_surv_df$age_at_death == "90 or Older"] <- "90"
patient_surv_df$age_at_death <- as.numeric(patient_surv_df$age_at_death)
patient_surv_df$age_at_onset <- as.numeric(patient_surv_df$age_at_onset)
patient_surv_df$disease_duration <- as.numeric(patient_surv_df$disease_duration_in_months)

patient_surv_df$expansion_status <- genetic_test_df$C9orf72_status[match(patient_surv_df$subject_id, genetic_test_df$external_subject_id)]
patient_surv_df$expansion_status <- factor(patient_surv_df$expansion_status, levels = c("Unknown", "Positive", "Negative"))


```

- Prep data

```{r}

als_surv_df <- patient_surv_df[(patient_surv_df$condition %in% c("als")),]
als_surv_df$sex_genotype <- factor(als_surv_df$sex_genotype, levels = c("XX", "XY"))

# als_surv_df$condition <- factor(als_surv_df$condition, levels = c("ctrl", "als"))
als_surv_df$status <- 1
als_surv_df$status[is.na(als_surv_df$age_at_onset) | is.na(als_surv_df$disease_duration)] <- 0

rci_related_columns <- c("rci_liability", "intron_expression_rate")

spinal_cord_df <- als_surv_df[als_surv_df$tissue_group == "spinal_cord",c("subject_id", "disease_duration", "age_at_onset", "age_at_death", "status", "condition", "expansion_status", rci_related_columns)]

brain_df <- als_surv_df[als_surv_df$tissue_group == "brain",c("subject_id", "disease_duration", "age_at_onset", "age_at_death", "status", "condition", "expansion_status", rci_related_columns)]
brain_df$condition <- factor(brain_df$condition, levels = c("ctrl", "als"))

```

- Make random survival forests for spinal cord


```{r}

require("randomForestSRC")
require("parallel")
# generate seeds

seed_file <- file.path(save_dir, "01-seed-file-for-comparing-RFS-survival-curves-between-als-and-control.rds")

if(!file.exists(seed_file)){
    seed_list <- sample.int(.Machine$integer.max, 1000)    
    saveRDS(object = seed_list, file = seed_file)
}else{
    seed_list <- readRDS(seed_file)
}

spinal_cord_liability_and_disease_models <- lapply(seed_list, function(x){
    set.seed(x)
    message(x)
    rsf_model <- rfsrc(Surv(disease_duration, status) ~ intron_expression_rate, 
        data = spinal_cord_df, ntree = 1000, importance = TRUE)
})  


```

- Get risk scores


```{r}


require("matrixStats")

non_na_spinal_cord_df <- spinal_cord_df[
    !is.na(spinal_cord_df$age_at_onset) & !is.na(spinal_cord_df$disease_duration), 
    c("subject_id", "disease_duration", "age_at_onset", "age_at_death", "status", "condition", "expansion_status", rci_related_columns)]

# get predicted risk 

spinal_cord_risk_score_matrix <- do.call(cbind, lapply(spinal_cord_liability_and_disease_models, function(x){
    message("Done...")
    rsf_pred <- predict(x, newdata = non_na_spinal_cord_df)
    rsf_pred$predicted
}))


non_na_spinal_cord_df$risk_score <- rowMedians(spinal_cord_risk_score_matrix)
median_risk <- median(non_na_spinal_cord_df$risk_score, na.rm = T)
non_na_spinal_cord_df$risk_group <- ifelse(non_na_spinal_cord_df$risk_score >= median_risk, "high", "low")

differential_survival <- survdiff(Surv(disease_duration, status) ~ risk_group, data = non_na_spinal_cord_df)
(differential_survival$obs[1] / differential_survival$exp[1]) / (differential_survival$obs[2] / differential_survival$exp[2])


```

- Plot survival curves

```{r}

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

fit <- survfit(Surv(disease_duration, status) ~ risk_group, data = non_na_spinal_cord_df)

the_plot <- ggsurvplot(fit, 
    data = non_na_spinal_cord_df,
    palette = c("#FF5666", "#FFCCC9"),
    pval = T, pval.method = T,
    conf.int = F,
    # surv.median.line = c("v"),
    risk.table = "absolute",
    ggtheme = theme_bw() + current_theme)$plot
the_plot <- the_plot + ggtitle(paste("random forest survival curve for disease duration"))

ggsave(width = 7, height = 7, file = file.path(plot_dir, "09-04-REBUTTAL-plot-random-forest-survival-curve-for-spinal-cord-ALS-disease-duration.pdf"), the_plot)


```

- Plot RCI emission rate for risk groups

```{r}

require("ggbeeswarm")
current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

# Plot the two ROC curves together
colours <- c("high" = "#FF5666", "low" = "#FFCCC9")
The_plot <- ggplot(non_na_spinal_cord_df, aes(x = risk_group, y = intron_expression_rate))
The_plot <- The_plot + geom_boxplot(aes(fill = risk_group), colour = "#000000", 
    width = 0.1, lwd = 0.5, outlier.shape = NA)
The_plot <- The_plot + geom_quasirandom(aes(fill = risk_group), shape = 21, colour = "#000000")
The_plot <- The_plot + scale_x_discrete("Risk group")
The_plot <- The_plot + scale_y_continuous("RCI emission rate")
The_plot <- The_plot + scale_fill_manual("Risk group", values = colours)
The_plot <- The_plot + labs(title = "Age at symptom onset")
# The_plot <- The_plot + facet_wrap(. ~ tissue_group, scales = "free")
The_plot <- The_plot + theme_bw() + current_theme
ggsave(width = 4, height = 7, file = file.path(plot_dir, "09-05-plot-rci-emission-rate-for-high-and-low-risk-groups-in-disease-duration.pdf"), The_plot)


wilcox.test(non_na_spinal_cord_df[non_na_spinal_cord_df$risk_group == "high", "intron_expression_rate"],
    non_na_spinal_cord_df[non_na_spinal_cord_df$risk_group == "low", "intron_expression_rate"])

# data:  non_na_spinal_cord_df[non_na_spinal_cord_df$risk_group == "high", "intron_expression_rate"] and non_na_spinal_cord_df[non_na_spinal_cord_df$risk_group == "low", "intron_expression_rate"]
# W = 3664, p-value = 0.04085
# alternative hypothesis: true location shift is not equal to 0


# Plot the two ROC curves together
colours <- c("high" = "#FF5666", "low" = "#FFCCC9")
The_plot <- ggplot(non_na_spinal_cord_df, aes(x = risk_group, y = age_at_onset))
The_plot <- The_plot + geom_boxplot(aes(fill = risk_group), colour = "#000000", 
    width = 0.1, lwd = 0.5, outlier.shape = NA)
The_plot <- The_plot + geom_quasirandom(aes(fill = risk_group), shape = 21, colour = "#000000")
The_plot <- The_plot + scale_x_discrete("Risk group")
The_plot <- The_plot + scale_y_continuous("age at onset")
The_plot <- The_plot + scale_fill_manual("Risk group", values = colours)
The_plot <- The_plot + labs(title = "Age at symptom onset")
# The_plot <- The_plot + facet_wrap(. ~ tissue_group, scales = "free")
The_plot <- The_plot + theme_bw() + current_theme
ggsave(width = 4, height = 7, file = file.path(plot_dir, "09-05-plot-age-at-onset-for-high-and-low-risk-groups-in-disease-duration.pdf"), The_plot)

wilcox.test(non_na_spinal_cord_df[non_na_spinal_cord_df$risk_group == "high", "age_at_onset"],
    non_na_spinal_cord_df[non_na_spinal_cord_df$risk_group == "low", "age_at_onset"])

#         Wilcoxon rank sum test with continuity correction

# data:  non_na_spinal_cord_df[non_na_spinal_cord_df$risk_group == "high", "age_at_onset"] and non_na_spinal_cord_df[non_na_spinal_cord_df$risk_group == "low", "age_at_onset"]
# W = 3927, p-value = 0.00297
# alternative hypothesis: true location shift is not equal to 0


```

- Make random forest models for brain

```{r}


brain_liability_and_disease_models <- lapply(seed_list, function(x){
    set.seed(x)
    message(x)
    rsf_model <- rfsrc(Surv(disease_duration, status) ~ intron_expression_rate, 
        data = brain_df, ntree = 1000, importance = TRUE)
})


```

- Get risk scores


```{r}


require("matrixStats")

non_na_brain_df <- brain_df[
    !is.na(brain_df$age_at_onset) & !is.na(brain_df$disease_duration), 
    c("subject_id", "disease_duration", "age_at_onset", "age_at_death", "status", "condition", "expansion_status", rci_related_columns)]

# get predicted risk 

brain_risk_score_matrix <- do.call(cbind, lapply(brain_liability_and_disease_models, function(x){
    message("Done...")
    rsf_pred <- predict(x, newdata = non_na_brain_df)
    rsf_pred$predicted
}))


non_na_brain_df$risk_score <- rowMedians(brain_risk_score_matrix)
median_risk <- median(non_na_brain_df$risk_score, na.rm = T)
non_na_brain_df$risk_group <- ifelse(non_na_brain_df$risk_score >= median_risk, "high", "low")

differential_survival <- survdiff(Surv(disease_duration, status) ~ risk_group, data = non_na_brain_df)
(differential_survival$obs[1] / differential_survival$exp[1]) / (differential_survival$obs[2] / differential_survival$exp[2])


```

- Plot survival curves

```{r}

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

fit <- survfit(Surv(disease_duration, status) ~ risk_group, data = non_na_brain_df)

the_plot <- ggsurvplot(fit, 
    data = non_na_brain_df,
    palette = c("#FF5666", "#FFCCC9"),
    pval = T, pval.method = T,
    conf.int = F,
    # surv.median.line = c("v"),
    risk.table = "absolute",
    ggtheme = theme_bw() + current_theme)$plot
the_plot <- the_plot + ggtitle(paste("random forest survival curve for disease duration"))

ggsave(width = 7, height = 7, file = file.path(plot_dir, "09-04-REBUTTAL-plot-random-forest-brain-for-spinal-cord-ALS-disease-duration.pdf"), the_plot)


```

- Plot RCI emission rate for risk groups

```{r}

require("ggbeeswarm")
current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

# Plot the two ROC curves together
colours <- c("high" = "#FF5666", "low" = "#FFCCC9")
The_plot <- ggplot(non_na_brain_df, aes(x = risk_group, y = intron_expression_rate))
The_plot <- The_plot + geom_boxplot(aes(fill = risk_group), colour = "#000000", 
    width = 0.1, lwd = 0.5, outlier.shape = NA)
The_plot <- The_plot + geom_quasirandom(aes(fill = risk_group), shape = 21, colour = "#000000")
The_plot <- The_plot + scale_x_discrete("Risk group")
The_plot <- The_plot + scale_y_continuous("RCI emission rate")
The_plot <- The_plot + scale_fill_manual("Risk group", values = colours)
The_plot <- The_plot + labs(title = "Age at symptom onset")
# The_plot <- The_plot + facet_wrap(. ~ tissue_group, scales = "free")
The_plot <- The_plot + theme_bw() + current_theme
ggsave(width = 4, height = 7, file = file.path(plot_dir, "09-05-plot-rci-emission-rate-for-brain-high-and-low-risk-groups-in-disease-duration.pdf"), The_plot)


wilcox.test(non_na_brain_df[non_na_brain_df$risk_group == "high", "intron_expression_rate"],
    non_na_brain_df[non_na_brain_df$risk_group == "low", "intron_expression_rate"])

# data:  non_na_spinal_cord_df[non_na_spinal_cord_df$risk_group == "high", "intron_expression_rate"] and non_na_spinal_cord_df[non_na_spinal_cord_df$risk_group == "low", "intron_expression_rate"]
# W = 3664, p-value = 0.04085
# alternative hypothesis: true location shift is not equal to 0


# Plot the two ROC curves together
colours <- c("high" = "#FF5666", "low" = "#FFCCC9")
The_plot <- ggplot(non_na_brain_df, aes(x = risk_group, y = age_at_onset))
The_plot <- The_plot + geom_boxplot(aes(fill = risk_group), colour = "#000000", 
    width = 0.1, lwd = 0.5, outlier.shape = NA)
The_plot <- The_plot + geom_quasirandom(aes(fill = risk_group), shape = 21, colour = "#000000")
The_plot <- The_plot + scale_x_discrete("Risk group")
The_plot <- The_plot + scale_y_continuous("age at onset")
The_plot <- The_plot + scale_fill_manual("Risk group", values = colours)
The_plot <- The_plot + labs(title = "Age at symptom onset")
# The_plot <- The_plot + facet_wrap(. ~ tissue_group, scales = "free")
The_plot <- The_plot + theme_bw() + current_theme
ggsave(width = 4, height = 7, file = file.path(plot_dir, "09-05-plot-age-at-onset-for-brain-high-and-low-risk-groups-in-disease-duration.pdf"), The_plot)

wilcox.test(non_na_spinal_cord_df[non_na_spinal_cord_df$risk_group == "high", "age_at_onset"],
    non_na_spinal_cord_df[non_na_spinal_cord_df$risk_group == "low", "age_at_onset"])


#         Wilcoxon rank sum test with continuity correction

# data:  non_na_spinal_cord_df[non_na_spinal_cord_df$risk_group == "high", "age_at_onset"] and non_na_spinal_cord_df[non_na_spinal_cord_df$risk_group == "low", "age_at_onset"]
# W = 3927, p-value = 0.00297
# alternative hypothesis: true location shift is not equal to 0

```

- save all data 

```{r}

non_na_brain_df$tissue_group <- "brain"
non_na_spinal_cord_df$tissue_group <- "spinal_cord"
non_na_all_samples_df <- rbind(non_na_brain_df,non_na_spinal_cord_df)

saveRDS(object = non_na_all_samples_df, file = file.path(save_dir, "04-01-save-patients-with-rsf-risk-based-on-disease-duration.rds"))

```