---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

library("here")
message("Project root:", here(), "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("ggplot2")
library(rlang)
library(riskRegression)
library(survival)
library(survminer)
library(pammtools)
library(gganimate)
library(transformr)
library(plotly)
library(reshape2)
library(knitr)
library(rmarkdown)
library(parallel)
require("contsurvplot")
require("RColorBrewer")
require("STRINGdb")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("final-figures","figure-3")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
tmp_dir <- file.path(analysis_path, "temp")


dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)
dir.create(tmp_dir, recursive = T)


```

### 3.0 Imports

- Load penetrance data

```{r}

nygc_penetrance_list <- readRDS(here("final-figures", "figure-2", "rdata_files", "02-save-nygc-rci-expression-summary-with-penetrance-matrix.rds"))
patient_df <- readRDS(here("13-nygc-consortium", "rdata_files", "03-05-nygc-patient-sampleinfo-datasheet.rds"))
clinical_metadata_df <- read.csv(here("0103-NYGC-consortium", "input_files", "samplesheet.csv"))

postmortem_list <- readRDS(here("13-nygc-consortium", "rdata_files", "03-02-list_of_counts_stat_samplesheet_from_nygc.rds"))
feature_ranges <- readRDS(here("05-simulated-reads-differential-profiling", "rdata_files", "01-all-merged-exonic-and-intronic-segments-in-genes.rds"))

disease_duration_rsf_df <- readRDS(here("final-figures", "figure-3", "rdata_files", "04-01-save-patients-with-rsf-risk-based-on-disease-duration.rds"))
age_of_onset_rsf_df <- readRDS(here("final-figures", "figure-3", "rdata_files", "02-summarised-patient-dataframe-with-rsf-risk-scores.rds"))

```

### Analysis


- Get gene level counts

```{r}

patient_df$tissue_group <- "spinal_cord"
patient_df$tissue_group[patient_df$tissue %in% c("Cortex", "Cortex_Motor", "Cerebellum")] <- "brain"
counts_matrix <- postmortem_list$counts

gene_level_counts_matrix <- do.call(rbind, mclapply(split(feature_ranges$hash_id[feature_ranges$type == "exon"], feature_ranges$gene_id[feature_ranges$type == "exon"]), function(hash_id){
    if(length(hash_id) == 1){
        return(counts_matrix[hash_id,])
    }
    return(colSums(counts_matrix[hash_id,]))
}, mc.cores = 8))

summarised_gene_count_matrix <- do.call(cbind, mclapply(split(patient_df$column_id, paste(patient_df$subject_id, patient_df$tissue_group)), function(column_ids){
    if(length(column_ids) == 1){
        return(gene_level_counts_matrix[,column_ids])
    }
    return(rowMeans(gene_level_counts_matrix[,column_ids]))
}, mc.cores = 8))


```

- Import gtf

```{r}

require("rtracklayer")
annotation_gtf <- here("Genomes","Homo_sapiens.GRCh38.104.chr_patch_hapl_scaff.gtf")
annotation_ranges <- import.gff2(annotation_gtf)
annotation_gene_ranges <- annotation_ranges[annotation_ranges$type == "gene"]

```

- Intersect risks


```{r}

age_of_onset_rsf_df$unique_id <- paste(age_of_onset_rsf_df$subject_id, age_of_onset_rsf_df$tissue_group)
disease_duration_rsf_df$unique_id <- paste(disease_duration_rsf_df$subject_id, disease_duration_rsf_df$tissue_group)

intersect_ids <- intersect(age_of_onset_rsf_df$unique_id, disease_duration_rsf_df$unique_id)

select_age_of_onset_rsf_df <- age_of_onset_rsf_df[age_of_onset_rsf_df$unique_id %in% intersect_ids,]
select_age_of_onset_rsf_df$disease_duration <- disease_duration_rsf_df$disease_duration[match(select_age_of_onset_rsf_df$unique_id, disease_duration_rsf_df$unique_id)]
select_age_of_onset_rsf_df$onset_risk <- select_age_of_onset_rsf_df$risk_group
select_age_of_onset_rsf_df$onset_risk[select_age_of_onset_rsf_df$onset_risk == "high"] <- "early"
select_age_of_onset_rsf_df$onset_risk[select_age_of_onset_rsf_df$onset_risk == "low"] <- "late"

select_age_of_onset_rsf_df$duration_risk <- disease_duration_rsf_df$risk_group[match(select_age_of_onset_rsf_df$unique_id, disease_duration_rsf_df$unique_id)]
select_age_of_onset_rsf_df$duration_risk[select_age_of_onset_rsf_df$duration_risk == "high"] <- "fast"
select_age_of_onset_rsf_df$duration_risk[select_age_of_onset_rsf_df$duration_risk == "low"] <- "slow"

```

- How many patients are in each group

```{r}

table(select_age_of_onset_rsf_df$onset_risk, select_age_of_onset_rsf_df$duration_risk)

  #      high low
  # high  107 106
  # low    61  61


```

- Perform differential expression profiling


```{r}

require("DESeq2")

selected_gene_counts_matrix <- summarised_gene_count_matrix[rowMeans(summarised_gene_count_matrix) > 1,]
deseq_object <- DESeqDataSetFromMatrix(countData = round(selected_gene_counts_matrix[,select_age_of_onset_rsf_df$unique_id[select_age_of_onset_rsf_df$tissue_group == "spinal_cord"]]), 
    colData = select_age_of_onset_rsf_df[select_age_of_onset_rsf_df$tissue_group == "spinal_cord",], 
    design = ~ onset_risk + duration_risk + onset_risk:duration_risk)
differential_object <- DESeq(deseq_object)


```

- Make contrasts to determine processes governing duration risk

```{r}

model_matrix <- model.matrix(design(differential_object), colData(differential_object))


```

- Identify factors governing onset risk


```{r}

onset <- colMeans(model_matrix[differential_object$onset_risk == "early",]) - colMeans(model_matrix[differential_object$onset_risk == "late",])
results_object <- results(differential_object, contrast = onset)
results_object$contrast <- "early_vs_late"
early_vs_late_df <- results_object
early_vs_late_df[early_vs_late_df$pvalue < 0.05 & !is.na(early_vs_late_df$pvalue),]


```


- Identify factors governing duration risk 

```{r}

duration <- colMeans(model_matrix[differential_object$duration_risk == "fast",]) - colMeans(model_matrix[differential_object$duration_risk == "slow",])
results_object <- results(differential_object, contrast = duration)
results_object$contrast <- "fast_vs_slow"
fast_vs_slow_df <- results_object
fast_vs_slow_df[fast_vs_slow_df$pvalue < 0.05 & !is.na(fast_vs_slow_df$pvalue),]

```

- Identify factors governing fast progressors in late onset patients

```{r}

late_fast_vs_late_slow <- colMeans(model_matrix[differential_object$onset_risk == "late" & differential_object$duration_risk == "fast",]) - colMeans(model_matrix[differential_object$onset_risk == "late" & differential_object$duration_risk == "slow",])

results_object <- results(differential_object, contrast = late_fast_vs_late_slow)
results_object$contrast <- "late_fast_vs_late_slow"
late_fast_vs_late_slow <- results_object
late_fast_vs_late_slow[late_fast_vs_late_slow$pvalue < 0.05 & !is.na(late_fast_vs_late_slow$pvalue),]


```

- Identify factors governing fast progressors in early onset patients

```{r}

# No genes detected

early_fast_vs_early_slow <- colMeans(model_matrix[differential_object$onset_risk == "early" & differential_object$duration_risk == "fast",]) - colMeans(model_matrix[differential_object$onset_risk == "early" & differential_object$duration_risk == "slow",])

results_object <- results(differential_object, contrast = early_fast_vs_early_slow)
results_object$contrast <- "early_fast_vs_early_slow"
early_fast_vs_early_slow <- results_object
early_fast_vs_early_slow[early_fast_vs_early_slow$pvalue < 0.05 & !is.na(early_fast_vs_early_slow$pvalue),]


```

- Identify factors governing early onset fast progressors vs late onset slow progressors

```{r}

# No genes detected

early_fast_vs_late_slow <- colMeans(model_matrix[differential_object$onset_risk == "early" & differential_object$duration_risk == "fast",]) - colMeans(model_matrix[differential_object$onset_risk == "late" & differential_object$duration_risk == "slow",])

results_object <- results(differential_object, contrast = early_fast_vs_late_slow)
results_object$contrast <- "early_fast_vs_late_slow"
early_fast_vs_late_slow <- results_object
early_fast_vs_late_slow[early_fast_vs_late_slow$pvalue < 0.05 & !is.na(early_fast_vs_late_slow$pvalue),]


```


- Perform gsea

```{r}

require("msigdbr")
require("fgsea")
require("parallel")

hallmark_gene_sets = msigdbr(species = "human", category = "H")

msigdbr_list = split(x = hallmark_gene_sets$gene_symbol, f = hallmark_gene_sets$gs_name)


contrast_df <- rbind(early_vs_late_df[early_vs_late_df$pvalue < 0.05 & !is.na(early_vs_late_df$pvalue),], 
    fast_vs_slow_df[fast_vs_slow_df$pvalue < 0.05 & !is.na(fast_vs_slow_df$pvalue),], 
    early_fast_vs_early_slow[early_fast_vs_early_slow$pvalue < 0.05 & !is.na(early_fast_vs_early_slow$pvalue),], 
    late_fast_vs_late_slow[late_fast_vs_late_slow$pvalue < 0.05 & !is.na(late_fast_vs_late_slow$pvalue),],
    early_fast_vs_late_slow[early_fast_vs_late_slow$pvalue < 0.05 & !is.na(early_fast_vs_late_slow$pvalue),])

contrast_df$gene_name <- annotation_gene_ranges$gene_name[match(rownames(contrast_df), annotation_gene_ranges$gene_id)]
contrast_split <- split(contrast_df, contrast_df$contrast)


# hallmark_gsea_df_list <- lapply(contrast_split, function(temp_df){
#     temp_non_na_df <- temp_df[!is.na(temp_df$gene_name) & !is.na(temp_df$log2FoldChange) & !is.na(temp_df$padj),]
#     temp_non_na_df$ranks <- rank(temp_non_na_df$log2FoldChange)
#     temp_non_na_df <- temp_non_na_df[order(temp_non_na_df$log2FoldChange, decreasing = T),]
#     a_vector <- temp_non_na_df$log2FoldChange
#     names(a_vector) <- temp_non_na_df$gene_name
#     fgsea(msigdbr_list, a_vector)
# })


# significant_go_bp_gsea_df_list <- lapply(hallmark_gsea_df_list, function(temp_df){
#     up_df <- temp_df[temp_df$padj < 0.05 & !is.na(temp_df$log2err) & temp_df$NES > 0,]
#     sig_up_df <- up_df[order(up_df$NES, decreasing = T),][1:min(nrow(up_df),10),]
#     sig_up_df$direction <- "up"

#     down_df <- temp_df[temp_df$padj < 0.05 & !is.na(temp_df$log2err) & temp_df$NES < 0,]
#     sig_down_df <- down_df[order(down_df$NES, decreasing = F),][1:min(nrow(down_df),10),]
#     sig_down_df$direction <- "down"
    
#     return(rbind(sig_up_df, sig_down_df))
# })

```

- Perform gene set enrichment

```{r}

library("enrichR")
dbs <- "GO_Biological_Process_2023"
directions <- c("up", "down")


contrast_df <- rbind(
    early_vs_late_df[early_vs_late_df$pvalue < 0.05 & !is.na(early_vs_late_df$pvalue),], 
    fast_vs_slow_df[fast_vs_slow_df$pvalue < 0.05 & !is.na(fast_vs_slow_df$pvalue),], 
    early_fast_vs_early_slow[early_fast_vs_early_slow$pvalue < 0.05 & !is.na(early_fast_vs_early_slow$pvalue),], 
    late_fast_vs_late_slow[late_fast_vs_late_slow$pvalue < 0.05 & !is.na(late_fast_vs_late_slow$pvalue),],
    early_fast_vs_late_slow[early_fast_vs_late_slow$pvalue < 0.05 & !is.na(early_fast_vs_late_slow$pvalue),])

contrast_df$gene_name <- annotation_gene_ranges$gene_name[match(rownames(contrast_df), annotation_gene_ranges$gene_id)]
contrast_df$direction <- ifelse(contrast_df$log2FoldChange > 0, "up", "down")

contrast_split <- split(contrast_df, paste(contrast_df$contrast, contrast_df$direction))
enrichment_result_df <- do.call(rbind, lapply(contrast_split, function(temp_df){
    db_df_list <- lapply(dbs, function(a_database){
        current_genes <- temp_df$gene_name[!is.na(temp_df$gene_name)]
        an_enrichment_df <- enrichr(current_genes,  a_database)[[1]]
        if(nrow(an_enrichment_df) == 0){
            return(NULL)
        }
        an_enrichment_df$contrast <- unique(temp_df$contrast)
        an_enrichment_df$direction <- unique(temp_df$direction)
        an_enrichment_df$num_gene <- length(current_genes)
        an_enrichment_df$db <- a_database
        Sys.sleep(5)
        return(an_enrichment_df)
    })
    db_df <- do.call(rbind, db_df_list)
}))


enrichment_result_df$gene_ratio <- vapply(enrichment_result_df$Overlap, function(x){
    eval(parse(text = x))
},1)
significant_gene_set_enrichment <- enrichment_result_df[enrichment_result_df$P.value < 0.05 &
    (enrichment_result_df$gene_ratio > 0.1),]
# significant_gene_set_enrichment$ids <- vapply(str_extract_all(string = significant_gene_set_enrichment$Term, pattern = "(?<=\\()(.*?)(?=\\))"), function(x){x[length(x)]},"1")
# # write.csv(significant_gene_set_enrichment, file = file.path(output_dir, "03-gene-set-enrichment-with-go-bp.csv"))
# # Perform enrichment smplify


selected_significant_gene_set_enrichment <- lapply(split(significant_gene_set_enrichment, paste(significant_gene_set_enrichment$contrast, significant_gene_set_enrichment$direction)), function(temp_df){
    return_df <- temp_df[order(temp_df$P.value, decreasing = F),][1:min(c(nrow(temp_df), 10)),]
    return_df$term_only <- vapply(str_split(return_df$Term, pattern = " "), function(x){ paste(x[-length(x)], collapse = " ")}, "1")
    return_df$term_only <- factor(return_df$term_only, levels = rev(return_df$term_only))
    return(return_df)
})


# selected_significant_gene_set_enrichment[selected_significant_gene_set_enrichment$contrast == "early_fast_vs_late_slow",c("Term", "direction")]


# contrast_split <- split(selected_significant_gene_set_enrichment, selected_significant_gene_set_enrichment$contrast)

lapply(selected_significant_gene_set_enrichment, function(temp_df){
    current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

    colours <- c("down" = "#065A82", "up" = "#E3655B")
    ThePlot <- ggplot(temp_df, aes(x = -log10(P.value), y = term_only))
    # ThePlot <- ThePlot + geom_vline(xintercept = 0, linetype = "dashed", colour = "#000000")
    ThePlot <- ThePlot + geom_point(aes(fill = direction, size = gene_ratio), colour = "#000000", shape = 21)
    ThePlot <- ThePlot + scale_x_continuous("-log10(p-value)")
    ThePlot <- ThePlot + scale_y_discrete("GO:BP enriched terms")
    ThePlot <- ThePlot + theme_bw()
    ThePlot <- ThePlot + scale_fill_manual("Direction", values = colours)
    ThePlot <- ThePlot + facet_wrap(. ~ direction, scales = "free", nrow = 1, ncol = 2)
    ThePlot <- ThePlot + ggtitle("gene set enrichment for repeat containing genes")
    ThePlot <- ThePlot + current_theme
    ggsave(height = 7, width = 5, file = file.path(plot_dir, paste(paste("09-08-plot-gene-set-enrichment-for",unique(temp_df$direction), unique(temp_df$contrast), sep = "-"), ".pdf", sep = "")), ThePlot)
})


```







