---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")
library(rlang)
library(riskRegression)
library(survival)
library(pammtools)
library(gganimate)
library(transformr)
library(plotly)
library(reshape2)
library(knitr)
library(rmarkdown)

require("contsurvplot")
require("RColorBrewer")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-3"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load penetrance data

```{r}

nygc_penetrance_list <- readRDS(file.path("final-figures/figure-2/rdata_files/02-save-nygc-rci-expression-summary-with-penetrance-matrix.rds"))
nygc_patient_liability_df <- nygc_penetrance_list$liability_df
nygc_penetrance_matrix <- nygc_penetrance_list$penetrance_matrix

```

### Analysis


- Prepare data for survival analysis

```{r}

# nygc_patient_liability_df$type <- factor(nygc_patient_liability_df$type, levels = c("control", "c9orf72-ALS", "other_mutation-fALS", "sporadic"))
# nygc_patient_liability_df$type_group <- as.character(nygc_patient_liability_df$type)
# nygc_patient_liability_df$type_group[nygc_patient_liability_df$type %in% c("other_mutation-fALS", "sporadic")] <- "sporadic+other_fALS"
# nygc_patient_liability_df$tissue_group <- "spinal_cord"
# nygc_patient_liability_df$tissue_group[nygc_patient_liability_df$tissue %in% c("Cortex", "Cortex_Motor", "Cerebellum")] <- "brain"


# patient_surv_df <- do.call(rbind, lapply(split(nygc_patient_liability_df, paste(nygc_patient_liability_df$subject_id, nygc_patient_liability_df$tissue_group)), function(temp_df){
#     message(unique(temp_df$subject_id))
#     data.frame(subject_id = unique(temp_df$subject_id),
#         # tissue = unique(temp_df$tissue),
#         tissue_group = unique(temp_df$tissue_group),
#         age_at_onset = unique(temp_df$age_at_symptom_onset),
#         age_at_death = unique(temp_df$age_at_death),
#         disease_duration_in_months = unique(temp_df$disease_duration_in_months),
#         sex_genotype = unique(temp_df$sex_genotype),
#         mnd_with_ftd = unique(temp_df$mnd_with_ftd),
#         condition = unique(temp_df$condition),
#         mnd_with_dementia = unique(temp_df$mnd_with_dementia),
#         type_group = unique(temp_df$type_group),
#         rci_liability = median(temp_df$rci_liability),
#         intron_expression_rate = median(temp_df$intron_expression_rate))
# }))

# patient_surv_df$age_at_death[patient_surv_df$age_at_death == "90 or Older"] <- "90"
# patient_surv_df$age_at_death <- as.numeric(patient_surv_df$age_at_death)
# patient_surv_df$age_at_onset <- as.numeric(patient_surv_df$age_at_onset)
# patient_surv_df$disease_duration <- as.numeric(patient_surv_df$disease_duration_in_months)

```

- Prepare data for brain and spinal cord age at death


```{r}

nygc_patient_liability_df$type <- factor(nygc_patient_liability_df$type, levels = c("control", "c9orf72-ALS", "other_mutation-fALS", "sporadic"))
nygc_patient_liability_df$type_group <- as.character(nygc_patient_liability_df$type)
nygc_patient_liability_df$type_group[nygc_patient_liability_df$type %in% c("other_mutation-fALS", "sporadic")] <- "sporadic+other_fALS"
nygc_patient_liability_df$tissue_group <- "spinal_cord"
nygc_patient_liability_df$tissue_group[nygc_patient_liability_df$tissue %in% c("Cortex", "Cortex_Motor", "Cerebellum")] <- "brain"

patient_surv_df <- do.call(rbind, lapply(split(nygc_patient_liability_df, paste(nygc_patient_liability_df$subject_id, nygc_patient_liability_df$tissue_group)), function(temp_df){
    message(unique(temp_df$subject_id))
    data.frame(subject_id = unique(temp_df$subject_id),
        # tissue = unique(temp_df$tissue),
        tissue_group = unique(temp_df$tissue_group),
        age_at_onset = unique(temp_df$age_at_symptom_onset),
        age_at_death = unique(temp_df$age_at_death),
        disease_duration_in_months = unique(temp_df$disease_duration_in_months),
        sex_genotype = unique(temp_df$sex_genotype),
        mnd_with_ftd = unique(temp_df$mnd_with_ftd),
        condition = unique(temp_df$condition),
        mnd_with_dementia = unique(temp_df$mnd_with_dementia),
        type_group = unique(temp_df$type_group),
        rci_liability = mean(temp_df$rci_liability),
        intron_expression_rate = mean(temp_df$intron_expression_rate))
}))

patient_surv_df$age_at_death[patient_surv_df$age_at_death == "90 or Older"] <- "90"
patient_surv_df$age_at_death <- as.numeric(patient_surv_df$age_at_death)
patient_surv_df$age_at_onset <- as.numeric(patient_surv_df$age_at_onset)
patient_surv_df$disease_duration <- as.numeric(patient_surv_df$disease_duration_in_months)


```


- Perform modelling of the age of onset survival function using rci liability

```{r}

als_surv_df <- patient_surv_df[!(patient_surv_df$condition %in% c("als+ftd")),]
als_surv_df$sex_genotype <- factor(als_surv_df$sex_genotype, levels = c("XX", "XY"))
als_surv_df$status <- 0
als_surv_df$status[als_surv_df$condition == "als"] <- 1

onset_vs_liability_model_output <- lapply(unique(als_surv_df$tissue_group), function(x){
    model <- coxph(Surv(age_at_death, status) ~ sex_genotype + rci_liability, 
        data=als_surv_df[als_surv_df$tissue_group == x,], x=TRUE)
    model
})
names(onset_vs_liability_model_output) <- unique(als_surv_df$tissue_group)

lapply(onset_vs_liability_model_output, function(x){
    summary(x)
})
# $Cortex
# [1] 0.1901537

# $Cortex_Motor
# [1] 0.287525

# $Spinal_Cord_Cervical
# [1] 0.001781006

# $Spinal_Cord_Lumbar
# [1] 0.0004423336

# $Cerebellum
# [1] 0.02153693

# $Spinal_Cord_Thoracic
# [1] 0.01095877

# $brain
# [1] 0.1414335

# $spinal_cord
# [1] 0.0005109276


lapply(names(onset_vs_liability_model_output), function(x){
    current_model <- onset_vs_liability_model_output[[x]]
    current_df <- als_surv_df[als_surv_df$tissue_group == x,]
    als_quantiles <- round(quantile(current_df$rci_liability, c(0.05, 0.25, 0.5, 0.75, 0.95)))
    the_plot <- plot_surv_lines(time = "age_at_death",
                    status = "status",
                    variable = "rci_liability",
                    data = current_df,
                    model = current_model,
                    cif = T,
                    custom_color = RColorBrewer::brewer.pal(n = 5, name = "Reds"),
                    horizon=als_quantiles)
    ggsave(width = 5, height = 5, file = file.path(plot_dir, paste("07-03",x,"cumulative-incidence-rci_liability-vs-age-at-death-figure-3D.pdf", sep = "-")), the_plot)
})



current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

# colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")


lapply(names(onset_vs_liability_model_output), function(x){
    current_df <- als_surv_df[als_surv_df$tissue_group == x,]
    als_quantiles <- round(quantile(current_df$rci_liability, c(0.05, 0.25, 0.5, 0.75, 0.95)))

    label_df <- data.frame(rci_liability = als_quantiles, group = factor(names(als_quantiles), levels = c("5%", "25%", "50%", "75%", "95%")))

    ThePlot <- ggplot(current_df, aes(x = rci_liability))
    ThePlot <- ThePlot + geom_density(fill = "#F6F4F3", colour = "#000000")
    ThePlot <- ThePlot + geom_vline(data = label_df, aes(xintercept = rci_liability, colour = group))
    ThePlot <- ThePlot + scale_x_continuous("ALS specific rci expression")
    ThePlot <- ThePlot + scale_y_continuous("density")
    ThePlot <- ThePlot + scale_colour_manual("Condition colour", values = RColorBrewer::brewer.pal(n = 5, name = "Reds"))
    # ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
    # ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
    ThePlot <- ThePlot + theme_bw()
    # ThePlot <- ThePlot + facet_grid(. ~ mnd_group)
    ThePlot <- ThePlot + ggtitle("RCI expression in PM")
    ThePlot <- ThePlot + current_theme
    ggsave(width = 5, height = 5, file = file.path(plot_dir, paste("07-03",x,"density-dist-rci-liability-vs-age-at-death-figure-3D.pdf", sep = "-")), ThePlot)
})

```

- Perform modelling of the survival function using RCI liability and age of onset


```{r}

nygc_patient_liability_df$type <- factor(nygc_patient_liability_df$type, levels = c("control", "c9orf72-ALS", "other_mutation-fALS", "sporadic"))
nygc_patient_liability_df$type_group <- as.character(nygc_patient_liability_df$type)
nygc_patient_liability_df$type_group[nygc_patient_liability_df$type %in% c("other_mutation-fALS", "sporadic")] <- "sporadic+other_fALS"
nygc_patient_liability_df$tissue_group <- "spinal_cord"
nygc_patient_liability_df$tissue_group[nygc_patient_liability_df$tissue %in% c("Cortex", "Cortex_Motor", "Cerebellum")] <- "brain"

patient_surv_df <- do.call(rbind, lapply(split(nygc_patient_liability_df, paste(nygc_patient_liability_df$subject_id, nygc_patient_liability_df$tissue)), function(temp_df){
    message(unique(temp_df$subject_id))
    data.frame(subject_id = unique(temp_df$subject_id),
        tissue = unique(temp_df$tissue),
        tissue_group = unique(temp_df$tissue_group),
        age_at_onset = unique(temp_df$age_at_symptom_onset),
        age_at_death = unique(temp_df$age_at_death),
        disease_duration_in_months = unique(temp_df$disease_duration_in_months),
        sex_genotype = unique(temp_df$sex_genotype),
        mnd_with_ftd = unique(temp_df$mnd_with_ftd),
        condition = unique(temp_df$condition),
        mnd_with_dementia = unique(temp_df$mnd_with_dementia),
        type_group = unique(temp_df$type_group),
        rci_liability = mean(temp_df$rci_liability),
        intron_expression_rate = mean(temp_df$intron_expression_rate))
}))

patient_surv_df$age_at_death[patient_surv_df$age_at_death == "90 or Older"] <- "90"
patient_surv_df$age_at_death <- as.numeric(patient_surv_df$age_at_death)
patient_surv_df$age_at_onset <- as.numeric(patient_surv_df$age_at_onset)
patient_surv_df$disease_duration <- as.numeric(patient_surv_df$disease_duration_in_months)


```

- Perform modelling of the disease duration survival function using RCI liability

```{r}

als_surv_df <- patient_surv_df[!(patient_surv_df$condition %in% c("ctrl", "als+ftd")),]
als_surv_df$sex_genotype <- factor(als_surv_df$sex_genotype, levels = c("XX", "XY"))
als_surv_df$status <- 0
als_surv_df$status[als_surv_df$disease_duration <= 36 & !is.na(als_surv_df$disease_duration)] <- 1
# whether patient survived beyond 36 months

liability_model_output <- lapply(unique(als_surv_df$tissue), function(x){
    model <- coxph(Surv(disease_duration, status) ~ age_at_onset + sex_genotype + rci_liability, 
        data=als_surv_df[als_surv_df$tissue == x,], x=TRUE)
    model
})
names(liability_model_output) <- unique(als_surv_df$tissue)

lapply(liability_model_output, function(x){
    summary(x)
})
# $Cortex
# [1] 0.7417383

# $Cortex_Motor
# [1] 0.8900708

# $Spinal_Cord_Cervical
# [1] 0.3936714

# $Spinal_Cord_Lumbar
# [1] 0.4826659

# $Cerebellum
# [1] 0.9885605

# $Spinal_Cord_Thoracic
# [1] 0.7476263



current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

# colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")


lapply(names(liability_model_output), function(x){
    current_df <- als_surv_df[als_surv_df$tissue == x,]
    als_quantiles <- round(quantile(current_df$rci_liability, c(0.05, 0.25, 0.5, 0.75, 0.95)))

    label_df <- data.frame(rci_liability = als_quantiles, group = factor(names(als_quantiles), levels = c("5%", "25%", "50%", "75%", "95%")))

    ThePlot <- ggplot(current_df, aes(x = rci_liability))
    ThePlot <- ThePlot + geom_density(fill = "#F6F4F3", colour = "#000000")
    ThePlot <- ThePlot + geom_vline(data = label_df, aes(xintercept = rci_liability, colour = group))
    ThePlot <- ThePlot + scale_x_continuous("ALS specific rci expression")
    ThePlot <- ThePlot + scale_y_continuous("density")
    ThePlot <- ThePlot + scale_colour_manual("Condition colour", values = RColorBrewer::brewer.pal(n = 5, name = "Reds"))
    # ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
    # ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
    ThePlot <- ThePlot + theme_bw()
    # ThePlot <- ThePlot + facet_grid(. ~ mnd_group)
    ThePlot <- ThePlot + ggtitle("RCI expression in PM")
    ThePlot <- ThePlot + current_theme
    ggsave(width = 5, height = 5, file = file.path(plot_dir, paste("07-01",x,"density-dist-rci-liability-vs-disease-duration.pdf", sep = "-")), ThePlot)
})


lapply(names(liability_model_output), function(x){
    current_model <- liability_model_output[[x]]
    current_df <- als_surv_df[als_surv_df$tissue == x,]
    als_quantiles <- round(quantile(current_df$rci_liability, c(0.05, 0.25, 0.5, 0.75, 0.95)))
    the_plot <- plot_surv_lines(time = "disease_duration",
                    status = "status",
                    variable = "rci_liability",
                    data = current_df,
                    model = current_model,
                    cif = F,
                    custom_color = RColorBrewer::brewer.pal(n = 5, name = "Reds"),
                    horizon=als_quantiles,
                    max_t = 36)
    ggsave(width = 5, height = 5, file = file.path(plot_dir, paste("07-01",x,"als-survival-lines-rci-liability-vs-disease-duration.pdf", sep = "-")), the_plot)
})


```

- Perform modelling of the disease duration survival function using rci emission rate

```{r}

als_surv_df <- patient_surv_df[!(patient_surv_df$condition %in% c("ctrl", "als+ftd")),]
als_surv_df$sex_genotype <- factor(als_surv_df$sex_genotype, levels = c("XX", "XY"))
als_surv_df$status <- 0
als_surv_df$status[als_surv_df$disease_duration <= 36 & !is.na(als_surv_df$disease_duration)] <- 1

emission_model_output <- lapply(unique(als_surv_df$tissue), function(x){
    model <- coxph(Surv(disease_duration, status) ~ age_at_onset + sex_genotype + intron_expression_rate, 
        data=als_surv_df[als_surv_df$tissue == x,], x=TRUE)
    model
})
names(emission_model_output) <- unique(als_surv_df$tissue)

rci_emission_rate_df <- do.call(rbind, lapply(unique(als_surv_df$tissue), function(x){
    current_df <- als_surv_df[als_surv_df$tissue == x,]
    als_quantiles <- round(quantile(current_df$intron_expression_rate, c(0.05, 0.25, 0.5, 0.75, 0.95)))
    curve_data_df <- curve_cont(data=current_df, variable="intron_expression_rate", 
        model=emission_model_output[[x]], horizon=als_quantiles, 
        times=seq(from = 1, by = 1, length.out = 36))
    curve_data_df$tissue <- x
    curve_data_df$als_q <- names(als_quantiles)[match(curve_data_df$cont, als_quantiles)]
    return(curve_data_df)
}))
# $Cortex
# [1] 0.5950327

# $Cortex_Motor
# [1] 0.1841539

# $Spinal_Cord_Cervical
# [1] 0.43191

# $Spinal_Cord_Lumbar
# [1] 0.001988699

# $Cerebellum
# [1] 0.9208218

# $Spinal_Cord_Thoracic
# [1] 0.8352768

lapply(emission_model_output, function(x){
    summary(x)$coefficients["intron_expression_rate",5]
})


lapply(names(liability_model_output), function(x){
    current_df <- als_surv_df[als_surv_df$tissue == x,]
    als_quantiles <- round(quantile(current_df$intron_expression_rate, c(0.05, 0.25, 0.5, 0.75, 0.95)))

    label_df <- data.frame(intron_expression_rate = als_quantiles, group = factor(names(als_quantiles), levels = c("5%", "25%", "50%", "75%", "95%")))

    ThePlot <- ggplot(current_df, aes(x = intron_expression_rate))
    ThePlot <- ThePlot + geom_density(fill = "#F6F4F3", colour = "#000000")
    ThePlot <- ThePlot + geom_vline(data = label_df, aes(xintercept = intron_expression_rate, colour = group))
    ThePlot <- ThePlot + scale_x_continuous("RCI emission rate")
    ThePlot <- ThePlot + scale_y_continuous("density")
    ThePlot <- ThePlot + scale_colour_manual("Condition colour", values = RColorBrewer::brewer.pal(n = 5, name = "Reds"))
    # ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
    # ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
    ThePlot <- ThePlot + theme_bw()
    # ThePlot <- ThePlot + facet_grid(. ~ mnd_group)
    ThePlot <- ThePlot + ggtitle("RCI expression in PM")
    ThePlot <- ThePlot + current_theme
    ggsave(width = 5, height = 5, file = file.path(plot_dir, paste("07-02",x,"density-dist-intron-expression-rate-vs-disease-duration.pdf", sep = "-")), ThePlot)
})



lapply(names(emission_model_output), function(x){
    current_model <- emission_model_output[[x]]
    current_df <- als_surv_df[als_surv_df$tissue == x,]
    als_quantiles <- round(quantile(current_df$intron_expression_rate, c(0.05, 0.25, 0.5, 0.75, 0.95)))
    the_plot <- plot_surv_lines(time = "disease_duration",
                    status = "status",
                    variable = "intron_expression_rate",
                    data = current_df,
                    model = current_model,
                    cif = F,
                    custom_color = RColorBrewer::brewer.pal(n = 5, name = "Reds"),
                    horizon=als_quantiles,
                    max_t = 36)
    ggsave(width = 5, height = 5, file = file.path(plot_dir, paste("07-02",x,"als-survival-lines-intron-expression-rate-vs-disease-duration.pdf", sep = "-")), the_plot)
})

```

