---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")
library(rlang)
library(riskRegression)
library(survival)
library(pammtools)
library(gganimate)
library(transformr)
library(plotly)
library(reshape2)
library(knitr)
library(rmarkdown)

require("contsurvplot")
require("RColorBrewer")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-3"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load penetrance data

```{r}

nygc_penetrance_list <- readRDS(file.path("final-figures/figure-2/rdata_files/02-save-nygc-rci-expression-summary-with-penetrance-matrix.rds"))
nygc_patient_liability_df <- nygc_penetrance_list$liability_df
nygc_penetrance_matrix <- nygc_penetrance_list$penetrance_matrix

```

- Prepare data

```{r}

nygc_patient_liability_df$tissue_group <- "spinal_cord"
nygc_patient_liability_df$tissue_group[nygc_patient_liability_df$tissue %in% c("Cortex", "Cortex_Motor", "Cerebellum")] <- "brain"
nygc_patient_liability_df$tissue <- factor(nygc_patient_liability_df$tissue, levels = c("Spinal_Cord_Cervical", "Spinal_Cord_Thoracic", "Spinal_Cord_Lumbar",
    "Cortex", "Cortex_Motor", "Cerebellum"))

nygc_patient_liability_df$type <- factor(nygc_patient_liability_df$type, levels = c("control", "c9orf72-ALS", "other_mutation-fALS", "sporadic"))
nygc_patient_liability_df$type_group <- as.character(nygc_patient_liability_df$type)
nygc_patient_liability_df$type_group[nygc_patient_liability_df$type %in% c("other_mutation-fALS", "sporadic")] <- "sporadic+other_fALS"

patient_surv_df <- do.call(rbind, lapply(split(nygc_patient_liability_df, paste(nygc_patient_liability_df$subject_id, nygc_patient_liability_df$tissue_group)), function(temp_df){
    message(unique(temp_df$subject_id))
    data.frame(subject_id = unique(temp_df$subject_id),
        # tissue = unique(temp_df$tissue),
        tissue_group = unique(temp_df$tissue_group),
        age_at_onset = unique(temp_df$age_at_symptom_onset),
        age_at_death = unique(temp_df$age_at_death),
        disease_duration_in_months = unique(temp_df$disease_duration_in_months),
        sex_genotype = unique(temp_df$sex_genotype),
        mnd_with_ftd = unique(temp_df$mnd_with_ftd),
        condition = unique(temp_df$condition),
        mnd_with_dementia = unique(temp_df$mnd_with_dementia),
        type_group = unique(temp_df$type_group),
        rci_liability = mean(temp_df$rci_liability),
        intron_expression_rate = mean(temp_df$intron_expression_rate))
}))

patient_surv_df$age_at_death[patient_surv_df$age_at_death == "90 or Older"] <- "90"
patient_surv_df$age_at_death <- as.numeric(patient_surv_df$age_at_death)
patient_surv_df$age_at_onset <- as.numeric(patient_surv_df$age_at_onset)
patient_surv_df$disease_duration <- as.numeric(patient_surv_df$disease_duration_in_months)

```

- Find patients who had contributed both brain and spinal cord segments


```{r}

patient_freq <- table(patient_surv_df$subject_id)

keep_patient_df <- patient_surv_df[patient_surv_df$subject_id %in% names(patient_freq)[patient_freq == 2] & 
    patient_surv_df$condition %in% c("ctrl", "als"),]

scatter_df <- do.call(rbind, lapply(split(keep_patient_df, keep_patient_df$subject_id), function(temp_df){
    message(unique(temp_df$subject_id))
    data.frame(subject_id = unique(temp_df$subject_id),
        age_at_onset = unique(temp_df$age_at_onset),
        age_at_death = unique(temp_df$age_at_death),
        disease_duration_in_months = unique(temp_df$disease_duration_in_months),
        sex_genotype = unique(temp_df$sex_genotype),
        mnd_with_ftd = unique(temp_df$mnd_with_ftd),
        condition = unique(temp_df$condition),
        mnd_with_dementia = unique(temp_df$mnd_with_dementia),
        type_group = unique(temp_df$type_group),
        brain_emission_rate = temp_df$intron_expression_rate[temp_df$tissue_group == "brain"],
        spinal_cord_emission_rate = temp_df$intron_expression_rate[temp_df$tissue_group == "spinal_cord"])
}))


scatter_df$condition <- factor(scatter_df$condition, levels = c("ctrl", "als"))
summary(lm(data = scatter_df[scatter_df$condition == "ctrl",], formula = spinal_cord_emission_rate ~ brain_emission_rate))
summary(lm(data = scatter_df[scatter_df$condition == "als",], formula = spinal_cord_emission_rate ~ brain_emission_rate))

```


```{r}

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

colours <- c("ctrl" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
fill_colours <- c("ctrl" = "#F6F4F3", 
    "c9orf72" = "#169873", "sporadic" = "#F49FBC",
    "tardbp" = "#FFD3BA", "mutation" = "#9EBD6E",
    "none" = "#805D93")

colours <- c("ctrl" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
ThePlot <- ggplot(scatter_df, 
    aes(x = brain_emission_rate, y = spinal_cord_emission_rate))
ThePlot <- ThePlot + geom_point(aes(fill = condition), colour = "#000000", shape = 21)
# ThePlot <- ThePlot + geom_point(aes(fill = condition), shape = 21, colour = "#000000", position = position_jitterdodge(dodge.width = 0.9))
ThePlot <- ThePlot + geom_smooth(method = "lm", formula = y ~ x, aes(colour = condition))
# ThePlot <- ThePlot + stat_summary(aes(colour = condition))
ThePlot <- ThePlot + scale_x_continuous("brain RCI emission rate")
ThePlot <- ThePlot + scale_y_continuous("spinal cord RCI emission rate")
ThePlot <- ThePlot + scale_colour_manual("Condition", values = colours)
ThePlot <- ThePlot + scale_fill_manual("Condition", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_wrap(. ~ condition, scales = "free")
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression across tissues")
ThePlot <- ThePlot + current_theme
ggsave(width = 10, height = 5, file = file.path(plot_dir, "09-01-plot-pairwise-scatterplot-for-spinal-cord-vs-brain-rci-emission-rate.pdf"), ThePlot)

```