---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

library("here")
message("Project root:", here(), "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("rtracklayer")
require("ggplot2")
require("matrixStats")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("41-magen-et-al-miRNA-seq")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS(here("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds"))
feature_ranges <- readRDS(here("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds"))
sig_result_list <- readRDS(here("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds"))
mirbase_gff_ranges <- import.gff3(here("41-magen-et-al-miRNA-seq/input_files/hsa.gff3"))

magen_count_matrix <- read.table(here("41-magen-et-al-miRNA-seq/input_files/GSE168714_All_samples_enrolment.txt"))
replication_set_samples <- read.csv(here("41-magen-et-al-miRNA-seq/input_files/GSE168714_Data_file_related_to_fig_3_5_replication.csv"))

```

- Fetch the significantly differentially expressed rcis

```{r}

sig_result_df <- sig_result_list$significant_events

repeat_containing_introns_df <- sig_result_df[sig_result_df$status == "repeat_containing" & 
    sig_result_df$region == "intron",]

repeat_containing_introns_df$cohort <- "public-data"
repeat_containing_introns_df$cohort[grepl(x = repeat_containing_introns_df$study, pattern = "nygc")] <- "nygc"
repeat_containing_introns_df$cohort[grepl(x = repeat_containing_introns_df$study, pattern = "answerals")] <- "answerals"

repeat_containing_introns_df$tissue_type <- "iPSCs"
repeat_containing_introns_df$tissue_type[grepl(x = repeat_containing_introns_df$study, pattern = "nygc")] <- "postmortem"

```

- Load repeat and rci object

```{r}

repeat_ranges <- unique(repeat_ranges_list$repeat_ranges)

feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges
de_feature_ranges_with_repeats <- feature_ranges_with_repeats[match(unique(repeat_containing_introns_df$feature_ids), feature_ranges_with_repeats$feature_id)]

```

- Fetch the magen et al discovery set

```{r}

magen_sampleinfo_df <- as.data.frame(t(magen_count_matrix[seq(1,6),]))
colnames(magen_sampleinfo_df) <- magen_sampleinfo_df[1,]
magen_sampleinfo_df <- magen_sampleinfo_df[-1,]

magen_sampleinfo_df[,2] <- as.numeric(magen_sampleinfo_df[,2])
# magen_sampleinfo_df <- magen_sampleinfo_df[,-6]

tested_mirna <- magen_count_matrix[-seq(1,6),1]
raw_count_matrix <- magen_count_matrix[-seq(1,6),-1]
rownames(raw_count_matrix) <- tested_mirna
colnames(raw_count_matrix) <- magen_sampleinfo_df$sample

replication_sampleinfo_df <- magen_sampleinfo_df[magen_sampleinfo_df$sample %in% replication_set_samples$index.case,]


```

- Subset the miRNA matrix

```{r}

name_intersect <- intersect(mirbase_gff_ranges$Name,rownames(raw_count_matrix))
length(name_intersect) / nrow(raw_count_matrix)
# 0.9815918
keep_count_matrix <- as.matrix(raw_count_matrix[name_intersect,replication_sampleinfo_df$sample])
class(keep_count_matrix) <- "numeric"
mirna_ranges <- mirbase_gff_ranges[match(name_intersect, mirbase_gff_ranges$Name)]
seqlevelsStyle(mirna_ranges) <- "Ensembl"

```

- Generate CPM matrix

```{r}

cpm_matrix <- t(t(keep_count_matrix) / (colSums(keep_count_matrix) / 1e6))

```

- Fetch the miRNA counts for rcis

```{r}

ol_object <- findOverlaps(mirna_ranges,de_feature_ranges_with_repeats)

overlapping_rci_ranges <- de_feature_ranges_with_repeats[subjectHits(ol_object)]

hash_pairs_df <- unique(data.frame(queries = queryHits(ol_object), hash_id = overlapping_rci_ranges$hash_id))
hash_pairs_df$weight <- 1 / table(hash_pairs_df$queries)[as.character(hash_pairs_df$queries)]
hash_pairs_df$mirna_id <- mirna_ranges$Name[hash_pairs_df$queries]
# weight is always 1

overlapping_rci_mirna_count_matrix <- do.call(rbind, lapply(split(hash_pairs_df, hash_pairs_df$hash_id), function(x){
    current_row <- x
    if(nrow(x) == 1){
        return(cpm_matrix[x$queries,] * x$weight)
    }
    return(colSums(cpm_matrix[x$queries,] * rep(x$weight, times = ncol(cpm_matrix))))
}))

saveRDS(object = list(
	rci_counts = overlapping_rci_mirna_count_matrix, 
	sampleinfo_df = replication_sampleinfo_df,
	discovery_clinical_df = replication_set_samples,
	mirna_ranges = mirna_ranges,
	mirna_rci_overlap = hash_pairs_df,
	rci_ranges = overlapping_rci_ranges), 
	file = file.path(save_dir, "01-06-rci-counts-for-replication-set.rds"))

```
