---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

library("here")
message("Project root:", here(), "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("rtracklayer")
require("ggplot2")
require("matrixStats")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("41-magen-et-al-miRNA-seq")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
feature_ranges <- readRDS("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds")
sig_result_list <- readRDS("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds")
mirbase_gff_ranges <- import.gff3("41-magen-et-al-miRNA-seq/input_files/hsa.gff3")

```

- Fetch the significantly differentially expressed rcis

```{r}

sig_result_df <- sig_result_list$significant_events

repeat_containing_introns_df <- sig_result_df[sig_result_df$status == "repeat_containing" & 
    sig_result_df$region == "intron",]

repeat_containing_introns_df$cohort <- "public-data"
repeat_containing_introns_df$cohort[grepl(x = repeat_containing_introns_df$study, pattern = "nygc")] <- "nygc"
repeat_containing_introns_df$cohort[grepl(x = repeat_containing_introns_df$study, pattern = "answerals")] <- "answerals"

repeat_containing_introns_df$tissue_type <- "iPSCs"
repeat_containing_introns_df$tissue_type[grepl(x = repeat_containing_introns_df$study, pattern = "nygc")] <- "postmortem"

```

- Load repeat and rci object

```{r}

repeat_ranges <- unique(repeat_ranges_list$repeat_ranges)

feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges
de_feature_ranges_with_repeats <- feature_ranges_with_repeats[match(unique(repeat_containing_introns_df$feature_ids), feature_ranges_with_repeats$feature_id)]

```

- Reformat the longitudinal data

```{r}

magen_longitudinal_count_matrix <- read.table("41-magen-et-al-miRNA-seq/input_files/GSE168714_Longitudinal_raw_counts_22_samples.txt")

magen_longitudinal_sampleinfo_df <- as.data.frame(t(magen_longitudinal_count_matrix[seq(1,4),]))
colnames(magen_longitudinal_sampleinfo_df) <- magen_longitudinal_sampleinfo_df[1,]
magen_longitudinal_sampleinfo_df <- magen_longitudinal_sampleinfo_df[-1,]
magen_longitudinal_sampleinfo_df$time <- factor(magen_longitudinal_sampleinfo_df$time, levels = c("A","B","C","D"))

tested_mirna <- magen_longitudinal_count_matrix[-seq(1,4),1]
raw_count_matrix <- magen_longitudinal_count_matrix[-seq(1,4),-1]
rownames(raw_count_matrix) <- tested_mirna
colnames(raw_count_matrix) <- magen_longitudinal_sampleinfo_df$sample

name_intersect <- intersect(mirbase_gff_ranges$Name, rownames(raw_count_matrix))
# length(name_intersect)
# length(name_intersect) / nrow(raw_count_matrix)
# 0.9805777

keep_count_matrix <- as.matrix(raw_count_matrix[name_intersect,])
class(keep_count_matrix) <- "numeric"

mirna_ranges <- mirbase_gff_ranges[match(name_intersect, mirbase_gff_ranges$Name)]
seqlevelsStyle(mirna_ranges) <- "Ensembl"

```

- Generate CPM matrix

```{r}

cpm_matrix <- t(t(keep_count_matrix) / (colSums(keep_count_matrix) / 1e6))

```

- Plot the venn diagram 

```{r}

require("ggvenn")
ol_object <- findOverlaps(mirna_ranges,de_feature_ranges_with_repeats)
overlapping_rci_ranges <- de_feature_ranges_with_repeats[subjectHits(ol_object)]

pair_list <- list(miRNAs = mirna_ranges$ID, 
	RCIs = de_feature_ranges_with_repeats$hash_id)
pair_list[["RCIs"]][subjectHits(ol_object)] <- mirna_ranges$ID[queryHits(ol_object)]

colours <- c("#FFE74C", "#FF5964")

pdf(file.path(plot_dir, "01-intersection-between-miRNAs-and-rcis-in-magen-et-al.pdf"), width = 7, height = 7)

ggvenn(pair_list, c("miRNAs", "RCIs"), fill_color = colours)

dev.off()


```

- Fetch the miRNA counts for rcis

```{r}


hash_pairs_df <- unique(data.frame(queries = queryHits(ol_object), hash_id = overlapping_rci_ranges$hash_id))
hash_pairs_df$weight <- 1 / table(hash_pairs_df$queries)[as.character(hash_pairs_df$queries)]
hash_pairs_df$mirna_id <- mirna_ranges$Name[hash_pairs_df$queries]
# weight is always 1

overlapping_rci_mirna_count_matrix <- do.call(rbind, lapply(split(hash_pairs_df, hash_pairs_df$hash_id), function(x){
    current_row <- x
    if(nrow(x) == 1){
        return(cpm_matrix[x$queries,] * x$weight)
    }
    return(colSums(cpm_matrix[x$queries,] * rep(x$weight, times = ncol(cpm_matrix))))
}))

saveRDS(object = list(
	rci_counts = overlapping_rci_mirna_count_matrix, 
	sampleinfo_df = magen_longitudinal_sampleinfo_df,
	mirna_ranges = mirna_ranges,
	mirna_rci_overlap = hash_pairs_df,
	rci_ranges = overlapping_rci_ranges), file = file.path(save_dir, "01-04-rci-counts-with-longitudinal-sampleinfo.rds"))

```

