---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("rtracklayer")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")
require("matrixStats")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "41-magen-et-al-miRNA-seq"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

longitudinal_data_list <- readRDS("41-magen-et-al-miRNA-seq/rdata_files/01-04-rci-counts-with-longitudinal-sampleinfo.rds")


magen_longitudinal_sampleinfo_df <- longitudinal_data_list$sampleinfo_df
overlapping_rci_mirna_count_matrix <- longitudinal_data_list$rci_counts
hash_pairs_df <- longitudinal_data_list$mirna_rci_overlap

```

- Investigate the longitudinal variation

```{r}

t4_matrix <- overlapping_rci_mirna_count_matrix[,magen_longitudinal_sampleinfo_df$time == "D"]
t1_matrix <- overlapping_rci_mirna_count_matrix[,magen_longitudinal_sampleinfo_df$time == "A"]

fc_matrix <- t4_matrix / t1_matrix
fc_matrix[is.na(fc_matrix) | is.infinite(fc_matrix) | is.nan(fc_matrix)] <- 0

average_change <- rowMeans(log2(fc_matrix+1), na.rm = T)
zscore_of_rcis <- (average_change - mean(average_change)) / sd(average_change)

standard_errors <- vapply(seq_len(nrow(fc_matrix)), function(x){
    a_row <- fc_matrix[x,]
    n <- length(a_row[!is.na(a_row)])
    std_err <- sd(a_row, na.rm = T) / sqrt(n)
    return(std_err)
},1)

plot_df <- data.frame(rci = names(zscore_of_rcis), zscore = zscore_of_rcis, std_err = log2(standard_errors))

```

- Plot variability of small RNAs from RCIs


```{r}

plot_df$group <- "none"
plot_df$group[plot_df$std_err > -0.2] <- "above -0.2"
plot_df$group[plot_df$std_err <= -0.2] <- "below -0.2"
table(plot_df$group)


current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 10, angle = 90),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))


# above <- #007CBE
#
# below <- #FFA3AF
colours <- c("above -0.2" = "#007CBE", "below -0.2" = "#CC3363")
ThePlot <- ggplot(plot_df, aes(x = zscore, y = std_err))
ThePlot <- ThePlot + geom_hline(yintercept = -0.2, linetype = "dashed")
ThePlot <- ThePlot + geom_point(aes(fill = group), colour = "#000000", shape = 21)
ThePlot <- ThePlot + scale_x_continuous("standardised change in miRNA associated RCIs (log2 t4/t1)")
ThePlot <- ThePlot + scale_y_continuous("variability of miRNA reads in RCIs (log2 t4/t1)")
# ThePlot <- ThePlot + scale_colour_manual("MND status", values = colours)
ThePlot <- ThePlot + scale_fill_manual("magen et al stability threshold", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_wrap(. ~ mnd_group, scales = "free", nrow = 1, ncol = 2)
ThePlot <- ThePlot + ggtitle("variability of miRNAs coming from RCIs")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 7, file = file.path(plot_dir, "04-02-plot-variability-of-miRNAs-originating-from-RCIs.pdf"), ThePlot)

```

- Get the pair with highest and lowest stability 

```{r}

highest_stability <- plot_df[which.max(plot_df$std_err),]
lowest_stability <- plot_df[which.min(plot_df$std_err),]

plot_fc_df <- reshape2::melt(fc_matrix)
colnames(plot_fc_df) <- c("rci", "sample", "foldchange")
plot_fc_df$rci <- factor(plot_fc_df$rci, levels = plot_df$rci[order(plot_df$std_err, decreasing = F)])
plot_fc_df$std_err <- plot_df$std_err[match(plot_fc_df$rci, plot_df$rci)]
keep_plot_fc_df <- plot_fc_df[plot_fc_df$rci %in% c(highest_stability$rci, lowest_stability$rci),]

# above <- #007CBE
#
# below <- #FFA3AF

require(ggbeeswarm)
colours <- c("above -0.2" = "#007CBE", "below -0.2" = "#CC3363")
ThePlot <- ggplot(keep_plot_fc_df, aes(x = rci, y = log2(foldchange + 1)))
ThePlot <- ThePlot + geom_boxplot(aes(fill = std_err), colour = "#000000", 
    width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9))
ThePlot <- ThePlot + geom_quasirandom(aes(fill = std_err), colour = "#000000", shape = 21)
ThePlot <- ThePlot + scale_x_discrete("RCIs with highest and lowest stability")
ThePlot <- ThePlot + scale_y_continuous("log2 t4/t1 values in 22 patients")
# ThePlot <- ThePlot + scale_colour_manual("MND status", values = colours)
ThePlot <- ThePlot + scale_fill_distiller("stability", palette = "RdBu")
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_wrap(. ~ mnd_group, scales = "free", nrow = 1, ncol = 2)
ThePlot <- ThePlot + ggtitle("t4/t1 values from 22 patients")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 7, file = file.path(plot_dir, "04-03-plot-t4-over-t1-values-for-miRNAs-originating-from-RCIs.pdf"), ThePlot)


hash_pairs_df[hash_pairs_df$hash_id %in% c(highest_stability$rci, lowest_stability$rci),]
#    queries  hash_id weight       mirna_id
# 44     889 4122ac4f      1    hsa-miR-657
# 45     890 4122ac4f      1 hsa-miR-338-5p
# 46     891 4122ac4f      1 hsa-miR-338-3p
# 67    1341 4d680749      1   hsa-miR-3938

```

- Do any of these rcis overlap the set of previously characterised miRNAs by magen et al.

```{r}

highly_stable_mirnas_used_by_magen <- read.csv("41-magen-et-al-miRNA-seq/input_files/highly-stable-mirna-used-by-magen-for-discovery.csv")
hash_pairs_df$overlap_with_magen_et_al <- hash_pairs_df$mirna_id %in% highly_stable_mirnas_used_by_magen$mirna

# > any(hash_pairs_df$overlap_with_magen_et_al)
# [1] FALSE

```

- We did not find any overlap with the set of miRNAs previously characterised by magen et al.


- Finally, save the highly stable rcis

```{r}

saveRDS(object = plot_df, file = file.path(save_dir, "01-05-rcis-with-stability-levels.rds"))

```
