---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("rtracklayer")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")
require("matrixStats")
require("glmnet")
source("41-magen-et-al-miRNA-seq/scripts/sources/models.R")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "41-magen-et-al-miRNA-seq"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Analyse association between increasing intron expression rates and miRNA production

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

discovery_clinical_df <- readRDS(file.path(save_dir, "01-07-discovery-clinical-with-mirna-counts-in-rcis.rds"))
# cutpoint_analysis_results <- readRDS(file.path(save_dir, "02-01-cutpoint-analysis-results-for-enrollment-and-onset.rds"))

discovery_data_list <- readRDS("41-magen-et-al-miRNA-seq/rdata_files/01-05-rci-counts-for-discovery-set.rds")
hash_pair_df <- discovery_data_list$mirna_rci_overlap
rci_set <- unique(hash_pair_df$hash_id[hash_pair_df$hash_id %in% colnames(discovery_clinical_df)])

```

- Perform 1k ridge regressions and compute the median score

```{r}

rci_only_ridge_composite_score_df_list <- return_iterated_ridge_composite_score(discovery_clinical_df, 
  rcis = rci_set, 
  iterate_n = 1000, 
  concat_col_name = "rci_only_ridge")
rci_only_ridge_composite_score_df <- rci_only_ridge_composite_score_df_list$prediction
seed_list <- rci_only_ridge_composite_score_df_list$seed_list


rci_and_nfl_ridge_composite_score_df_list <- return_iterated_ridge_composite_score(discovery_clinical_df[!is.na(discovery_clinical_df$NfL_numeric),], 
  rcis = c(rci_set, "NfL_numeric"), 
  iterate_n = 1000, 
  concat_col_name = "rci_and_nfl_ridge", 
  seed_list = seed_list)
rci_and_nfl_ridge_composite_score_df <- rci_and_nfl_ridge_composite_score_df_list$prediction

rci_and_mir181_ridge_composite_score_df_list <- return_iterated_ridge_composite_score(discovery_clinical_df, 
  rcis = c(rci_set, "miR_181_numeric"), 
  iterate_n = 1000, 
  concat_col_name = "rci_and_mir181_ridge", 
  seed_list = seed_list)
rci_and_mir181_ridge_composite_score_df <- rci_and_mir181_ridge_composite_score_df_list$prediction

rci_and_nfl_and_mir181_ridge_composite_score_df_list <- return_iterated_ridge_composite_score(discovery_clinical_df[!is.na(discovery_clinical_df$NfL_numeric),], 
  rcis = c(rci_set, "NfL_numeric", "miR_181_numeric"), 
  iterate_n = 1000, 
  concat_col_name = "rci_and_nfl_and_mir181_ridge", 
  seed_list = seed_list)
rci_and_nfl_and_mir181_ridge_composite_score_df <- rci_and_nfl_and_mir181_ridge_composite_score_df_list$prediction

```

- Assemble all of the different scores together

```{r}

score_list <- list(
  rci_only = rci_only_ridge_composite_score_df,
  rci_and_nfl = rci_and_nfl_ridge_composite_score_df,
  rci_and_mir181 = rci_and_mir181_ridge_composite_score_df,
  rci_and_nfl_and_mir181 = rci_and_nfl_and_mir181_ridge_composite_score_df)

variable_list <- list(
  rci_only = c("rci_only_ridge_log_odds", "rci_only_ridge_prob"),
  rci_and_nfl = c("rci_and_nfl_ridge_log_odds", "rci_and_nfl_ridge_prob"),
  rci_and_mir181 = c("rci_and_mir181_ridge_log_odds", "rci_and_mir181_ridge_prob"),
  rci_and_nfl_and_mir181 = c("rci_and_nfl_and_mir181_ridge_log_odds", "rci_and_nfl_and_mir181_ridge_prob"))

```

- Evaluate cutpoints for onset

```{r}

require("evaluateCutpoints")

onset_hr_df_list <- lapply(names(score_list), function(x){
  scores_to_test <- variable_list[[x]]
  test_df <- score_list[[x]]
  test_df_onset <- test_df[,c("survival_onset", "status", scores_to_test)]
  
  evaluateCutpoints(
    mainDir = "/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/41-magen-et-al-miRNA-seq/output_files/",
    resultsDirName = "onset-composite-score-analysis-results",
    df = test_df_onset,
    biomarkerList = scores_to_test,
    time = "survival_onset",
    event = "status",
    setCutpoint = NULL,
    setCutpoint2 = NULL)
  setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
  
  cutpoint_dir <- "/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/41-magen-et-al-miRNA-seq/output_files/onset-composite-score-analysis-results"
  current_date <- "2025-04-10"
  rci_onset_reported_hr_values_df <-  do.call(rbind, lapply(scores_to_test, function(x){
    return_df <- read.csv(file.path(cutpoint_dir, current_date, x, "2", "maxstat", "MaxStatResults.csv"))[,-1]
    return_df <- as.data.frame(t(return_df))
    colnames(return_df) <- return_df[1,]
    return_df <- cbind(data.frame(molecule = x), return_df[-1,])
    return_df[,"P-value"] <- as.numeric(return_df[,"P-value"])
    return_df[,"Cutpoint"] <- as.numeric(return_df[,"Cutpoint"])
    return_df[,"HR"] <- as.numeric(return_df[,"HR"])
    return(return_df)
  }))
  rci_onset_reported_hr_values_df$sig <- "NS"
  rci_onset_reported_hr_values_df$sig[rci_onset_reported_hr_values_df[,"P-value"] < 0.1] <- "."
  rci_onset_reported_hr_values_df$sig[rci_onset_reported_hr_values_df[,"P-value"] < 0.05] <- "*"
  rci_onset_reported_hr_values_df$sig[rci_onset_reported_hr_values_df[,"P-value"] < 0.01] <- "**"
  rci_onset_reported_hr_values_df$sig[rci_onset_reported_hr_values_df[,"P-value"] < 0.001] <- "***"
  return(rci_onset_reported_hr_values_df)
})
onset_hr_df <- do.call(rbind, onset_hr_df_list)


```

- Evaluate cutpoints for enrollment


```{r}

enrollment_hr_df_list <- lapply(names(score_list), function(x){
  scores_to_test <- variable_list[[x]]
  test_df <- score_list[[x]]

  test_df_enrollment <- test_df[,c("survival_enrolment", "status", scores_to_test)]
  # colnames(test_df) <- c("time", "event", names(biomarkers))
  evaluateCutpoints(
    mainDir = "/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/41-magen-et-al-miRNA-seq/output_files/",
    resultsDirName = "enrollment-composite-score-analysis-results",
    df = test_df_enrollment,
    biomarkerList = scores_to_test,
    time = "survival_enrolment",
    event = "status",
    setCutpoint = NULL,
    setCutpoint2 = NULL)

  setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
  cutpoint_dir <- "/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/41-magen-et-al-miRNA-seq/output_files/enrollment-composite-score-analysis-results"
  current_date <- "2025-04-10"
  rci_enrollment_reported_hr_values_df <-  do.call(rbind, lapply(scores_to_test, function(x){
    return_df <- read.csv(file.path(cutpoint_dir, current_date, x, "2", "maxstat", "MaxStatResults.csv"))[,-1]
    return_df <- as.data.frame(t(return_df))
    colnames(return_df) <- return_df[1,]
    return_df <- cbind(data.frame(molecule = x), return_df[-1,])
    return_df[,"P-value"] <- as.numeric(return_df[,"P-value"])
    return_df[,"Cutpoint"] <- as.numeric(return_df[,"Cutpoint"])
    return_df[,"HR"] <- as.numeric(return_df[,"HR"])
    return(return_df)
  }))

  rci_enrollment_reported_hr_values_df$sig <- "NS"
  rci_enrollment_reported_hr_values_df$sig[rci_enrollment_reported_hr_values_df[,"P-value"] < 0.1] <- "."
  rci_enrollment_reported_hr_values_df$sig[rci_enrollment_reported_hr_values_df[,"P-value"] < 0.05] <- "*"
  rci_enrollment_reported_hr_values_df$sig[rci_enrollment_reported_hr_values_df[,"P-value"] < 0.01] <- "**"
  rci_enrollment_reported_hr_values_df$sig[rci_enrollment_reported_hr_values_df[,"P-value"] < 0.001] <- "***"

  return(rci_enrollment_reported_hr_values_df)
})
enrollment_hr_df <- do.call(rbind, enrollment_hr_df_list)


```


- Make AUC plot for onset

```{r}

# require("pROC")
# require("ggplot2")
# # roc_logistic <- roc(test_df$status, test_df$logistic_prob)
# # auc(roc_logistic)
# # Area under the curve: 1
# roc_lasso <- roc(test_df_enrollment$status, test_df_enrollment$lasso_prob)
# auc(roc_lasso)
# # Area under the curve: 0.8454
# roc_ridge <- roc(test_df_enrollment$status, test_df_enrollment$ridge_prob)
# auc(roc_ridge)
# # Area under the curve: 0.8268
# roc_elastic_net <- roc(test_df_enrollment$status, test_df_enrollment$elastic_net_prob)
# auc(roc_elastic_net)
# # Area under the curve: 0.8487

# roc_df <- ggroc(list(
#   # Logistic = roc_logistic,
#   Lasso = roc_lasso,
#   ElasticNet = roc_elastic_net,
#   Ridge = roc_ridge))$data

# current_theme <- theme(
#     legend.position = "bottom",
#     text = element_text(size = 12),
#     axis.text.x = element_text(size = 12),
#     axis.text.y = element_text(size = 10, angle = 90),
#     legend.text = element_text(size = 10),
#     legend.title = element_text(size = 10),
#     plot.background = element_blank(),
#     panel.grid.minor = element_blank(),
#     panel.grid.major = element_line(colour = "azure3",
#     linetype = "dashed",
#     size = 0.2))


# the_plot <- ggplot(roc_df, aes(x = 1 - specificity, y = sensitivity, color = name)) +
#   geom_line(size = 1.2) +
#   geom_abline(linetype = "dashed", color = "gray") +
#   theme_minimal(base_size = 14) +
#   labs(
#     title = "ROC Curves - Model Comparison",
#     x = "1 - Specificity (False Positive Rate)",
#     y = "Sensitivity (True Positive Rate)",
#     color = "Model"
#   ) +
#   scale_color_manual(values = c("Lasso" = "#7D82B8", "ElasticNet" = "#613F75", "Ridge" = "#313715")) +
#   current_theme
# ggsave(width = 7, height = 7, file = file.path(plot_dir, "05-04-plot-ROC-curves-for-all-models-rci-only.pdf"), the_plot)

```

- Save all models for creating a composite score

```{r}

saveRDS(object = list(rci_only = rci_only_ridge_composite_score_df_list,
  rci_and_nfl = rci_and_nfl_ridge_composite_score_df_list,
  rci_and_mir181 = rci_and_mir181_ridge_composite_score_df_list,
  rci_and_nfl_and_mir181 = rci_and_nfl_and_mir181_ridge_composite_score_df_list), 
  file = file.path(save_dir, "02-02-save-all-iterated-models-for-creating-a-composite-score-using-rcis-nfl-and-mir181.rds"))

```
