---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("rtracklayer")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")
require("matrixStats")
require("glmnet")
source("41-magen-et-al-miRNA-seq/scripts/sources/models.R")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "41-magen-et-al-miRNA-seq"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Analyse association between increasing intron expression rates and miRNA production

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}


replication_clinical_df <- readRDS(file.path(save_dir, "01-08-replication-clinical-with-mirna-counts-in-rcis.rds"))
clinical_df_in_use <- replication_clinical_df

replication_data_list <- readRDS(file.path(save_dir, "01-06-rci-counts-for-replication-set.rds"))
hash_pair_df <- replication_data_list$mirna_rci_overlap
rci_set <- unique(hash_pair_df$hash_id[hash_pair_df$hash_id %in% colnames(clinical_df_in_use)])
all_models_list <- readRDS(file.path(save_dir, "02-02-save-all-iterated-models-for-creating-a-composite-score-using-rcis-nfl-and-mir181.rds"))

rci_only_ridge_composite_score_df_list <- all_models_list$rci_only
seed_list <- rci_only_ridge_composite_score_df_list$seed_list

rci_and_nfl_ridge_composite_score_df_list <- all_models_list$rci_and_nfl
rci_and_mir181_ridge_composite_score_df_list <- all_models_list$rci_and_mir181
rci_and_nfl_and_mir181_ridge_composite_score_df_list <- all_models_list$rci_and_nfl_and_mir181


```

- Fetch scores rcis, rcis + nfl, rcis + mir181 and rcis + nfl + mir181

```{r}

variable_list <- list(
  rci_only = rci_set,
  rci_and_nfl = c(rci_set, "NfL_numeric"),
  rci_and_mir181 = c(rci_set, "miR_181_numeric"),
  rci_and_nfl_and_mir181 = c(rci_set, "NfL_numeric", "miR_181_numeric"))

concat_list <- list(
  rci_only = "rci_only_ridge",
  rci_and_nfl = "rci_and_nfl_ridge",
  rci_and_mir181 = "rci_and_mir181_ridge",
  rci_and_nfl_and_mir181 = "rci_and_nfl_and_mir181_ridge")

model_list <- list(
  rci_only = rci_only_ridge_composite_score_df_list$model_list,
  rci_and_nfl = rci_and_nfl_ridge_composite_score_df_list$model_list,
  rci_and_mir181 = rci_and_mir181_ridge_composite_score_df_list$model_list,
  rci_and_nfl_and_mir181 = rci_and_nfl_and_mir181_ridge_composite_score_df_list$model_list)

all_rci_score_list <- lapply(names(variable_list), function(x){
  current_var <- variable_list[[x]]
  current_concat <- concat_list[[x]]
  current_models <- model_list[[x]]
  current_clinical_df <- clinical_df_in_use
  if(grepl(x = x, pattern = "nfl")){
    current_clinical_df <- clinical_df_in_use[!is.na(clinical_df_in_use$NfL_numeric),]
  }
  all_seed_output_list <- lapply(seq_along(seed_list), function(y){
    message(y)
    current_score_df <- predict_ridge_composite_score_using_model(current_clinical_df, 
      rcis = current_var, 
      model = current_models[[y]],
      seed = seed_list[y],
      concat_col_name = current_concat)
      return(list(log_odds = current_score_df[,paste(current_concat, "log_odds", sep = "_")],
      prob = current_score_df[,paste(current_concat, "prob", sep = "_")]))
  })
  ridge_prob_values <- rowMedians(do.call(cbind, lapply(all_seed_output_list, function(y){
      y$prob
  })))
  ridge_log_odds_values <- rowMedians(do.call(cbind, lapply(all_seed_output_list, function(y){
      y$log_odds
  })))
  current_clinical_df[,paste(current_concat, "log_odds", sep = "_")] <- ridge_log_odds_values
  current_clinical_df[,paste(current_concat, "prob", sep = "_")] <- ridge_prob_values
  return(current_clinical_df)
})
names(all_rci_score_list) <- names(variable_list)


```

- Group all patients on the basis of cutpoints and compute log-rank stats

```{r}

require("survival")
require("survminer")
require("patchwork")


current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

score_list <- list(
  rci_only = c("rci_only_ridge_prob"),
  rci_and_nfl = c("rci_and_nfl_ridge_prob"),
  rci_and_mir181 = c("rci_and_mir181_ridge_prob"),
  rci_and_nfl_and_mir181 = c("rci_and_nfl_and_mir181_ridge_prob"))

onset_cutpoint_analysis_df_list <- lapply(names(score_list), function(x){
  scores_to_test <- score_list[[x]]

  current_clinical_df <- all_rci_score_list[[x]]
  test_df_onset <- current_clinical_df[,c("survival_onset", "status", scores_to_test)]

  cutpoint_dir <- file.path(output_dir, "onset-composite-score-analysis-results")
  current_date <- "2025-04-10"
  rci_onset_reported_hr_values_df <-  do.call(rbind, lapply(scores_to_test, function(x1){
    return_df <- read.csv(file.path(cutpoint_dir, current_date, x1, "2", "maxstat", "MaxStatResults.csv"))[,-1]
    return_df <- as.data.frame(t(return_df))
    colnames(return_df) <- return_df[1,]
    return_df <- cbind(data.frame(molecule = x, score = x1), return_df[-1,])
    return_df[,"P-value"] <- as.numeric(return_df[,"P-value"])
    return_df[,"Cutpoint"] <- as.numeric(return_df[,"Cutpoint"])
    return_df[,"HR"] <- as.numeric(return_df[,"HR"])
    return(return_df)
  }))
  rci_onset_reported_hr_values_df$sig <- "NS"
  rci_onset_reported_hr_values_df$sig[rci_onset_reported_hr_values_df[,"P-value"] < 0.1] <- "."
  rci_onset_reported_hr_values_df$sig[rci_onset_reported_hr_values_df[,"P-value"] < 0.05] <- "*"
  rci_onset_reported_hr_values_df$sig[rci_onset_reported_hr_values_df[,"P-value"] < 0.01] <- "**"
  rci_onset_reported_hr_values_df$sig[rci_onset_reported_hr_values_df[,"P-value"] < 0.001] <- "***"

  log_rank_stat_df <-  do.call(rbind, lapply(scores_to_test, function(x1){
    current_cutpoint_df <- rci_onset_reported_hr_values_df[rci_onset_reported_hr_values_df$score == x1,]
    test_df_onset$current_group <- ifelse(test_df_onset[,x1] > current_cutpoint_df[,"Cutpoint"], "High", "Low")
    num_groups <- length(unique(test_df_onset$current_group)) 
    if(num_groups == 1){
      diff_surv_summary_df <- data.frame(
        molecule = x,
        score = x1,
        logrank_stat = NA,
        pval = NA,
        hr = NA,
        median_high = NA,
        median_low = NA)
      return(diff_surv_summary_df)
    }
    diff_surv <- survdiff(Surv(survival_onset, status) ~ current_group, data = test_df_onset)
    diff_surv_summary_df <- data.frame(
      molecule = x, 
      score = x1,
      logrank_stat = diff_surv$chisq,
      pval = diff_surv$pvalue,
      hr = (diff_surv$obs[1] / diff_surv$exp[1]) / (diff_surv$obs[2] / diff_surv$exp[2]),
      median_high = median(test_df_onset$survival_onset[test_df_onset$current_group == "High"]),
      median_low = median(test_df_onset$survival_onset[test_df_onset$current_group == "Low"]))
    return(diff_surv_summary_df)
  }))
  log_rank_stat_df$sig <- "NS"
  log_rank_stat_df$sig[log_rank_stat_df[,"pval"] < 0.1] <- "."
  log_rank_stat_df$sig[log_rank_stat_df[,"pval"] < 0.05] <- "*"
  log_rank_stat_df$sig[log_rank_stat_df[,"pval"] < 0.01] <- "**"
  log_rank_stat_df$sig[log_rank_stat_df[,"pval"] < 0.001] <- "***"

  plot_list <- lapply(scores_to_test, function(x1){
    current_cutpoint_df <- rci_onset_reported_hr_values_df[rci_onset_reported_hr_values_df$score == x1,]
    current_logrank_df <- log_rank_stat_df[log_rank_stat_df$score == x1,]
    test_df_onset$current_group <- ifelse(test_df_onset[,x1] > current_cutpoint_df[,"Cutpoint"], "High", "Low")

    fit <- survfit(Surv(survival_onset, status) ~ current_group, data = test_df_onset)

    the_plot <- ggsurvplot(fit, 
    data = test_df_onset,
    palette = c("#EF767A", "#7D7ABC"),
    pval = T, pval.method = T,
    conf.int = F,
    title = paste(x, " LRT=", round(current_logrank_df$logrank_stat, 2), ", p=", round(current_logrank_df$pval, 3), sep = ""),
    # surv.median.line = c("v"),
    risk.table = "absolute",
    ggtheme = theme_bw() + current_theme)$plot
  })
  names(plot_list) <- scores_to_test
  return(list(cutpoint_df = rci_onset_reported_hr_values_df, log_rank = log_rank_stat_df, plots = plot_list))
})
names(onset_cutpoint_analysis_df_list) <- names(score_list)


selected_plots <- lapply(onset_cutpoint_analysis_df_list, function(x){
  x$plots[vapply(names(x$plots), function(y){
    grepl(x = y, pattern = "ridge_prob")
  },T)][[1]]
})

final_plot <- wrap_plots(selected_plots, nrow = 2, ncol = 2)

ggsave(file = file.path(plot_dir, paste("06-05-plot-onset-survival-curves-using-iterated-median-composite-score-for-onset-in-replication-set.pdf", sep = "")), final_plot, width = 10, height = 10)


```

- Fetch log-rank stats

```{r}

logrank_stat_df <- do.call(rbind, lapply(onset_cutpoint_analysis_df_list, function(x){
  x$log_rank
}))
#                                      molecule                             score
# rci_only                             rci_only               rci_only_ridge_prob
# rci_and_nfl                       rci_and_nfl            rci_and_nfl_ridge_prob
# rci_and_mir181                 rci_and_mir181         rci_and_mir181_ridge_prob
# rci_and_nfl_and_mir181 rci_and_nfl_and_mir181 rci_and_nfl_and_mir181_ridge_prob
#                        logrank_stat       pval       hr median_high median_low
# rci_only                   3.289350 0.06973060 1.598967          32      47.20
# rci_and_nfl                2.868814 0.09031142 1.569353          32      46.95
# rci_and_mir181             3.522072 0.06055669 1.644268          32      49.15
# rci_and_nfl_and_mir181     2.148165 0.14274036 1.410942          32      42.35
#                        sig
# rci_only                 .
# rci_and_nfl              .
# rci_and_mir181           .
# rci_and_nfl_and_mir181  NS

```



- Plot survival curves for enrollment

```{r}

require("survival")
require("survminer")
require("patchwork")

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

score_list <- list(
  rci_only = c("rci_only_ridge_prob"),
  rci_and_nfl = c("rci_and_nfl_ridge_prob"),
  rci_and_mir181 = c("rci_and_mir181_ridge_prob"),
  rci_and_nfl_and_mir181 = c("rci_and_nfl_and_mir181_ridge_prob"))

enrollment_cutpoint_analysis_df_list <- lapply(names(score_list), function(x){
  scores_to_test <- score_list[[x]]

  current_clinical_df <- all_rci_score_list[[x]]
  test_df_enrollment <- current_clinical_df[,c("survival_enrolment", "status", scores_to_test)]

  cutpoint_dir <- file.path(output_dir, "enrollment-composite-score-analysis-results")
  current_date <- "2025-04-10"
  rci_enrollment_reported_hr_values_df <-  do.call(rbind, lapply(scores_to_test, function(x1){
    return_df <- read.csv(file.path(cutpoint_dir, current_date, x1, "2", "maxstat", "MaxStatResults.csv"))[,-1]
    return_df <- as.data.frame(t(return_df))
    colnames(return_df) <- return_df[1,]
    return_df <- cbind(data.frame(molecule = x, score = x1), return_df[-1,])
    return_df[,"P-value"] <- as.numeric(return_df[,"P-value"])
    return_df[,"Cutpoint"] <- as.numeric(return_df[,"Cutpoint"])
    return_df[,"HR"] <- as.numeric(return_df[,"HR"])
    return(return_df)
  }))
  rci_enrollment_reported_hr_values_df$sig <- "NS"
  rci_enrollment_reported_hr_values_df$sig[rci_enrollment_reported_hr_values_df[,"P-value"] < 0.1] <- "."
  rci_enrollment_reported_hr_values_df$sig[rci_enrollment_reported_hr_values_df[,"P-value"] < 0.05] <- "*"
  rci_enrollment_reported_hr_values_df$sig[rci_enrollment_reported_hr_values_df[,"P-value"] < 0.01] <- "**"
  rci_enrollment_reported_hr_values_df$sig[rci_enrollment_reported_hr_values_df[,"P-value"] < 0.001] <- "***"

  log_rank_stat_df <-  do.call(rbind, lapply(scores_to_test, function(x1){
    current_cutpoint_df <- rci_enrollment_reported_hr_values_df[rci_enrollment_reported_hr_values_df$score == x1,]
    test_df_enrollment$current_group <- ifelse(test_df_enrollment[,x1] > current_cutpoint_df[,"Cutpoint"], "High", "Low")
    num_groups <- length(unique(test_df_enrollment$current_group)) 
    if(num_groups == 1){
      diff_surv_summary_df <- data.frame(
        molecule = x,
        score = x1,
        logrank_stat = NA,
        pval = NA,
        hr = NA,
        median_high = NA,
        median_low = NA)
      return(diff_surv_summary_df)
    }
    diff_surv <- survdiff(Surv(survival_enrolment, status) ~ current_group, data = test_df_enrollment)
    diff_surv_summary_df <- data.frame(
      molecule = x, 
      score = x1,
      logrank_stat = diff_surv$chisq,
      pval = diff_surv$pvalue,
      hr = (diff_surv$obs[1] / diff_surv$exp[1]) / (diff_surv$obs[2] / diff_surv$exp[2]),
      median_high = median(test_df_enrollment$survival_enrolment[test_df_enrollment$current_group == "High"]),
      median_low = median(test_df_enrollment$survival_enrolment[test_df_enrollment$current_group == "Low"]))
    return(diff_surv_summary_df)
  }))
  log_rank_stat_df$sig <- "NS"
  log_rank_stat_df$sig[log_rank_stat_df[,"pval"] < 0.1] <- "."
  log_rank_stat_df$sig[log_rank_stat_df[,"pval"] < 0.05] <- "*"
  log_rank_stat_df$sig[log_rank_stat_df[,"pval"] < 0.01] <- "**"
  log_rank_stat_df$sig[log_rank_stat_df[,"pval"] < 0.001] <- "***"

  plot_list <- lapply(scores_to_test, function(x1){
    current_cutpoint_df <- rci_enrollment_reported_hr_values_df[rci_enrollment_reported_hr_values_df$score == x1,]
    current_logrank_df <- log_rank_stat_df[log_rank_stat_df$score == x1,]
    test_df_enrollment$current_group <- ifelse(test_df_enrollment[,x1] > current_cutpoint_df[,"Cutpoint"], "High", "Low")

    fit <- survfit(Surv(survival_enrolment, status) ~ current_group, data = test_df_enrollment)

    the_plot <- ggsurvplot(fit, 
    data = test_df_enrollment,
    palette = c("#EF767A", "#7D7ABC"),
    pval = T, pval.method = T,
    conf.int = F,
    title = paste(x, " LRT=", round(current_logrank_df$logrank_stat, 2), ", p=", round(current_logrank_df$pval, 3), sep = ""),
    # surv.median.line = c("v"),
    risk.table = "absolute",
    ggtheme = theme_bw() + current_theme)$plot
  })
  names(plot_list) <- scores_to_test
  return(list(cutpoint_df = rci_enrollment_reported_hr_values_df, log_rank = log_rank_stat_df, plots = plot_list))
})
names(enrollment_cutpoint_analysis_df_list) <- names(score_list)


selected_plots <- lapply(enrollment_cutpoint_analysis_df_list, function(x){
  x$plots[vapply(names(x$plots), function(y){
    grepl(x = y, pattern = "ridge_prob")
  },T)][[1]]
})

final_plot <- wrap_plots(selected_plots, nrow = 2, ncol = 2)

ggsave(file = file.path(plot_dir, "06-06-plot-enrollment-survival-curves-using-iterated-median-composite-score-for-enrollment-in-replication-set.pdf"), final_plot, width = 10, height = 10)


```

- Fetch log-rank statistics

```{r}

enrollment_logrank_stat_df <- do.call(rbind, lapply(enrollment_cutpoint_analysis_df_list, function(x){
  x$log_rank
}))

#                                      molecule                             score
# rci_only                             rci_only               rci_only_ridge_prob
# rci_and_nfl                       rci_and_nfl            rci_and_nfl_ridge_prob
# rci_and_mir181                 rci_and_mir181         rci_and_mir181_ridge_prob
# rci_and_nfl_and_mir181 rci_and_nfl_and_mir181 rci_and_nfl_and_mir181_ridge_prob
#                        logrank_stat        pval       hr median_high median_low
# rci_only                   7.642855 0.005699762 2.015705        11.8       26.9
# rci_and_nfl               10.102304 0.001480840 1.883500        11.2       22.4
# rci_and_mir181             8.122319 0.004372360 2.091858        11.9       28.5
# rci_and_nfl_and_mir181     7.344527 0.006726721 1.864093        12.0       25.2
#                        sig
# rci_only                **
# rci_and_nfl             **
# rci_and_mir181          **
# rci_and_nfl_and_mir181  **

```