---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

library("here")
proj_root <- here()
message("Project root:", proj_root, "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("rtracklayer")
require("ggplot2")
require("matrixStats")
require("glmnet")
source("41-magen-et-al-miRNA-seq/scripts/sources/models.R")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("41-magen-et-al-miRNA-seq")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Analyse association between increasing intron expression rates and miRNA production

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}


replication_clinical_df <- readRDS(file.path(save_dir, "01-08-replication-clinical-with-mirna-counts-in-rcis.rds"))
discovery_clinical_df <- readRDS(file.path(save_dir, "01-07-discovery-clinical-with-mirna-counts-in-rcis.rds"))
combined_clinical_df <- rbind(replication_clinical_df, discovery_clinical_df)

clinical_df_in_use <- combined_clinical_df

replication_data_list <- readRDS(file.path(save_dir, "01-06-rci-counts-for-replication-set.rds"))
hash_pair_df <- replication_data_list$mirna_rci_overlap
rci_set <- unique(hash_pair_df$hash_id[hash_pair_df$hash_id %in% colnames(clinical_df_in_use)])
all_models_list <- readRDS(file.path(save_dir, "02-02-save-all-iterated-models-for-creating-a-composite-score-using-rcis-nfl-and-mir181.rds"))

rci_only_ridge_composite_score_df_list <- all_models_list$rci_only
seed_list <- rci_only_ridge_composite_score_df_list$seed_list

rci_and_nfl_ridge_composite_score_df_list <- all_models_list$rci_and_nfl
rci_and_mir181_ridge_composite_score_df_list <- all_models_list$rci_and_mir181
rci_and_nfl_and_mir181_ridge_composite_score_df_list <- all_models_list$rci_and_nfl_and_mir181


```

- Fetch scores rcis, rcis + nfl, rcis + mir181 and rcis + nfl + mir181

```{r}

variable_list <- list(
  rci_only = rci_set,
  rci_and_nfl = c(rci_set, "NfL_numeric"),
  rci_and_mir181 = c(rci_set, "miR_181_numeric"),
  rci_and_nfl_and_mir181 = c(rci_set, "NfL_numeric", "miR_181_numeric"))

concat_list <- list(
  rci_only = "rci_only_ridge",
  rci_and_nfl = "rci_and_nfl_ridge",
  rci_and_mir181 = "rci_and_mir181_ridge",
  rci_and_nfl_and_mir181 = "rci_and_nfl_and_mir181_ridge")

model_list <- list(
  rci_only = rci_only_ridge_composite_score_df_list$model_list,
  rci_and_nfl = rci_and_nfl_ridge_composite_score_df_list$model_list,
  rci_and_mir181 = rci_and_mir181_ridge_composite_score_df_list$model_list,
  rci_and_nfl_and_mir181 = rci_and_nfl_and_mir181_ridge_composite_score_df_list$model_list)

all_rci_score_list <- lapply(names(variable_list), function(x){
  current_var <- variable_list[[x]]
  current_concat <- concat_list[[x]]
  current_models <- model_list[[x]]
  current_clinical_df <- clinical_df_in_use
  if(grepl(x = x, pattern = "nfl")){
    current_clinical_df <- clinical_df_in_use[!is.na(clinical_df_in_use$NfL_numeric),]
  }
  all_seed_output_list <- lapply(seq_along(seed_list), function(y){
    message(y)
    current_score_df <- predict_ridge_composite_score_using_model(current_clinical_df, 
      rcis = current_var, 
      model = current_models[[y]],
      seed = seed_list[y],
      concat_col_name = current_concat)
      return(list(log_odds = current_score_df[,paste(current_concat, "log_odds", sep = "_")],
      prob = current_score_df[,paste(current_concat, "prob", sep = "_")]))
  })
  ridge_prob_values <- rowMedians(do.call(cbind, lapply(all_seed_output_list, function(y){
      y$prob
  })))
  ridge_log_odds_values <- rowMedians(do.call(cbind, lapply(all_seed_output_list, function(y){
      y$log_odds
  })))
  current_clinical_df[,paste(current_concat, "log_odds", sep = "_")] <- ridge_log_odds_values
  current_clinical_df[,paste(current_concat, "prob", sep = "_")] <- ridge_prob_values
  return(current_clinical_df)
})
names(all_rci_score_list) <- names(variable_list)


```

- Group all patients on the basis of cutpoints and compute log-rank stats

```{r}

require("survival")
require("survminer")
require("patchwork")


current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

score_list <- list(
  rci_only = c("rci_only_ridge_prob"),
  rci_and_nfl = c("rci_and_nfl_ridge_prob"),
  rci_and_mir181 = c("rci_and_mir181_ridge_prob"),
  rci_and_nfl_and_mir181 = c("rci_and_nfl_and_mir181_ridge_prob"))

  onset_file <- file.path(save_dir, "03-07-saved-combined-cohort-onset-cutpoint-analysis-data.rds")
  onset_cutpoint_analysis_df_list <- readRDS(onset_file)


selected_plots <- lapply(onset_cutpoint_analysis_df_list, function(x){
  x$plots[vapply(names(x$plots), function(y){
    grepl(x = y, pattern = "ridge_prob")
  },T)][[1]]
})

final_plot <- wrap_plots(selected_plots, nrow = 2, ncol = 2)

ggsave(file = file.path(plot_dir, paste("06-07-plot-onset-survival-curves-using-iterated-median-composite-score-for-onset-in-combined-set.pdf", sep = "")), final_plot, width = 10, height = 10)


```

- Fetch log-rank stats

```{r}

logrank_stat_df <- do.call(rbind, lapply(onset_cutpoint_analysis_df_list, function(x){
  x$log_rank
}))
#                                      molecule                             score
# rci_only                             rci_only               rci_only_ridge_prob
# rci_and_nfl                       rci_and_nfl            rci_and_nfl_ridge_prob
# rci_and_mir181                 rci_and_mir181         rci_and_mir181_ridge_prob
# rci_and_nfl_and_mir181 rci_and_nfl_and_mir181 rci_and_nfl_and_mir181_ridge_prob
#                        logrank_stat        pval       hr median_high median_low
# rci_only                   6.987868 0.008206404 1.732878      33.035      48.60
# rci_and_nfl                6.334079 0.011843959 1.722453      33.220      46.95
# rci_and_mir181             8.545650 0.003463488 1.852817      33.035      50.55
# rci_and_nfl_and_mir181     6.559542 0.010432322 1.600553      32.900      46.95
#                        sig
# rci_only                **
# rci_and_nfl              *
# rci_and_mir181          **
# rci_and_nfl_and_mir181   *

```



- Plot survival curves for enrollment in combined data

```{r}

require("survival")
require("survminer")
require("patchwork")

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

score_list <- list(
  rci_only = c("rci_only_ridge_prob"),
  rci_and_nfl = c("rci_and_nfl_ridge_prob"),
  rci_and_mir181 = c("rci_and_mir181_ridge_prob"),
  rci_and_nfl_and_mir181 = c("rci_and_nfl_and_mir181_ridge_prob"))

  enrollment_file <- file.path(save_dir, "03-08-saved-combined-cohort-enrollment-cutpoint-analysis-data.rds")
  enrollment_cutpoint_analysis_df_list <- readRDS(enrollment_file)

selected_plots <- lapply(enrollment_cutpoint_analysis_df_list, function(x){
  x$plots[vapply(names(x$plots), function(y){
    grepl(x = y, pattern = "ridge_prob")
  },T)][[1]]
})

final_plot <- wrap_plots(selected_plots, nrow = 2, ncol = 2)

ggsave(file = file.path(plot_dir, "06-08-plot-enrollment-survival-curves-using-iterated-median-composite-score-for-enrollment-in-combined-set.pdf"), final_plot, width = 10, height = 10)


```

- Fetch log-rank statistics

```{r}

enrollment_logrank_stat_df <- do.call(rbind, lapply(enrollment_cutpoint_analysis_df_list, function(x){
  x$log_rank
}))

#                                      molecule                             score
# rci_only                             rci_only               rci_only_ridge_prob
# rci_and_nfl                       rci_and_nfl            rci_and_nfl_ridge_prob
# rci_and_mir181                 rci_and_mir181         rci_and_mir181_ridge_prob
# rci_and_nfl_and_mir181 rci_and_nfl_and_mir181 rci_and_nfl_and_mir181_ridge_prob
#                        logrank_stat         pval       hr median_high
# rci_only                   13.02933 0.0003066497 2.097071       12.20
# rci_and_nfl                15.51110 0.0000820220 1.852008       12.08
# rci_and_mir181             15.09843 0.0001020481 2.242394       12.20
# rci_and_nfl_and_mir181     14.50326 0.0001399173 1.991108       12.20
#                        median_low sig
# rci_only                   28.450 ***
# rci_and_nfl                23.030 ***
# rci_and_mir181             29.605 ***
# rci_and_nfl_and_mir181     25.585 ***

```