---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

library("here")
proj_root <- here()
message("Project root:", proj_root, "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("rtracklayer")
require("ggplot2")
require("matrixStats")
require("glmnet")
source("41-magen-et-al-miRNA-seq/scripts/sources/models.R")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("41-magen-et-al-miRNA-seq")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Analyse association between increasing intron expression rates and miRNA production

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

discovery_clinical_df <- readRDS(file.path(save_dir, "01-07-discovery-clinical-with-mirna-counts-in-rcis.rds"))
# cutpoint_analysis_results <- readRDS(file.path(save_dir, "02-01-cutpoint-analysis-results-for-enrollment-and-onset.rds"))

discovery_data_list <- readRDS(here("41-magen-et-al-miRNA-seq/rdata_files/01-05-rci-counts-for-discovery-set.rds"))
hash_pair_df <- discovery_data_list$mirna_rci_overlap
rci_set <- unique(hash_pair_df$hash_id[hash_pair_df$hash_id %in% colnames(discovery_clinical_df)])
all_models_list <- readRDS(file.path(save_dir, "02-02-save-all-iterated-models-for-creating-a-composite-score-using-rcis-nfl-and-mir181.rds"))

rci_only_ridge_composite_score_df_list <- all_models_list$rci_only
seed_list <- rci_only_ridge_composite_score_df_list$seed_list

rci_and_nfl_ridge_composite_score_df_list <- all_models_list$rci_and_nfl
rci_and_mir181_ridge_composite_score_df_list <- all_models_list$rci_and_mir181
rci_and_nfl_and_mir181_ridge_composite_score_df_list <- all_models_list$rci_and_nfl_and_mir181

```

- Fetch all individual scores

```{r}

variable_list <- list(
  rci_only = rci_set,
  rci_and_nfl = c(rci_set, "NfL_numeric"),
  rci_and_mir181 = c(rci_set, "miR_181_numeric"),
  rci_and_nfl_and_mir181 = c(rci_set, "NfL_numeric", "miR_181_numeric"))

concat_list <- list(
  rci_only = "rci_only_ridge",
  rci_and_nfl = "rci_and_nfl_ridge",
  rci_and_mir181 = "rci_and_mir181_ridge",
  rci_and_nfl_and_mir181 = "rci_and_nfl_and_mir181_ridge")

all_rci_score_list <- lapply(names(variable_list), function(x){
  current_var <- variable_list[[x]]
  current_concat <- concat_list[[x]]
  current_clinical_df <- discovery_clinical_df
  if(grepl(x = x, pattern = "nfl")){
    current_clinical_df <- discovery_clinical_df[!is.na(discovery_clinical_df$NfL_numeric),]
  }
  all_seed_output_list <- lapply(seq_along(seed_list), function(y){
    message(y)
    current_score_df <- return_ridge_composite_score(current_clinical_df, 
      rcis = current_var, 
      concat_col_name = current_concat, 
      seed = seed_list[y])
    return(list(log_odds = current_score_df[,paste(current_concat, "log_odds", sep = "_")],
      prob = current_score_df[,paste(current_concat, "prob", sep = "_")]))
  })
  ridge_prob_matrix <- do.call(cbind, lapply(all_seed_output_list, function(y){
    y$prob
  }))
  ridge_log_odds_matrix <- do.call(cbind, lapply(all_seed_output_list, function(y){
    y$log_odds
  }))
  colnames(ridge_prob_matrix) <- paste(current_concat, "prob", "seed", seed_list, sep = "_")
  colnames(ridge_log_odds_matrix) <- paste(current_concat, "log_odds", "seed", seed_list, sep = "_")
  return(list(log_odds = ridge_log_odds_matrix, prob = ridge_prob_matrix))
})
names(all_rci_score_list) <- names(variable_list)

```

- Evaluate cutpoints for onset

```{r}

require("evaluateCutpoints")
current_date <- "2025-04-11"
result_dir <- "individual-seed-onset-composite-score-analysis-results"
require("parallel")
onset_cutpoint_analysis_df_list <- lapply(names(variable_list), function(x){
  message(x)
  current_var <- variable_list[[x]]
  current_concat <- concat_list[[x]]
  rci_score_list <- all_rci_score_list[[x]]
  current_clinical_df <- discovery_clinical_df
  cutpoint_dir <- file.path(output_dir, result_dir)
  if(grepl(x = x, pattern = "nfl")){
    current_clinical_df <- discovery_clinical_df[!is.na(discovery_clinical_df$NfL_numeric),]
  }
  current_clinical_df <- cbind(current_clinical_df, rci_score_list$prob)

  scores_to_test <- colnames(rci_score_list$prob)
  test_df_onset <- current_clinical_df[,c("survival_onset", "status", scores_to_test)]
  # Check if these scores have been computed already 
  keep_scores_to_test <- scores_to_test[vapply(scores_to_test, function(x1){
      !file.exists(file.path(cutpoint_dir, current_date, x1, "2", "maxstat", "MaxStatResults.csv"))
  },T)]

  if(length(keep_scores_to_test) > 0){
    evaluateCutpoints(
      mainDir = output_dir,
      resultsDirName = result_dir,
      df = test_df_onset,
      biomarkerList = keep_scores_to_test,
      time = "survival_onset",
      event = "status",
      setCutpoint = NULL,
      setCutpoint2 = NULL)    
  }

  setwd(proj_root)

  message("Fetching cutpoints...")
  rci_onset_reported_hr_values_df <-  do.call(rbind, lapply(scores_to_test, function(x1){
    return_df <- read.csv(file.path(cutpoint_dir, current_date, x1, "2", "maxstat", "MaxStatResults.csv"))[,-1]
    return_df <- as.data.frame(t(return_df))
    colnames(return_df) <- return_df[1,]
    return_df <- cbind(data.frame(molecule = x, score = x1), return_df[-1,])
    return_df[,"P-value"] <- as.numeric(return_df[,"P-value"])
    return_df[,"Cutpoint"] <- as.numeric(return_df[,"Cutpoint"])
    return_df[,"HR"] <- as.numeric(return_df[,"HR"])
    return(return_df)
  }))
  rci_onset_reported_hr_values_df$sig <- "NS"
  rci_onset_reported_hr_values_df$sig[rci_onset_reported_hr_values_df[,"P-value"] < 0.1] <- "."
  rci_onset_reported_hr_values_df$sig[rci_onset_reported_hr_values_df[,"P-value"] < 0.05] <- "*"
  rci_onset_reported_hr_values_df$sig[rci_onset_reported_hr_values_df[,"P-value"] < 0.01] <- "**"
  rci_onset_reported_hr_values_df$sig[rci_onset_reported_hr_values_df[,"P-value"] < 0.001] <- "***"
  message("Performing differential survival...")
  log_rank_stat_df <-  do.call(rbind, lapply(scores_to_test, function(x1){
      current_cutpoint_df <- rci_onset_reported_hr_values_df[rci_onset_reported_hr_values_df$score == x1,]
      test_df_onset$current_group <- ifelse(test_df_onset[,x1] > current_cutpoint_df[,"Cutpoint"], "High", "Low")
      num_groups <- length(unique(test_df_onset$current_group)) 
      if(num_groups == 1){
        diff_surv_summary_df <- data.frame(
          molecule = x,
          score_type = x1,
          logrank_stat = NA,
          pval = NA,
          hr = NA)
        return(diff_surv_summary_df)
      }
      diff_surv <- survdiff(Surv(survival_onset, status) ~ current_group, data = test_df_onset)
      diff_surv_summary_df <- data.frame(
        molecule = x, 
        score_type = x1,
        logrank_stat = diff_surv$chisq,
        pval = diff_surv$pvalue,
        hr = (diff_surv$obs[1] / diff_surv$exp[1]) / (diff_surv$obs[2] / diff_surv$exp[2]))
    return(diff_surv_summary_df)
  }))
  log_rank_stat_df$sig <- "NS"
  log_rank_stat_df$sig[log_rank_stat_df[,"pval"] < 0.1] <- "."
  log_rank_stat_df$sig[log_rank_stat_df[,"pval"] < 0.05] <- "*"
  log_rank_stat_df$sig[log_rank_stat_df[,"pval"] < 0.01] <- "**"
  log_rank_stat_df$sig[log_rank_stat_df[,"pval"] < 0.001] <- "***"
  return(list(cutpoint_df = rci_onset_reported_hr_values_df, log_rank = log_rank_stat_df))
})

names(onset_cutpoint_analysis_df_list) <- names(variable_list)

```

- Get log-rank statistic

```{r}

diff_surv_df <- do.call(rbind, lapply(onset_cutpoint_analysis_df_list, function(x){
  x$log_rank
}))

summarised_diff_surv_df <- do.call(rbind, lapply(split(diff_surv_df, diff_surv_df$molecule), function(y){
  non_na_y <- y[!is.na(y$logrank_stat),]

  logrank_stat_margin <- qt(0.975,df=nrow(non_na_y)-1) * sd(non_na_y$logrank_stat)/sqrt(nrow(non_na_y))
  pval_margin <- qt(0.975,df=nrow(non_na_y)-1) * sd(non_na_y$pval)/sqrt(nrow(non_na_y))
  hr_margin <- qt(0.975,df=nrow(non_na_y)-1) * sd(non_na_y$hr)/sqrt(nrow(non_na_y))

  return_df <- data.frame(molecule = unique(non_na_y$molecule), 
    type = c("logrank_stat", "pval", "hazard_ratio"),
    lower = c(mean(non_na_y$logrank_stat) - logrank_stat_margin,
      -log10(mean(non_na_y$pval) - pval_margin),
      mean(non_na_y$hr) - hr_margin),
    mean = c(mean(non_na_y$logrank_stat), -log10(mean(non_na_y$pval)), mean(non_na_y$hr)),
    upper = c(mean(non_na_y$logrank_stat) + logrank_stat_margin,
      -log10(mean(non_na_y$pval) + pval_margin),
      mean(non_na_y$hr) + hr_margin),
    intercepts = c(NA,-log10(0.05),1))
  return(return_df)
}))

rownames(summarised_diff_surv_df) <- NULL


```

- Plot forest plot of logstat, hazard ratios and pvalues

```{r}

summarised_diff_surv_df$molecule <- factor(summarised_diff_surv_df$molecule, 
  levels = rev(c("rci_only", "rci_and_nfl", "rci_and_mir181", "rci_and_nfl_and_mir181")))

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

ThePlot <- ggplot(summarised_diff_surv_df, aes(y = mean, x = molecule))
ThePlot <- ThePlot + geom_pointrange(aes(ymin = lower, ymax = upper))
ThePlot <- ThePlot + geom_hline(aes(yintercept = intercepts), lty = "dashed", colour = "#000000")
ThePlot <- ThePlot + scale_x_discrete("composite scores")
ThePlot <- ThePlot + scale_y_continuous("values")
ThePlot <- ThePlot + coord_flip()
# ThePlot <- ThePlot + scale_colour_manual("MND status", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_grid(. ~ type, scales = "free")
ThePlot <- ThePlot + ggtitle("forest plot of hazard ratios")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 5, file = file.path(plot_dir, "06-01-plot-onset-forest-plot-of-lrt-pval-hr-for-all-1k-models.pdf"), ThePlot)


```


- Evaluate cutpoints for enrollment

```{r}

require("evaluateCutpoints")
current_date <- "2025-04-14"
result_dir <- "individual-seed-enrollment-composite-score-analysis-results"
enrollment_cutpoint_analysis_df_list <- mclapply(names(variable_list), function(x){
  message(x)
  cutpoint_dir <- file.path(output_dir, result_dir)
  current_var <- variable_list[[x]]
  current_concat <- concat_list[[x]]
  rci_score_list <- all_rci_score_list[[x]]
  current_clinical_df <- discovery_clinical_df
  if(grepl(x = x, pattern = "nfl")){
    current_clinical_df <- discovery_clinical_df[!is.na(discovery_clinical_df$NfL_numeric),]
  }
  current_clinical_df <- cbind(current_clinical_df, rci_score_list$prob)

  scores_to_test <- colnames(rci_score_list$prob)
  test_df_enrollment <- current_clinical_df[,c("survival_enrolment", "status", scores_to_test)]
  # Check if these scores have been computed already 
  keep_scores_to_test <- scores_to_test[vapply(scores_to_test, function(x1){
      !file.exists(file.path(cutpoint_dir, current_date, x1, "2", "maxstat", "MaxStatResults.csv"))
  },T)]

  if(length(keep_scores_to_test) > 0){
    evaluateCutpoints(
      mainDir = output_dir,
      resultsDirName = result_dir,
      df = test_df_enrollment,
      biomarkerList = keep_scores_to_test,
      time = "survival_enrolment",
      event = "status",
      setCutpoint = NULL,
      setCutpoint2 = NULL)    
  }

  setwd(proj_root)

  message("Fetching cutpoints...")
  rci_enrollment_reported_hr_values_df <-  do.call(rbind, lapply(scores_to_test, function(x1){
    return_df <- read.csv(file.path(cutpoint_dir, current_date, x1, "2", "maxstat", "MaxStatResults.csv"))[,-1]
    return_df <- as.data.frame(t(return_df))
    colnames(return_df) <- return_df[1,]
    return_df <- cbind(data.frame(molecule = x, score = x1), return_df[-1,])
    return_df[,"P-value"] <- as.numeric(return_df[,"P-value"])
    return_df[,"Cutpoint"] <- as.numeric(return_df[,"Cutpoint"])
    return_df[,"HR"] <- as.numeric(return_df[,"HR"])
    return(return_df)
  }))
  rci_enrollment_reported_hr_values_df$sig <- "NS"
  rci_enrollment_reported_hr_values_df$sig[rci_enrollment_reported_hr_values_df[,"P-value"] < 0.1] <- "."
  rci_enrollment_reported_hr_values_df$sig[rci_enrollment_reported_hr_values_df[,"P-value"] < 0.05] <- "*"
  rci_enrollment_reported_hr_values_df$sig[rci_enrollment_reported_hr_values_df[,"P-value"] < 0.01] <- "**"
  rci_enrollment_reported_hr_values_df$sig[rci_enrollment_reported_hr_values_df[,"P-value"] < 0.001] <- "***"
  message("Performing differential survival...")
  log_rank_stat_df <-  do.call(rbind, lapply(scores_to_test, function(x1){
      current_cutpoint_df <- rci_enrollment_reported_hr_values_df[rci_enrollment_reported_hr_values_df$score == x1,]
      test_df_enrollment$current_group <- ifelse(test_df_enrollment[,x1] > current_cutpoint_df[,"Cutpoint"], "High", "Low")
      num_groups <- length(unique(test_df_enrollment$current_group)) 
      if(num_groups == 1){
        diff_surv_summary_df <- data.frame(
          molecule = x,
          score_type = x1,
          logrank_stat = NA,
          pval = NA,
          hr = NA)
        return(diff_surv_summary_df)
      }
      diff_surv <- survdiff(Surv(survival_enrolment, status) ~ current_group, data = test_df_enrollment)
      diff_surv_summary_df <- data.frame(
        molecule = x, 
        score_type = x1,
        logrank_stat = diff_surv$chisq,
        pval = diff_surv$pvalue,
        hr = (diff_surv$obs[1] / diff_surv$exp[1]) / (diff_surv$obs[2] / diff_surv$exp[2]))
    return(diff_surv_summary_df)
  }))
  log_rank_stat_df$sig <- "NS"
  log_rank_stat_df$sig[log_rank_stat_df[,"pval"] < 0.1] <- "."
  log_rank_stat_df$sig[log_rank_stat_df[,"pval"] < 0.05] <- "*"
  log_rank_stat_df$sig[log_rank_stat_df[,"pval"] < 0.01] <- "**"
  log_rank_stat_df$sig[log_rank_stat_df[,"pval"] < 0.001] <- "***"
  return(list(cutpoint_df = rci_enrollment_reported_hr_values_df, log_rank = log_rank_stat_df))
}, mc.cores = 4)

names(enrollment_cutpoint_analysis_df_list) <- names(variable_list)

saveRDS(object = list(onset = onset_cutpoint_analysis_df_list, enrollment = enrollment_cutpoint_analysis_df_list), 
  file = file.path(save_dir, "02-03-onset-and-enrollment-cutpoint-analysis-results-for-each-individual-seed-list.rds"))

```

- Get log-rank statistic for enrollment

```{r}

enrollment_diff_surv_df <- do.call(rbind, lapply(enrollment_cutpoint_analysis_df_list, function(x){
  x$log_rank
}))

summarised_enrollment_diff_surv_df <- do.call(rbind, lapply(split(enrollment_diff_surv_df, enrollment_diff_surv_df$molecule), function(y){
  non_na_y <- y[!is.na(y$logrank_stat),]

  logrank_stat_margin <- qt(0.975,df=nrow(non_na_y)-1) * sd(non_na_y$logrank_stat)/sqrt(nrow(non_na_y))
  pval_margin <- qt(0.975,df=nrow(non_na_y)-1) * sd(non_na_y$pval)/sqrt(nrow(non_na_y))
  hr_margin <- qt(0.975,df=nrow(non_na_y)-1) * sd(non_na_y$hr)/sqrt(nrow(non_na_y))

  return_df <- data.frame(molecule = unique(non_na_y$molecule), 
    type = c("logrank_stat", "pval", "hazard_ratio"),
    lower = c(mean(non_na_y$logrank_stat) - logrank_stat_margin,
      -log10(mean(non_na_y$pval) - pval_margin),
      mean(non_na_y$hr) - hr_margin),
    mean = c(mean(non_na_y$logrank_stat), -log10(mean(non_na_y$pval)), mean(non_na_y$hr)),
    upper = c(mean(non_na_y$logrank_stat) + logrank_stat_margin,
      -log10(mean(non_na_y$pval) + pval_margin),
      mean(non_na_y$hr) + hr_margin),
    intercepts = c(NA,-log10(0.05),1))
  return(return_df)
}))

rownames(summarised_enrollment_diff_surv_df) <- NULL


```



- Plot forest plot of logstat, hazard ratios and pvalues for enrollment

```{r}

summarised_enrollment_diff_surv_df$molecule <- factor(summarised_enrollment_diff_surv_df$molecule, 
  levels = rev(c("rci_only", "rci_and_nfl", "rci_and_mir181", "rci_and_nfl_and_mir181")))

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

ThePlot <- ggplot(summarised_enrollment_diff_surv_df, aes(y = mean, x = molecule))
ThePlot <- ThePlot + geom_pointrange(aes(ymin = lower, ymax = upper))
ThePlot <- ThePlot + geom_hline(aes(yintercept = intercepts), lty = "dashed", colour = "#000000")
ThePlot <- ThePlot + scale_x_discrete("composite scores")
ThePlot <- ThePlot + scale_y_continuous("values")
ThePlot <- ThePlot + coord_flip()
# ThePlot <- ThePlot + scale_colour_manual("MND status", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_grid(. ~ type, scales = "free")
ThePlot <- ThePlot + ggtitle("forest plot of hazard ratios")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 5, file = file.path(plot_dir, "06-02-plot-enrollment-forest-plot-of-lrt-pval-hr-for-all-1k-models.pdf"), ThePlot)

```