---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current project root directory and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

library("here")
proj_root <- here()
message("Project root:", proj_root, "\n")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("rtracklayer")
require("ggplot2")
require("matrixStats")
require("glmnet")
source("41-magen-et-al-miRNA-seq/scripts/sources/models.R")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- here("41-magen-et-al-miRNA-seq")
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Analyse association between increasing intron expression rates and miRNA production

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}


replication_clinical_df <- readRDS(file.path(save_dir, "01-08-replication-clinical-with-mirna-counts-in-rcis.rds"))
discovery_clinical_df <- readRDS(file.path(save_dir, "01-07-discovery-clinical-with-mirna-counts-in-rcis.rds"))
combined_clinical_df <- rbind(replication_clinical_df, discovery_clinical_df)

clinical_df_in_use <- combined_clinical_df

replication_data_list <- readRDS(file.path(save_dir, "01-06-rci-counts-for-replication-set.rds"))
hash_pair_df <- replication_data_list$mirna_rci_overlap
rci_set <- unique(hash_pair_df$hash_id[hash_pair_df$hash_id %in% colnames(clinical_df_in_use)])

keep_hash_pair_df <- hash_pair_df[hash_pair_df$hash_id %in% rci_set,]

```

- Add gene id to hash pair data.frame

```{r}

feature_ranges <- readRDS(here("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds"))
keep_hash_pair_df$gene_id <- feature_ranges$gene_id[match(keep_hash_pair_df$hash_id, feature_ranges$hash_id)]

```

- Investigate overlap between predictors and miRNAs

```{r}

hpo_terms_with_cluster_df <- readRDS(file.path("23-final-gene-set-enrichments/rdata_files/19-hpo-interesting-terms-with-group-assignment-and-plot-order.rds"))

predictor_genes <- lapply(split(str_split(hpo_terms_with_cluster_df$inputGenes, pattern = ","), hpo_terms_with_cluster_df$group), function(x){
    unique(do.call(c, x))
})
predictor_genes$moderate_predictors <- predictor_genes$moderate_predictors[!(predictor_genes$moderate_predictors %in% predictor_genes$strong_predictors)]
predictor_genes$modest_predictors <- predictor_genes$modest_predictors[!(predictor_genes$modest_predictors %in% predictor_genes$moderate_predictors) & 
    !(predictor_genes$modest_predictors %in% predictor_genes$strong_predictors)]

predictor_df <- data.frame(gene_id = do.call(c, predictor_genes),
  predictor = rep(names(predictor_genes), vapply(predictor_genes, length, 1)))

keep_hash_pair_df$predictor <- predictor_df[match(keep_hash_pair_df$gene_id, predictor_df$gene_id), "predictor"]

unique_hash_df <- unique(keep_hash_pair_df[,c("hash_id","predictor")])

frequency_df <- as.data.frame(table(unique_hash_df$predictor))
colnames(frequency_df) <- c("predictor_group", "frequency")
frequency_df$proportion <- frequency_df$frequency / nrow(unique_hash_df)
frequency_df <- rbind(frequency_df, data.frame(predictor_group = "modest_predictors", frequency = 0, proportion = 0))

frequency_df$predictor_group <- factor(frequency_df$predictor_group, levels = c("strong_predictors", "moderate_predictors", "modest_predictors"))

```

- Plot frequency

```{r}


current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

Colours <- c("strong_predictors" = "#f36f48", "moderate_predictors" = "#5688C7", "modest_predictors" = "#FDE2E2")
ThePlot <- ggplot(frequency_df, aes(x = predictor_group, y = proportion))
ThePlot <- ThePlot + geom_col(aes(fill = predictor_group), shape = 21, colour = "#000000")
ThePlot <- ThePlot + geom_text(aes(label = frequency, y = proportion / 2))
ThePlot <- ThePlot + scale_x_discrete("predictor groups")
ThePlot <- ThePlot + scale_y_continuous("proportion of miR-RCIs identified in each predictor group")
ThePlot <- ThePlot + scale_fill_manual("predictors", values = Colours)
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(. ~ type, scales = "free")
ThePlot <- ThePlot + ggtitle("Proportion of rcis producing miR-RCIs")
ThePlot <- ThePlot + current_theme
ggsave(width = 5, height = 10, file = file.path(plot_dir, "06-12-plot-proportion-of-miR-RCIs-in-each-predictor-groups.pdf"), ThePlot)


```
