---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
require("rtracklayer")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")
require("matrixStats")
require("glmnet")
source("41-magen-et-al-miRNA-seq/scripts/sources/models.R")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "41-magen-et-al-miRNA-seq"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Analyse association between increasing intron expression rates and miRNA production

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

discovery_clinical_df <- readRDS(file.path(save_dir, "01-07-discovery-clinical-with-mirna-counts-in-rcis.rds"))
clinical_df_in_use <- discovery_clinical_df

discovery_data_list <- readRDS("41-magen-et-al-miRNA-seq/rdata_files/01-06-rci-counts-for-replication-set.rds")
hash_pair_df <- discovery_data_list$mirna_rci_overlap
rci_set <- unique(hash_pair_df$hash_id[hash_pair_df$hash_id %in% colnames(clinical_df_in_use)])
keep_hash_pair_df <- hash_pair_df[hash_pair_df$hash_id %in% rci_set,]

```

- Add gene id to hash pair data.frame

```{r}

feature_ranges <- readRDS("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds")
keep_hash_pair_df$gene_id <- feature_ranges$gene_id[match(keep_hash_pair_df$hash_id, feature_ranges$hash_id)]

annotation_gtf <- "/nemo/lab/patanir/home/users/palk/Genomes/ENSEMBL_hg38/Homo_sapiens.GRCh38.104.chr_patch_hapl_scaff.gtf"
annotation_ranges <- import.gff2(annotation_gtf)

```

- Investigate overlap between predictors and miRNAs

```{r}

hpo_terms_with_cluster_df <- readRDS(file.path("23-final-gene-set-enrichments/rdata_files/19-hpo-interesting-terms-with-group-assignment-and-plot-order.rds"))

predictor_genes <- lapply(split(str_split(hpo_terms_with_cluster_df$inputGenes, pattern = ","), hpo_terms_with_cluster_df$group), function(x){
    unique(do.call(c, x))
})
predictor_genes$moderate_predictors <- predictor_genes$moderate_predictors[!(predictor_genes$moderate_predictors %in% predictor_genes$strong_predictors)]
predictor_genes$modest_predictors <- predictor_genes$modest_predictors[!(predictor_genes$modest_predictors %in% predictor_genes$moderate_predictors) & 
    !(predictor_genes$modest_predictors %in% predictor_genes$strong_predictors)]

predictor_df <- data.frame(gene_id = do.call(c, predictor_genes),
  predictor = rep(names(predictor_genes), vapply(predictor_genes, length, 1)))

keep_hash_pair_df$predictor <- predictor_df[match(keep_hash_pair_df$gene_id, predictor_df$gene_id), "predictor"]


```


- Fetch coefficients for rcis

```{r}

all_models_list <- readRDS(file.path(save_dir, "02-02-save-all-iterated-models-for-creating-a-composite-score-using-rcis-nfl-and-mir181.rds"))
rci_only_ridge_composite_score_df_list <- all_models_list$rci_only
model_list <- rci_only_ridge_composite_score_df_list$model_list

all_coef <- do.call(rbind, lapply(model_list, function(x){
  coef(x, s = "lambda.min")[rci_set,]
}))

all_coef_df <- reshape2::melt(all_coef)
colnames(all_coef_df) <- c("seq", "rci", "coef")


```

- Isolate predictor hashes

```{r}

predictor_hash_pair_df <- keep_hash_pair_df[!is.na(keep_hash_pair_df$predictor),]

all_coef_df$predictor_group <- predictor_hash_pair_df$predictor[match(all_coef_df$rci, predictor_hash_pair_df$hash_id)]
# all_coef_df$rescaled_weights <- scales::rescale(all_coef_df$coef, c(0,1))
keep_coef_df <- all_coef_df[!is.na(all_coef_df$predictor_group),]
keep_coef_df$predictor_group <- factor(keep_coef_df$predictor_group, levels = c("strong_predictors", "moderate_predictors", "modest_predictors"))

wilcox.test(keep_coef_df$coef[keep_coef_df$predictor_group == "strong_predictors"], keep_coef_df$coef[keep_coef_df$predictor_group == "moderate_predictors"])$p.value


current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

Colours <- c("strong_predictors" = "#f36f48", "moderate_predictors" = "#5688C7", "modest_predictors" = "#FDE2E2")

ThePlot <- ggplot(keep_coef_df, aes(x = predictor_group, y = coef))
ThePlot <- ThePlot + geom_boxplot(aes(fill = predictor_group), colour = "#000000", 
    width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9))
ThePlot <- ThePlot + scale_x_discrete("predictor groups")
ThePlot <- ThePlot + scale_y_continuous("ridge regression coefficients of miR-RCIs stratified by predictor groups")
ThePlot <- ThePlot + scale_fill_manual("predictors", values = Colours)
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(. ~ type, scales = "free")
ThePlot <- ThePlot + ggtitle("Proportion of rcis producing miR-RCIs")
ThePlot <- ThePlot + current_theme
ggsave(width = 5, height = 10, file = file.path(plot_dir, "06-13-plot-ridge-regression-weights-of-miR-RCIs-in-each-predictor-groups.pdf"), ThePlot)


```


- Which genes contribute the most towards the score


```{r}


annotation_ranges[annotation_ranges$type == "gene" & annotation_ranges$gene_id %in% feature_ranges$gene_id[feature_ranges$hash_id %in% unique(keep_coef_df[keep_coef_df$coef > 0, 2])]]

# GRanges object with 2 ranges and 22 metadata columns:
#       seqnames              ranges strand |         source     type     score
#          <Rle>           <IRanges>  <Rle> |       <factor> <factor> <numeric>
#   [1]        3 124033369-124726325      + | ensembl_havana     gene        NA
#   [2]       16     5239802-7713340      + | ensembl_havana     gene        NA
#           phase         gene_id gene_version   gene_name    gene_source
#       <integer>     <character>  <character> <character>    <character>
#   [1]      <NA> ENSG00000160145           16       KALRN ensembl_havana
#   [2]      <NA> ENSG00000078328           23      RBFOX1 ensembl_havana
#         gene_biotype transcript_id transcript_version transcript_name
#          <character>   <character>        <character>     <character>
#   [1] protein_coding          <NA>               <NA>            <NA>
#   [2] protein_coding          <NA>               <NA>            <NA>
#       transcript_source transcript_biotype         tag     ccds_id
#             <character>        <character> <character> <character>
#   [1]              <NA>               <NA>        <NA>        <NA>
#   [2]              <NA>               <NA>        <NA>        <NA>
#       transcript_support_level exon_number     exon_id exon_version  protein_id
#                    <character> <character> <character>  <character> <character>
#   [1]                     <NA>        <NA>        <NA>         <NA>        <NA>
#   [2]                     <NA>        <NA>        <NA>         <NA>        <NA>
#       protein_version
#           <character>
#   [1]            <NA>
#   [2]            <NA>

```
