---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-2"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
feature_ranges <- readRDS("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds")
sig_result_list <- readRDS("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds")
nygc_object_list <- readRDS("13-nygc-consortium/rdata_files/03-02-list_of_counts_stat_samplesheet_from_nygc.rds")
patient_df <- readRDS("13-nygc-consortium/rdata_files/03-05-nygc-patient-sampleinfo-datasheet.rds")
# patient_liability_df <- readRDS(file.path(save_dir, "01-save-nygc-rci-liabilities-in-patient-samples.rds"))

```

- Load repeat and rci object

```{r}

repeat_ranges <- unique(repeat_ranges_list$repeat_ranges)
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges

```

### Analysis


- Generate TPM matrix

```{r}

feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges
repeat_ranges <- repeat_ranges_list$repeat_ranges

sig_result_df <- sig_result_list$significant_events


repeat_containing_introns_df <- sig_result_df[sig_result_df$status == "repeat_containing" & 
    sig_result_df$region == "intron",]

repeat_containing_introns_df$cohort <- "public-data"
repeat_containing_introns_df$cohort[grepl(x = repeat_containing_introns_df$study, pattern = "nygc")] <- "nygc"
repeat_containing_introns_df$cohort[grepl(x = repeat_containing_introns_df$study, pattern = "answerals")] <- "answerals"

repeat_containing_introns_df$tissue_type <- "iPSCs"
repeat_containing_introns_df$tissue_type[grepl(x = repeat_containing_introns_df$study, pattern = "nygc")] <- "postmortem"

counts_matrix <- nygc_object_list$counts
temp_matrix <- counts_matrix / (width(feature_ranges)[match(rownames(counts_matrix), feature_ranges$hash_id)] / 1e3)
tpm_matrix <- t(t(temp_matrix) / (colSums(temp_matrix) / 1e6))


keep_hash_ids <- feature_ranges_with_repeats$hash_id[match(unique(repeat_containing_introns_df$feature_ids), feature_ranges_with_repeats$feature_id)]
gene_ids <- feature_ranges_with_repeats$gene_id[match(keep_hash_ids, feature_ranges_with_repeats$hash_id)]
keep_tpm_matrix <- tpm_matrix[keep_hash_ids,colnames(tpm_matrix) %in% patient_df$column_id]

```

- Generate Intron expression rates

```{r}

intronic_expression_rate <- do.call(rbind, lapply(keep_hash_ids, function(a_hash_id){
    exonic_hash_id <- feature_ranges$hash_id[feature_ranges$gene_id == gene_ids[keep_hash_ids == a_hash_id] & feature_ranges$type == "exon"]
    if(length(exonic_hash_id) == 1){
        exonic_counts <- temp_matrix[exonic_hash_id,] / 1e6
    }else{
        exonic_counts <- colSums(temp_matrix[exonic_hash_id,]) / 1e6
    }

    intronic_hash_id <- a_hash_id
    if(length(intronic_hash_id) == 1){
        intronic_counts <- temp_matrix[intronic_hash_id,]
    }else{
        intronic_counts <- colSums(temp_matrix[intronic_hash_id,])
    }
    intronic_emission_per_million_exons <- intronic_counts / exonic_counts
    return(intronic_emission_per_million_exons)
}))
rownames(intronic_expression_rate) <- keep_hash_ids

repeat_lengths <- feature_ranges_with_repeats$repeat_width[match(rownames(keep_tpm_matrix), feature_ranges_with_repeats$hash_id)]
total_repeat_width <- vapply(split(width(unique(repeat_ranges)), unique(repeat_ranges)$span_group), function(x){
    sum(x) 
},1)

expected_distribution_of_repeats <- total_repeat_width / sum(width(reduce(unique(feature_ranges_with_repeats), ignore.strand = T)))

```

- Generate penetrant RCIs

```{r}

coldata_df_split <- split(patient_df, paste(patient_df$subject_id, patient_df$tissue))
patient_penetrance_matrix <- do.call(rbind, lapply(coldata_df_split, function(temp_df){
    current_columns <- temp_df$column_id
    
    current_tpm_vector <- Reduce("+", lapply(current_columns, function(x){
        keep_tpm_matrix[,x]  
    })) / length(current_columns)

    current_threshold <- 1
    if(any(temp_df$tissue %in% c("Cortex", "Cortex_Motor", "Cerebellum"))){
        current_threshold <- 0.1
    }
    a_vector <- rep(0, length(unique(feature_ranges_with_repeats$hash_id)))
    names(a_vector) <- unique(feature_ranges_with_repeats$hash_id)
    a_vector[names(current_tpm_vector[current_tpm_vector > current_threshold])] <- 1
    return(a_vector)
}))

```

- Generate liability data

```{r}

coldata_df_split <- split(patient_df, paste(patient_df$subject_id, patient_df$tissue))
patient_liability_df <- do.call(rbind, lapply(coldata_df_split, function(temp_df){
    return_df <- unique(temp_df[,-c(1,2,3,4,5,6,7,8,9,11,19,21)])
    current_columns <- temp_df$column_id
    if(nrow(return_df) > 1){
        if("N/A" %in% return_df$mnd_with_ftd & "Yes" %in% return_df$mnd_with_ftd){
            return_df <- return_df[return_df$mnd_with_ftd == "Yes",]
        }else if("N/A" %in% return_df$mnd_with_ftd & "No" %in% return_df$mnd_with_ftd){
            return_df <- return_df[return_df$mnd_with_ftd == "No",]
        }else if("tardbp" %in% return_df$type & "sporadic" %in% return_df$type){
            return_df <- return_df[return_df$type == "tardbp",]
        }else if("fus" %in% return_df$mutation & "none" %in% return_df$mutation){
            return_df <- return_df[return_df$mutation == "fus",]
        }else{
            stop(message(paste(return_df$subject_id, return_df$tissue)))
        }
    }
    
    current_tpm_vector <- Reduce("+", lapply(current_columns, function(x){
        keep_tpm_matrix[,x]  
    })) / length(current_columns)

    current_rate_vector <- Reduce("+", lapply(current_columns, function(x){
        intronic_expression_rate[,x]  
    })) / length(current_columns)

    all_introns <- feature_ranges_with_repeats 
    all_introns$signal <- current_tpm_vector[all_introns$hash_id]
    all_introns$weighted.signal <- all_introns$signal / (table(all_introns$hash_id)[all_introns$hash_id])
    all_introns$weighted_signal_per_expansion <- all_introns$weighted.signal / all_introns$expansion_size
    all_introns <- all_introns[!is.na(all_introns$signal)]

    expected_split <- split(all_introns$weighted.signal, all_introns$span_group)
    total_expected_signal <- vapply(expected_split, sum, 1)
    prop_expected <- total_expected_signal / sum(total_expected_signal)

    current_threshold <- 1
    if(return_df$tissue %in% c("Cortex", "Cortex_Motor", "Cerebellum")){
        current_threshold <- 0.1
    }

    return_df$rci_liability <- sum(current_tpm_vector[current_tpm_vector > current_threshold])
    return_df$median_rci_liability <- median(current_tpm_vector[current_tpm_vector > current_threshold])
    return_df$n_of_rcis <- length(current_tpm_vector[current_tpm_vector > current_threshold])
    return_df$intron_expression_rate <- median(current_rate_vector[
        current_tpm_vector > current_threshold 
        & !is.infinite(current_rate_vector)], na.rm = T)

    current_introns <- all_introns[all_introns$hash_id %in% names(current_tpm_vector[current_tpm_vector > current_threshold])]

    width_split <- split(current_introns$repeat_width, current_introns$span_group)
    total_width <- sum(current_introns$repeat_width)
    prop_width <- vapply(width_split, sum, 1) / total_width

    signal_split <- split(current_introns$weighted.signal, current_introns$span_group)

    observed_proportion <- vapply(c("trinucleotide", "quadnucleotide", "pentanucleotide", "hexanucleotide"), function(x){
        if(x %in% names(signal_split)){
            return(sum(signal_split[[x]]) / sum(current_introns$weighted.signal))
        }else{
            return(0)
        }
    },1)
    enrichment <- observed_proportion / prop_expected[names(observed_proportion)]

    overall_mean_signal <- mean(current_introns$weighted.signal)
    mean_weighted_signal <- vapply(c("trinucleotide", "quadnucleotide", "pentanucleotide", "hexanucleotide"), function(x){
        if(x %in% current_introns$span_group){
            return(mean(current_introns$weighted.signal[current_introns$span_group == x]))
        }else{
            return(0)
        }
    },1)

    share_of_signal <- vapply(c("trinucleotide", "quadnucleotide", "pentanucleotide", "hexanucleotide"), function(x){
        if(x %in% current_introns$span_group){
            prop <- sum(width(current_introns[current_introns$span_group == x])) / sum(width(current_introns))
            return(prop  * (mean_weighted_signal[x] / overall_mean_signal))
        }else{
            return(0)
        }
    },1)

    within_group_theil <- vapply(c("trinucleotide", "quadnucleotide", "pentanucleotide", "hexanucleotide"), function(x){
        if(x %in% current_introns$span_group){
            current_signal <- current_introns$weighted.signal[current_introns$span_group == x]
            inner_ratio <- current_signal / mean_weighted_signal[x]
            return(sum(inner_ratio * log(inner_ratio)))
        }else{
            return(0)
        }
    },1)

    overall_ratios <- vapply(c("trinucleotide", "quadnucleotide", "pentanucleotide", "hexanucleotide"), function(x){
        if(x %in% current_introns$span_group){
            return(mean_weighted_signal[x] / overall_mean_signal)
        }else{
            return(0)
        }
    },1)
    all_between_repeat_component <- log(overall_ratios) * share_of_signal
    between_repeat_component <- sum(all_between_repeat_component)


    theil_t <- sum(share_of_signal * within_group_theil) + between_repeat_component

    names(enrichment) <- paste(names(enrichment), "enrichment", sep = "_")
    names(observed_proportion) <- paste(names(observed_proportion), "observed", sep = "_")
    names(all_between_repeat_component) <- paste(names(all_between_repeat_component), "between_theil", sep = "_")

    new_return_df <- return_df
    new_return_df$entropy <- theil_t
    return(cbind(new_return_df[1,], t(as.matrix(enrichment)), t(as.matrix(observed_proportion)), t(as.matrix(all_between_repeat_component))))
}))

patient_liability_df$penetrance_row <- seq_len(nrow(patient_penetrance_matrix))

```

- Save NYGC penetrance data
 
```{r}

saveRDS(object = list(liability_df = patient_liability_df, penetrance_matrix = patient_penetrance_matrix), file = file.path(save_dir, "02-save-nygc-rci-expression-summary-with-penetrance-matrix.rds"))

```
