---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")
require("ggbeeswarm")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-3"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")
source(file.path("13-03-public-data/scripts", "src", "experiment_matrix_functions.R"))

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
feature_ranges <- readRDS("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds")
sig_result_list <- readRDS("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds")
publicdata_object_list <- readRDS("13-03-public-data/rdata_files/02-list_of_counts_stat_samplesheet_from_public_data.rds")
nygc_penetrance_list <- readRDS(file.path("final-figures/figure-2/rdata_files/02-save-nygc-rci-expression-summary-with-penetrance-matrix.rds"))

```

- Load repeat and rci object

```{r}


repeat_ranges <- unique(repeat_ranges_list$repeat_ranges)
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges


```

- Get count matrix

```{r}

sample_df <- publicdata_object_list$sample_df
counts_matrix <- publicdata_object_list$counts


```

### Analysis


- Generate TPM matrix

```{r}


sig_result_df <- sig_result_list$significant_events

repeat_containing_introns_df <- sig_result_df[sig_result_df$status == "repeat_containing" & 
    sig_result_df$region == "intron",]


repeat_containing_introns_df$cohort <- "public-data"
repeat_containing_introns_df$cohort[grepl(x = repeat_containing_introns_df$study, pattern = "nygc")] <- "nygc"
repeat_containing_introns_df$cohort[grepl(x = repeat_containing_introns_df$study, pattern = "answerals")] <- "answerals"

repeat_containing_introns_df$tissue_type <- "iPSCs"
repeat_containing_introns_df$tissue_type[grepl(x = repeat_containing_introns_df$study, pattern = "nygc")] <- "postmortem"


temp_matrix <- counts_matrix / (width(feature_ranges)[match(rownames(counts_matrix), feature_ranges$hash_id)] / 1e3)
tpm_matrix <- t(t(temp_matrix) / (colSums(temp_matrix) / 1e6))

keep_hash_ids <- feature_ranges_with_repeats$hash_id[match(unique(repeat_containing_introns_df$feature_ids), feature_ranges_with_repeats$feature_id)]
keep_tpm_matrix <- tpm_matrix[keep_hash_ids,]

```

- Fetch NYGC penetrant rcis

```{r}

nygc_patient_liability_df <- nygc_penetrance_list$liability_df
nygc_penetrance_matrix <- nygc_penetrance_list$penetrance_matrix

nygc_patient_df_split <- split(nygc_patient_liability_df, paste(nygc_patient_liability_df$condition, nygc_patient_liability_df$tissue_group))
nygc_ordered_patient_penetrance_matrix_list <- lapply(nygc_patient_df_split, function(temp_df){
    nygc_penetrance_matrix[temp_df$penetrance_row[order(temp_df$repeat_intron_expr_tpm, decreasing = F)],]
})

nygc_penetrant_rcis <- lapply(nygc_ordered_patient_penetrance_matrix_list, function(x){
    temp_vector <- colSums(x)
    names(temp_vector[temp_vector > 0])
})

brain_rcis <- nygc_penetrant_rcis[grepl(x = names(nygc_penetrant_rcis), pattern = "brain")]
spinal_cord_rcis <- nygc_penetrant_rcis[grepl(x = names(nygc_penetrant_rcis), pattern = "spinal_cord")]

als_unique_spinal_cord_rcis <- unique(c(spinal_cord_rcis[["als spinal_cord"]], spinal_cord_rcis[["als+ftd spinal_cord"]]))
als_unique_spinal_cord_rcis <- als_unique_spinal_cord_rcis[!(als_unique_spinal_cord_rcis %in% spinal_cord_rcis[["ctrl spinal_cord"]])]

als_unique_brain_rcis <- unique(c(brain_rcis[["als brain"]], brain_rcis[["als+ftd brain"]]))
als_unique_brain_rcis <- als_unique_brain_rcis[!(als_unique_brain_rcis %in% brain_rcis[["ctrl brain"]])]

```


- Prepare experiment data.frame


```{r}


all_functions_list <- return_function_list()


all_experiment_df <- do.call(rbind, lapply(seq_along(all_functions_list), function(x){
    sample_name <- names(all_functions_list)[[x]]
    message(sample_name)
    FUN <- all_functions_list[[sample_name]]
    experiment_df <- FUN(sample_df, study_specific_design = F)$experiment
}))

all_experiment_df$condition[all_experiment_df$condition == "vcp mutant"] <- "vcp"
all_experiment_df$condition[all_experiment_df$condition == "vcp knock-in"] <- "iso"
all_experiment_df$condition[all_experiment_df$condition == "tardbp mutant"] <- "tardbp"
all_experiment_df$condition[all_experiment_df$condition == "wildtype"] <- "ctrl"
all_experiment_df$als_group <- all_experiment_df$condition
all_experiment_df$als_group[all_experiment_df$condition == "ctrl"] <- "ctrl"
all_experiment_df$als_group[!(all_experiment_df$condition %in% c("ctrl", "iso", "tdp43kd"))] <- "als"
all_experiment_df$als_group <- factor(all_experiment_df$als_group, levels = c("ctrl", "iso", "als", "tdp43kd"))


whole_cell_experiment_df <- all_experiment_df[all_experiment_df$fraction %in% c("who", "cytoplasm", "cyt") & 
    all_experiment_df$als_group != "tdp43kd" & 
    !(all_experiment_df$study %in% c("hnrnpa2b1-markmiller-2021", "sod1-bhinge-2017", "sod1-moccia-2014", 
        "sod1-wang-2017", "tardbp-kd-brown-2022", "tardbp-kd-kapeli-2016", "tardbp-kd-klim-2022")),]
whole_cell_experiment_df$study <- factor(whole_cell_experiment_df$study)
whole_cell_experiment_df$instrument <- factor(whole_cell_experiment_df$instrument)
whole_cell_experiment_df$library_layout <- factor(whole_cell_experiment_df$library_layout)
whole_cell_experiment_df <- whole_cell_experiment_df[whole_cell_experiment_df$als_group %in% c("ctrl", "als"),]
rownames(whole_cell_experiment_df) <- NULL

```

- Compute liabilities

```{r}

lower_threshold <- 0.1
higher_threshold <- 1
lower_liabilities <- do.call(c, lapply(whole_cell_experiment_df$samplename, function(x){
    current_tpm_vector <- keep_tpm_matrix[,x]
    sum(current_tpm_vector[current_tpm_vector > lower_threshold])
}))

higher_liabilities <- do.call(c, lapply(whole_cell_experiment_df$samplename, function(x){
    current_tpm_vector <- keep_tpm_matrix[,x]
    sum(current_tpm_vector[current_tpm_vector > higher_threshold])
}))

spinal_cord_liability <- do.call(c, lapply(whole_cell_experiment_df$samplename, function(x){
    current_tpm_vector <- keep_tpm_matrix[,x]
    sum(current_tpm_vector[names(current_tpm_vector) %in% als_unique_spinal_cord_rcis])
}))

brain_liability <- do.call(c, lapply(whole_cell_experiment_df$samplename, function(x){
    current_tpm_vector <- keep_tpm_matrix[,x]
    sum(current_tpm_vector[names(current_tpm_vector) %in% als_unique_brain_rcis])
}))


lower_liability_df <- cbind(whole_cell_experiment_df, liability = lower_liabilities, threshold = "tpm > 0.1")
higher_liability_df <- cbind(whole_cell_experiment_df, liability = higher_liabilities, threshold = "tpm > 1")
spinal_cord_liability_df <- cbind(whole_cell_experiment_df, liability = spinal_cord_liability, threshold = "spinal_cord")
brain_liability_df <- cbind(whole_cell_experiment_df, liability = brain_liability, threshold = "brain")

liability_df <- rbind(lower_liability_df, higher_liability_df, spinal_cord_liability_df, brain_liability_df)

```

- Plot rci liabilities across studies

```{r}


current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12, angle = 90, vjust = 1),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))


colours <- c("ctrl" = "#F6F4F3", "als" = "#F87575")

ThePlot <- ggplot(liability_df[liability_df$threshold == "spinal_cord",], 
    aes(x = study, y = log10(liability)))
ThePlot <- ThePlot + geom_boxplot(aes(fill = als_group), colour = "#000000", 
    width = 0.1, lwd = 0.5, position = position_dodge(width = 0.6))
ThePlot <- ThePlot + scale_x_discrete("mnd group")
ThePlot <- ThePlot + scale_y_continuous("RCI TPM")
ThePlot <- ThePlot + scale_colour_manual("Condition colour", values = colours)
ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_wrap(. ~ tissue, scales = "free")
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression in AnswerALS iPSCs")
ThePlot <- ThePlot + current_theme
ggsave(width = 15, height = 7, file = file.path(plot_dir, "07-01-plot-RCI-liability-in-polyA+-iPSC-MNs-supplementary-figure-7D.pdf"), ThePlot)

```

- Get p-values

```{r}

wilcox.test(liability_df$liability[liability_df$als_group == "als" & liability_df$threshold == "spinal_cord"], liability_df$liability[liability_df$als_group == "ctrl" & liability_df$threshold == "spinal_cord"], alternative = "greater")$p.value

vapply(split(liability_df, liability_df$study), function(x){
    wilcox.test(x$liability[x$als_group == "als" & x$threshold == "spinal_cord"], x$liability[x$als_group == "ctrl" & x$threshold == "spinal_cord"], alternative = "greater")$p.value
},1)


```

