---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-3"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
feature_ranges <- readRDS("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds")
sig_result_list <- readRDS("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds")
answerals_object_list <- readRDS("13-01-answerals/rdata_files/01-02-list_of_counts_stat_samplesheet_from_answerals.rds")
nygc_penetrance_list <- readRDS(file.path("final-figures/figure-2/rdata_files/02-save-nygc-rci-expression-summary-with-penetrance-matrix.rds"))
sampleinfo_df <- read.csv("/nemo/project/proj-luscombn-patani/working/answerals/sample-details/samplesheet.csv")

```

- Load repeat and rci object

```{r}


repeat_ranges <- unique(repeat_ranges_list$repeat_ranges)
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges


```


### Analysis


- Generate TPM matrix

```{r}

sampleinfo_df$sample_name <- str_replace_all(sampleinfo_df$participant_id, pattern = "-", replacement = "_")

sig_result_df <- sig_result_list$significant_events

repeat_containing_introns_df <- sig_result_df[sig_result_df$status == "repeat_containing" & 
    sig_result_df$region == "intron",]


repeat_containing_introns_df$cohort <- "public-data"
repeat_containing_introns_df$cohort[grepl(x = repeat_containing_introns_df$study, pattern = "nygc")] <- "nygc"
repeat_containing_introns_df$cohort[grepl(x = repeat_containing_introns_df$study, pattern = "answerals")] <- "answerals"

repeat_containing_introns_df$tissue_type <- "iPSCs"
repeat_containing_introns_df$tissue_type[grepl(x = repeat_containing_introns_df$study, pattern = "nygc")] <- "postmortem"

counts_matrix <- answerals_object_list$counts
patient_df <- answerals_object_list$sample_df
temp_matrix <- counts_matrix / (width(feature_ranges)[match(rownames(counts_matrix), feature_ranges$hash_id)] / 1e3)
tpm_matrix <- t(t(temp_matrix) / (colSums(temp_matrix) / 1e6))

keep_hash_ids <- feature_ranges_with_repeats$hash_id[match(unique(repeat_containing_introns_df$feature_ids), feature_ranges_with_repeats$feature_id)]
keep_tpm_matrix <- tpm_matrix[keep_hash_ids,colnames(tpm_matrix) %in% patient_df$column_id]

repeat_lengths <- feature_ranges_with_repeats$repeat_width[match(rownames(keep_tpm_matrix), feature_ranges_with_repeats$hash_id)]

```

- Fetch NYGC penetrant rcis

```{r}

nygc_patient_liability_df <- nygc_penetrance_list$liability_df
nygc_penetrance_matrix <- nygc_penetrance_list$penetrance_matrix

nygc_patient_df_split <- split(nygc_patient_liability_df, paste(nygc_patient_liability_df$condition, nygc_patient_liability_df$tissue_group))
nygc_ordered_patient_penetrance_matrix_list <- lapply(nygc_patient_df_split, function(temp_df){
    nygc_penetrance_matrix[temp_df$penetrance_row[order(temp_df$repeat_intron_expr_tpm, decreasing = F)],]
})

nygc_penetrant_rcis <- lapply(nygc_ordered_patient_penetrance_matrix_list, function(x){
    temp_vector <- colSums(x)
    names(temp_vector[temp_vector > 0])
})

brain_rcis <- nygc_penetrant_rcis[grepl(x = names(nygc_penetrant_rcis), pattern = "brain")]
spinal_cord_rcis <- nygc_penetrant_rcis[grepl(x = names(nygc_penetrant_rcis), pattern = "spinal_cord")]

als_unique_spinal_cord_rcis <- unique(c(spinal_cord_rcis[["als spinal_cord"]], spinal_cord_rcis[["als+ftd spinal_cord"]]))
als_unique_spinal_cord_rcis <- als_unique_spinal_cord_rcis[!(als_unique_spinal_cord_rcis %in% spinal_cord_rcis[["ctrl spinal_cord"]])]

als_unique_brain_rcis <- unique(c(brain_rcis[["als brain"]], brain_rcis[["als+ftd brain"]]))
als_unique_brain_rcis <- als_unique_brain_rcis[!(als_unique_brain_rcis %in% brain_rcis[["ctrl brain"]])]

```

- Generate liability data


```{r}

# threshold for reliable express 
# threshold <- 0.1

patient_liability_df <- do.call(rbind, lapply(seq_len(nrow(patient_df)), function(x1){
    temp_df <- patient_df[x1,]
    return_df <- temp_df
    current_columns <- temp_df$column_id
    
    current_tpm_vector <- Reduce("+", lapply(current_columns, function(x){
        keep_tpm_matrix[,x]  
    })) / length(current_columns)

    all_introns <- feature_ranges_with_repeats 
    all_introns$signal <- current_tpm_vector[all_introns$hash_id]
    all_introns$weighted.signal <- all_introns$signal / (table(all_introns$hash_id)[all_introns$hash_id])
    all_introns$weighted_signal_per_expansion <- all_introns$weighted.signal / all_introns$expansion_size
    all_introns <- all_introns[!is.na(all_introns$signal)]

    expected_split <- split(all_introns$weighted.signal, all_introns$span_group)
    total_expected_signal <- vapply(expected_split, sum, 1)
    prop_expected <- total_expected_signal / sum(total_expected_signal)

    return_df$spinal_cord_liability <- sum(current_tpm_vector[current_tpm_vector > 0.1 & names(current_tpm_vector) %in% als_unique_spinal_cord_rcis])
    return_df$brain_liability <- sum(current_tpm_vector[current_tpm_vector > 0.1 & names(current_tpm_vector) %in% als_unique_brain_rcis])
    if(any(is.na(return_df$repeat_intron_expr_tpm))){
        stop(x1)
    }
    # return_df$median_repeat_intron_expr_tpm <- median(current_tpm_vector[current_tpm_vector > threshold])

    current_introns <- all_introns[all_introns$hash_id %in% c(als_unique_spinal_cord_rcis, als_unique_brain_rcis)]

    width_split <- split(current_introns$repeat_width, current_introns$span_group)
    total_width <- sum(current_introns$repeat_width)
    prop_width <- vapply(width_split, sum, 1) / total_width

    signal_split <- split(current_introns$weighted.signal, current_introns$span_group)

    observed_proportion <- vapply(c("trinucleotide", "quadnucleotide", "pentanucleotide", "hexanucleotide"), function(x){
        if(x %in% names(signal_split)){
            return(sum(signal_split[[x]]) / sum(current_introns$weighted.signal))
        }else{
            return(0)
        }
    },1)
    enrichment <- observed_proportion / prop_expected[names(observed_proportion)]

    overall_mean_signal <- mean(current_introns$weighted.signal)
    mean_weighted_signal <- vapply(c("trinucleotide", "quadnucleotide", "pentanucleotide", "hexanucleotide"), function(x){
        if(x %in% current_introns$span_group){
            return(mean(current_introns$weighted.signal[current_introns$span_group == x]))
        }else{
            return(0)
        }
    },1)

    share_of_signal <- vapply(c("trinucleotide", "quadnucleotide", "pentanucleotide", "hexanucleotide"), function(x){
        if(x %in% current_introns$span_group){
            prop <- sum(width(current_introns[current_introns$span_group == x])) / sum(width(current_introns))
            return(prop  * (mean_weighted_signal[x] / overall_mean_signal))
        }else{
            return(0)
        }
    },1)

    within_group_theil <- vapply(c("trinucleotide", "quadnucleotide", "pentanucleotide", "hexanucleotide"), function(x){
        if(x %in% current_introns$span_group){
            current_signal <- current_introns$weighted.signal[current_introns$span_group == x]
            inner_ratio <- current_signal / mean_weighted_signal[x]
            return(sum(inner_ratio * log(inner_ratio)))
        }else{
            return(0)
        }
    },1)

    overall_ratios <- vapply(c("trinucleotide", "quadnucleotide", "pentanucleotide", "hexanucleotide"), function(x){
        if(x %in% current_introns$span_group){
            return(mean_weighted_signal[x] / overall_mean_signal)
        }else{
            return(0)
        }
    },1)
    all_between_repeat_component <- log(overall_ratios) * share_of_signal
    between_repeat_component <- sum(all_between_repeat_component)


    theil_t <- sum(share_of_signal * within_group_theil) + between_repeat_component

    names(enrichment) <- paste(names(enrichment), "enrichment", sep = "_")
    names(observed_proportion) <- paste(names(observed_proportion), "observed", sep = "_")
    names(all_between_repeat_component) <- paste(names(all_between_repeat_component), "between_theil", sep = "_")

    new_return_df <- return_df
    new_return_df$entropy <- theil_t
    return(cbind(new_return_df[1,], t(as.matrix(enrichment)), t(as.matrix(observed_proportion)), t(as.matrix(all_between_repeat_component))))
}))

```


- Attach metadata

```{r}

sample_name_matches <- match(patient_liability_df$sample_name, sampleinfo_df$sample_name)

patient_liability_df$di_m_ns_line_name <- sampleinfo_df$di_m_ns_line_name[sample_name_matches]

dimn_matches <- match(patient_liability_df$di_m_ns_line_name, non_na_marker_df[,"diMNs-line"])
patient_liability_df$marker_data_available <- !is.na(dimn_matches)
patient_liability_df[,"SMI32"] <- non_na_marker_df[dimn_matches,"SMI32"]
patient_liability_df[,"ISL1"] <- non_na_marker_df[dimn_matches,"ISL1"]
patient_liability_df[,"NKX6.1"] <- non_na_marker_df[dimn_matches,"NKX6.1"]
patient_liability_df[,"TUJ1"] <- non_na_marker_df[dimn_matches,"TUJ1"]
patient_liability_df[,"S100b"] <- non_na_marker_df[dimn_matches,"S100b"]
patient_liability_df$nygc_cgnd <- sampleinfo_df$nygc_cgnd_id[sample_name_matches]
patient_liability_df$progression <- sampleinfo_df$progression[sample_name_matches]
patient_liability_df$age_onset <- sampleinfo_df$age_onset[sample_name_matches]
patient_liability_df$age_at_collection <- sampleinfo_df$age_onset[sample_name_matches]
patient_liability_df$early_late_onset <- sampleinfo_df$early_late_onset[sample_name_matches]
patient_liability_df$early_late_death <- sampleinfo_df$early_late_death[sample_name_matches]
patient_liability_df$family_history <- sampleinfo_df$family_history[sample_name_matches]
patient_liability_df$mnd_group <- sampleinfo_df$mnd_group[sample_name_matches]
patient_liability_df$mutation <- sampleinfo_df$mutation[sample_name_matches]
patient_liability_df$expansion_group <- ifelse(patient_liability_df$mutation == "c9orf72", "c9orf72_expansion", "lacking_expansion")
patient_liability_df$expansion_group[patient_liability_df$mnd_group == "healthy"] <- "control"

```

- Plot nygc specific liability values

```{r}

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))
colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")

ThePlot <- ggplot(patient_liability_df[patient_liability_df$mnd_group %in% c("als", "healthy"),], 
    aes(x = mnd_group, y = spinal_cord_liability))
ThePlot <- ThePlot + geom_boxplot(aes(fill = mnd_group), colour = "#000000", 
    width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9))
ThePlot <- ThePlot + geom_quasirandom(aes(fill = mnd_group), colour = "#000000", shape = 21)
ThePlot <- ThePlot + scale_x_discrete("mnd group")
ThePlot <- ThePlot + scale_y_continuous("RCI liability")
ThePlot <- ThePlot + scale_colour_manual("Condition colour", values = colours)
ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_wrap(. ~ tissue, scales = "free")
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression in AnswerALS iPSCs")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 10, file = file.path(plot_dir, "07-02-plot-NYGC-spinal_cord-specific-RCI-liability-in-healthy-and-ALS-supplementary-figure-7D.pdf"), ThePlot)



ThePlot <- ggplot(patient_liability_df[patient_liability_df$mnd_group %in% c("als", "healthy"),], 
    aes(x = mnd_group, y = brain_liability))
ThePlot <- ThePlot + geom_boxplot(aes(fill = mnd_group), colour = "#000000", 
    width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9))
ThePlot <- ThePlot + geom_quasirandom(aes(fill = mnd_group), colour = "#000000", shape = 21)
ThePlot <- ThePlot + scale_x_discrete("mnd group")
ThePlot <- ThePlot + scale_y_continuous("RCI liability")
ThePlot <- ThePlot + scale_colour_manual("Condition colour", values = colours)
ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_wrap(. ~ tissue, scales = "free")
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression in AnswerALS iPSCs")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 10, file = file.path(plot_dir, "07-03-plot-NYGC-brain-specific-RCI-liability-in-healthy-and-ALS-supplementary-figure-7D.pdf"), ThePlot)


```
