---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")
require("ggbeeswarm")
require("parallel")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-3"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
feature_ranges <- readRDS("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds")
sig_result_list <- readRDS("20-import-significant-events/rdata_files/02-significantly-differentially-expressed-events.rds")
answerals_object_list <- readRDS("13-01-answerals/rdata_files/01-02-list_of_counts_stat_samplesheet_from_answerals.rds")
sampleinfo_df <- read.csv("/nemo/project/proj-luscombn-patani/working/answerals/sample-details/samplesheet.csv")
cell_line_markers_csv <- read.csv("13-01-answerals/input_files/Baxi-et-al-IF-raw-data-file-for-Koustav-Pal-confirmed-data.csv")
patient_liability_df <- readRDS(file.path(save_dir, "02-save-answerals-rci-expression-summary.rds"))

```

### Analysis

- Keep only those lines which have a corresponding marker signal

```{r}

keep_patient_liability_df <- patient_liability_df[patient_liability_df$marker_data_available,]
table(patient_liability_df$marker_data_available)
# FALSE  TRUE
#   108   179

vapply(split(keep_patient_liability_df, keep_patient_liability_df$mnd_group), function(temp_df){
	mean(temp_df$repeat_intron_expr_tpm)
},1)
 #      als   carrier   healthy other_mnd
 # 156.1175  176.9971  167.0744  169.3146

```

- Plot RCI liability vs marker signal

```{r}

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")

plot_data_df <- do.call(rbind, lapply(c("SMI32", "ISL1", "NKX6.1", "TUJ1", "S100b"), function(x){
    keep_patient_liability_df$marker <- x
    keep_patient_liability_df$marker_signal <- keep_patient_liability_df[,x]
    return(keep_patient_liability_df)
}))


colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")

ThePlot <- ggplot(plot_data_df[plot_data_df$mnd_group %in% c("healthy", "als"),], 
    aes(x = marker_signal))
ThePlot <- ThePlot + geom_density(fill = "#F6F4F3")
ThePlot <- ThePlot + scale_x_continuous("Marker percentage")
ThePlot <- ThePlot + scale_y_continuous("density")
ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_wrap(. ~ marker, scales = "free")
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression in AnswerALS iPSCs")
ThePlot <- ThePlot + current_theme
ggsave(width = 10, height = 5, file = file.path(plot_dir, "03-01-supplementary-figure-7A.pdf"), ThePlot)


```

- Plot marker vs liability plots

```{r}


ThePlot <- ggplot(plot_data_df[plot_data_df$mnd_group %in% c("healthy", "als"),],
    aes(x = marker_signal, y = repeat_intron_expr_tpm))
ThePlot <- ThePlot + geom_point(aes(fill = mnd_group), colour = "#000000", shape = 21)
ThePlot <- ThePlot + geom_smooth(method = "lm", formula = y ~ x, colour = "#000000")
ThePlot <- ThePlot + scale_x_continuous("Marker percentage")
ThePlot <- ThePlot + scale_y_continuous("RCI liability")
ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_wrap(. ~ marker, scales = "free")
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression in AnswerALS iPSCs")
ThePlot <- ThePlot + current_theme
ggsave(width = 10, height = 5, file = file.path(plot_dir, "03-02-supplementary-figure-7A.pdf"), ThePlot)


```

- Plot patient liability on the basis of marker groups 

```{r}


patient_liability_df$marker_group <- ifelse(patient_liability_df$marker_data_available, "marker_available", "marker_not_available")

ThePlot <- ggplot(patient_liability_df[patient_liability_df$mnd_group %in% c("als", "healthy"),], 
    aes(x = mnd_group, y = repeat_intron_expr_tpm))
ThePlot <- ThePlot + geom_boxplot(aes(fill = mnd_group), colour = "#000000", 
    width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9))
ThePlot <- ThePlot + geom_quasirandom(aes(fill = mnd_group), colour = "#000000", shape = 21)
ThePlot <- ThePlot + scale_x_discrete("mnd group")
ThePlot <- ThePlot + scale_y_continuous("RCI liability")
ThePlot <- ThePlot + scale_colour_manual("Condition colour", values = colours)
ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_wrap(. ~ marker_group)
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression in AnswerALS iPSCs")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 10, file = file.path(plot_dir, "03-03-plot-RCI-liability-in-healthy-and-ALS-stratifying-by-markers-supplementary-figure-7A.pdf"), ThePlot)


lapply(split(patient_liability_df, patient_liability_df$mnd_group)[c("als", "healthy")], function(x){
	wilcox.test(x$repeat_intron_expr_tpm[x$marker_group == "marker_available"], x$repeat_intron_expr_tpm[x$marker_group == "marker_not_available"])$p.value
})

# $als
# [1] 0.02352673

# $healthy
# [1] 0.007593504

```


- Investigate association between gene expression and marker signal

```{r}

counts_matrix <- answerals_object_list$counts

# select exons 
exonic_ranges <- feature_ranges[feature_ranges$type == "exon"]
exons_split_by_gene_id <- split(exonic_ranges$hash_id, exonic_ranges$gene_id)

gene_expression_matrix <- do.call(rbind, mclapply(exons_split_by_gene_id, function(x){
	if(length(x) == 1){
		return(counts_matrix[x,])
	}
	return(colSums(counts_matrix[x,]))
}, mc.cores = 8))




estimate_starting_effect <- function(liability_df, gene_expression_df, predictor = "liability", marker_col){
	starting_gene_effect <- do.call(rbind, lapply(seq_len(ncol(gene_expression_df)), function(x2){
		var_name <- paste("gene",x2, sep = "_")
		construct_df <- cbind(data.frame(liability = liability_df$repeat_intron_expr_tpm,
					mnd_group = liability_df[,"mnd_group"],
					marker = liability_df[,marker_col]), gene_expression_df[,x2])
		construct_df$mnd_group <- factor(construct_df$mnd_group, levels = c("healthy", "carrier", "other_mnd", "als"))
		colnames(construct_df) <- c("liability", "mnd_group", "marker", var_name)
		current_formula <- paste(predictor , paste("mnd_group", paste(var_name, collapse = " + "), sep = " + "), sep = "~")
		lm_summary <- summary(lm(data = construct_df, formula = formula(current_formula)))
		coefficients_vector <- lm_summary$coefficients
		c(rsq = lm_summary$r.squared, pval = coefficients_vector[var_name,4])
	}))
	rownames(starting_gene_effect) <- colnames(gene_expression_df)
	return(starting_gene_effect)
}


return_lm_results <- function(linear_model){
	lm_summary <- summary(linear_model)
	coefficients_vector <- lm_summary$coefficients
	if(nrow(coefficients_vector) == 1){
		return(NULL)
	}
	coef_vector <- do.call(cbind, lapply(rownames(coefficients_vector), function(x){
		ret_vector <- data.frame("est" = coefficients_vector[x,1], "pval" = coefficients_vector[x,4])
		colnames(ret_vector) <- c(paste(x,"est", sep = "_"), paste(x,"pval", sep = "_"))
		return(ret_vector)
	}))
	return_vector <- data.frame(rsq = lm_summary$r.squared, 
		RSS = crossprod(lm_summary$residuals),
		MSE = crossprod(lm_summary$residuals) / length(lm_summary$residuals),
		RMSE = sqrt(crossprod(lm_summary$residuals) / length(lm_summary$residuals)))
	return(cbind(return_vector, coef_vector))
}



liability_effect_list <- lapply(c("SMI32", "ISL1", "NKX6.1", "TUJ1", "S100b"), function(x1){
	message(x1)
	patient_columns <- keep_patient_liability_df$column_id
	marker_signal <- keep_patient_liability_df[,x1]
	liability_signal <- keep_patient_liability_df$repeat_intron_expr_tpm
	aggregate_df <- do.call(rbind, mclapply(seq_len(nrow(gene_expression_matrix)), function(x2){
		# message(x2)
		current_vector <- gene_expression_matrix[x2,patient_columns]
		temp_df <- data.frame(expr = current_vector, marker = marker_signal)
		model <- lm(data = temp_df, formula = marker ~ expr)
		return_results <- return_lm_results(model)
		if(!is.null(return_results)){
			return(cbind(data.frame(gene_id = rownames(gene_expression_matrix)[x2], expr = mean(current_vector)), return_results))	
		}else{
			return(NULL)
		}
	}, mc.cores = 8))
	aggregate_df$marker <- x1
	aggregate_df <- aggregate_df[!is.na(aggregate_df$expr_pval),]
	aggregate_df$expr_padj <- p.adjust(aggregate_df$expr_pval, method = "fdr")
	significant_df <- aggregate_df[aggregate_df$expr_padj < 0.05,]

	if(nrow(significant_df) == 0){
		return(NULL)
	}

	liability_result_df <- do.call(rbind, lapply(significant_df$gene_id, function(a_gene_id){
		current_vector <- gene_expression_matrix[a_gene_id,patient_columns]
		temp_df <- data.frame(expr = current_vector, liability = liability_signal)
		model <- lm(data = temp_df, formula = liability ~ expr)
		return_results <- return_lm_results(model)
		if(!is.null(return_results)){
			return(cbind(data.frame(gene_id = a_gene_id, expr = mean(current_vector)), return_results))	
		}else{
			return(NULL)
		}
	}))
	significant_df$liability_est <- liability_result_df$expr_est[match(significant_df$gene_id, liability_result_df$gene_id)]
	significant_df$liability_pval <- liability_result_df$expr_pval[match(significant_df$gene_id, liability_result_df$gene_id)]
	significant_df$liability_padj <- p.adjust(significant_df$liability_pval, method = "fdr")

	significant_liability_df <- significant_df[significant_df$liability_padj < 0.05,]
	if(nrow(significant_liability_df) > 100){
		significant_liability_df <- significant_df[order(significant_df$liability_padj, decreasing = F),][1:100,]
	}
	
	current_gene_expr_df <- do.call(cbind, lapply(significant_liability_df$gene_id, function(x3){
		gene_expression_matrix[x3,patient_columns]
	}))
	colnames(current_gene_expr_df) <- significant_liability_df$gene_id

	#estimate starting gene effect on liability
	liability_starting_effect <- as.data.frame(estimate_starting_effect(keep_patient_liability_df, current_gene_expr_df, marker_col = x1, predictor = "liability"))
	marker_starting_effect <- as.data.frame(estimate_starting_effect(keep_patient_liability_df, current_gene_expr_df, marker_col = x1, predictor = "marker"))

	liability_starting_effect$marker_rsq <- marker_starting_effect[,"rsq"]
	liability_starting_effect$marker_pval <- marker_starting_effect[,"pval"]
	liability_starting_effect$combined_rsq <- liability_starting_effect$rsq + liability_starting_effect$marker_rsq
	liability_starting_effect$combined_pval <- liability_starting_effect$pval + liability_starting_effect$marker_pval

	return(liability_starting_effect[order(liability_starting_effect$combined_rsq, decreasing = T),])

	# current_gene_expr_df <- current_gene_expr_df[,rownames(liability_starting_effect)[order(liability_starting_effect$combined_rsq, decreasing = T)]]

	# all_gene_effect <- do.call(rbind, lapply(seq_len(ncol(current_gene_expr_df)), function(x2){
	# 	message(x2)
	# 	construct_df <- cbind(data.frame(liability = keep_patient_liability_df$repeat_intron_expr_tpm,
	# 				mnd_group = keep_patient_liability_df[,"mnd_group"],
	# 				marker = keep_patient_liability_df[,x1]), current_gene_expr_df[,1:x2])
	# 	construct_df$mnd_group <- factor(construct_df$mnd_group, levels = c("healthy", "carrier", "other_mnd", "als"))
	# 	colnames(construct_df) <- c("liability", "mnd_group", "marker", paste("gene",1:x2, sep = "_"))

	# 	current_formula <- paste("liability" , paste("mnd_group", "marker", paste(paste(paste("gene",1:x2, sep = "_")), collapse = " + "), sep = " + "), sep = "~")
	# 	lm_summary <- summary(lm(data = construct_df, formula = formula(current_formula)))
	# 	coefficients_vector <- lm_summary$coefficients
	# 	c(rsq = lm_summary$r.squared, 
	# 		RSS = crossprod(lm_summary$residuals),
	# 		MSE = crossprod(lm_summary$residuals) / length(lm_summary$residuals),
	# 		RMSE = sqrt(crossprod(lm_summary$residuals) / length(lm_summary$residuals)),
	# 		# marker_est = coefficients_vector["marker",1], 
	# 		# marker_pval = coefficients_vector["marker",4], 
	# 		als_est = coefficients_vector["mnd_groupals",1], 
	# 		als_pval = coefficients_vector["mnd_groupals",4])
	# }))
	# rownames(all_gene_effect) <- colnames(current_gene_expr_df)
	# return(all_gene_effect)
})


names(liability_effect_list) <- c("SMI32", "ISL1", "NKX6.1", "TUJ1", "S100b")

gene_list <- unique(do.call(c, lapply(liability_effect_list, function(x){
	rownames(x)
})))

all_gene_starting_effects_on_liability <- as.data.frame(do.call(rbind, lapply(gene_list, function(x){
	do.call(cbind, lapply(liability_effect_list, function(y){
		if(x %in% rownames(y)){
			return(y[x,"rsq"])
		}
		return(0)
	}))
})))
rownames(all_gene_starting_effects_on_liability) <- gene_list
all_gene_starting_effects_on_liability$mean_rsq <- rowMeans(all_gene_starting_effects_on_liability)
all_gene_starting_effects_on_liability <- all_gene_starting_effects_on_liability[order(all_gene_starting_effects_on_liability$mean_rsq, decreasing = T),]

all_gene_expression_df <- do.call(cbind, lapply(rownames(all_gene_starting_effects_on_liability), function(x3){
		gene_expression_matrix[x3,keep_patient_liability_df$column_id]
}))
colnames(all_gene_expression_df) <- rownames(all_gene_starting_effects_on_liability)


all_gene_effect <- do.call(rbind, lapply(seq_len(ncol(all_gene_expression_df)), function(x2){
	# message(x2)
	construct_df <- cbind(data.frame(liability = keep_patient_liability_df$repeat_intron_expr_tpm,
				mnd_group = keep_patient_liability_df[,"mnd_group"]), all_gene_expression_df[,1:x2])
	construct_df$mnd_group <- factor(construct_df$mnd_group, levels = c("healthy", "carrier", "other_mnd", "als"))
	colnames(construct_df) <- c("liability", "mnd_group", paste("gene",1:x2, sep = "_"))

	current_formula <- paste("liability" , paste("mnd_group", paste(paste(paste("gene",1:x2, sep = "_")), collapse = " + "), sep = " + "), sep = "~")
	lm_summary <- summary(lm(data = construct_df, formula = formula(current_formula)))
	coefficients_vector <- lm_summary$coefficients
	c(rsq = lm_summary$r.squared, 
		RSS = crossprod(lm_summary$residuals),
		MSE = crossprod(lm_summary$residuals) / length(lm_summary$residuals),
		RMSE = sqrt(crossprod(lm_summary$residuals) / length(lm_summary$residuals)),
		# marker_est = coefficients_vector["marker",1], 
		# marker_pval = coefficients_vector["marker",4], 
		als_est = coefficients_vector["mnd_groupals",1], 
		als_pval = coefficients_vector["mnd_groupals",4])
}))
rownames(all_gene_effect) <- colnames(all_gene_expression_df)

most_predictive_features <- rownames(all_gene_effect)[all_gene_effect[,"als_pval"] < 0.05 & !is.na(all_gene_effect[,"als_pval"])]

marker_fit_matrix <- do.call(rbind, lapply(c("SMI32", "ISL1", "NKX6.1", "TUJ1", "S100b"), function(x1){
	construct_df <- cbind(data.frame(liability = keep_patient_liability_df$repeat_intron_expr_tpm,
		marker = keep_patient_liability_df[,x1],
		mnd_group = keep_patient_liability_df[,"mnd_group"]), all_gene_expression_df[,most_predictive_features])
	construct_df$mnd_group <- factor(construct_df$mnd_group, levels = c("healthy", "carrier", "other_mnd", "als"))
	colnames(construct_df) <- c("liability", "marker", "mnd_group", paste("gene",seq_along(most_predictive_features), sep = "_"))

	current_formula <- paste("marker" , paste("mnd_group", paste(paste(paste("gene",seq_along(most_predictive_features), sep = "_")), collapse = " + "), sep = " + "), sep = "~")
	lm_summary <- summary(lm(data = construct_df, formula = formula(current_formula)))
	coefficients_vector <- lm_summary$coefficients
	data.frame(marker = x1,
		rsq = lm_summary$r.squared, 
		RSS = crossprod(lm_summary$residuals),
		MSE = crossprod(lm_summary$residuals) / length(lm_summary$residuals),
		RMSE = sqrt(crossprod(lm_summary$residuals) / length(lm_summary$residuals)),
		# marker_est = coefficients_vector["marker",1], 
		# marker_pval = coefficients_vector["marker",4], 
		als_est = coefficients_vector["mnd_groupals",1], 
		als_pval = coefficients_vector["mnd_groupals",4])
}))
#   marker       rsq      RSS      MSE     RMSE     als_est  als_pval
# 1  SMI32 0.5462867 7650.936 42.74266 6.537787 -2.58966157 0.3583354
# 2   ISL1 0.7612091 1834.331 10.24766 3.201196  1.09369463 0.4279120
# 3 NKX6.1 0.7123302 4212.320 23.53251 4.851032  3.08140721 0.1422971
# 4   TUJ1 0.6407933 4762.709 26.60731 5.158228  0.94219519 0.6712711
# 5  S100b 0.7503876 7361.459 41.12547 6.412914 -0.02373435 0.9931352

```

- Use these predictive features to project across all data

```{r}

test_patient_liability_df <- patient_liability_df[!(patient_liability_df$marker_data_available),]
predictive_expr_matrix <- t(gene_expression_matrix[most_predictive_features,test_patient_liability_df$column_id])

test_gene_effect <- do.call(rbind, lapply(seq_len(ncol(predictive_expr_matrix)), function(x2){
	# message(x2)
	construct_df <- cbind(data.frame(liability = test_patient_liability_df$repeat_intron_expr_tpm,
				mnd_group = test_patient_liability_df[,"mnd_group"]), predictive_expr_matrix[,1:x2])
	construct_df$mnd_group <- factor(construct_df$mnd_group, levels = c("healthy", "carrier", "other_mnd", "als"))
	colnames(construct_df) <- c("liability", "mnd_group", paste("gene",1:x2, sep = "_"))

	current_formula <- paste("liability" , paste("mnd_group", paste(paste(paste("gene",1:x2, sep = "_")), collapse = " + "), sep = " + "), sep = "~")
	lm_summary <- summary(lm(data = construct_df, formula = formula(current_formula)))
	coefficients_vector <- lm_summary$coefficients
	c(rsq = lm_summary$r.squared, 
		RSS = crossprod(lm_summary$residuals),
		MSE = crossprod(lm_summary$residuals) / length(lm_summary$residuals),
		RMSE = sqrt(crossprod(lm_summary$residuals) / length(lm_summary$residuals)),
		# marker_est = coefficients_vector["marker",1], 
		# marker_pval = coefficients_vector["marker",4], 
		als_est = coefficients_vector["mnd_groupals",1], 
		als_pval = coefficients_vector["mnd_groupals",4])
}))

```


- Get fit for linear model and reduced model 


```{r}

all_patients_gene_expr_df <- t(gene_expression_matrix[most_predictive_features,patient_liability_df$column_id])
construct_liability_df <- cbind(data.frame(
	liability = patient_liability_df$repeat_intron_expr_tpm,
	mnd_group = factor(patient_liability_df$mnd_group, levels = c("healthy", "carrier", "other_mnd", "als"))), all_patients_gene_expr_df)
colnames(construct_liability_df) <- c("liability", "mnd_group", paste("gene",seq_along(most_predictive_features), sep = "_"))

current_formula <- paste("liability" , paste("mnd_group", paste(paste(paste("gene",seq_along(most_predictive_features), sep = "_")), collapse = " + "), sep = " + "), sep = "~")
linear_model <- lm(data = construct_liability_df, formula = formula(current_formula))

reduced_model <- lm(data = construct_liability_df, formula = liability ~ mnd_group)

experimental_effect <- linear_model$fitted.values - reduced_model$fitted.values

patient_liability_df$adjusted_liability <- patient_liability_df$repeat_intron_expr_tpm - experimental_effect

```

- Plot adjusted liability 

```{r}

colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
ThePlot <- ggplot(patient_liability_df[patient_liability_df$mnd_group %in% c("als", "healthy"),], 
    aes(x = mnd_group, y = adjusted_liability))
ThePlot <- ThePlot + geom_boxplot(aes(fill = mnd_group), colour = "#000000", 
    width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9))
ThePlot <- ThePlot + geom_quasirandom(aes(fill = mnd_group), colour = "#000000", shape = 21)
ThePlot <- ThePlot + scale_x_discrete("mnd group")
ThePlot <- ThePlot + scale_y_continuous("adjusted RCI liability")
ThePlot <- ThePlot + scale_colour_manual("Condition colour", values = colours)
ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_grid(. ~ marker_group, scales = "free")
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression in AnswerALS iPSCs")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 10, file = file.path(plot_dir, "03-04-plot-adjusted-RCI-liability-in-mnd-groups-in-AnswerALS-supplementary-figure-7A.pdf"), ThePlot)

```
