---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-2"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}



repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
feature_ranges <- readRDS("05-simulated-reads-differential-profiling/rdata_files/01-all-merged-exonic-and-intronic-segments-in-genes.rds")
liability_list <- readRDS("final-figures/figure-2/rdata_files/02-save-nygc-rci-expression-summary-with-penetrance-matrix.rds")
# nygc_object_list <- readRDS("13-nygc-consortium/rdata_files/03-02-list_of_counts_stat_samplesheet_from_nygc.rds")
patient_df <- readRDS("13-nygc-consortium/rdata_files/03-05-nygc-patient-sampleinfo-datasheet.rds")
clinical_metadata_df <- read.csv("0103-NYGC-consortium/input_files/samplesheet.csv")
patient_liability_df <- liability_list$liability_df
penetrance_matrix <- liability_list$penetrance_matrix


```

### Analysis

- Compute expected number of repeats per kb of intron

```{r}

repeat_ranges <- repeat_ranges_list$repeat_ranges
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges
feature_ranges_with_repeats$class_specific_width <- width(feature_ranges_with_repeats) / table(feature_ranges_with_repeats$hash_id)[feature_ranges_with_repeats$hash_id]

total_intron_length_per_class <- vapply(split(feature_ranges_with_repeats$class_specific_width, feature_ranges_with_repeats$span_group), sum, 1) / 1000
total_repeat_length <- vapply(split(width(repeat_ranges), repeat_ranges$span_group), sum, 1)

expected_number_of_repeats_per_kb_of_intron <- total_repeat_length / total_intron_length_per_class[names(total_repeat_length)]


```

- Summarised penetrance matrix

```{r}

split_liability_df <- split(patient_liability_df, paste(patient_liability_df$subject_id, patient_liability_df$tissue_group))

summary_df <- do.call(rbind, lapply(split_liability_df, function(temp_df){
    return_df <- unique(temp_df[,c("subject_id", "tissue_group", "type", "condition")])
    return_df$repeat_intron_expr_tpm <- mean(temp_df$repeat_intron_expr_tpm)
    return(return_df)
}))

summarised_penetrance_matrix <- do.call(rbind, lapply(split_liability_df, function(temp_df){
    if(length(temp_df$penetrance_row) > 1){
        sum_values <- colSums(penetrance_matrix[temp_df$penetrance_row,])
        as.numeric(sum_values > 0)        
    }else{
        penetrance_matrix[temp_df$penetrance_row,]
    }
}))

summary_df$penetrance_row <- match(paste(summary_df$subject_id, summary_df$tissue_group), rownames(summarised_penetrance_matrix))
summary_df$c9_repeat_size <- vapply(clinical_metadata_df$c9_repeat_size[match(summary_df$subject_id, clinical_metadata_df$external_subject_id)], function(x){
    if(x %in% c("", "N/A")){
        return(NA)
    }
    split_text <- as.numeric(str_split(x, pattern = "/")[[1]])
    return(max(split_text))
},1)

als_patient_summary_df <- summary_df[summary_df$condition == "als",]

```

- Compute enrichment of repeat classes


```{r}

enrichment_matrix <- do.call(rbind, lapply(summary_df$penetrance_row, function(x){
    a_vector <- summarised_penetrance_matrix[x,]
    current_introns <- names(a_vector)[a_vector > 0]
    current_intron_ranges <- feature_ranges_with_repeats[feature_ranges_with_repeats$hash_id %in% current_introns]

    current_intron_length_per_class <- vapply(split(current_intron_ranges$class_specific_width, current_intron_ranges$span_group), sum, 1) / 1000
    current_repeat_length <- vapply(split(width(repeat_ranges[current_intron_ranges$which_repeat]), repeat_ranges[current_intron_ranges$which_repeat]$span_group), sum, 1)

    observed_number_of_repeats_per_kb_of_intron <- current_repeat_length / current_intron_length_per_class[names(current_repeat_length)]
    enrichment_vector <- vapply(names(expected_number_of_repeats_per_kb_of_intron), function(y){
        if(y %in% names(observed_number_of_repeats_per_kb_of_intron)){
            return( observed_number_of_repeats_per_kb_of_intron[y] / expected_number_of_repeats_per_kb_of_intron[y])
        }else{
            return(0)
        }
    },1)
    names(enrichment_vector) <- names(expected_number_of_repeats_per_kb_of_intron)
    return(enrichment_vector)
}))

summary_df <- cbind(summary_df, enrichment_matrix)

summary_df[!is.na(summary_df$c9_repeat_size) & summary_df$tissue_group == "spinal_cord" & summary_df$c9_repeat_size < 20,]

```

- Similarity between patients

```{r}

# library(factoextra)
# library(NbClust)

# current_penetrance_matrix <- summarised_penetrance_matrix[als_patient_summary_df$penetrance_row[!is.na(als_patient_summary_df$c9_repeat_size)],]
# pair_matrix <- cbind(rep(seq_len(nrow(current_penetrance_matrix)), each = nrow(current_penetrance_matrix)),
#     rep(seq_len(nrow(current_penetrance_matrix)), times = nrow(current_penetrance_matrix)))

# all_rcis <- colnames(current_penetrance_matrix)[colSums(current_penetrance_matrix) > 0]

# similarity_df <- do.call(rbind, lapply(seq_len(nrow(pair_matrix)), function(x){
#     current_x_row <- current_penetrance_matrix[pair_matrix[x,1],all_rcis]
#     current_x <- names(current_x_row)[current_x_row == 1]
#     current_y_row <- current_penetrance_matrix[pair_matrix[x,2],all_rcis]
#     current_y <- names(current_y_row)[current_y_row == 1]
#     jaccard_index <- length(intersect(current_x, current_y)) / length(union(current_x, current_y))
#     return(data.frame(x = rownames(current_penetrance_matrix)[pair_matrix[x,1]],
#                 y = rownames(current_penetrance_matrix)[pair_matrix[x,2]],
#                 ji = jaccard_index))
# }))

# similarity_matrix <- reshape2::acast(data = similarity_df, formula = x ~ y, value.var = "ji")
# dissimilarity_matrix <- 1 - similarity_matrix
# distance_matrix <- as.dist(dissimilarity_matrix)

# elbow_plot <- fviz_nbclust(similarity_matrix, FUNcluster = kmeans, method = "gap_stat", k.max = 15, nboot = 500) + 
#     geom_vline(xintercept = 5, linetype = 2) + 
#     labs(subtitle = "Elbow method")
#     ggsave(width = 10, height = 10, file = file.path(plot_dir, "03-01-plot-optimal-number-of-clusters-of-patients-supplementary-figure-7B.pdf"), elbow_plot)

# kclustering <- kmeans(x = similarity_matrix, centers = 9)

```

- Fetch average correlation value between cluster members

```{r}

# clusters <- kclustering$cluster
# clusters_split <- split(names(clusters), clusters)

# vapply(clusters_split, function(x){
#     current_matrix <- similarity_matrix[x,x]
#     mean(current_matrix)
# },1)

#         1         2         3         4         5         6         7         8
# 0.7417916 0.7119200 0.8255938 0.8473417 0.6890778 0.8015838 0.7891163 0.7839250
#         9        10        11
# 0.8826417 0.8051600 0.7655091

```

- Fetch cluster containing mostly c9 patients


```{r}

# lapply(clusters_split, function(x){
#     current_types <- summary_df$type[paste(summary_df$subject_id, summary_df$tissue_group) %in% x]
#     table(current_types)
# })

# lapply(clusters_split, function(x){
#     current_liability <- summary_df$repeat_intron_expr_tpm[paste(summary_df$subject_id, summary_df$tissue_group) %in% x]
#     mean(current_liability) 
# })

# lapply(clusters_split, function(x){
#     current_c9_size <- summary_df$c9_repeat_size[paste(summary_df$subject_id, summary_df$tissue_group) %in% x]
#     median(current_c9_size[!is.na(current_c9_size)])
# })


```

- Correlation between c9 and liability


```{r}


c9_als_patients_df <- als_patient_summary_df[!is.na(als_patient_summary_df$c9_repeat_size) & als_patient_summary_df$c9_repeat_size > 20,]
c9_als_patients_split <- split(c9_als_patients_df, paste(c9_als_patients_df$subject_id))

c9_summary_df <- do.call(rbind, lapply(c9_als_patients_split, function(temp_df){
    return_df <- unique(temp_df[,c("subject_id", "c9_repeat_size")])
    return_df$repeat_intron_expr_tpm <- mean(temp_df$repeat_intron_expr_tpm)
    return(return_df)
}))


```

- Plot C9 repeat size vs liability

```{r}

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))
colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
ThePlot <- ggplot(c9_summary_df, 
    aes(x = log10(c9_repeat_size), y = log10(repeat_intron_expr_tpm)))
ThePlot <- ThePlot + geom_point(fill = "#F87575", colour = "#000000")
ThePlot <- ThePlot + geom_smooth(method = "lm", formula = y ~ x)
ThePlot <- ThePlot + scale_x_continuous("C9 repeat length")
ThePlot <- ThePlot + scale_y_continuous("rci liability")
# ThePlot <- ThePlot + scale_fill_manual("Repeat colours", values = repeat_colour_groups)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + ggtitle("C9 repeat length versus rci liability")
ThePlot <- ThePlot + current_theme
ggsave(width = 3, height = 7, file = file.path(plot_dir, "05-01-Supplementary-figure-1.pdf"), ThePlot)

```
