---
title:Microsatellite summary analysis
output:pdf_document
---


### Main questions

This document contains all of the data required for analysing the AnswerALS data.


### 1.0 Package requirements

- change to current working directory `/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript` and include required packages 

```{r include = TRUE, message = FALSE, warning = FALSE}

setwd("/nemo/lab/patanir/home/users/palk/Projects/0006-phase-condensation-of-repeats-in-neurons/analysis/00-manuscript/")
require("HiCBricks")
require("GenomicRanges")
require("stringr")
# require("Biostrings")
# require("rtracklayer")
# require("GenomeInfoDb")
# require("GenomicFeatures")
require("ggplot2")
require("ggvenn")
require("ggbeeswarm")

```

### 2.0 Include experiment matrix

- Let's first import the counts.

```{r include = FALSE, message = FALSE, warning = FALSE}

analysis_path <- "final-figures/figure-3"
save_dir <- file.path(analysis_path, "rdata_files")
plot_dir <- file.path(analysis_path, "plots")
input_dir <- file.path(analysis_path, "input_files")
output_dir <- file.path(analysis_path, "output_files")

dir.create(save_dir, recursive = T)
dir.create(plot_dir, recursive = T)
dir.create(input_dir, recursive = T)
dir.create(output_dir, recursive = T)

```

### 3.0 Imports

- Load data files 

```{r include = TRUE, message = FALSE, warning = FALSE}

repeat_ranges_list <- readRDS("final-figures/figure-1/rdata_files/00-repeat-ranges-introns-exons-with-mapping-to-exons-and-introns.rds")
repeat_ranges <- repeat_ranges_list$repeat_ranges
feature_ranges_with_repeats <- repeat_ranges_list$repeat_containing_ranges
patient_penetrance_list <- readRDS(file.path(save_dir, "03-save-answerals-rci-expression-summary-with-penetrance-matrix.rds"))
nygc_penetrance_list <- readRDS(file.path("final-figures/figure-2/rdata_files/02-save-nygc-rci-expression-summary-with-penetrance-matrix.rds"))

patient_liability_df <- patient_penetrance_list$liability_df
penetrance_matrix <- patient_penetrance_list$penetrance_matrix
rownames(penetrance_matrix) <- patient_liability_df$sample_name


nygc_patient_liability_df <- nygc_penetrance_list$liability_df
nygc_penetrance_matrix <- nygc_penetrance_list$penetrance_matrix


```

- Compute penetrance values for answerals and nygc


```{r}

answerals_sample_split <- split(patient_liability_df$sample_name, patient_liability_df$mnd_group)[c("healthy", "als")]

penetrant_rcis <- lapply(answerals_sample_split, function(x){
	current_matrix <- penetrance_matrix[x,]
	colnames(current_matrix)[colSums(current_matrix) > 0]
})

answerals_penetrance_df <- do.call(rbind, lapply(names(answerals_sample_split), function(condition){
	x <- answerals_sample_split[[condition]]
	current_matrix <- penetrance_matrix[x,]
	keep_matrix <- current_matrix[,colSums(current_matrix) > 0]
	penetrance_values <- colSums(keep_matrix) / nrow(keep_matrix)
	data.frame(hash_id = names(penetrance_values), 
		condition = condition, 
		penetrance = penetrance_values)
}))

answerals_penetrance_df$answerals_group[answerals_penetrance_df$hash_id %in% penetrant_rcis[["als"]] & answerals_penetrance_df$hash_id %in% penetrant_rcis[["healthy"]]] <- "ALS+Control"
answerals_penetrance_df$answerals_group[answerals_penetrance_df$hash_id %in% penetrant_rcis[["als"]] & !(answerals_penetrance_df$hash_id %in% penetrant_rcis[["healthy"]])] <- "ALS-specific"
answerals_penetrance_df$answerals_group[!(answerals_penetrance_df$hash_id %in% penetrant_rcis[["als"]]) & answerals_penetrance_df$hash_id %in% penetrant_rcis[["healthy"]]] <- "Control-specific"

```

- Plot penetrance distribution


```{r}

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
answerals_penetrance_df$condition <- factor(answerals_penetrance_df$condition, levels = c("healthy", "als", "other_mnd", "carrier"))
ThePlot <- ggplot(answerals_penetrance_df, 
    aes(x = penetrance))
# ThePlot <- ThePlot + geom_violin(aes(fill = condition), colour = "#000000")
# ThePlot <- ThePlot + geom_boxplot(aes(fill = condition), colour = "#000000", 
#     width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9))
# ThePlot <- ThePlot + geom_quasirandom(aes(fill = condition), colour = "#000000", shape = 21)
ThePlot <- ThePlot + geom_density(aes(fill = condition), colour = "#000000", alpha = 0.6)
# ThePlot <- ThePlot + geom_smooth(method = "lm", aes(colour = mnd_group), formula = y ~ x)
ThePlot <- ThePlot + scale_x_continuous("penetrance")
# ThePlot <- ThePlot + scale_y_continuous("")
ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
# ThePlot <- ThePlot + facet_grid(. ~ repeat_classes)
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression in AnswerALS iPSCs")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 10, file = file.path(plot_dir, "06-07-plot-RCI-penetrance-between-healthy-and-als-figure-2E.pdf"), ThePlot)


```

- Get repeat classes penentrance

```{r}

repeat_classes <- split(feature_ranges_with_repeats$hash_id, feature_ranges_with_repeats$span_group)
repeat_classes_penetrance_df <- do.call(rbind, lapply(names(repeat_classes), function(x){
	current_df <- answerals_penetrance_df[answerals_penetrance_df$hash_id %in% repeat_classes[[x]],]
	current_df$repeat_class <- x
	return(current_df)
}))


colours <- c("healthy" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
answerals_penetrance_df$condition <- factor(answerals_penetrance_df$condition, levels = c("healthy", "als", "other_mnd", "carrier"))
ThePlot <- ggplot(repeat_classes_penetrance_df, 
    aes(x = penetrance))
# ThePlot <- ThePlot + geom_violin(aes(fill = condition), colour = "#000000")
# ThePlot <- ThePlot + geom_boxplot(aes(fill = condition), colour = "#000000", 
#     width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9))
# ThePlot <- ThePlot + geom_quasirandom(aes(fill = condition), colour = "#000000", shape = 21)
ThePlot <- ThePlot + geom_density(aes(fill = condition), colour = "#000000", alpha = 0.6)
# ThePlot <- ThePlot + geom_smooth(method = "lm", aes(colour = mnd_group), formula = y ~ x)
ThePlot <- ThePlot + scale_x_continuous("penetrance")
# ThePlot <- ThePlot + scale_y_continuous("")
ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_grid(. ~ repeat_class)
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression in AnswerALS iPSCs")
ThePlot <- ThePlot + current_theme
ggsave(width = 7, height = 10, file = file.path(plot_dir, "06-07-plot-RCI-penetrance-between-healthy-and-als-group-by-repeat-class-figure-2E.pdf"), ThePlot)



```


- Compute penetrance values for anygc


```{r}


summary_df <- do.call(rbind, lapply(split(nygc_patient_liability_df, paste(nygc_patient_liability_df$subject_id, nygc_patient_liability_df$tissue_group)), function(temp_df){
    return_df <- unique(temp_df[,c("subject_id", "tissue_group", "condition")])
    return_df$repeat_intron_expr_tpm <- mean(temp_df$repeat_intron_expr_tpm)
    return(return_df)
}))

summarised_penetrance_matrix <- do.call(rbind, lapply(split(nygc_patient_liability_df, paste(nygc_patient_liability_df$subject_id, nygc_patient_liability_df$tissue_group)), function(temp_df){
    current_rows <- temp_df$penetrance_row
    if(length(current_rows) == 1){
    	return(nygc_penetrance_matrix[current_rows,])
    }
    return_vector <- colSums(nygc_penetrance_matrix[current_rows,]) 
    return_vector[return_vector > 0] <- 1
    return(return_vector)
}))
summary_df$penetrance_row <- seq_len(nrow(summarised_penetrance_matrix))


nygc_rci_split <- split(summary_df$penetrance_row, paste(summary_df$tissue_group, summary_df$condition))
nygc_penetrance_df <- do.call(rbind, lapply(names(nygc_rci_split), function(a_group){
	x <- nygc_rci_split[[a_group]]
	current_matrix <- summarised_penetrance_matrix[x,]
	keep_matrix <- current_matrix[,colSums(current_matrix) > 0]
	penetrance_values <- colSums(keep_matrix) / nrow(keep_matrix)
	data.frame(hash_id = names(penetrance_values), 
		condition = str_split(a_group, pattern = " ", simplify = T)[,2], 
		tissue_group = str_split(a_group, pattern = " ", simplify = T)[,1],
		penetrance = penetrance_values)
}))


```

- Plot penetrance distribution


```{r}

current_theme <- theme(
    legend.position = "bottom",
    text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "azure3",
    linetype = "dashed",
    size = 0.2))

colours <- c("ctrl" = "#F6F4F3", "als" = "#F87575", "als+ftd" = "#FFA9A3")
nygc_penetrance_df$condition <- factor(nygc_penetrance_df$condition, levels = c("ctrl", "als"))
ThePlot <- ggplot(nygc_penetrance_df[nygc_penetrance_df$condition %in% c("ctrl", "als"),], 
    aes(x = penetrance))
# ThePlot <- ThePlot + geom_violin(aes(fill = condition), colour = "#000000")
# ThePlot <- ThePlot + geom_boxplot(aes(fill = condition), colour = "#000000", 
#     width = 0.1, lwd = 0.5, position = position_dodge(width = 0.9))
# ThePlot <- ThePlot + geom_quasirandom(aes(fill = condition), colour = "#000000", shape = 21)
ThePlot <- ThePlot + geom_density(aes(fill = condition), colour = "#000000", alpha = 0.6)
# ThePlot <- ThePlot + geom_smooth(method = "lm", aes(colour = mnd_group), formula = y ~ x)
ThePlot <- ThePlot + scale_x_continuous("penetrance")
# ThePlot <- ThePlot + scale_y_continuous("")
ThePlot <- ThePlot + scale_fill_manual("Condition colour", values = colours)
# ThePlot <- ThePlot + scale_fill_brewer("genotypes", palette = "Set1")
ThePlot <- ThePlot + theme_bw()
ThePlot <- ThePlot + facet_grid(. ~ tissue_group)
ThePlot <- ThePlot + ggtitle("Repeat containing intron expression in AnswerALS iPSCs")
ThePlot <- ThePlot + current_theme
ggsave(width = 10, height = 7, file = file.path(plot_dir, "06-08-plot-nygc-RCI-penetrance-between-healthy-and-als-figure-2E.pdf"), ThePlot)


```


